---
title: "DS4I_Assignment_2"
author: "Hope Hennessy"
date: "2025-09-28"
output: pdf_document
---




Build an ensemble recommender system for book recommendations using a modified "Book-Crossing" dataset containing ratings (0-10 scale) from 10,000 users on 150 books.
Core Requirements

1. Build Four Types of Recommender Systems:

* Item-based collaborative filtering (code from scratch)
* User-based collaborative filtering (code from scratch)
* Matrix factorization-based collaborative filtering
* Neural network-based collaborative filtering

2. System Capabilities:

Recommend books to existing users
Handle new users (assuming they provide ratings for ≤5 books initially)

3. Evaluation and Analysis:

* Compare accuracy across all four methods using cross-validation
* Investigate the relationship between dataset size and accuracy (e.g., how does accuracy change with 5 vs 50 vs 100 titles?)
* Determine if there's a point where adding more titles doesn't improve accuracy

4. Data Analysis:

* Conduct exploratory data analysis (EDA)
* Use findings to inform train/test data splitting




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 5, fig.height = 5, 
                      fig.align = "center", warning = FALSE, message = FALSE, 
                      fig.show = 'hold', out.width = '70%')


library(tidyverse)
library(patchwork)
library(caret)
library(kableExtra)
library(recosystem)
```

```{r}
load("book_ratings.Rdata")

# Data structure
head(book_info)
str(book_info)
dim(book_info)
head(book_ratings)
str(book_ratings)
dim(book_ratings)
head(user_info) # don't need Age to build recommender, but can include this info if want to go further
str(user_info)
dim(user_info)


# print("Missing values in ratings:")
sum(is.na(book_ratings$Book.Rating))
print("Unique users:")
length(unique(book_ratings$User.ID))
print("Unique books:")
length(unique(book_ratings$ISBN))


# Check for missing values 
colSums(is.na(book_info))
colSums(is.na(book_ratings))
colSums(is.na(user_info))  # 12098 missing Age values

```

```{r}
# Merging book_ratings with book_info 
data <- book_ratings %>%
  left_join(book_info, by = "ISBN")

summary(data) # can clearly see age has some impossible outliers
head(data)
dim(data)

sapply(data, function(x) if(is.numeric(x)) range(x, na.rm = TRUE)) # check var ranges

table(data$Book.Rating) # Zero means no rating


length(unique(data$User.ID))
length(data$User.ID)

```

```{r}
# Group and count the number of records per User.ID
hist_data <- data %>%
  group_by(User.ID) %>%
  count(name = "count")
hist_data

# Check the maximum count
max(hist_data$count)

# Plot using ggplot
ggplot(hist_data, aes(x = User.ID, y = count)) +
  geom_col() +
  labs(x = "User ID", y = "Count", title = "Counts per User ID") +
  theme_minimal()

```



### Adding Age to dataset

```{r}
# Merge with user_info to include age 
full_data <- data %>%
  left_join(user_info, by = "User.ID")

boxplot(full_data$Age) # Age has some very large outliers
full_data <- full_data %>% filter(full_data$Age < 110) 
# data %>% filter(Age < 5)
```




```{r}
hist(data$Book.Rating, main = "Distribution of Jester Ratings",
     col = "yellow", xlab = "Ratings")
```





# EDA
...



# User-Based Collaborative Filtering (UBCF)


```{r}
# Create small sample for testing code
set.seed(123) 
sample_users <- sample(unique(book_ratings$User.ID), 5000)
sample_data <- data %>% filter(User.ID %in% sample_users)


# Convert 0 ratings to NA (unrated) 
book_ratings_clean <- sample_data %>%
  mutate(Book.Rating = ifelse(Book.Rating == 0, NA, Book.Rating))

# Convert to wide format 
user_item_matrix <- book_ratings_clean %>% 
  select(User.ID, ISBN, Book.Rating) %>%
  pivot_wider(names_from = ISBN, values_from = Book.Rating, values_fill = NA)

user_item_matrix

# Convert to matrix
user_ids <- user_item_matrix$User.ID
user_item_matrix <- as.matrix(user_item_matrix[, -1])
rownames(user_item_matrix) <- user_ids


# --------------------------
# Cosine Similarity Function 
# --------------------------

cosine_similarity <- function(a, b) {
  valid <- !is.na(a) & !is.na(b)  # TRUE only when both users rated the same book
  if (sum(valid) == 0) return(0)  # no books in common - similarity of 0 
  
  a <- a[valid]
  b <- b[valid]
  crossprod(a, b) / sqrt(crossprod(a) * crossprod(b))
}

# --------------------------
# Similarity matrix  
# --------------------------

n_users <- nrow(user_item_matrix)
user_similarity_matrix <- matrix(0, nrow = n_users, ncol = n_users)
rownames(user_similarity_matrix) <- rownames(user_item_matrix)
colnames(user_similarity_matrix) <- rownames(user_item_matrix)


for (i in 1:n_users) {
  for (j in i:n_users) {
    
    if (i == j) {
      user_similarity_matrix[i, j] <- 0  # user's similarity to themselves = 0 
    } else {
      sim <- cosine_similarity(user_item_matrix[i, ], user_item_matrix[j, ])
      user_similarity_matrix[i, j] <- sim
      user_similarity_matrix[j, i] <- sim
    }
  }
}

View(user_similarity_matrix)

# --------------------------
# Predictions 
# --------------------------

# Predict a single book rating for a user using standardized weighted similarities
predict_rating <- function(target_user, book_isbn, user_item_matrix, user_sim_matrix) {
  target_user <- as.character(target_user)
  if (!target_user %in% rownames(user_item_matrix) || 
      !book_isbn %in% colnames(user_item_matrix)) return(NA)
  
  sims <- user_sim_matrix[target_user, ]
  rated_users <- !is.na(user_item_matrix[, book_isbn])
  sims <- sims * rated_users
  
  if (sum(sims) == 0) return(NA)
  
  weights <- sims / sum(sims)
  sum(weights * user_item_matrix[, book_isbn], na.rm = TRUE)
}

# Predict top N unread books for a user
predict_user_ratings <- function(target_user, user_item_matrix, user_sim_matrix, book_info, n_recommendations = 10) {
  
  unread_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  
  preds <- sapply(unread_books, function(book) {
    predict_rating(target_user, book, user_item_matrix, user_sim_matrix)
  })
  
  top_books <- sort(preds, decreasing = TRUE)[1:min(n_recommendations, length(preds))]
  
  recommendations <- data.frame(ISBN = names(top_books), Predicted_Rating = top_books) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# --------------------------
# Explain Recommendations
# --------------------------
explain_recommendations <- function(target_user, user_item_matrix, user_sim_matrix, book_info, top_n_neighbors = 5) {
  
  target_user <- as.character(target_user)
  sims <- user_sim_matrix[target_user, ]
  
  neighbors <- data.frame(User.ID = names(sims), Similarity = sims) %>%
    filter(User.ID != target_user, Similarity > 0) %>%
    arrange(desc(Similarity)) %>%
    head(top_n_neighbors)
  
  cat("Top similar users to", target_user, ":\n")
  print(neighbors)
  
  cat("\nBooks highly rated by these users:\n")
  for (i in 1:nrow(neighbors)) {
    neighbor_id <- neighbors$User.ID[i]
    high_rated <- which(user_item_matrix[neighbor_id, ] >= 7 & !is.na(user_item_matrix[neighbor_id, ]))
    if (length(high_rated) > 0) {
      book_isbns <- colnames(user_item_matrix)[high_rated[1:min(5, length(high_rated))]]
      book_titles <- book_info$Book.Title[match(book_isbns, book_info$ISBN)]
      cat("\nUser", neighbor_id, "(similarity:", round(neighbors$Similarity[i], 3), "):\n")
      for (title in book_titles) cat(" -", title, "\n")
    }
  }
}

# --------------
# Example Usage
# ---------------

sample_user <- rownames(user_item_matrix)[1]
cat("\nTop 5 recommendations for User", sample_user, ":\n")
recs <- predict_user_ratings(sample_user, user_item_matrix, user_similarity_matrix, book_info, n_recommendations = 5)
print(recs)

# Explain why these books were recommended
explain_recommendations(sample_user, user_item_matrix, user_similarity_matrix, book_info)
```



```{r}
# How many book pairs do users share?
overlap_matrix <- matrix(0, nrow = nrow(user_item_matrix), ncol = nrow(user_item_matrix))
rownames(overlap_matrix) <- rownames(user_item_matrix)
colnames(overlap_matrix) <- rownames(user_item_matrix)

for (i in 1:nrow(user_item_matrix)) {
  for (j in 1:nrow(user_item_matrix)) {
    overlap_matrix[i, j] <- sum(!is.na(user_item_matrix[i, ]) & !is.na(user_item_matrix[j, ]))
  }
}

# Check overlap distribution
cat("Overlap statistics (excluding diagonal):\n")
diag(overlap_matrix) <- NA
summary(as.vector(overlap_matrix))
hist(overlap_matrix, main = "Number of Books Rated by Both Users", 
     xlab = "Overlap Count", breaks = 20)
```


```{r}
# Check user 276925 and their most similar user (638)
target_user <- "276925"
similar_user <- "638"

# Get their ratings
target_ratings <- user_item_matrix[target_user, ]
similar_ratings <- user_item_matrix[similar_user, ]

# Find books both rated
common_books <- !is.na(target_ratings) & !is.na(similar_ratings)

cat("Books both users rated:", sum(common_books), "\n\n")

if (sum(common_books) > 0) {
  cat("Common book ratings:\n")
  comparison <- data.frame(
    ISBN = colnames(user_item_matrix)[common_books],
    User_276925 = target_ratings[common_books],
    User_638 = similar_ratings[common_books]
  )
  print(comparison)
  
  cat("\nAre all ratings identical?", all(comparison$User_276925 == comparison$User_638), "\n")
}
```




# Item-Based Collaborative Filtering (IBCF)


















