---
title: "Ensemble Recommender System for Book Recommendations: A Comparative Analysis of Collaborative Filtering Approaches"
author: "Hope Hennessy"
date: "2025-10-01"
output:
  pdf_document:
    latex_engine: xelatex
---

\newpage

# Abstract

This study presents a comprehensive comparative analysis of four collaborative filtering approaches for book recommendation systems using a modified Book-Crossing dataset. We implement item-based collaborative filtering, user-based collaborative filtering, matrix factorization, and neural network-based methods to build an ensemble recommender system. Our analysis includes cross-validation performance evaluation, cold start problem handling, and investigation of dataset size effects on predictive accuracy. The results demonstrate the relative strengths and limitations of each approach, providing insights for practical recommendation system deployment.

## Assignment Overview

Build an ensemble recommender system for book recommendations using a modified "Book-Crossing" dataset containing ratings (0-10 scale) from 10,000 users on 145 books.

### Core Requirements

1. **Build Four Types of Recommender Systems:**
   - Item-based collaborative filtering (code from scratch)
   - User-based collaborative filtering (code from scratch)
   - Matrix factorization-based collaborative filtering
   - Neural network-based collaborative filtering

2. **System Capabilities:**
   - Recommend books to existing users
   - Handle new users (assuming they provide ratings for ≤5 books initially)

3. **Evaluation and Analysis:**
   - Compare accuracy across all four methods using cross-validation
   - Investigate the relationship between dataset size and accuracy
   - Determine if there's a point where adding more titles doesn't improve accuracy

4. **Data Analysis:**
   - Conduct exploratory data analysis (EDA)
   - Use findings to inform train/test data splitting

# 1. Introduction




```{r setup, include=FALSE}
# ============================================================================
# GLOBAL SETTINGS AND CONFIGURATION
# ============================================================================

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6,
  fig.show = "hold",
  out.width = "80%",
  dpi = 300,
  crop = NULL
)

# Set random seed for reproducibility
set.seed(123)

# ============================================================================
# LIBRARY IMPORTS
# ============================================================================

library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(cowplot)

library(knitr)
library(kableExtra)

library(caret)
library(recosystem)
library(h2o)

options(
  lintr.linter_file = "none",
  lintr.exclude_linters = c(
    "object_usage_linter",
    "object_name_linter",
    "cyclocomp_linter",
    "line_length_linter"
  )
)


# ============================================================================
# HELPER FUNCTIONS FOR TABLE FORMATTING
# ============================================================================

# Function to truncate long text for table display
truncate_text <- function(text, max_length = 50) {
  ifelse(nchar(text) > max_length, 
         paste0(substr(text, 1, max_length - 3), "..."), 
         text)
}

# Function to format recommendation tables consistently
format_recommendation_table <- function(data, caption, method_name) {
  data %>%
    mutate(Rank = row_number()) %>%
    mutate(
      Book.Title = truncate_text(Book.Title, 50),
      Book.Author = truncate_text(Book.Author, 25)
    ) %>%
    select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
    kable(
      format = "latex",
      booktabs = TRUE,
      caption = paste(method_name, caption),
      digits = 2,
      align = c("c", "l", "l", "c"),
      col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
    ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "scale_down"),
      font_size = 10
    )
}

# Load saved results
load('nn_tuning_results.Rdata')
load('mf_tuning_results.Rdata')
load('mf_final_model_mf.Rdata')
load('mf_tuning_results.Rdata')
load('all_cv_results.Rdata')
load('dataset_size_analysis_complete.Rdata')

```

# Data Preparation and Exploratory Analysis

## Data Loading and Initial Inspection

```{r data-loading, include=FALSE}
# ============================================================================
# DATA LOADING
# ============================================================================

load("book_ratings.Rdata")

head(book_info)
str(book_info)
dim(book_info)
head(book_ratings)
str(book_ratings)
dim(book_ratings)
head(user_info) # will not be using age in this report
str(user_info)
dim(user_info)

# Check for missing values
missing_summary <- data.frame(
  Dataset = c("Book Info", "Book Ratings", "User Info"),
  Missing_Values = c(
    sum(is.na(book_info)),
    sum(is.na(book_ratings)),
    sum(is.na(user_info))   # missing age values
  )
)

```

## Data Cleaning and Duplicate Handling

```{r data-cleaning}
# ============================================================================
# DUPLICATE DETECTION AND HANDLING
# ============================================================================

# Initial data merging
initial_data <- book_ratings %>%
  left_join(book_info, by = "ISBN")

# Clean book titles and authors for duplicate detection
book_info_clean <- book_info %>%
  mutate(
    clean_title = Book.Title %>%
      tolower() %>%
      gsub('\\(.*?\\)', '', .) %>%
      gsub('\\s*:\\s*.*$', '', .) %>%
      gsub('^(the|a|an)\\s+', '', .) %>%
      gsub('[^a-z0-9\\s-]', '', .) %>%
      gsub('\\s+', ' ', .) %>%
      trimws(),
    clean_author = Book.Author %>%
      tolower() %>%
      gsub('\\s+jr\\.*$', '', .) %>%
      gsub('\\s+sr\\.*$', '', .) %>%
      gsub('\\s+(iii|ii)$', '', .) %>%
      gsub('\\s+', ' ', .) %>%
      trimws()
  )

# Identify duplicates
duplicates <- book_info_clean %>%
  group_by(clean_title, clean_author) %>%
  filter(n() > 1) %>%
  arrange(clean_title, clean_author, ISBN) %>%
  ungroup()

# ============================================================================
# MERGE RATINGS FUNCTION
# ============================================================================

merge_ratings <- function(ratings_data, book_info_data, duplicates_data) {
  
  if (nrow(duplicates_data) > 0) {
    # Create canonical ISBN mapping
    duplicate_mapping <- duplicates_data %>%
      group_by(clean_title, clean_author) %>%
      arrange(ISBN) %>%
      mutate(canonical_ISBN = first(ISBN)) %>%
      ungroup() %>%
      select(ISBN, canonical_ISBN)
    
    # Create complete mapping
    all_books_mapping <- book_info_data %>%
      select(ISBN) %>%
      left_join(duplicate_mapping, by = "ISBN") %>%
      mutate(canonical_ISBN = ifelse(is.na(canonical_ISBN), ISBN, canonical_ISBN))
    
    # Apply mapping to ratings
    ratings_mapped <- ratings_data %>%
      left_join(all_books_mapping, by = "ISBN") %>%
      mutate(
        original_ISBN = ISBN,
        ISBN = canonical_ISBN
      ) %>%
      select(-canonical_ISBN, -original_ISBN)
    
    # Keep highest rating per user-book combination
    ratings_merged <- ratings_mapped %>%
      group_by(User.ID, ISBN) %>%
      slice_max(Book.Rating, n = 1, with_ties = FALSE) %>%
      ungroup()
    
    # Update book info with cleaned titles and authors
    book_info_updated <- book_info_data %>%
      left_join(all_books_mapping, by = "ISBN") %>%
      filter(ISBN == canonical_ISBN) %>%
      mutate(
        Book.Title = Book.Title %>%
          gsub('\\(.*?\\)', '', .) %>%  # Remove parenthetical info
          gsub('\\s*:\\s*.*$', '', .) %>%  # Remove subtitle after colon
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws(),  # Remove leading/trailing whitespace
        Book.Author = Book.Author %>%
          gsub('\\s+jr\\.*$', '', .) %>%  # Remove "Jr." suffixes
          gsub('\\s+sr\\.*$', '', .) %>%  # Remove "Sr." suffixes
          gsub('\\s+(iii|ii)$', '', .) %>%  # Remove "II", "III" suffixes
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws()  # Remove leading/trailing whitespace
      ) %>%
      select(ISBN, Book.Title, Book.Author)
    
  } else {
    ratings_merged <- ratings_data
    # Clean all book titles and authors even when no duplicates
    book_info_updated <- book_info_data %>%
      mutate(
        Book.Title = Book.Title %>%
          gsub('\\(.*?\\)', '', .) %>%  # Remove parenthetical info
          gsub('\\s*:\\s*.*$', '', .) %>%  # Remove subtitle after colon
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws(),  # Remove leading/trailing whitespace
        Book.Author = Book.Author %>%
          gsub('\\s+jr\\.*$', '', .) %>%  # Remove "Jr." suffixes
          gsub('\\s+sr\\.*$', '', .) %>%  # Remove "Sr." suffixes
          gsub('\\s+(iii|ii)$', '', .) %>%  # Remove "II", "III" suffixes
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws()  # Remove leading/trailing whitespace
      ) %>%
      select(ISBN, Book.Title, Book.Author)
    all_books_mapping <- NULL
  }
  
  return(list(
    ratings_merged = ratings_merged,
    book_info_updated = book_info_updated,
    mapping = all_books_mapping
  ))
}

# Execute merge
merged_result <- merge_ratings(book_ratings, book_info, duplicates)

# Create final clean dataset
data <- merged_result$ratings_merged %>%
  left_join(merged_result$book_info_updated, by = "ISBN") %>%
  select(User.ID, ISBN, Book.Rating, Book.Title, Book.Author)

book_info <- merged_result$book_info_updated
```


Duplicate book entries (same title and author with different ISBNs) were identified through normalized string matching and merged by mapping all variants to a canonical ISBN. 
When a user had rated multiple versions of the same book, the highest rating was retained.




## Exploratory Data Analysis

```{r eda-summary}
# ============================================================================
# DATASET SUMMARY STATISTICS
# ============================================================================

# Basic statistics
data_summary <- data.frame(
  Metric = c(
    "Total Users",
    "Total Books",
    "Total Ratings",
    "Average Rating",
    "Rating Range",
    "Matrix Sparsity"
  ),
  Value = c(
    length(unique(data$User.ID)),
    length(unique(data$ISBN)),
    nrow(data),
    round(mean(data$Book.Rating, na.rm = TRUE), 2),
    paste(min(data$Book.Rating), "-", max(data$Book.Rating)),
    paste0(
      round(
        (1 - nrow(data) / 
          (length(unique(data$User.ID)) * length(unique(data$ISBN)))) * 100,
        2
      ),
      "%"
    )
  )
)

kable(data_summary,
      format = "latex",
      booktabs = TRUE,
      caption = "Dataset Overview",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

```




```{r eda-visualizations, fig.cap="Key Dataset Characteristics for Recommender Systems. Panel A: Distribution of explicit ratings (1-10 scale) showing user rating patterns, with red dashed line indicating the mean rating. Panel B: User activity distribution showing the sparsity challenge in collaborative filtering, where most users interact with few books (long tail) while few users are highly active. Panel C: Relationship between book popularity and average rating quality, with red line showing the trend to understand rating biases and inform cold start problem solutions. Panel D: Proportion of explicit vs implicit feedback in the dataset, where explicit ratings (1-10) are used for model training while implicit ratings (0, read but not rated) are used with confidence weighting."}

  
# ============================================================================
# FOCUSED VISUALIZATIONS: MOST INFORMATIVE FOR RECOMMENDER SYSTEMS
# ============================================================================

# 1. EXPLICIT RATING DISTRIBUTION (Most Important)
explicit_data <- data %>% filter(Book.Rating > 0)

p1 <- ggplot(explicit_data, aes(x = Book.Rating)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(explicit_data$Book.Rating), 
             linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = mean(explicit_data$Book.Rating), y = max(table(explicit_data$Book.Rating)) * 0.9, 
           label = paste("Mean =", round(mean(explicit_data$Book.Rating), 1)), 
           color = "red", hjust = -0.1, size = 3.5) +
  labs(
    title = "Distribution of Explicit Ratings",
    x = "Rating",
    y = "Frequency"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 2. SPARSITY ANALYSIS (Critical for CF)
sparsity_data <- data %>%
  group_by(User.ID) %>%
  summarise(
    total_books_rated = n(),
    explicit_ratings = sum(Book.Rating > 0),
    implicit_ratings = sum(Book.Rating == 0),
    .groups = "drop"
  ) %>%
  mutate(
    explicit_ratio = explicit_ratings / total_books_rated,
    user_type = case_when(
      explicit_ratio == 1 ~ "Only Explicit",
      explicit_ratio == 0 ~ "Only Implicit", 
      explicit_ratio > 0.5 ~ "Mostly Explicit",
      TRUE ~ "Mostly Implicit"
    )
  )

p2 <- ggplot(sparsity_data, aes(x = total_books_rated)) +
  geom_histogram(binwidth = 1, fill = "lightcoral", color = "white", alpha = 0.8) +
  scale_x_continuous(limits = c(0, quantile(sparsity_data$total_books_rated, 0.95))) +
  labs(
    title = "User Activity Distribution",
    x = "Total Books per User",
    y = "Number of Users"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 3. BOOK POPULARITY (Important for Cold Start)
book_popularity <- data %>%
  group_by(ISBN) %>%
  summarise(
    total_interactions = n(),
    explicit_ratings = sum(Book.Rating > 0),
    avg_rating = mean(Book.Rating[Book.Rating > 0], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(avg_rating))  # Only books with explicit ratings

p3 <- ggplot(book_popularity, aes(x = explicit_ratings, y = avg_rating)) +
  geom_point(alpha = 0.6, color = "darkgreen", size = 2) +
  labs(
    title = "Book Popularity vs Average Rating",
    x = "Number of Explicit Ratings",
    y = "Average Rating"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 4. EXPLICIT VS IMPLICIT RATIO (Key Insight)
interaction_summary <- data %>%
  summarise(
    total_interactions = n(),
    explicit_count = sum(Book.Rating > 0),
    implicit_count = sum(Book.Rating == 0),
    explicit_pct = round(mean(Book.Rating > 0) * 100, 1),
    implicit_pct = round(mean(Book.Rating == 0) * 100, 1)
  )

# Create a simple bar chart for the ratio
ratio_data <- data.frame(
  Type = c("Explicit Ratings", "Implicit Ratings"),
  Count = c(interaction_summary$explicit_count, interaction_summary$implicit_count),
  Percentage = c(interaction_summary$explicit_pct, interaction_summary$implicit_pct)
)

p4 <- ggplot(ratio_data, aes(x = Type, y = Count, fill = Type)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(Percentage, "%")), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Explicit Ratings" = "steelblue", "Implicit Ratings" = "lightcoral")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Explicit vs Implicit Feedback",
    x = "Feedback Type",
    y = "Count"
  ) +
  theme_cowplot() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )

# Create the combined plot with legend
combined_plot <- plot_grid(
  p1, p2,
  p3, p4,
  ncol = 2, nrow = 2,
  labels = c("A", "B", "C", "D"),
  label_size = 12
)

# Display the plot
combined_plot
```

The dataset exhibits characteristics typical of recommendation system challenges (Figure 1). 
The distribution of explicit ratings (Panel A) shows a right-skewed pattern with a mean of 7.5, indicating users tend to rate books they enjoy rather than books they dislike. 
User activity follows a long-tail distribution (Panel B), where the majority of users have rated fewer than 10 books while a small subset of highly active users have rated 20 or more—highlighting the sparsity problem central to collaborative filtering. 
Book popularity analysis (Panel C) reveals no strong correlation between the number of ratings a book receives and its average rating quality, suggesting popularity bias is not a major concern for this dataset. 
Finally, the dataset contains both explicit ratings (1-10 scale, 65.2%) and implicit feedback (rating of 0 indicating "read but not rated", 34.8%) as shown in Panel D. 
To handle this mixed feedback, we implement confidence weighting in our similarity calculations, assigning lower confidence to implicit ratings while still leveraging them to reduce sparsity. 
After applying minimum rating thresholds, 58% of users were excluded for having fewer than 3 ratings, while all 145 books in the dataset met the minimum threshold of 5 ratings per book, resulting in a final dataset with 99.2% sparsity across 4,200 users and 145 books.




"After applying minimum rating thresholds, 58% of users were excluded for having fewer than 3 ratings, while all 145 books in the dataset met the minimum threshold of 5 ratings per book."



# Methodology

## User-Item Matrix Construction

```{r matrix-construction}
# ============================================================================
# USER-ITEM MATRIX CREATION FUNCTION
# ============================================================================

create_user_item_matrix <- function(ratings_data,
                                     min_ratings_per_book = 3,
                                     min_ratings_per_user = 2,
                                     implicit_rating = 4,
                                     implicit_confidence = 0.5) {
  
  # Handle implicit ratings (rating = 0)
  ratings_clean <- ratings_data %>%
    mutate(
      is_implicit = (Book.Rating == 0),
      original_rating = Book.Rating,
      Book.Rating = ifelse(Book.Rating == 0, implicit_rating, Book.Rating),
      confidence = ifelse(is_implicit, implicit_confidence, 1.0)
    )
  
  # Create matrices
  user_item_matrix <- ratings_clean %>%
    select(User.ID, ISBN, Book.Rating) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = Book.Rating,
      values_fill = NA
    )
  
  confidence_matrix <- ratings_clean %>%
    select(User.ID, ISBN, confidence) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = confidence,
      values_fill = NA
    )
  
  original_ratings_matrix <- ratings_clean %>%
    select(User.ID, ISBN, original_rating) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = original_rating,
      values_fill = NA
    )
  
  # Convert to matrix format
  user_ids <- user_item_matrix$User.ID
  user_item_matrix <- as.matrix(user_item_matrix[, -1])
  confidence_matrix <- as.matrix(confidence_matrix[, -1])
  original_ratings_matrix <- as.matrix(original_ratings_matrix[, -1])
  
  rownames(user_item_matrix) <- user_ids
  rownames(confidence_matrix) <- user_ids
  rownames(original_ratings_matrix) <- user_ids
  
  # Filter by minimum ratings
  books_to_keep <- colSums(!is.na(user_item_matrix)) >= min_ratings_per_book
  user_item_matrix <- user_item_matrix[, books_to_keep]
  confidence_matrix <- confidence_matrix[, books_to_keep]
  original_ratings_matrix <- original_ratings_matrix[, books_to_keep]
  
  users_to_keep <- rowSums(!is.na(user_item_matrix)) >= min_ratings_per_user
  user_item_matrix <- user_item_matrix[users_to_keep, ]
  confidence_matrix <- confidence_matrix[users_to_keep, ]
  original_ratings_matrix <- original_ratings_matrix[users_to_keep, ]
  
  return(list(
    ratings = user_item_matrix,
    confidence = confidence_matrix,
    original_ratings = original_ratings_matrix
  ))
}

# ============================================================================
# EVALUATION FUNCTION
# ============================================================================

evaluate_predictions <- function(pred, actual, original_ratings) {
  # Separate explicit and implicit ratings
  is_implicit <- (original_ratings == 0)
  explicit_mask <- !is_implicit & !is.na(actual)
  
  # Calculate RMSE on explicit ratings only
  rmse_explicit <- sqrt(mean((pred[explicit_mask] - actual[explicit_mask])^2, na.rm = TRUE))
  rmse_overall <- sqrt(mean((pred - actual)^2, na.rm = TRUE))
  
  return(list(
    rmse_explicit = rmse_explicit,
    rmse_overall = rmse_overall,
    n_explicit = sum(explicit_mask),
    n_implicit = sum(is_implicit)
  ))
}

# ============================================================================
# UNIFIED MATRIX CREATION FOR ALL METHODS
# ============================================================================

# Create matrices once with consistent parameters
matrix_result <- create_user_item_matrix(
  data,
  min_ratings_per_book = 5,
  min_ratings_per_user = 3,
  implicit_rating = 4,
  implicit_confidence = 0.5
)

user_item_matrix <- matrix_result$ratings
confidence_matrix <- matrix_result$confidence
original_ratings_matrix <- matrix_result$original_ratings

```


## Common Functions for All Methods

```{r common-functions}
# ============================================================================
# NORMALIZATION FUNCTIONS
# ============================================================================

# User-mean normalization (for UBCF and NN)
normalize_matrix_user <- function(user_item_matrix) {
  user_means <- rowMeans(user_item_matrix, na.rm = TRUE)
  user_item_matrix_normalized <- sweep(user_item_matrix, 1, user_means, FUN = "-")
  
  return(list(
    normalized = user_item_matrix_normalized,
    means = user_means
  ))
}

# Item-mean normalization (for IBCF)
normalize_matrix_item <- function(user_item_matrix) {
  item_means <- colMeans(user_item_matrix, na.rm = TRUE)
  user_item_matrix_normalized <- sweep(user_item_matrix, 2, item_means, FUN = "-")
  
  return(list(
    normalized = user_item_matrix_normalized,
    means = item_means
  ))
}
```



# Collaborative Filtering Implementations

## User-Based Collaborative Filtering (UBCF)

### Similarity Computation

```{r ubcf-similarity}
# ============================================================================
# UBCF: USER SIMILARITY COMPUTATION
# ============================================================================

compute_user_similarity_matrix <- function(user_item_matrix_normalized,
                                            confidence_matrix) {
  
  # Prepare matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  
  # Weighted ratings
  weighted_mat <- mat * conf
  
  # Compute cosine similarity
  numerator <- weighted_mat %*% t(weighted_mat)
  magnitudes <- sqrt(rowSums(weighted_mat^2))
  denominator <- outer(magnitudes, magnitudes)
  
  user_similarity_matrix <- numerator / denominator
  user_similarity_matrix[is.nan(user_similarity_matrix)] <- 0
  diag(user_similarity_matrix) <- 0
  
  rownames(user_similarity_matrix) <- rownames(user_item_matrix_normalized)
  colnames(user_similarity_matrix) <- rownames(user_item_matrix_normalized)
  
  return(user_similarity_matrix)
}
```

### Recommendation Functions

```{r ubcf-recommendations}
# ============================================================================
# UBCF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_ubcf <- function(target_user,
                                     user_item_matrix,
                                     user_item_matrix_normalized,
                                     confidence_matrix,
                                     user_sim_matrix,
                                     user_means,
                                     book_info,
                                     n_recommendations = 10,
                                     k = NULL) {
  
  target_user <- as.character(target_user)
  
  # Find unrated books
  unrated_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get similarities
  sims <- user_sim_matrix[target_user, ]
  sims[is.na(sims) | is.nan(sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < length(sims)) {
    non_zero_count <- sum(sims != 0)
    if (non_zero_count > 0) {
      k_actual <- min(k, non_zero_count)
      top_k_users <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
      sims_filtered <- rep(0, length(sims))
      names(sims_filtered) <- names(sims)
      sims_filtered[top_k_users] <- sims[top_k_users]
      sims <- sims_filtered
    }
  }
  
  if (sum(abs(sims) > 0) == 0) return(data.frame())
  
  # Prepare weighted matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  weighted_mat <- mat * conf
  
  # Calculate predictions
  weighted_ratings <- t(weighted_mat[, unrated_books, drop = FALSE]) %*% sims
  
  rated_mask <- !is.na(confidence_matrix[, unrated_books, drop = FALSE])
  conf_weighted_sims <- sweep(
    conf[, unrated_books, drop = FALSE] * rated_mask,
    1,
    abs(sims),
    "*"
  )
  sum_weighted_sims <- colSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + user_means[target_user]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Get top recommendations
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  top_books <- sort(preds_valid, decreasing = TRUE)[1:min(n_recommendations, length(preds_valid))]
  
  recommendations <- data.frame(
    ISBN = names(top_books),
    Predicted_Rating = as.numeric(top_books)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# UBCF: RECOMMENDATION FOR NEW USERS (COLD START)
# ============================================================================

recommend_for_new_user_ubcf <- function(new_user_ratings,
                                         user_item_matrix,
                                         user_item_matrix_normalized,
                                         confidence_matrix,
                                         user_means,
                                         book_info,
                                         n_recommendations = 10,
                                         k = NULL) {
  
  if (length(new_user_ratings) == 0) return(data.frame())
  
  # Create new user vector
  new_user_vector <- rep(NA, ncol(user_item_matrix))
  names(new_user_vector) <- colnames(user_item_matrix)
  
  matched_books <- intersect(names(new_user_ratings), names(new_user_vector))
  if (length(matched_books) == 0) return(data.frame())
  
  new_user_vector[matched_books] <- new_user_ratings[matched_books]
  
  # Normalize
  new_user_mean <- mean(new_user_vector, na.rm = TRUE)
  new_user_normalized <- new_user_vector - new_user_mean
  new_user_normalized[is.na(new_user_normalized)] <- 0
  new_user_conf <- ifelse(is.na(new_user_vector), 0, 1)
  new_user_weighted <- new_user_normalized * new_user_conf
  
  # Prepare matrices
  mat_normalized <- user_item_matrix_normalized
  mat_normalized[is.na(mat_normalized)] <- 0
  conf <- confidence_matrix
  conf[is.na(conf)] <- 0
  mat_weighted <- mat_normalized * conf
  
  # Compute similarities
  existing_magnitudes <- sqrt(rowSums(mat_weighted^2))
  new_user_magnitude <- sqrt(sum(new_user_weighted^2))
  
  if (new_user_magnitude == 0) return(data.frame())
  
  new_user_sims <- as.vector(
    (mat_weighted %*% new_user_weighted) /
      (existing_magnitudes * new_user_magnitude)
  )
  new_user_sims[!is.finite(new_user_sims)] <- 0
  names(new_user_sims) <- rownames(user_item_matrix_normalized)
  
  # Apply k-NN
  if (!is.null(k) && k > 0 && k < length(new_user_sims)) {
    threshold <- sort(abs(new_user_sims), decreasing = TRUE)[min(k, sum(new_user_sims != 0))]
    new_user_sims[abs(new_user_sims) < threshold] <- 0
  }
  
  if (all(new_user_sims == 0)) return(data.frame())
  
  # Make predictions
  unrated_books <- names(new_user_vector)[is.na(new_user_vector)]
  if (length(unrated_books) == 0) return(data.frame())
  
  weighted_ratings <- as.vector(t(mat_weighted[, unrated_books, drop = FALSE]) %*% new_user_sims)
  
  rated_mask <- !is.na(user_item_matrix[, unrated_books, drop = FALSE])
  conf_for_unrated <- conf[, unrated_books, drop = FALSE] * rated_mask
  sum_weighted_sims <- as.vector(t(conf_for_unrated) %*% abs(new_user_sims))
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- (weighted_ratings / sum_weighted_sims) + new_user_mean
  preds <- pmin(pmax(preds, 1), 10)
  names(preds) <- unrated_books
  
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  n_actual <- min(n_recommendations, length(preds_valid))
  top_indices <- order(preds_valid, decreasing = TRUE)[1:n_actual]
  
  recommendations <- data.frame(
    ISBN = names(preds_valid)[top_indices],
    Predicted_Rating = round(preds_valid[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```



### Cross-Validation

```{r ubcf-cv}
# ============================================================================
# UBCF: CROSS-VALIDATION
# ============================================================================

cross_validate_ubcf <- function(user_item_matrix,
                                 confidence_matrix,
                                 original_ratings_matrix,
                                 n_folds = 5,
                                 k = 30) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    test_obs <- observed[test_indices, , drop = FALSE]
    
    # Create training matrices
    train_matrix <- user_item_matrix
    train_confidence <- confidence_matrix
    train_matrix[test_obs] <- NA
    train_confidence[test_obs] <- NA
    
    # Normalize and compute similarity
    norm_result <- normalize_matrix_user(train_matrix)
    train_normalized <- norm_result$normalized
    user_means <- norm_result$means
    user_sim_matrix <- compute_user_similarity_matrix(train_normalized, train_confidence)
    
    # Prepare for prediction
    train_norm_filled <- train_normalized
    train_norm_filled[is.na(train_norm_filled)] <- 0
    train_conf_filled <- train_confidence
    train_conf_filled[is.na(train_conf_filled)] <- 0
    
    # Make predictions
    predictions <- numeric(nrow(test_obs))
    
    for (i in 1:nrow(test_obs)) {
      user_idx <- test_obs[i, 1]
      item_idx <- test_obs[i, 2]
      
      sims <- user_sim_matrix[user_idx, ]
      sims[is.na(sims) | is.nan(sims)] <- 0
      
      # Apply k-NN
      if (k < length(sims) && sum(abs(sims) > 0) > 0) {
        k_actual <- min(k, sum(abs(sims) > 0))
        top_k_idx <- order(abs(sims), decreasing = TRUE)[1:k_actual]
        sims_filtered <- rep(0, length(sims))
        sims_filtered[top_k_idx] <- sims[top_k_idx]
        sims <- sims_filtered
      }
      
      rated_mask <- !is.na(train_matrix[, item_idx])
      
      if (sum(abs(sims[rated_mask])) == 0) {
        predictions[i] <- user_means[user_idx]
        next
      }
      
      # Confidence-weighted prediction
      weighted_ratings <- train_norm_filled[rated_mask, item_idx] *
        train_conf_filled[rated_mask, item_idx]
      weighted_sims <- abs(sims[rated_mask]) *
        train_conf_filled[rated_mask, item_idx]
      
      weighted_sum <- sum(sims[rated_mask] * weighted_ratings)
      sum_weighted_sims <- sum(weighted_sims)
      
      if (sum_weighted_sims > 0) {
        predictions[i] <- (weighted_sum / sum_weighted_sims) + user_means[user_idx]
      } else {
        predictions[i] <- user_means[user_idx]
      }
    }
    
    predictions <- pmin(pmax(predictions, 1), 10)
    actual <- user_item_matrix[test_obs]
    original_ratings <- original_ratings_matrix[test_obs]
    metrics <- evaluate_predictions(predictions, actual, original_ratings)
    
    cv_results[fold, ] <- list(fold, metrics$rmse_explicit)
  }
  
  return(cv_results)
}
```

### Implementation

```{r ubcf-implementation}
# ============================================================================
# UBCF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Normalize and compute similarity
normalized_result_ubcf <- normalize_matrix_user(user_item_matrix)
user_item_matrix_normalized_ubcf <- normalized_result_ubcf$normalized
user_means_ubcf <- normalized_result_ubcf$means

user_similarity_matrix <- compute_user_similarity_matrix(
  user_item_matrix_normalized_ubcf,
  confidence_matrix
)

# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ubcf <- recommend_for_user_ubcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_sim_matrix = user_similarity_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

recs_ubcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "UBCF: Top 10 Recommendations for Existing User",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Generate recommendations for new user (cold start)
sample_books <- colnames(user_item_matrix)[1:5]
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), sample_books)

new_user_recs_ubcf <- recommend_for_new_user_ubcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

new_user_recs_ubcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "UBCF: Top 10 Recommendations for New User (Cold Start)",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )


```



## Item-Based Collaborative Filtering (IBCF)

### Similarity Computation

```{r ibcf-similarity}
# ============================================================================
# IBCF: ITEM SIMILARITY COMPUTATION
# ============================================================================

compute_item_similarity_matrix <- function(user_item_matrix_normalized,
                                            confidence_matrix) {
  
  # Prepare matrices
  mat <- user_item_matrix_normalized
  mat[is.na(mat)] <- 0
  conf <- confidence_matrix
  conf[is.na(conf)] <- 0
  
  # Transpose for item-item similarity
  mat_t <- t(mat)
  conf_t <- t(conf)
  
  # Weighted ratings
  weighted_mat_t <- mat_t * conf_t
  
  # Compute cosine similarity
  numerator <- weighted_mat_t %*% t(weighted_mat_t)
  magnitudes <- sqrt(rowSums(weighted_mat_t^2))
  denominator <- outer(magnitudes, magnitudes)
  
  item_similarity_matrix <- numerator / denominator
  item_similarity_matrix[is.nan(item_similarity_matrix)] <- 0
  diag(item_similarity_matrix) <- 0
  
  rownames(item_similarity_matrix) <- colnames(user_item_matrix_normalized)
  colnames(item_similarity_matrix) <- colnames(user_item_matrix_normalized)
  
  return(item_similarity_matrix)
}
```

### Recommendation Functions

```{r ibcf-recommendations}
# ============================================================================
# IBCF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_ibcf <- function(target_user,
                                     user_item_matrix,
                                     user_item_matrix_normalized,
                                     confidence_matrix,
                                     item_sim_matrix,
                                     item_means,
                                     book_info,
                                     n_recommendations = 10,
                                     k = NULL) {
  
  target_user <- as.character(target_user)
  
  # Find unrated books
  unrated_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get item similarities
  item_sims <- item_sim_matrix[unrated_books, , drop = FALSE]
  item_sims[is.na(item_sims) | is.nan(item_sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < ncol(item_sims)) {
    for (i in 1:nrow(item_sims)) {
      sims <- item_sims[i, ]
      non_zero_count <- sum(sims != 0)
      if (non_zero_count > 0) {
        k_actual <- min(k, non_zero_count)
        top_k_items <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
        sims_filtered <- rep(0, length(sims))
        names(sims_filtered) <- names(sims)
        sims_filtered[top_k_items] <- sims[top_k_items]
        item_sims[i, ] <- sims_filtered
      }
    }
  }
  
  if (sum(abs(item_sims) > 0) == 0) return(data.frame())
  
  # Prepare weighted matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  weighted_mat <- mat * conf
  
  # Calculate predictions
  weighted_ratings <- item_sims %*% weighted_mat[target_user, ]
  
  rated_mask <- !is.na(user_item_matrix[target_user, ])
  rated_item_mask <- matrix(
    rated_mask,
    nrow = nrow(item_sims),
    ncol = length(rated_mask),
    byrow = TRUE
  )
  
  conf_weighted_sims <- rated_item_mask * abs(item_sims) * conf[target_user, ]
  sum_weighted_sims <- rowSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + item_means[unrated_books]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Get top recommendations
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  top_books <- sort(preds_valid, decreasing = TRUE)[1:min(n_recommendations, length(preds_valid))]
  
  recommendations <- data.frame(
    ISBN = names(top_books),
    Predicted_Rating = as.numeric(top_books)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# IBCF: RECOMMENDATION FOR NEW USERS (COLD START)
# ============================================================================

recommend_for_new_user_ibcf <- function(new_user_ratings,
                                         user_item_matrix,
                                         user_item_matrix_normalized,
                                         confidence_matrix,
                                         item_means,
                                         item_sim_matrix,
                                         book_info,
                                         n_recommendations = 10,
                                         k = NULL) {
  
  if (length(new_user_ratings) == 0) return(data.frame())
  
  # Create new user vector
  new_user_vector <- rep(NA, ncol(user_item_matrix))
  names(new_user_vector) <- colnames(user_item_matrix)
  
  matched_books <- intersect(names(new_user_ratings), names(new_user_vector))
  if (length(matched_books) == 0) return(data.frame())
  
  new_user_vector[matched_books] <- new_user_ratings[matched_books]
  
  # Normalize
  new_user_normalized <- new_user_vector - item_means
  new_user_normalized[is.na(new_user_normalized)] <- 0
  new_user_conf <- ifelse(is.na(new_user_vector), 0, 1)
  new_user_weighted <- new_user_normalized * new_user_conf
  
  # Find unrated books
  unrated_books <- names(new_user_vector)[is.na(new_user_vector)]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get item similarities
  item_sims <- item_sim_matrix[unrated_books, , drop = FALSE]
  item_sims[is.na(item_sims) | is.nan(item_sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < ncol(item_sims)) {
    for (i in 1:nrow(item_sims)) {
      sims <- item_sims[i, ]
      non_zero_count <- sum(sims != 0)
      if (non_zero_count > 0) {
        k_actual <- min(k, non_zero_count)
        top_k_items <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
        sims_filtered <- rep(0, length(sims))
        names(sims_filtered) <- names(sims)
        sims_filtered[top_k_items] <- sims[top_k_items]
        item_sims[i, ] <- sims_filtered
      }
    }
  }
  
  if (sum(abs(item_sims) > 0) == 0) return(data.frame())
  
  # Make predictions
  weighted_ratings <- item_sims %*% new_user_weighted
  
  rated_mask <- !is.na(new_user_vector)
  rated_item_mask <- matrix(
    rated_mask,
    nrow = nrow(item_sims),
    ncol = length(rated_mask),
    byrow = TRUE
  )
  
  conf_weighted_sims <- rated_item_mask * abs(item_sims) * new_user_conf
  sum_weighted_sims <- rowSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + item_means[unrated_books]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  n_actual <- min(n_recommendations, length(preds_valid))
  top_indices <- order(preds_valid, decreasing = TRUE)[1:n_actual]
  
  recommendations <- data.frame(
    ISBN = names(preds_valid)[top_indices],
    Predicted_Rating = round(preds_valid[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```

### Cross-Validation

```{r ibcf-cv}
# ============================================================================
# IBCF: CROSS-VALIDATION
# ============================================================================

cross_validate_ibcf <- function(user_item_matrix,
                                 confidence_matrix,
                                 original_ratings_matrix,
                                 n_folds = 5,
                                 k = 30) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    test_obs <- observed[test_indices, , drop = FALSE]
    
    # Create training matrices
    train_matrix <- user_item_matrix
    train_confidence <- confidence_matrix
    train_matrix[test_obs] <- NA
    train_confidence[test_obs] <- NA
    
    # Normalize and compute similarity
    norm_result <- normalize_matrix_item(train_matrix)
    train_normalized <- norm_result$normalized
    item_means <- norm_result$means
    item_sim_matrix <- compute_item_similarity_matrix(train_normalized, train_confidence)
    
    # Prepare for prediction
    train_norm_filled <- train_normalized
    train_norm_filled[is.na(train_norm_filled)] <- 0
    train_conf_filled <- train_confidence
    train_conf_filled[is.na(train_conf_filled)] <- 0
    
    # Make predictions
    predictions <- numeric(nrow(test_obs))
    
    for (i in 1:nrow(test_obs)) {
      user_idx <- test_obs[i, 1]
      item_idx <- test_obs[i, 2]
      
      rated_items <- which(!is.na(train_matrix[user_idx, ]))
      
      if (length(rated_items) == 0) {
        predictions[i] <- item_means[item_idx]
        next
      }
      
      sims <- item_sim_matrix[item_idx, rated_items]
      sims[is.na(sims) | is.nan(sims)] <- 0
      
      # Apply k-NN
      if (sum(abs(sims) > 0) > 0 && k < length(sims)) {
        k_actual <- min(k, sum(abs(sims) > 0))
        top_k_idx <- order(abs(sims), decreasing = TRUE)[1:k_actual]
        sims_filtered <- rep(0, length(sims))
        sims_filtered[top_k_idx] <- sims[top_k_idx]
        sims <- sims_filtered
      }
      
      # Confidence-weighted prediction
      if (sum(abs(sims)) > 0) {
        weighted_ratings <- train_norm_filled[user_idx, rated_items] *
          train_conf_filled[user_idx, rated_items]
        weighted_sims <- abs(sims) * train_conf_filled[user_idx, rated_items]
        
        weighted_sum <- sum(sims * weighted_ratings)
        sum_weighted_sims <- sum(weighted_sims)
        
        if (sum_weighted_sims > 0) {
          predictions[i] <- (weighted_sum / sum_weighted_sims) + item_means[item_idx]
        } else {
          predictions[i] <- item_means[item_idx]
        }
      } else {
        predictions[i] <- item_means[item_idx]
      }
    }
    
    predictions <- pmin(pmax(predictions, 1), 10)
    actual <- user_item_matrix[test_obs]
    original_ratings <- original_ratings_matrix[test_obs]
    metrics <- evaluate_predictions(predictions, actual, original_ratings)
    
    cv_results[fold, ] <- list(fold, metrics$rmse_explicit)
    # cat("RMSE:", round(metrics$rmse_explicit, 4), "\n")
  }
  
  return(cv_results)
}
```

### Implementation

```{r ibcf-implementation}
# ============================================================================
# IBCF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Normalize and compute similarity
normalized_result_ibcf <- normalize_matrix_item(user_item_matrix)
user_item_matrix_normalized_ibcf <- normalized_result_ibcf$normalized
item_means_ibcf <- normalized_result_ibcf$means

item_similarity_matrix <- compute_item_similarity_matrix(
  user_item_matrix_normalized_ibcf,
  confidence_matrix
)

# cat("Similarity matrix computed\n\n")

# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ibcf <- recommend_for_user_ibcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_sim_matrix = item_similarity_matrix,
  item_means = item_means_ibcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

recs_ibcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "IBCF: Top 10 Recommendations for Existing User",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Generate recommendations for new user (cold start)

new_user_recs_ibcf <- recommend_for_new_user_ibcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_means = item_means_ibcf,
  item_sim_matrix = item_similarity_matrix,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

new_user_recs_ibcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "IBCF: Top 10 Recommendations for New User (Cold Start)",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

```


## Matrix Factorization

### Data Preparation

```{r mf-data-prep}
# ============================================================================
# MF: DATA PREPARATION (EXPLICIT RATINGS ONLY)
# ============================================================================

create_explicit_matrix <- function(ratings_data,
                                    min_ratings_per_book = 3,
                                    min_ratings_per_user = 2) {
  
  # Filter explicit ratings only
  explicit_ratings <- ratings_data %>%
    filter(Book.Rating > 0) %>%
    select(User.ID, ISBN, Book.Rating)
  
  # Create user-item matrix
  user_item_matrix <- explicit_ratings %>%
    pivot_wider(
      names_from = ISBN,
      values_from = Book.Rating,
      values_fill = NA
    )
  
  # Convert to matrix
  user_ids <- user_item_matrix$User.ID
  user_item_matrix <- as.matrix(user_item_matrix[, -1])
  rownames(user_item_matrix) <- user_ids
  
  # Filter by minimum ratings
  books_to_keep <- colSums(!is.na(user_item_matrix)) >= min_ratings_per_book
  user_item_matrix <- user_item_matrix[, books_to_keep]
  
  users_to_keep <- rowSums(!is.na(user_item_matrix)) >= min_ratings_per_user
  user_item_matrix <- user_item_matrix[users_to_keep, ]
  
  return(user_item_matrix)
}

prepare_recosystem_data <- function(user_item_matrix, original_ratings_matrix) {
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  
  train_data <- data.frame(
    user_index = observed[, 1] - 1,
    item_index = observed[, 2] - 1,
    rating = user_item_matrix[observed],
    original_rating = original_ratings_matrix[observed]
  )
  
  # Filter explicit ratings only
  train_data_explicit <- train_data[train_data$original_rating != 0, ]
  
  user_ids <- data.frame(
    user_index = 0:(nrow(user_item_matrix) - 1),
    user_id = rownames(user_item_matrix)
  )
  
  item_ids <- data.frame(
    item_index = 0:(ncol(user_item_matrix) - 1),
    item_id = colnames(user_item_matrix)
  )
  
  list(
    train_data = train_data_explicit,
    user_ids = user_ids,
    item_ids = item_ids,
    n_users = nrow(user_item_matrix),
    n_items = ncol(user_item_matrix)
  )
}
```

### Model Training

```{r mf-training}
# ============================================================================
# MF: HYPERPARAMETER TUNING
# ============================================================================

tune_reco_model <- function(train_data,
                             n_factors = c(10, 20, 30),
                             learning_rate = c(0.1, 0.05, 0.01),
                             costp_l2 = c(0.01, 0.1),
                             costq_l2 = c(0.01, 0.1),
                             n_iter = 50,
                             verbose = TRUE) {
  
  train_set <- data_memory(
    user_index = train_data$user_index,
    item_index = train_data$item_index,
    rating = train_data$rating,
    index1 = FALSE
  )
  
  rs <- Reco()
  
  opts <- rs$tune(train_set, opts = list(
    dim = n_factors,
    lrate = learning_rate,
    costp_l2 = costp_l2,
    costq_l2 = costq_l2,
    niter = n_iter,
    nthread = 4,
    verbose = verbose
  ))
  
  return(opts)
}

# ============================================================================
# MF: MODEL TRAINING
# ============================================================================

train_reco_model <- function(train_data,
                              n_factors = 20,
                              learning_rate = 0.1,
                              costp_l2 = 0.01,
                              costq_l2 = 0.01,
                              n_iter = 100,
                              n_threads = 4,
                              verbose = TRUE) {
  
  train_set <- data_memory(
    user_index = train_data$user_index,
    item_index = train_data$item_index,
    rating = train_data$rating,
    index1 = FALSE
  )
  
  model <- Reco()
  model$train(train_set, opts = list(
    dim = n_factors,
    lrate = learning_rate,
    costp_l2 = costp_l2,
    costq_l2 = costq_l2,
    niter = n_iter,
    nthread = n_threads,
    verbose = verbose
  ))
  
  return(model)
}
```

### Recommendation Functions

```{r mf-recommendations}
# ============================================================================
# MF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_mf <- function(model,
                                   user_item_matrix,
                                   user_id,
                                   n_recommendations = 10,
                                   user_ids_map,
                                   item_ids_map,
                                   book_info) {
  
  if (!user_id %in% user_ids_map$user_id) stop("User not found")
  
  user_idx <- user_ids_map$user_index[user_ids_map$user_id == user_id]
  user_row <- user_item_matrix[as.character(user_id), ]
  unrated_items <- which(is.na(user_row))
  
  if (length(unrated_items) == 0) return(data.frame())
  
  item_indices <- unrated_items - 1
  pred_set <- data_memory(
    user_index = rep(user_idx, length(item_indices)),
    item_index = item_indices,
    index1 = FALSE
  )
  
  pred_ratings <- model$predict(pred_set, out_memory())
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  recommended_items <- colnames(user_item_matrix)[unrated_items[top_indices]]
  
  data.frame(
    ISBN = recommended_items,
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
}

# ============================================================================
# MF: RECOMMENDATION FOR NEW USERS (COLD START)
# ============================================================================

recommend_for_new_user_mf <- function(model,
                                       new_user_ratings,
                                       user_item_matrix,
                                       item_ids_map,
                                       book_info,
                                       n_recommendations = 10) {
  
  # Hybrid approach for cold start
  ratings_df <- if (is.numeric(new_user_ratings) && !is.null(names(new_user_ratings))) {
    data.frame(item_id = names(new_user_ratings), rating = as.numeric(new_user_ratings))
  } else {
    new_user_ratings
  }
  
  ratings_df <- ratings_df[ratings_df$item_id %in% item_ids_map$item_id, ]
  if (nrow(ratings_df) == 0) stop("No rated items exist in training data")
  
  # Calculate item means for fallback
  item_means <- colMeans(user_item_matrix, na.rm = TRUE)
  
  # Get unrated items
  all_items <- item_ids_map$item_id
  rated_items <- ratings_df$item_id
  unrated_items <- setdiff(all_items, rated_items)
  
  if (length(unrated_items) == 0) return(data.frame())
  
  # Use item means adjusted by user bias
  user_mean_rating <- mean(ratings_df$rating)
  global_mean_rating <- mean(user_item_matrix, na.rm = TRUE)
  user_bias <- user_mean_rating - global_mean_rating
  
  pred_ratings <- item_means[unrated_items] + user_bias
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  recommended_items <- names(pred_ratings)[top_indices]
  
  data.frame(
    ISBN = recommended_items,
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
}
```

### Cross-Validation

```{r mf-cv}
# ============================================================================
# MF: CROSS-VALIDATION
# ============================================================================

cross_validate_mf <- function(user_item_matrix,
                               original_ratings_matrix,
                               optimal_params,
                               n_folds = 5,
                               n_iter = 50,
                               verbose = TRUE) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  if (n_ratings < n_folds * 2) {
    stop("Not enough ratings to perform cross-validation.")
  }
  
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    Fold = integer(n_folds),
    RMSE = numeric(n_folds)
  )
  
  if (verbose) {
    # cat("Starting", n_folds, "-fold cross-validation...\n")
  }
  
  for (fold in 1:n_folds) {
    # if (verbose) cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_idx <- which(fold_indices == fold)
    train_idx <- which(fold_indices != fold)
    
    train_obs <- observed[train_idx, , drop = FALSE]
    test_obs <- observed[test_idx, , drop = FALSE]
    
    # Prepare training data (explicit only)
    train_data <- data.frame(
      user_index = train_obs[, 1] - 1,
      item_index = train_obs[, 2] - 1,
      rating = user_item_matrix[train_obs],
      original_rating = original_ratings_matrix[train_obs]
    )
    
    train_data_explicit <- train_data[train_data$original_rating != 0, ]
    
    # Prepare test data (explicit only)
    test_data <- data.frame(
      user_index = test_obs[, 1] - 1,
      item_index = test_obs[, 2] - 1,
      rating = user_item_matrix[test_obs],
      original_rating = original_ratings_matrix[test_obs]
    )
    
    test_data_explicit <- test_data[test_data$original_rating != 0, ]
    
    if (nrow(test_data_explicit) == 0) {
      cv_results[fold, ] <- c(fold, NA)
      next
    }
    
    # Train model
    train_set <- data_memory(
      user_index = train_data_explicit$user_index,
      item_index = train_data_explicit$item_index,
      rating = train_data_explicit$rating,
      index1 = FALSE
    )
    
    test_set <- data_memory(
      user_index = test_data_explicit$user_index,
      item_index = test_data_explicit$item_index,
      rating = test_data_explicit$rating,
      index1 = FALSE
    )
    
    r <- Reco()
    r$train(train_set, opts = list(
      dim = optimal_params$min$dim,
      lrate = optimal_params$min$lrate,
      costp_l2 = optimal_params$min$costp_l2,
      costq_l2 = optimal_params$min$costq_l2,
      niter = n_iter,
      nthread = 4,
      verbose = FALSE
    ))
    
    preds <- r$predict(test_set, out_memory())
    preds <- pmax(pmin(preds, 10), 1)
    
    # Calculate RMSE
    actual <- test_data_explicit$rating
    original <- test_data_explicit$original_rating
    metrics <- evaluate_predictions(preds, actual, original)
    
    cv_results[fold, ] <- c(fold, metrics$rmse_explicit)
    # cat("RMSE:", round(metrics$rmse_explicit, 4), "\n")
  }
  
  return(cv_results)
}
```

### Implementation

```{r mf-implementation}
# ============================================================================
# MF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Prepare data for recosystem
prepared <- prepare_recosystem_data(user_item_matrix, original_ratings_matrix)

# Hyperparameter tuning
# optimal_params_mf <- tune_reco_model(
#   train_data = prepared$train_data,
#   n_factors = c(10, 20, 30),
#   learning_rate = c(0.1, 0.05, 0.01),
#   costp_l2 = c(0.01, 0.1),
#   costq_l2 = c(0.01, 0.1),
#   n_iter = 50,
#   verbose = FALSE
# )

# save(optimal_params_mf, file = 'mf_tuning_results.Rdata')

# Train final model
# model_mf <- train_reco_model(
#   train_data = prepared$train_data,
#   n_factors = optimal_params_mf$min$dim,
#   learning_rate = optimal_params_mf$min$lrate,
#   costp_l2 = optimal_params_mf$min$costp_l2,
#   costq_l2 = optimal_params_mf$min$costq_l2,
#   n_iter = 100,
#   n_threads = 4,
#   verbose = FALSE
# )

# save(model_mf, file = 'mf_final_model_mf.Rdata')


# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_mf <- recommend_for_user_mf(
  model = model_mf,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids_map = prepared$user_ids,
  item_ids_map = prepared$item_ids,
  book_info = book_info
)

recs_mf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF: Top 10 Recommendations for Existing User",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Generate recommendations for new user (cold start)

recs_new_mf <- recommend_for_new_user_mf(
  model = model_mf,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  item_ids_map = prepared$item_ids,
  book_info = book_info,
  n_recommendations = 10
)

recs_new_mf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF: Top 10 Recommendations for New User (Cold Start)",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Run cross-validation
cv_results_mf <- cross_validate_mf(
  user_item_matrix = user_item_matrix,
  original_ratings_matrix = original_ratings_matrix,
  optimal_params = optimal_params_mf,
  n_folds = 5,
  n_iter = 50,
  verbose = TRUE
)

cv_results_mf %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF: Cross-Validation Results",
    digits = 4,
    col.names = c("Fold", "RMSE"),
    align = c("c", "c")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# cat("\n=== MF Performance Summary ===\n")
# cat("Mean RMSE:", round(mean(cv_results_mf$RMSE, na.rm = TRUE), 4), "\n")
# cat("SD RMSE:", round(sd(cv_results_mf$RMSE, na.rm = TRUE), 4), "\n")
```



## Neural Network-Based Collaborative Filtering

### Data Preparation

```{r nn-data-prep}
# ============================================================================
# NN: DATA PREPARATION
# ============================================================================

prepare_h2o_data <- function(user_item_matrix, original_ratings_matrix = NULL) {
  
  # Apply user-mean normalization
  user_means <- rowMeans(user_item_matrix, na.rm = TRUE)
  normalized_matrix <- sweep(user_item_matrix, 1, user_means, FUN = "-")
  
  # Get observed ratings
  observed <- which(!is.na(normalized_matrix), arr.ind = TRUE)
  
  # Create training data
  train_data <- data.frame(
    user_id = rownames(user_item_matrix)[observed[, 1]],
    book_id = colnames(user_item_matrix)[observed[, 2]],
    rating = as.numeric(normalized_matrix[observed])
  )
  
  # Add original rating information
  if (!is.null(original_ratings_matrix)) {
    train_data$original_rating <- as.numeric(original_ratings_matrix[observed])
    train_data$is_implicit <- (train_data$original_rating == 0)
  } else {
    train_data$original_rating <- train_data$rating + user_means[observed[, 1]]
    train_data$is_implicit <- FALSE
  }
  
  return(list(
    train_data = train_data,
    user_ids = data.frame(user_id = rownames(user_item_matrix)),
    book_ids = data.frame(book_id = colnames(user_item_matrix)),
    n_users = nrow(user_item_matrix),
    n_items = ncol(user_item_matrix),
    user_means = user_means,
    original_matrix = user_item_matrix,
    original_ratings_matrix = original_ratings_matrix
  ))
}
```

### Hyperparameter Tuning

```{r nn-tuning}
# ============================================================================
# NN: HYPERPARAMETER TUNING WITH GRID SEARCH
# ============================================================================

# Initialize H2O cluster
suppressMessages(h2o.init(ip = "localhost", port = 54321))
h2o.no_progress()

# # Prepare data
# h2o_data_tuning <- prepare_h2o_data(user_item_matrix, original_ratings_matrix)
# 
# # Filter explicit ratings only
# train_data_explicit <- h2o_data_tuning$train_data[!h2o_data_tuning$train_data$is_implicit, ]
# 
# # Convert to H2O frame
# train_h2o_tuning <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
# train_h2o_tuning$user_id <- as.factor(train_h2o_tuning$user_id)
# train_h2o_tuning$book_id <- as.factor(train_h2o_tuning$book_id)
# 
# # Define hyperparameter grid
# hyper_params <- list(
#   activation = c("Tanh", "Rectifier"),
#   hidden = list(
#     c(16),
#     c(32, 16),
#     c(64, 32),
#     c(64, 32, 16),
#     c(128, 64, 32)
#   ),
#   l1 = c(0.00001, 0.0001),
#   l2 = c(0.00001, 0.0001)
# )
# 
# cat("Total models to train:", 
#     length(hyper_params$activation) * length(hyper_params$hidden) * 
#     length(hyper_params$l1) * length(hyper_params$l2), "\n\n")
# 
# set.seed(123)

# # Run grid search
# model_grid <- h2o.grid(
#   algorithm = "deeplearning",
#   grid_id = "nn_grid_book_recommendations",
#   hyper_params = hyper_params,
#   x = c("user_id", "book_id"),
#   y = "rating",
#   training_frame = train_h2o_tuning,
#   nfolds = 5,
#   stopping_metric = "RMSE",
#   stopping_rounds = 5,
#   stopping_tolerance = 0.001,
#   seed = 123,
#   adaptive_rate = TRUE,
#   epochs = 50,
#   max_w2 = 10,
#   initial_weight_distribution = "UniformAdaptive"
# )
# 
# # Get results
# grid_results <- h2o.getGrid(
#   "nn_grid_book_recommendations",
#   sort_by = "rmse",
#   decreasing = FALSE)

# all_models_df <- as.data.frame(grid_results@summary_table)

# Select best model
# best_model_id <- grid_results@model_ids[[1]]
# best_model <- h2o.getModel(best_model_id)
# cv_rmse <- h2o.rmse(best_model, xval = TRUE)

# best_nn_results <- data.frame(
#   Activation = best_model@allparameters$activation,
#   Hidden_Layers = paste(best_model@allparameters$hidden, collapse = ", "),
#   N_Layers = length(best_model@allparameters$hidden),
#   L1_Regularization = best_model@allparameters$l1,
#   L2_Regularization = best_model@allparameters$l2,
#   CV_RMSE = round(cv_rmse, 4),
#   Data_Type = "Explicit Only"
# )


# Grid search results
kable(all_models_df %>% select(-model_ids),
      format = "latex",
      booktabs = TRUE,
      caption = "Neural Network Grid Search Results",
      align = c("c", "c", "c", "c", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Best model results
kable(best_nn_results,
      format = "latex",
      booktabs = TRUE,
      caption = "Optimal Neural Network Configuration",
      align = c("c", "c", "c", "c", "c", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )



# save(best_nn_results, grid_results, all_models_df, file = 'nn_tuning_results.Rdata') 
```

### Model Training

```{r nn-training}
# ============================================================================
# NN: MODEL TRAINING FUNCTION
# ============================================================================

train_h2o_model <- function(train_data,
                             hidden = c(64, 32),
                             epochs = 30,
                             activation = "Tanh",
                             l1 = 0.0001,
                             l2 = 0.0001,
                             seed = 123,
                             verbose = FALSE) {
  
  if (!verbose) h2o.no_progress()
  
  model <- h2o.deeplearning(
    x = c("user_id", "book_id"),
    y = "rating",
    training_frame = train_data,
    hidden = hidden,
    epochs = epochs,
    activation = activation,
    l1 = l1,
    l2 = l2,
    adaptive_rate = TRUE,
    seed = seed,
    stopping_rounds = 5,
    stopping_metric = "RMSE",
    stopping_tolerance = 0.001,
    max_w2 = 10,
    initial_weight_distribution = "UniformAdaptive",
    initial_weight_scale = 0.5
  )
  
  return(model)
}
```

### Recommendation Functions

```{r nn-recommendations}
# ============================================================================
# NN: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_nn <- function(model,
                                   user_item_matrix,
                                   user_id,
                                   n_recommendations = 10,
                                   user_ids,
                                   book_ids,
                                   user_means,
                                   book_info) {
  
  if (!user_id %in% user_ids$user_id) {
    stop("User ID not found in training data")
  }
  
  # Find unrated books
  user_row <- user_item_matrix[as.character(user_id), , drop = TRUE]
  unrated_books <- which(is.na(user_row))
  
  if (length(unrated_books) == 0) {
    message("User has rated all available books")
    return(data.frame())
  }
  
  # Prepare prediction input
  pred_data <- data.frame(
    user_id = rep(user_id, length(unrated_books)),
    book_id = colnames(user_item_matrix)[unrated_books]
  )
  
  # Convert to H2O frame
  pred_h2o <- as.h2o(pred_data)
  pred_h2o$user_id <- as.factor(pred_h2o$user_id)
  pred_h2o$book_id <- as.factor(pred_h2o$book_id)
  
  # Predict ratings
  predictions <- h2o.predict(model, pred_h2o)
  predictions_df <- as.data.frame(predictions)
  
  # Denormalize predictions
  user_mean <- user_means[as.character(user_id)]
  pred_ratings <- predictions_df$predict + user_mean
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  # Select top recommendations
  n_to_recommend <- min(n_recommendations, length(pred_ratings))
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:n_to_recommend]
  
  recommendations <- data.frame(
    ISBN = pred_data$book_id[top_indices],
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  )
  
  recommendations <- recommendations %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# NN: RECOMMENDATION FOR NEW USERS (COLD START)
# ============================================================================

recommend_for_new_user_nn <- function(model,
                                       new_user_ratings,
                                       user_item_matrix,
                                       book_ids,
                                       book_info,
                                       n_recommendations = 10,
                                       user_means = NULL) {
  
  if (length(new_user_ratings) == 0) {
    stop("No ratings provided for new user")
  }
  
  rated_books <- names(new_user_ratings)
  valid_books <- rated_books[rated_books %in% book_ids$book_id]
  
  if (length(valid_books) == 0) {
    stop("None of the rated books exist in training data")
  }
  
  # Find similar users based on rated books
  new_user_mean <- mean(new_user_ratings)
  new_user_normalized <- new_user_ratings - new_user_mean
  
  user_similarities <- numeric(nrow(user_item_matrix))
  names(user_similarities) <- rownames(user_item_matrix)
  
  for (i in 1:nrow(user_item_matrix)) {
    user_ratings <- user_item_matrix[i, valid_books]
    valid_overlap <- !is.na(user_ratings)
    
    if (sum(valid_overlap) == 0) {
      user_similarities[i] <- 0
      next
    }
    
    overlap_books <- valid_books[valid_overlap]
    existing_mean <- mean(user_ratings[valid_overlap], na.rm = TRUE)
    existing_normalized <- user_ratings[valid_overlap] - existing_mean
    new_normalized <- new_user_normalized[overlap_books]
    
    # Cosine similarity
    numerator <- sum(existing_normalized * new_normalized)
    denominator <- sqrt(sum(existing_normalized^2)) * sqrt(sum(new_normalized^2))
    
    if (denominator > 0) {
      user_similarities[i] <- numerator / denominator
    } else {
      user_similarities[i] <- 0
    }
  }
  
  # Use top-K similar users
  k <- min(30, sum(user_similarities > 0))
  
  if (k == 0) {
    # Fallback to item popularity
    item_means <- colMeans(user_item_matrix, na.rm = TRUE)
    all_books <- book_ids$book_id
    unrated_books <- setdiff(all_books, rated_books)
    pred_ratings <- item_means[unrated_books] + (new_user_mean - mean(user_means))
  } else {
    top_k_users <- names(sort(user_similarities, decreasing = TRUE)[1:k])
    similarities <- user_similarities[top_k_users]
    
    all_books <- book_ids$book_id
    unrated_books <- setdiff(all_books, rated_books)
    
    # Weighted average of similar users' ratings
    weighted_preds <- numeric(length(unrated_books))
    names(weighted_preds) <- unrated_books
    
    for (book in unrated_books) {
      similar_user_ratings <- user_item_matrix[top_k_users, book]
      valid_ratings <- !is.na(similar_user_ratings)
      
      if (sum(valid_ratings) > 0) {
        weighted_sum <- sum(similarities[valid_ratings] * similar_user_ratings[valid_ratings])
        sum_weights <- sum(abs(similarities[valid_ratings]))
        weighted_preds[book] <- weighted_sum / sum_weights
      } else {
        weighted_preds[book] <- mean(user_item_matrix[, book], na.rm = TRUE)
      }
    }
    
    pred_ratings <- weighted_preds + new_user_mean - mean(user_means)
  }
  
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  
  recommendations <- data.frame(
    ISBN = names(pred_ratings)[top_indices],
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```

### Cross-Validation

```{r nn-cv}
# ============================================================================
# NN: CROSS-VALIDATION
# ============================================================================

cross_validate_nn <- function(user_item_matrix,
                               original_ratings_matrix = NULL,
                               n_folds = 5,
                               seed = 123,
                               hidden = c(64, 32),
                               epochs = 20,
                               activation = "Tanh",
                               l1 = 0.0001,
                               l2 = 0.0001) {
  
  set.seed(seed)
  h2o.no_progress()
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  if (n_ratings < n_folds * 2) {
    stop("Not enough ratings to perform cross-validation")
  }
  
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  # cat("Starting", n_folds, "-fold cross-validation for Neural Network...\n")
  # cat("Training on EXPLICIT ratings only\n\n")
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    train_indices <- which(fold_indices != fold)
    
    test_obs <- observed[test_indices, , drop = FALSE]
    train_obs <- observed[train_indices, , drop = FALSE]
    
    # Create training matrix
    train_matrix <- user_item_matrix
    train_matrix[test_obs] <- NA
    
    # Normalize
    train_user_means <- rowMeans(train_matrix, na.rm = TRUE)
    train_normalized <- sweep(train_matrix, 1, train_user_means, FUN = "-")
    
    # Create training data
    train_data <- data.frame(
      user_id = rownames(user_item_matrix)[train_obs[, 1]],
      book_id = colnames(user_item_matrix)[train_obs[, 2]],
      rating = as.numeric(train_normalized[train_obs])
    )
    
    # Filter explicit ratings
    if (!is.null(original_ratings_matrix)) {
      train_data$original_rating <- original_ratings_matrix[train_obs]
      train_data_explicit <- train_data[train_data$original_rating != 0, ]
    } else {
      train_data_explicit <- train_data
    }
    
    if (nrow(train_data_explicit) == 0) {
      cv_results[fold, ] <- c(fold, NA)
      # cat("No explicit ratings - skipping\n")
      next
    }
    
    # Convert to H2O
    train_h2o <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
    train_h2o$user_id <- as.factor(train_h2o$user_id)
    train_h2o$book_id <- as.factor(train_h2o$book_id)
    
    # Train model
    model <- h2o.deeplearning(
      x = c("user_id", "book_id"),
      y = "rating",
      training_frame = train_h2o,
      hidden = hidden,
      epochs = epochs,
      activation = activation,
      l1 = l1,
      l2 = l2,
      adaptive_rate = TRUE,
      seed = seed,
      stopping_rounds = 3,
      stopping_tolerance = 0.001,
      max_w2 = 10,
      initial_weight_distribution = "UniformAdaptive",
      initial_weight_scale = 0.5,
      verbose = FALSE
    )
    
    # Prepare test data (explicit only)
    test_user_means <- train_user_means[rownames(user_item_matrix)[test_obs[, 1]]]
    test_data <- data.frame(
      user_id = rownames(user_item_matrix)[test_obs[, 1]],
      book_id = colnames(user_item_matrix)[test_obs[, 2]],
      rating = as.numeric(user_item_matrix[test_obs]) - test_user_means
    )
    
    if (!is.null(original_ratings_matrix)) {
      test_data$original_rating <- original_ratings_matrix[test_obs]
      test_data_explicit <- test_data[test_data$original_rating != 0, ]
      test_user_means_explicit <- test_user_means[test_data$original_rating != 0]
      actual_explicit <- user_item_matrix[test_obs][test_data$original_rating != 0]
    } else {
      test_data_explicit <- test_data
      test_user_means_explicit <- test_user_means
      actual_explicit <- user_item_matrix[test_obs]
    }
    
    # Convert to H2O
    test_h2o <- as.h2o(test_data_explicit[, c("user_id", "book_id", "rating")])
    test_h2o$user_id <- as.factor(test_h2o$user_id)
    test_h2o$book_id <- as.factor(test_h2o$book_id)
    
    # Predict
    predictions <- as.data.frame(h2o.predict(model, test_h2o))
    predictions_denorm <- predictions$predict + test_user_means_explicit
    predictions_denorm <- pmax(pmin(predictions_denorm, 10), 1)
    
    # Calculate RMSE
    rmse_value <- sqrt(mean((predictions_denorm - actual_explicit)^2, na.rm = TRUE))
    
    cv_results[fold, ] <- list(fold, rmse_value)
    # cat("RMSE:", round(rmse_value, 4), "\n")
    
    # Cleanup
    h2o.rm(model)
    h2o.rm(train_h2o)
    h2o.rm(test_h2o)
    gc()
  }
  
  return(cv_results)
}
```

### Implementation

```{r nn-implementation}
# ============================================================================
# NN: MODEL TRAINING AND EVALUATION
# ============================================================================

# Initialize H2O cluster
suppressMessages(h2o.init(ip = "localhost", port = 54321))
h2o.no_progress()

# Prepare data for final model
prepared <- prepare_h2o_data(user_item_matrix, original_ratings_matrix)

optimal_params_nn <- list(
  hidden = as.numeric(unlist(strsplit(as.character(best_nn_results$Hidden_Layers), ", "))),
  epochs = 30,  # Default value
  activation = as.character(best_nn_results$Activation),
  l1 = as.numeric(best_nn_results$L1_Regularisation),
  l2 = as.numeric(best_nn_results$L2_Regularisation)
)

# Train final model
train_data_explicit <- prepared$train_data[!prepared$train_data$is_implicit, ]

train_h2o <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
train_h2o$user_id <- as.factor(train_h2o$user_id)
train_h2o$book_id <- as.factor(train_h2o$book_id)


# Train new model
model_nn <- train_h2o_model(
  train_data = train_h2o,
  hidden = optimal_params_nn$hidden,
  epochs = optimal_params_nn$epochs,
  activation = optimal_params_nn$activation,
  l1 = optimal_params_nn$l1,
  l2 = optimal_params_nn$l2,
  verbose = FALSE)
  
# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_nn <- recommend_for_user_nn(
  model = model_nn,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids = prepared$user_ids,
  book_ids = prepared$book_ids,
  user_means = prepared$user_means,
  book_info = book_info
)

recs_nn %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "NN: Top 10 Recommendations for Existing User",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Generate recommendations for new user (cold start)


new_user_recs_nn <- recommend_for_new_user_nn(
  model = model_nn,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  book_ids = prepared$book_ids,
  book_info = book_info,
  n_recommendations = 10,
  user_means = prepared$user_means
)

new_user_recs_nn %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "NN: Top 10 Recommendations for New User (Cold Start)",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

```




# Results and Analysis

## Cross-Validation Performance Comparison

The comprehensive evaluation of all four collaborative filtering methods reveals distinct performance characteristics across the book recommendation domain. Table 1 presents the five-fold cross-validation results, demonstrating clear performance differences between approaches.

```{r comprehensive-comparison}
# ============================================================================
# COMPREHENSIVE CROSS-VALIDATION COMPARISON
# ============================================================================

run_comprehensive_cv_comparison <- function(user_item_matrix,
                                            confidence_matrix,
                                            original_ratings_matrix,
                                            optimal_params_mf,
                                            optimal_params_nn,
                                            n_folds = 5,
                                            k = 50) {
  
  
  all_results <- data.frame(
    Method = character(4),
    CV_RMSE_Mean = numeric(4),
    CV_RMSE_SD = numeric(4),
    stringsAsFactors = FALSE
  )
  
  # User-Based CF
  # cat("=== Running User-Based CF ===\n")
  cv_results_ubcf <- cross_validate_ubcf(
    user_item_matrix,
    confidence_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    k = k
  )
  
  all_results[1, ] <- list(
    "User-Based CF",
    round(mean(cv_results_ubcf$rmse), 3),
    round(sd(cv_results_ubcf$rmse), 3)
  )
  
  # Item-Based CF
  # cat("\n=== Running Item-Based CF ===\n")
  cv_results_ibcf <- cross_validate_ibcf(
    user_item_matrix,
    confidence_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    k = k
  )
  
  all_results[2, ] <- list(
    "Item-Based CF",
    round(mean(cv_results_ibcf$rmse), 3),
    round(sd(cv_results_ibcf$rmse), 3)
  )
  
  # Matrix Factorization
  # cat("\n=== Running Matrix Factorization ===\n")
  cv_results_mf <- cross_validate_mf(
    user_item_matrix,
    original_ratings_matrix,
    optimal_params_mf,
    n_folds = n_folds,
    n_iter = 50,
    verbose = TRUE
  )
  
  all_results[3, ] <- list(
    "Matrix Factorization",
    round(mean(cv_results_mf$RMSE, na.rm = TRUE), 3),
    round(sd(cv_results_mf$RMSE, na.rm = TRUE), 3)
  )
  
  # Neural Network
  # cat("\n=== Running Neural Network ===\n")
  cv_results_nn <- cross_validate_nn(
    user_item_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    hidden = optimal_params_nn$hidden,
    epochs = optimal_params_nn$epochs,
    activation = optimal_params_nn$activation,
    l1 = optimal_params_nn$l1,
    l2 = optimal_params_nn$l2
  )
  
  all_results[4, ] <- list(
    "Neural Network",
    round(mean(cv_results_nn$rmse, na.rm = TRUE), 3),
    round(sd(cv_results_nn$rmse, na.rm = TRUE), 3)
  )
  

  # Return all results including individual CV results
  return(list(
    summary = all_results,
    cv_results_ubcf = cv_results_ubcf,
    cv_results_ibcf = cv_results_ibcf,
    cv_results_mf = cv_results_mf,
    cv_results_nn = cv_results_nn
  ))
}

# # Run comprehensive comparison
# comparison_results <- run_comprehensive_cv_comparison(
#   user_item_matrix = user_item_matrix,
#   confidence_matrix = confidence_matrix,
#   original_ratings_matrix = original_ratings_matrix,
#   optimal_params_mf = optimal_params_mf,
#   optimal_params_nn = optimal_params_nn,
#   n_folds = 5,
#   k = 50
# )
# 
# # Extract results
# cv_comparison_results <- comparison_results$summary
# cv_results_ubcf <- comparison_results$cv_results_ubcf
# cv_results_ibcf <- comparison_results$cv_results_ibcf
# cv_results_mf <- comparison_results$cv_results_mf
# cv_results_nn <- comparison_results$cv_results_nn
# 
# save(cv_comparison_results, cv_results_ubcf, cv_results_ibcf, cv_results_mf, cv_results_nn, 
#      file = "all_cv_results.Rdata")

# Display results
kable(
  cv_comparison_results,
  format = "latex",
  booktabs = TRUE,
  caption = "Cross-Validation Performance Comparison - All Methods",
  col.names = c("Method", "Mean RMSE", "SD RMSE"),
  align = c("l", "c", "c")
) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )



```

```{r results-visualization, fig.cap="Performance Comparison Across Methods"}
# ============================================================================
# VISUALIZATION: PERFORMANCE COMPARISON
# ============================================================================

# Ensure H2O is available for any remaining operations
if (!h2o::h2o.clusterIsUp()) {
  suppressMessages(h2o::h2o.init(ip = "localhost", port = 54321))
  h2o::h2o.no_progress()
}

# Create bar plot for RMSE comparison
ggplot(cv_comparison_results, aes(x = reorder(Method, CV_RMSE_Mean), y = CV_RMSE_Mean)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_errorbar(
    aes(ymin = CV_RMSE_Mean - CV_RMSE_SD, ymax = CV_RMSE_Mean + CV_RMSE_SD),
    width = 0.2
  ) +
  geom_text(aes(label = round(CV_RMSE_Mean, 3)), vjust = -0.5) +
  labs(
    title = "Cross-Validation Performance Comparison",
    subtitle = "Lower RMSE indicates better performance",
    x = "Method",
    y = "Mean RMSE"
  ) +
  theme_cowplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```

The cross-validation results demonstrate a clear performance hierarchy among the four collaborative filtering approaches. **Matrix Factorization** achieved the best performance with a mean RMSE of 1.820 ± 0.020, followed closely by **Neural Networks** at 1.899 ± 0.053. **Item-Based Collaborative Filtering** performed moderately well with an RMSE of 2.746 ± 0.018, while **User-Based Collaborative Filtering** showed the highest error rate at 3.182 ± 0.047.

The superior performance of model-based methods (MF and NN) over memory-based approaches (UBCF and IBCF) reflects their ability to capture latent patterns in user-item interactions through learned representations. The relatively low standard deviations across all methods (ranging from 0.018 to 0.053) indicate consistent performance across different data folds, suggesting robust model behavior.

# Dataset Size Impact Analysis




```{r}
# ============================================================================
# DATASET SIZE IMPACT ANALYSIS - FULLY CORRECTED
# ============================================================================
# Research Question: How many book titles does a recommender system need
# to achieve good prediction accuracy?
# 
# Methodology:
# 1. For each catalog size (5, 10, 20, 30, 50, 75, 100, 145 books)
# 2. Select TOP N most popular books
# 3. For EACH USER: Split their ratings 80/20 into train/test
# 4. Train MF model on 80%, test on 20%
# 5. Run 5 times with different random seeds for robustness
# 6. CRITICAL: Only include users who have rated enough books to split fairly
#
# Key: This ensures the same books appear in both train and test,
# and each user contributes proportionally to both sets
# ============================================================================

analyze_catalog_sizes_fair <- function(data, 
                                      catalog_sizes = c(5, 10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100, 110, 120, 130, 145),
                                      optimal_params_mf,
                                      test_fraction = 0.2,
                                      min_user_ratings = 3,  # Minimum ratings per user to participate
                                      n_seeds = 5,
                                      base_seed = 123) {
  
  set.seed(base_seed)
  all_results <- data.frame()
  
  # Get book popularity rankings
  book_popularity <- data %>%
    group_by(ISBN) %>%
    summarise(n_ratings = n(), .groups = "drop") %>%
    arrange(desc(n_ratings))
  
  # cat("=== Fair Catalog Size Analysis ===\n")
  # cat("Method: Per-user 80/20 split on TOP N popular books\n")
  # cat("Minimum ratings per user:", min_user_ratings, "\n")
  # cat("Testing catalog sizes:", paste(catalog_sizes, collapse = ", "), "\n")
  # cat("Total catalog sizes to test:", length(catalog_sizes), "\n\n")
  
  # Verify we have enough books
  if (nrow(book_popularity) < max(catalog_sizes)) {
    stop("Not enough books in dataset. Available: ", nrow(book_popularity), 
         ", Requested: ", max(catalog_sizes))
  }
  
  # Process each catalog size
  for (n_books in catalog_sizes) {
    # cat("=== Catalog Size:", n_books, "books ===\n")
    
    # Select TOP N most popular books
    selected_books <- head(book_popularity$ISBN, n_books)
    data_subset <- data %>% filter(ISBN %in% selected_books)
    
    # Count ratings per user in this subset
    user_rating_counts <- data_subset %>%
      group_by(User.ID) %>%
      summarise(n_ratings = n(), .groups = "drop")
    
    # Only keep users with enough ratings to split
    eligible_users <- user_rating_counts %>%
      filter(n_ratings >= min_user_ratings) %>%
      pull(User.ID)
    
    data_subset_filtered <- data_subset %>%
      filter(User.ID %in% eligible_users)
    
    n_eligible <- length(eligible_users)
    n_total_ratings <- nrow(data_subset_filtered)
    
    # cat("Selected", length(selected_books), "books\n")
    # cat("Eligible users (>=", min_user_ratings, "ratings):", n_eligible, "\n")
    # cat("Total ratings from eligible users:", n_total_ratings, "\n")
    
    if (n_eligible < 10) {
      # cat("SKIP: Not enough eligible users\n\n")
      
      all_results <- rbind(all_results, data.frame(
        catalog_size = n_books,
        seed = NA,
        n_users = 0,
        n_items = length(selected_books),
        n_ratings_total = n_total_ratings,
        n_explicit = NA,
        sparsity_pct = NA,
        n_train = NA,
        n_test = NA,
        rmse = NA,
        reason = "Insufficient eligible users"
      ))
      next
    }
    
    # Run multiple seeds
    for (seed_num in 1:n_seeds) {
      current_seed <- base_seed + seed_num
      set.seed(current_seed)
      
      # cat("  Seed", seed_num, "of", n_seeds, "...")
      
      # ================================================================
      # CRITICAL: PER-USER 80/20 SPLIT
      # ================================================================
      # For each user, randomly select 20% of their ratings for testing
      # This ensures EVERY user contributes to both train and test sets
      
      train_data_list <- list()
      test_data_list <- list()
      
      for (user in eligible_users) {
        user_ratings <- data_subset_filtered %>% 
          filter(User.ID == user)
        
        n_user_ratings <- nrow(user_ratings)
        n_test_user <- max(1, floor(test_fraction * n_user_ratings))
        
        # Randomly select test indices for this user
        test_indices <- sample(1:n_user_ratings, n_test_user)
        
        test_data_list[[as.character(user)]] <- user_ratings[test_indices, ]
        train_data_list[[as.character(user)]] <- user_ratings[-test_indices, ]
      }
      
      # Combine all users' train and test data
      train_data_final <- bind_rows(train_data_list)
      test_data_final <- bind_rows(test_data_list)
      
      # cat("Train:", nrow(train_data_final), "ratings, ",
      #     "Test:", nrow(test_data_final), "ratings...")
      
      # ================================================================
      # CREATE MATRICES
      # ================================================================
      
      # Create FULL matrix first (to get consistent dimensions)
      matrix_result_full <- create_user_item_matrix(
        data_subset_filtered,
        min_ratings_per_book = 1,  # Don't filter - we already selected books
        min_ratings_per_user = 1,  # Don't filter - we already selected users
        implicit_rating = 4,
        implicit_confidence = 0.5
      )
      
      full_matrix <- matrix_result_full$ratings
      full_original <- matrix_result_full$original_ratings
      
      # Create TRAIN matrix (mask test ratings)
      train_matrix <- full_matrix
      train_original <- full_original
      
      # Mask out test ratings
      for (i in 1:nrow(test_data_final)) {
        user_id <- as.character(test_data_final$User.ID[i])
        book_id <- test_data_final$ISBN[i]
        
        if (user_id %in% rownames(train_matrix) && 
            book_id %in% colnames(train_matrix)) {
          train_matrix[user_id, book_id] <- NA
          train_original[user_id, book_id] <- NA
        }
      }
      
      # Calculate metrics
      n_users <- nrow(full_matrix)
      n_items <- ncol(full_matrix)
      n_ratings_total <- sum(!is.na(full_matrix))
      n_explicit_total <- sum(full_original[!is.na(full_original)] > 0)
      sparsity <- mean(is.na(full_matrix)) * 100
      
      # Prepare train data for MF (explicit only)
      train_observed <- which(!is.na(train_matrix), arr.ind = TRUE)
      
      train_data_mf <- data.frame(
        user_index = train_observed[, 1] - 1,
        item_index = train_observed[, 2] - 1,
        rating = train_matrix[train_observed],
        original_rating = train_original[train_observed]
      )
      
      # Filter to explicit ratings only
      train_data_explicit <- train_data_mf %>%
        filter(original_rating > 0)
      
      n_train_explicit <- nrow(train_data_explicit)
      
      if (n_train_explicit < 10) {
        # cat(" SKIP (only", n_train_explicit, "explicit training ratings)\n")
        
        all_results <- rbind(all_results, data.frame(
          catalog_size = n_books,
          seed = seed_num,
          n_users = n_users,
          n_items = n_items,
          n_ratings_total = n_ratings_total,
          n_explicit = n_explicit_total,
          sparsity_pct = sparsity,
          n_train = n_train_explicit,
          n_test = NA,
          rmse = NA,
          reason = "Insufficient training data"
        ))
        next
      }
      
      # ================================================================
      # TRAIN MODEL
      # ================================================================
      
      # Create mapping for users and items
      user_ids_map <- data.frame(
        user_index = 0:(n_users - 1),
        user_id = rownames(train_matrix)
      )
      
      item_ids_map <- data.frame(
        item_index = 0:(n_items - 1),
        item_id = colnames(train_matrix)
      )
      
      # Train MF model
      train_set <- data_memory(
        user_index = train_data_explicit$user_index,
        item_index = train_data_explicit$item_index,
        rating = train_data_explicit$rating,
        index1 = FALSE
      )
      
      model <- Reco()
      suppressWarnings({
        model$train(train_set, opts = list(
          dim = optimal_params_mf$min$dim,
          lrate = optimal_params_mf$min$lrate,
          costp_l2 = optimal_params_mf$min$costp_l2,
          costq_l2 = optimal_params_mf$min$costq_l2,
          niter = 50,
          nthread = 4,
          verbose = FALSE
        ))
      })
      
      # ================================================================
      # MAKE PREDICTIONS ON TEST SET
      # ================================================================
      
      predictions <- numeric(nrow(test_data_final))
      test_actuals <- numeric(nrow(test_data_final))
      test_original_ratings <- numeric(nrow(test_data_final))
      valid_predictions <- rep(TRUE, nrow(test_data_final))
      
      for (i in 1:nrow(test_data_final)) {
        user_id <- as.character(test_data_final$User.ID[i])
        book_id <- test_data_final$ISBN[i]
        
        # Check if user and book exist in matrices
        if (!(user_id %in% user_ids_map$user_id) || 
            !(book_id %in% item_ids_map$item_id)) {
          valid_predictions[i] <- FALSE
          next
        }
        
        # Get indices
        user_idx <- user_ids_map$user_index[user_ids_map$user_id == user_id]
        item_idx <- item_ids_map$item_index[item_ids_map$item_id == book_id]
        
        # Make prediction
        pred_set <- data_memory(
          user_index = user_idx,
          item_index = item_idx,
          index1 = FALSE
        )
        
        pred <- model$predict(pred_set, out_memory())
        predictions[i] <- pmax(pmin(pred, 10), 1)
        
        # Get actual rating
        test_actuals[i] <- full_matrix[user_id, book_id]
        test_original_ratings[i] <- full_original[user_id, book_id]
      }
      
      # Filter to valid predictions and explicit ratings only
      valid_mask <- valid_predictions & (test_original_ratings > 0)
      n_test_explicit <- sum(valid_mask)
      
      if (n_test_explicit == 0) {
        # cat(" SKIP (no valid explicit test ratings)\n")
        
        all_results <- rbind(all_results, data.frame(
          catalog_size = n_books,
          seed = seed_num,
          n_users = n_users,
          n_items = n_items,
          n_ratings_total = n_ratings_total,
          n_explicit = n_explicit_total,
          sparsity_pct = sparsity,
          n_train = n_train_explicit,
          n_test = 0,
          rmse = NA,
          reason = "No valid test ratings"
        ))
        next
      }
      
      # Calculate RMSE on explicit test ratings
      pred_explicit <- predictions[valid_mask]
      actual_explicit <- test_actuals[valid_mask]
      
      rmse <- sqrt(mean((pred_explicit - actual_explicit)^2, na.rm = TRUE))
      
      # Store results
      all_results <- rbind(all_results, data.frame(
        catalog_size = n_books,
        seed = seed_num,
        n_users = n_users,
        n_items = n_items,
        n_ratings_total = n_ratings_total,
        n_explicit = n_explicit_total,
        sparsity_pct = sparsity,
        n_train = n_train_explicit,
        n_test = n_test_explicit,
        rmse = rmse,
        reason = "Success"
      ))
      
      # cat(" RMSE =", round(rmse, 4), "\n")
    }
    
    # cat("\n")
  }
  
  # cat("=== Analysis Complete ===\n\n")
  
  # ================================================================
  # AGGREGATE RESULTS
  # ================================================================
  
  summary_results <- all_results %>%
    group_by(catalog_size) %>%
    summarise(
      n_valid_runs = sum(!is.na(rmse)),
      rmse_mean = mean(rmse, na.rm = TRUE),
      rmse_sd = sd(rmse, na.rm = TRUE),
      rmse_se = sd(rmse, na.rm = TRUE) / sqrt(sum(!is.na(rmse))),
      rmse_min = min(rmse, na.rm = TRUE),
      rmse_max = max(rmse, na.rm = TRUE),
      n_users_mean = round(mean(n_users, na.rm = TRUE), 0),
      n_items_mean = round(mean(n_items, na.rm = TRUE), 0),
      sparsity_mean = round(mean(sparsity_pct, na.rm = TRUE), 1),
      n_explicit_mean = round(mean(n_explicit, na.rm = TRUE), 0),
      n_train_mean = round(mean(n_train, na.rm = TRUE), 0),
      n_test_mean = round(mean(n_test, na.rm = TRUE), 0),
      .groups = "drop"
    ) %>%
    arrange(catalog_size) %>%
    mutate(
      rmse_change = rmse_mean - lag(rmse_mean),
      improvement = -rmse_change,
      percent_improvement = -rmse_change / lag(rmse_mean) * 100
    )
  
  return(list(
    detailed = all_results,
    summary = summary_results
  ))
}

# ============================================================================
# RUN THE ANALYSIS
# ============================================================================

# catalog_analysis_results <- analyze_catalog_sizes_fair(
#   data = data,
#   catalog_sizes = c(5, 10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100, 110, 120, 130, 145),
#   optimal_params_mf = optimal_params_mf,
#   test_fraction = 0.2,
#   min_user_ratings = 3,  
#   n_seeds = 5,
#   base_seed = 123
# )
# 
# # Extract results
# size_impact_results <- catalog_analysis_results$summary
# size_impact_results_detailed <- catalog_analysis_results$detailed
# 
# save(size_impact_results, size_impact_results_detailed, 
#      file = "dataset_size_analysis_complete.Rdata")


kable(
  size_impact_results %>% 
    select(catalog_size, rmse_mean, rmse_se, sparsity_mean),
  format = "latex",
  booktabs = TRUE,
  caption = "Impact of Catalog Size on Prediction Accuracy",
  digits = 4,
  col.names = c("Catalog Size", "Mean RMSE", "SE", "Sparsity %"),
  align = c("c", "c", "c", "c")
) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```


```{r dataset-size-visualization, fig.cap="Dataset Size Impact Analysis: RMSE trends across catalog sizes with error bars and optimal size highlighted."}

# RMSE Trend with Error Bars
ggplot(size_impact_results, aes(x = catalog_size, y = rmse_mean)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  geom_errorbar(aes(ymin = rmse_mean - rmse_se, ymax = rmse_mean + rmse_se),
                width = 3, color = "steelblue", alpha = 0.7) +
  geom_point(data = size_impact_results[which.min(size_impact_results$rmse_mean), ],
             color = "red", size = 5, shape = 18) +
  labs(title = "A. RMSE by Catalog Size",
       subtitle = "Error bars show ±1 SE across 5 seeds. Red diamond indicates optimal size.",
       x = "Number of Books", y = "Mean RMSE") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

```

The analysis reveals a **non-monotonic relationship** between catalog size and prediction accuracy. Performance initially improves dramatically from 5 to 10 books (RMSE decreases from 2.027 to 1.878), then shows fluctuations around the optimal range. The **optimal catalog size** occurs at **50 books** with an RMSE of 1.815 ± 0.027, representing a 10.4% improvement over the smallest catalog size.

Beyond 50 books, performance begins to degrade slightly, with RMSE increasing to 1.833 ± 0.024 at the full catalog size of 145 books. This suggests that **diminishing returns** set in after approximately 50 books, where additional titles introduce noise rather than improving prediction quality. The optimal range appears to be between 40-60 books, where the balance between data richness and model complexity is most favorable.







# Discussion

## Key Findings and Performance Analysis

The comprehensive evaluation reveals several important insights about collaborative filtering methods in the book recommendation domain. The clear performance hierarchy—Matrix Factorization (RMSE: 1.820) > Neural Networks (RMSE: 1.899) > Item-Based CF (RMSE: 2.746) > User-Based CF (RMSE: 3.182)—demonstrates the superiority of model-based approaches over memory-based methods in this sparse dataset.

### Model-Based vs Memory-Based Performance Gap

The significant performance gap between model-based and memory-based methods (approximately 0.9-1.4 RMSE points) highlights the importance of learning latent representations in sparse recommendation scenarios. Matrix Factorization's superior performance can be attributed to its ability to capture underlying user preferences and item characteristics through low-dimensional embeddings, effectively addressing the sparsity challenge that plagues memory-based approaches.

The Neural Network's close second-place performance (RMSE: 1.899) demonstrates the potential of deep learning for collaborative filtering, though the additional complexity did not translate to improved performance over the simpler matrix factorization approach. This suggests that for this dataset size and sparsity level, the linear relationships captured by matrix factorization may be sufficient.

### Dataset Size Impact and Diminishing Returns

The dataset size analysis reveals a critical finding: **optimal performance occurs at 50 books rather than the full catalog of 145 books**. This non-monotonic relationship challenges the common assumption that more data always leads to better performance. The 10.4% improvement from 5 to 50 books demonstrates the importance of sufficient data density, while the subsequent degradation suggests that additional books introduce noise rather than signal.

This finding has significant practical implications for recommendation system deployment. Rather than maintaining large catalogs, systems may benefit from curating smaller, high-quality item sets that maximize user-item interaction density. The optimal range of 40-60 books provides a practical guideline for catalog management in similar recommendation scenarios.

### Cold Start Problem Solutions

All four methods successfully addressed the cold start problem, though through different mechanisms. Memory-based methods leveraged user similarity calculations based on limited rating profiles, while model-based methods employed hybrid approaches combining user similarity with item popularity. The confidence weighting approach proved particularly effective for memory-based methods, allowing them to utilize implicit feedback while maintaining prediction quality.

The successful cold start implementations demonstrate the practical applicability of all methods for real-world deployment, where new users are common and initial rating profiles are sparse.

## Method-Specific Insights

### User-Based Collaborative Filtering Limitations

UBCF's poor performance (RMSE: 3.182) reflects the method's sensitivity to user similarity calculations in sparse environments. With users rating only a small fraction of available books, finding reliable user similarities becomes challenging. The method's performance could potentially improve with larger user bases or more active users, but for this dataset, the sparsity proved too challenging.

### Item-Based Collaborative Filtering Stability

IBCF's moderate performance (RMSE: 2.746) demonstrates its stability advantage over UBCF. Item-item relationships tend to be more stable than user-user relationships, as items have more consistent characteristics across users. This stability makes IBCF more robust in sparse environments, though still inferior to model-based approaches.

### Matrix Factorization Dominance

Matrix Factorization's superior performance (RMSE: 1.820) demonstrates the power of learning latent representations. The method's ability to capture underlying user preferences and item characteristics through 20-dimensional embeddings effectively addresses the sparsity challenge. The optimal hyperparameters (20 factors, learning rate 0.1, L2 regularization 0.01) suggest a balance between model complexity and regularization that prevents overfitting while capturing meaningful patterns.

### Neural Network Potential

The Neural Network's second-place performance (RMSE: 1.899) showcases the potential of deep learning for collaborative filtering. The optimal architecture (Tanh activation, 64-32 neurons, minimal regularization) suggests that complex non-linear patterns exist in the data, though the additional complexity did not provide significant benefits over matrix factorization for this dataset size.

## Limitations and Constraints

### Dataset Characteristics

The analysis is constrained by the specific characteristics of the Book-Crossing dataset. The extreme sparsity (99.2%) and limited catalog size (145 books) may not generalize to other recommendation domains. The findings may be more applicable to niche recommendation scenarios rather than large-scale commercial systems.

### Evaluation Methodology

The focus on RMSE as the primary evaluation metric provides a narrow view of recommendation system performance. Real-world systems often prioritize ranking quality, diversity, and novelty over rating prediction accuracy. The cross-validation approach, while rigorous, may not fully capture the temporal dynamics of user preferences.

### Computational Considerations

The analysis did not extensively evaluate computational efficiency, which is crucial for real-world deployment. Model-based methods, while more accurate, require more computational resources for training and inference compared to memory-based approaches.

## Implications for Practice

### Method Selection Guidelines

For practitioners, the results suggest clear guidelines for method selection:

- **High accuracy requirements**: Use Matrix Factorization or Neural Networks
- **Real-time systems**: Consider Item-Based CF for its stability and scalability
- **Cold start scenarios**: Implement hybrid approaches combining collaborative and content-based filtering
- **Resource constraints**: Balance accuracy against computational requirements

### Catalog Management Insights

The dataset size analysis provides important insights for catalog management. Rather than maintaining large catalogs, systems may benefit from curating smaller, high-quality item sets that maximize user-item interaction density. The optimal range of 40-60 books provides a practical guideline for similar recommendation scenarios.

### Confidence Weighting Effectiveness

The success of confidence weighting in handling implicit feedback demonstrates its value for mixed-feedback recommendation systems. This approach allows systems to leverage implicit signals while maintaining prediction quality, providing a practical solution for real-world scenarios where explicit feedback is limited.

## Future Research Directions

### Hybrid Approaches

The complementary strengths of different methods suggest opportunities for hybrid approaches. Combining Matrix Factorization's accuracy with Item-Based CF's stability could provide robust, high-performance recommendation systems.

### Deep Learning Enhancements

While Neural Networks showed promise, their performance did not significantly exceed Matrix Factorization. Future research could explore more sophisticated architectures, attention mechanisms, or multi-task learning approaches to unlock the full potential of deep learning for collaborative filtering.

### Temporal Dynamics

The current analysis treats user preferences as static. Future work should investigate temporal dynamics, user preference evolution, and the impact of recency on recommendation quality.

### Evaluation Metrics

Expanding beyond RMSE to include ranking metrics, diversity measures, and user satisfaction indicators would provide a more comprehensive evaluation framework for recommendation systems.

\newpage

# Conclusion

This study successfully implemented and compared four collaborative filtering approaches for book recommendations using a modified Book-Crossing dataset. The comprehensive evaluation provides valuable insights for both researchers and practitioners in the recommendation systems domain.

## Key Contributions

### Empirical Performance Rankings

The rigorous cross-validation analysis established a clear performance hierarchy: **Matrix Factorization** (RMSE: 1.820) emerged as the superior method, followed by **Neural Networks** (RMSE: 1.899), **Item-Based Collaborative Filtering** (RMSE: 2.746), and **User-Based Collaborative Filtering** (RMSE: 3.182). This ranking demonstrates the clear advantage of model-based approaches over memory-based methods in sparse recommendation scenarios.

### Dataset Size Optimization

A critical finding emerged from the dataset size analysis: **optimal performance occurs at 50 books rather than the full catalog of 145 books**. This non-monotonic relationship challenges conventional wisdom about data quantity, revealing that beyond a certain point, additional items introduce noise rather than signal. The 10.4% improvement from 5 to 50 books, followed by performance degradation, provides practical guidance for catalog management in recommendation systems.

### Cold Start Solutions

All four methods successfully addressed the cold start problem through distinct approaches. Memory-based methods leveraged user similarity calculations, while model-based methods employed hybrid strategies combining user similarity with item popularity. The confidence weighting approach proved particularly effective for handling mixed explicit-implicit feedback, demonstrating practical applicability for real-world deployment.

### Confidence Weighting Innovation

The implementation of confidence weighting for implicit feedback represents a significant methodological contribution. By assigning confidence scores of 0.5 to implicit ratings (converted to rating 4) and 1.0 to explicit ratings, the approach successfully leveraged implicit signals while maintaining prediction quality—a crucial capability for sparse recommendation environments.

## Practical Implications

### Method Selection Guidelines

For practitioners, the results provide clear decision criteria:

- **Maximum Accuracy**: Implement Matrix Factorization with 20 latent factors, learning rate 0.1, and L2 regularization 0.01
- **Real-time Requirements**: Consider Item-Based CF for its stability and computational efficiency
- **Cold Start Scenarios**: Deploy hybrid approaches combining collaborative and content-based filtering
- **Resource Constraints**: Balance accuracy against computational requirements based on specific use cases

### Catalog Management Strategy

The dataset size analysis suggests a paradigm shift from "bigger is better" to "optimal size is best." Rather than maintaining large catalogs, systems should focus on curating smaller, high-quality item sets (40-60 items) that maximize user-item interaction density. This approach can improve both prediction accuracy and computational efficiency.

### Implementation Considerations

The confidence weighting methodology provides a practical solution for mixed-feedback systems, enabling effective utilization of implicit signals while preserving explicit rating integrity. This approach is particularly valuable for systems where explicit feedback is limited but implicit signals are abundant.

## Research Contributions

### Methodological Rigor

The study employed consistent evaluation protocols across all methods, ensuring fair comparison through identical train-test splits, cross-validation procedures, and evaluation metrics. This methodological rigor provides a reliable foundation for future comparative studies in collaborative filtering.

### Comprehensive Implementation

The complete implementation of four distinct collaborative filtering approaches, including custom-coded memory-based methods and optimized model-based approaches, provides a comprehensive framework for understanding method trade-offs and implementation challenges.

### Empirical Evidence

The empirical evidence challenges several assumptions in the recommendation systems literature, particularly regarding the relationship between dataset size and performance, and the effectiveness of different approaches in sparse environments.

## Limitations and Future Work

### Dataset Constraints

The analysis is limited by the specific characteristics of the Book-Crossing dataset, including extreme sparsity (99.2%) and limited catalog size (145 books). Future work should validate findings across diverse datasets and domains to establish generalizability.

### Evaluation Scope

The focus on RMSE as the primary metric provides a narrow view of recommendation system performance. Future research should incorporate ranking metrics, diversity measures, and user satisfaction indicators to provide a more comprehensive evaluation framework.

### Temporal Dynamics

The current analysis treats user preferences as static. Future work should investigate temporal dynamics, user preference evolution, and the impact of recency on recommendation quality.

## Final Recommendations

For researchers, this study provides a solid foundation for understanding collaborative filtering trade-offs in sparse environments. The clear performance rankings and methodological insights offer valuable guidance for future research directions.

For practitioners, the study offers practical guidelines for method selection, catalog management, and implementation strategies. The confidence weighting approach and dataset size optimization findings provide immediate value for real-world recommendation system deployment.

The findings demonstrate that **Matrix Factorization achieves superior prediction accuracy** while each approach offers unique advantages depending on deployment constraints. The confidence-weighted similarity approach effectively handles implicit feedback, improving prediction quality across memory-based methods.

This research provides a solid foundation for developing production-ready book recommendation systems, with extensible code and methodologies applicable to other domains. The comprehensive evaluation framework and practical insights contribute to both the academic understanding and practical implementation of collaborative filtering systems.






```{r cleanup, include=FALSE}
# Clean up H2O cluster at the end
if (exists("h2o") && h2o::h2o.clusterIsUp()) {
  h2o::h2o.shutdown(prompt = FALSE)
}
```





# writing

Exploratory Data Analysis
The dataset exhibits characteristics typical of recommendation system challenges (Figure 1). The distribution of explicit ratings (Panel A) shows a right-skewed pattern with a mean of 7.5, indicating users tend to rate books they enjoy rather than books they dislike. User activity follows a long-tail distribution (Panel B), where the majority of users have rated fewer than 10 books while a small subset of highly active users have rated 20 or more—highlighting the sparsity problem central to collaborative filtering. Book popularity analysis (Panel C) reveals no strong correlation between the number of ratings a book receives and its average rating quality, suggesting popularity bias is not a major concern for this dataset. Finally, the dataset contains both explicit ratings (1-10 scale, 65.2%) and implicit feedback (rating of 0 indicating "read but not rated", 34.8%) as shown in Panel D. To handle this mixed feedback, we implement confidence weighting in our similarity calculations, assigning lower confidence to implicit ratings while still leveraging them to reduce sparsity. After applying minimum rating thresholds, 58% of users were excluded for having fewer than 3 ratings, while all 145 books in the dataset met the minimum threshold of 5 ratings per book, resulting in a final dataset with 99.2% sparsity across 4,200 users and 145 books.



Alternative shorter version (if you want it more concise):
The dataset exhibits key challenges for collaborative filtering (Figure 1). 
Explicit ratings (Panel A) show a right-skewed distribution (mean = 7.5), reflecting user preference to rate enjoyed books. User activity (Panel B) follows a long-tail pattern, with most users rating fewer than 10 books—the core sparsity challenge. Book popularity (Panel C) shows no strong relationship with average ratings, mitigating popularity bias concerns. The dataset contains 65.2% explicit ratings and 34.8% implicit feedback (Panel D), which we handle through confidence weighting in similarity calculations. After filtering, the final dataset comprises 4,200 users and 145 books with 99.2% sparsity.




# Appendix: Detailed Exploratory Data Analysis
```{r appendix-detailed-eda}
# ============================================================================
# DETAILED USER AND BOOK ACTIVITY ANALYSIS
# ============================================================================

# User activity analysis
user_activity <- data %>%
  group_by(User.ID) %>%
  summarise(
    num_ratings = n(),
    avg_rating = mean(Book.Rating, na.rm = TRUE),
    .groups = "drop"
  )

# Book popularity analysis
book_popularity <- data %>%
  group_by(ISBN) %>%
  summarise(
    num_ratings = n(),
    avg_rating = mean(Book.Rating, na.rm = TRUE),
    .groups = "drop"
  )

# User activity summary table
user_activity_summary <- data.frame(
  Metric = c(
    "Total Users",
    "Min Ratings per User", 
    "Max Ratings per User",
    "Mean Ratings per User",
    "Users with 1-2 Ratings",
    "Users with 3-5 Ratings", 
    "Users with 6-10 Ratings",
    "Users with 11-20 Ratings",
    "Users with >20 Ratings"
  ),
  Value = c(
    nrow(user_activity),
    min(user_activity$num_ratings),
    max(user_activity$num_ratings),
    round(mean(user_activity$num_ratings), 2),
    sum(user_activity$num_ratings <= 2),
    sum(user_activity$num_ratings >= 3 & user_activity$num_ratings <= 5),
    sum(user_activity$num_ratings >= 6 & user_activity$num_ratings <= 10),
    sum(user_activity$num_ratings >= 11 & user_activity$num_ratings <= 20),
    sum(user_activity$num_ratings > 20)
  )
)

kable(user_activity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed User Activity Distribution",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Book popularity summary table
book_popularity_summary <- data.frame(
  Metric = c(
    "Total Books",
    "Min Ratings per Book",
    "Max Ratings per Book", 
    "Mean Ratings per Book",
    "Books with 1-4 Ratings",
    "Books with 5-9 Ratings",
    "Books with 10-19 Ratings", 
    "Books with 20-49 Ratings",
    "Books with ≥50 Ratings"
  ),
  Value = c(
    nrow(book_popularity),
    min(book_popularity$num_ratings),
    max(book_popularity$num_ratings),
    round(mean(book_popularity$num_ratings), 2),
    sum(book_popularity$num_ratings <= 4),
    sum(book_popularity$num_ratings >= 5 & book_popularity$num_ratings <= 9),
    sum(book_popularity$num_ratings >= 10 & book_popularity$num_ratings <= 19),
    sum(book_popularity$num_ratings >= 20 & book_popularity$num_ratings <= 49),
    sum(book_popularity$num_ratings >= 50)
  )
)

kable(book_popularity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Book Popularity Distribution",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Rating distribution (numeric breakdown)
rating_dist <- table(data$Book.Rating)
rating_dist_df <- data.frame(
  Rating = names(rating_dist),
  Count = as.numeric(rating_dist),
  Percentage = round(as.numeric(rating_dist) / sum(rating_dist) * 100, 1)
)

kable(rating_dist_df, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Rating Distribution",
      align = c("c", "c", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Dataset sparsity analysis
total_possible_ratings <- length(unique(data$User.ID)) * length(unique(data$ISBN))
actual_ratings <- nrow(data)
sparsity_percentage <- round((1 - actual_ratings / total_possible_ratings) * 100, 2)

sparsity_summary <- data.frame(
  Metric = c(
    "Total Possible User-Book Pairs",
    "Actual Ratings",
    "Sparsity Percentage"
  ),
  Value = c(
    format(total_possible_ratings, big.mark = ","),
    format(actual_ratings, big.mark = ","),
    paste0(sparsity_percentage, "%")
  )
)

kable(sparsity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Dataset Sparsity Analysis",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```
