---
title: "Ensemble Recommender System for Book Recommendations: A Comparative Analysis of Collaborative Filtering Approaches"
author: "Hope Hennessy"
date: "2025-10-01"
output:
  pdf_document:
    latex_engine: xelatex
---

# Abstract

This study compares four collaborative filtering approaches for book recommendation systems using a modified Book-Crossing dataset containing 45,370 ratings from 10,000 users across 145 books. Item-based (IBCF) and user-based (UBCF) collaborative filtering models were implemented from scratch, together with matrix factorization (MF) and neural network (NN) models. To address the challenge of mixed feedback in a highly sparse environment with 99.2% sparsity, confidence-weighted similarity measures were employed to integrate both explicit ratings on a 1–10 scale and implicit feedback represented by zero ratings. Cross-validation revealed clear performance differences, with MF achieving the lowest RMSE of 1.820 (SD = 0.020), followed by NN at 1.899 (SD = 0.053), IBCF at 2.746 (SD = 0.018), and UBCF at 3.182 (SD = 0.047). Model-based approaches demonstrated greater robustness to sparsity and achieved a 43 percent reduction in prediction error relative to memory-based methods. Cold-start analysis showed an 82.5 percent divergence in recommendations across algorithms, reflecting fundamentally different preference modeling strategies. Catalog size analysis indicated that predictive accuracy improved up to about 50 books, after which performance gains became marginal. These results provide practical guidance for designing efficient and accurate recommendation systems in data-sparse environments.




# 1. Introduction


The rapid expansion of digital content has created an environment of overwhelming information, making it increasingly difficult for users to discover relevant material. Recommender systems address this challenge by predicting user preferences and presenting personalized suggestions, thereby improving engagement and satisfaction. Among the various approaches developed, collaborative filtering (CF) remains one of the most widely adopted and successful paradigms. CF infers user interests from historical patterns of user–item interactions and underpins many large-scale platforms such as Amazon, Netflix, and Goodreads, where personalized recommendations play a central role in driving user retention and commercial success.


This study utilizes a modified version of the Book-Crossing dataset, which contains 45,370 ratings from 10,000 users across 145 books. Ratings are expressed on a 0–10 scale (inclusive), where scores from 1 to 10 represent explicit feedback and zeros indicate implicit feedback, signifying that a book was read but not explicitly rated. The presence of both feedback types, combined with extreme sparsity (99.2%), presents unique modeling challenges. This dual nature of feedback introduces additional complexity, as implicit data provide weaker preference signals yet contribute valuable behavioral information that can enhance model performance. The dataset's extreme sparsity means that the vast majority of potential user–book interactions are unobserved, presenting major challenges for collaborative filtering, particularly when constructing accurate similarity measures and learning stable latent representations.


The primary objective of this study is to build and evaluate an ensemble of collaborative filtering models capable of predicting user ratings and generating personalized book recommendations. The analysis implements and assesses four collaborative filtering algorithms — user-based (UBCF) and item-based (IBCF) models developed from scratch, alongside matrix factorization (MF) and neural network (NN) models implemented using efficient frameworks — comparing their predictive performance through cross-validation to analyze their relative strengths and weaknesses. Additionally, the study examines the cold-start problem by evaluating algorithmic behavior when limited user feedback is available and investigates how catalog size influences predictive accuracy, identifying the threshold beyond which additional titles yield diminishing returns in model performance. Through these objectives, the study seeks to generate evidence-based insights on the trade-offs between accuracy, scalability, and interpretability in real-world recommendation systems.


Collaborative filtering techniques can be broadly classified into memory-based and model-based approaches. Memory-based algorithms, including UBCF and IBCF, compute user–user or item–item similarities directly from the user–item matrix and predict ratings using weighted averages of neighbors. These methods are simple, interpretable, and well-suited for smaller datasets, though they tend to perform poorly in highly sparse or large-scale environments.
Model-based methods, represented here by matrix factorization and neural network models, learn latent representations that capture hidden preference structures within lower-dimensional spaces. These techniques are more robust to sparsity and can generalize better to unseen data, albeit with higher computational costs and reduced interpretability.

To mitigate the limitations of memory-based methods in sparse environments, this study incorporates confidence-weighted similarity measures within UBCF and IBCF to account for implicit feedback, thereby improving their ability to model uncertainty and leverage weaker signals. Model-based approaches (MF and NN) train exclusively on explicit ratings. 



# 2. Data Understanding and Preparation

## 2.1 Data Description

The dataset consists of three components: book_ratings, which records user–book interactions with variables User.ID, ISBN, and Book.Rating on a 0–10 scale; book_info, which provides metadata such as Book.Title and Book.Author; and user_info, which contains limited demographic details like age but is excluded from this analysis to focus on collaborative filtering. Ratings from 1–10 represent explicit feedback, where users actively rated books, while a rating of 0 indicates implicit feedback, meaning the user read the book but did not provide a score.



```{r setup, include=FALSE}
# ============================================================================
# GLOBAL SETTINGS AND CONFIGURATION
# ============================================================================

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6,
  fig.show = "hold",
  out.width = "80%",
  dpi = 300,
  crop = NULL
)

# Set random seed for reproducibility
set.seed(123)

# ============================================================================
# LIBRARY IMPORTS
# ============================================================================

library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(cowplot)

library(knitr)
library(kableExtra)

library(caret)
library(recosystem)
library(h2o)

options(
  lintr.linter_file = "none",
  lintr.exclude_linters = c(
    "object_usage_linter",
    "object_name_linter",
    "cyclocomp_linter",
    "line_length_linter"
  )
)


# ============================================================================
# HELPER FUNCTIONS FOR TABLE FORMATTING
# ============================================================================

# Function to truncate long text for table display
truncate_text <- function(text, max_length = 50) {
  ifelse(nchar(text) > max_length, 
         paste0(substr(text, 1, max_length - 3), "..."), 
         text)
}

# Function to format recommendation tables consistently
format_recommendation_table <- function(data, caption, method_name) {
  data %>%
    mutate(Rank = row_number()) %>%
    mutate(
      Book.Title = truncate_text(Book.Title, 50),
      Book.Author = truncate_text(Book.Author, 25)
    ) %>%
    select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
    kable(
      format = "latex",
      booktabs = TRUE,
      caption = paste(method_name, caption),
      digits = 2,
      align = c("c", "l", "l", "c"),
      col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
    ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "scale_down"),
      font_size = 10
    )
}

# Load saved results
load('nn_tuning_results.Rdata')
load('mf_tuning_results.Rdata')
load('mf_final_model_mf.Rdata')
load('mf_tuning_results.Rdata')
load('all_cv_results.Rdata')
load('dataset_size_analysis_complete.Rdata')

```

## 2.2 Exploratory Data Analysis

As shown in Table 1, the raw dataset contains 10,000 users, 145 books, and 45,370 total interactions, of which 18,299 (40.3%) are explicit ratings and 27,071 (59.7%) are implicit feedback. The mean for explicit ratings is 7.84 on the 1-10 scale, reflecting a positivity bias common in user-generated ratings. The initial user-item matrix exhibits 96.87% sparsity, meaning that only 3.13% of all possible user-book pairs have recorded interactions. This extreme sparsity is typical of recommendation system datasets and motivates the methodological choices described in subsequent sections.

The dataset exhibits characteristics typical of recommendation system challenges (Figure 1). Panel A shows the distribution of explicit ratings, which is right-skewed with a mean of 7.8, indicating that users tend to rate books they enjoy rather than books they dislike—a common positivity bias in explicit rating systems that can affect baseline predictions. Panel B reveals that user activity follows a long-tail distribution, with most users rating fewer than 10 books and a small subset of highly active users rating 20 or more, highlighting the sparsity problem central to collaborative filtering. Panel C examines book popularity and shows no strong correlation between the number of ratings a book receives and its average rating, suggesting that while users generally rate positively, this positivity does not systematically favor popular books over niche titles, thereby mitigating popularity bias concerns. Finally, Panel D indicates that 40.3% of the dataset consists of explicit ratings and 59.7% implicit feedback, presenting a mixed-feedback challenge that requires careful handling in model design.

Detailed statistical breakdowns of user activity, book popularity, and rating distributions are provided in Appendix A.

```{r data-loading, include=FALSE}
# ============================================================================
# DATA LOADING
# ============================================================================

load("book_ratings.Rdata")

head(book_info)
str(book_info)
dim(book_info)
head(book_ratings)
str(book_ratings)
dim(book_ratings)
head(user_info) # will not be using age in this report
str(user_info)
dim(user_info)

# Check for missing values
missing_summary <- data.frame(
  Dataset = c("Book Info", "Book Ratings", "User Info"),
  Missing_Values = c(
    sum(is.na(book_info)),
    sum(is.na(book_ratings)),
    sum(is.na(user_info))   # missing age values
  )
)

```

```{r data-cleaning}
# ============================================================================
# DUPLICATE DETECTION AND HANDLING
# ============================================================================

# Initial data merging
initial_data <- book_ratings %>%
  left_join(book_info, by = "ISBN")

# Clean book titles and authors for duplicate detection
book_info_clean <- book_info %>%
  mutate(
    clean_title = Book.Title %>%
      tolower() %>%
      gsub('\\(.*?\\)', '', .) %>%
      gsub('\\s*:\\s*.*$', '', .) %>%
      gsub('^(the|a|an)\\s+', '', .) %>%
      gsub('[^a-z0-9\\s-]', '', .) %>%
      gsub('\\s+', ' ', .) %>%
      trimws(),
    clean_author = Book.Author %>%
      tolower() %>%
      gsub('\\s+jr\\.*$', '', .) %>%
      gsub('\\s+sr\\.*$', '', .) %>%
      gsub('\\s+(iii|ii)$', '', .) %>%
      gsub('\\s+', ' ', .) %>%
      trimws()
  )

# Identify duplicates
duplicates <- book_info_clean %>%
  group_by(clean_title, clean_author) %>%
  filter(n() > 1) %>%
  arrange(clean_title, clean_author, ISBN) %>%
  ungroup()

# ============================================================================
# MERGE RATINGS FUNCTION
# ============================================================================

merge_ratings <- function(ratings_data, book_info_data, duplicates_data) {
  
  if (nrow(duplicates_data) > 0) {
    # Create canonical ISBN mapping
    duplicate_mapping <- duplicates_data %>%
      group_by(clean_title, clean_author) %>%
      arrange(ISBN) %>%
      mutate(canonical_ISBN = first(ISBN)) %>%
      ungroup() %>%
      select(ISBN, canonical_ISBN)
    
    # Create complete mapping
    all_books_mapping <- book_info_data %>%
      select(ISBN) %>%
      left_join(duplicate_mapping, by = "ISBN") %>%
      mutate(canonical_ISBN = ifelse(is.na(canonical_ISBN), ISBN, canonical_ISBN))
    
    # Apply mapping to ratings
    ratings_mapped <- ratings_data %>%
      left_join(all_books_mapping, by = "ISBN") %>%
      mutate(
        original_ISBN = ISBN,
        ISBN = canonical_ISBN
      ) %>%
      select(-canonical_ISBN, -original_ISBN)
    
    # Keep highest rating per user-book combination
    ratings_merged <- ratings_mapped %>%
      group_by(User.ID, ISBN) %>%
      slice_max(Book.Rating, n = 1, with_ties = FALSE) %>%
      ungroup()
    
    # Update book info with cleaned titles and authors
    book_info_updated <- book_info_data %>%
      left_join(all_books_mapping, by = "ISBN") %>%
      filter(ISBN == canonical_ISBN) %>%
      mutate(
        Book.Title = Book.Title %>%
          gsub('\\(.*?\\)', '', .) %>%  # Remove parenthetical info
          gsub('\\s*:\\s*.*$', '', .) %>%  # Remove subtitle after colon
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws(),  # Remove leading/trailing whitespace
        Book.Author = Book.Author %>%
          gsub('\\s+jr\\.*$', '', .) %>%  # Remove "Jr." suffixes
          gsub('\\s+sr\\.*$', '', .) %>%  # Remove "Sr." suffixes
          gsub('\\s+(iii|ii)$', '', .) %>%  # Remove "II", "III" suffixes
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws()  # Remove leading/trailing whitespace
      ) %>%
      select(ISBN, Book.Title, Book.Author)
    
  } else {
    ratings_merged <- ratings_data
    # Clean all book titles and authors even when no duplicates
    book_info_updated <- book_info_data %>%
      mutate(
        Book.Title = Book.Title %>%
          gsub('\\(.*?\\)', '', .) %>%  # Remove parenthetical info
          gsub('\\s*:\\s*.*$', '', .) %>%  # Remove subtitle after colon
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws(),  # Remove leading/trailing whitespace
        Book.Author = Book.Author %>%
          gsub('\\s+jr\\.*$', '', .) %>%  # Remove "Jr." suffixes
          gsub('\\s+sr\\.*$', '', .) %>%  # Remove "Sr." suffixes
          gsub('\\s+(iii|ii)$', '', .) %>%  # Remove "II", "III" suffixes
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws()  # Remove leading/trailing whitespace
      ) %>%
      select(ISBN, Book.Title, Book.Author)
    all_books_mapping <- NULL
  }
  
  return(list(
    ratings_merged = ratings_merged,
    book_info_updated = book_info_updated,
    mapping = all_books_mapping
  ))
}

# Execute merge
merged_result <- merge_ratings(book_ratings, book_info, duplicates)

# Create final clean dataset
data <- merged_result$ratings_merged %>%
  left_join(merged_result$book_info_updated, by = "ISBN") %>%
  select(User.ID, ISBN, Book.Rating, Book.Title, Book.Author)

book_info <- merged_result$book_info_updated
```

## 2.3 Data Preprocessing and Filtering

Duplicate book entries, where the same title and author appeared under different ISBNs, were identified using normalized string matching. Titles and author names were converted to lowercase, with parenthetical text, subtitles, special characters, and extra whitespace removed, and author suffixes standardized. All duplicates were mapped to a canonical ISBN (the first alphabetically), and when users rated multiple versions of the same book, the highest rating was retained. This cleaning process reduced ambiguity and ensured consistent, high-quality data for collaborative filtering. 

To ensure meaningful and reliable collaborative filtering, we applied a dual-threshold filtering strategy that balanced data quality with dataset size. Books were required to have at least five ratings to ensure adequate interaction data for similarity calculations, and users were required to have rated at least three books to prevent sparse profiles from introducing noise. This process excluded approximately 58% of users with insufficient rating histories, leaving 4,200 active users, while all 145 books met the inclusion threshold.

After applying these filtering criteria, the resulting user-item matrix retained 45,370 interactions but exhibited increased sparsity of 99.2% due to the reduced user base relative to the complete catalog. This sparsity level reflects the naturally sparse structure typical of real-world recommendation systems, where most users interact with only a small fraction of available items. Table 1 summarizes the final dataset statistics used for all subsequent model training and evaluation.


```{r eda-summary}
# ============================================================================
# DATASET SUMMARY STATISTICS
# ============================================================================
# Calculate explicit ratings statistics
explicit_ratings <- data %>% filter(Book.Rating > 0)

# Basic statistics
data_summary <- data.frame(
  Metric = c(
    "Total Users",
    "Total Books",
    "Total Interactions",
    "Explicit Ratings",
    "Implicit Ratings",
    "Average Explicit Rating",
    "Explicit Rating Range",
    "Matrix Sparsity"
  ),
  Value = c(
    length(unique(data$User.ID)),
    length(unique(data$ISBN)),
    nrow(data),
    nrow(explicit_ratings),
    sum(data$Book.Rating == 0),
    round(mean(explicit_ratings$Book.Rating, na.rm = TRUE), 2),
    paste(min(explicit_ratings$Book.Rating), "-", max(explicit_ratings$Book.Rating)),
    paste0(
      round(
        (1 - nrow(data) / 
          (length(unique(data$User.ID)) * length(unique(data$ISBN)))) * 100,
        2
      ),
      "%"
    )
  )
)

kable(data_summary,
      format = "latex",
      booktabs = TRUE,
      caption = "Dataset Overview: Summary statistics for book recommendation analysis including user count, book count, rating statistics (explicit only), and matrix sparsity",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

```{r eda-visualizations, fig.cap="Key Dataset Characteristics: (A) Explicit rating distribution (mean 7.8), (B) User activity patterns showing long-tail distribution, (C) Book popularity versus average rating, (D) Explicit (40.3%) versus implicit (59.7%) feedback proportions. Confidence weighting (1.0 for explicit, 0.5 for implicit) enables leveraging all interactions while appropriately weighting signal quality."}

  
# ============================================================================
# FOCUSED VISUALIZATIONS: MOST INFORMATIVE FOR RECOMMENDER SYSTEMS
# ============================================================================

# 1. EXPLICIT RATING DISTRIBUTION (Most Important)
explicit_data <- data %>% filter(Book.Rating > 0)

p1 <- ggplot(explicit_data, aes(x = Book.Rating)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(explicit_data$Book.Rating), 
             linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = mean(explicit_data$Book.Rating), y = max(table(explicit_data$Book.Rating)) * 0.9, 
           label = paste("Mean =", round(mean(explicit_data$Book.Rating), 1)), 
           color = "red", hjust = -0.1, size = 3.5) +
  labs(
    title = "Distribution of Explicit Ratings",
    x = "Rating",
    y = "Frequency"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 2. SPARSITY ANALYSIS (Critical for CF)
sparsity_data <- data %>%
  group_by(User.ID) %>%
  summarise(
    total_books_rated = n(),
    explicit_ratings = sum(Book.Rating > 0),
    implicit_ratings = sum(Book.Rating == 0),
    .groups = "drop"
  ) %>%
  mutate(
    explicit_ratio = explicit_ratings / total_books_rated,
    user_type = case_when(
      explicit_ratio == 1 ~ "Only Explicit",
      explicit_ratio == 0 ~ "Only Implicit", 
      explicit_ratio > 0.5 ~ "Mostly Explicit",
      TRUE ~ "Mostly Implicit"
    )
  )

p2 <- ggplot(sparsity_data, aes(x = total_books_rated)) +
  geom_histogram(binwidth = 1, fill = "lightcoral", color = "white", alpha = 0.8) +
  scale_x_continuous(limits = c(0, quantile(sparsity_data$total_books_rated, 0.95))) +
  labs(
    title = "User Activity Distribution",
    x = "Total Books per User",
    y = "Number of Users"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 3. BOOK POPULARITY (Important for Cold-Start)
book_popularity <- data %>%
  group_by(ISBN) %>%
  summarise(
    total_interactions = n(),
    explicit_ratings = sum(Book.Rating > 0),
    avg_rating = mean(Book.Rating[Book.Rating > 0], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(avg_rating))  # Only books with explicit ratings

p3 <- ggplot(book_popularity, aes(x = explicit_ratings, y = avg_rating)) +
  geom_point(alpha = 0.6, color = "darkgreen", size = 2) +
  labs(
    title = "Book Popularity vs Average Rating",
    x = "Number of Explicit Ratings",
    y = "Average Rating"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 4. EXPLICIT VS IMPLICIT RATIO (Key Insight)
interaction_summary <- data %>%
  summarise(
    total_interactions = n(),
    explicit_count = sum(Book.Rating > 0),
    implicit_count = sum(Book.Rating == 0),
    explicit_pct = round(mean(Book.Rating > 0) * 100, 1),
    implicit_pct = round(mean(Book.Rating == 0) * 100, 1)
  )

# Create a simple bar chart for the ratio
ratio_data <- data.frame(
  Type = c("Explicit Ratings", "Implicit Ratings"),
  Count = c(interaction_summary$explicit_count, interaction_summary$implicit_count),
  Percentage = c(interaction_summary$explicit_pct, interaction_summary$implicit_pct)
)

p4 <- ggplot(ratio_data, aes(x = Type, y = Count, fill = Type)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(Percentage, "%")), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Explicit Ratings" = "steelblue", "Implicit Ratings" = "lightcoral")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Explicit vs Implicit Feedback",
    x = "Feedback Type",
    y = "Count"
  ) +
  theme_cowplot() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )

# Create the combined plot with legend
combined_plot <- plot_grid(
  p1, p2,
  p3, p4,
  ncol = 2, nrow = 2,
  labels = c("A", "B", "C", "D"),
  label_size = 12
)

# Display the plot
combined_plot
```


# 3. Methodology

## 3.1 Evaluation Framework

Performance was evaluated using 5-fold cross-validation with root mean squared error (RMSE) as the primary metric. We selected RMSE as our primary metric because it penalizes large prediction errors more heavily than mean absolute error (MAE), which is critical in recommendation systems where large mispredictions can severely degrade user experience. Importantly, RMSE was calculated only on explicit ratings (1–10 scale), excluding implicit feedback (0 ratings), to measure true prediction accuracy on user-expressed preferences. 

For each fold, 80% of the observed ratings were assigned to the training set and 20% to the test set. Test ratings were temporarily hidden in the training matrices by setting them to NA. The model was then trained on the remaining 80% of ratings, and predictions were generated for the hidden 20%. This ensures that the model does not see the test ratings during training, providing an unbiased assessment of predictive accuracy. Confidence weighting was applied consistently across memory-based methods to allow fair comparison. 

## 3.2 User-Item Matrix Construction and Implicit Feedback

To address the challenge of mixed feedback types, three matrices were constructed. First, a ratings matrix converted implicit ratings (originally 0) to a value of 4, providing a neutral-positive signal for user engagement without explicit ratings. This value was chosen to represent moderate interest—lower than the explicit ratings mean (7.84) but acknowledging that implicit feedback indicates some level of engagement. Second, a confidence matrix assigned weights of 1.0 for explicit ratings (1–10) and 0.5 for implicit ratings, allowing implicit feedback to contribute to similarity calculations while appropriately down-weighting its influence. This confidence-weighting strategy was applied specifically to memory-based methods (UBCF and IBCF) during similarity computation. Third, the original ratings matrix, retaining raw 0–10 values, was preserved for evaluation purposes, ensuring RMSE calculations reflected performance on explicit ratings only. This multi-matrix approach addresses sparsity by leveraging all user-item interactions while respecting the differing informational content of explicit and implicit feedback.

```{r matrix-construction}
# ============================================================================
# USER-ITEM MATRIX CREATION FUNCTION
# ============================================================================

create_user_item_matrix <- function(ratings_data,
                                     min_ratings_per_book = 3,
                                     min_ratings_per_user = 2,
                                     implicit_rating = 4,
                                     implicit_confidence = 0.5) {
  
  # Handle implicit ratings (rating = 0)
  ratings_clean <- ratings_data %>%
    mutate(
      is_implicit = (Book.Rating == 0),
      original_rating = Book.Rating,
      Book.Rating = ifelse(Book.Rating == 0, implicit_rating, Book.Rating),
      confidence = ifelse(is_implicit, implicit_confidence, 1.0)
    )
  
  # Create matrices
  user_item_matrix <- ratings_clean %>%
    select(User.ID, ISBN, Book.Rating) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = Book.Rating,
      values_fill = NA
    )
  
  confidence_matrix <- ratings_clean %>%
    select(User.ID, ISBN, confidence) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = confidence,
      values_fill = NA
    )
  
  original_ratings_matrix <- ratings_clean %>%
    select(User.ID, ISBN, original_rating) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = original_rating,
      values_fill = NA
    )
  
  # Convert to matrix format
  user_ids <- user_item_matrix$User.ID
  user_item_matrix <- as.matrix(user_item_matrix[, -1])
  confidence_matrix <- as.matrix(confidence_matrix[, -1])
  original_ratings_matrix <- as.matrix(original_ratings_matrix[, -1])
  
  rownames(user_item_matrix) <- user_ids
  rownames(confidence_matrix) <- user_ids
  rownames(original_ratings_matrix) <- user_ids
  
  # Filter by minimum ratings
  books_to_keep <- colSums(!is.na(user_item_matrix)) >= min_ratings_per_book
  user_item_matrix <- user_item_matrix[, books_to_keep]
  confidence_matrix <- confidence_matrix[, books_to_keep]
  original_ratings_matrix <- original_ratings_matrix[, books_to_keep]
  
  users_to_keep <- rowSums(!is.na(user_item_matrix)) >= min_ratings_per_user
  user_item_matrix <- user_item_matrix[users_to_keep, ]
  confidence_matrix <- confidence_matrix[users_to_keep, ]
  original_ratings_matrix <- original_ratings_matrix[users_to_keep, ]
  
  return(list(
    ratings = user_item_matrix,
    confidence = confidence_matrix,
    original_ratings = original_ratings_matrix
  ))
}

# ============================================================================
# EVALUATION FUNCTION
# ============================================================================

evaluate_predictions <- function(pred, actual, original_ratings) {
  # Separate explicit and implicit ratings
  is_implicit <- (original_ratings == 0)
  explicit_mask <- !is_implicit & !is.na(actual)
  
  # Calculate RMSE on explicit ratings only
  rmse_explicit <- sqrt(mean((pred[explicit_mask] - actual[explicit_mask])^2, na.rm = TRUE))
  rmse_overall <- sqrt(mean((pred - actual)^2, na.rm = TRUE))
  
  return(list(
    rmse_explicit = rmse_explicit,
    rmse_overall = rmse_overall,
    n_explicit = sum(explicit_mask),
    n_implicit = sum(is_implicit)
  ))
}

# ============================================================================
# UNIFIED MATRIX CREATION FOR ALL METHODS
# ============================================================================

# Create matrices once with consistent parameters
matrix_result <- create_user_item_matrix(
  data,
  min_ratings_per_book = 5,
  min_ratings_per_user = 3,
  implicit_rating = 4,
  implicit_confidence = 0.5
)

user_item_matrix <- matrix_result$ratings
confidence_matrix <- matrix_result$confidence
original_ratings_matrix <- matrix_result$original_ratings

```

```{r common-functions}
# ============================================================================
# NORMALIZATION FUNCTIONS
# ============================================================================

# User-mean normalization (for UBCF and NN)
normalize_matrix_user <- function(user_item_matrix) {
  user_means <- rowMeans(user_item_matrix, na.rm = TRUE)
  user_item_matrix_normalized <- sweep(user_item_matrix, 1, user_means, FUN = "-")
  
  return(list(
    normalized = user_item_matrix_normalized,
    means = user_means
  ))
}

# Item-mean normalization (for IBCF)
normalize_matrix_item <- function(user_item_matrix) {
  item_means <- colMeans(user_item_matrix, na.rm = TRUE)
  user_item_matrix_normalized <- sweep(user_item_matrix, 2, item_means, FUN = "-")
  
  return(list(
    normalized = user_item_matrix_normalized,
    means = item_means
  ))
}
```



## 3.3 Collaborative Filtering Approaches

### 3.3.1 User-Based Collaborative Filtering (UBCF)

UBCF estimates a user's rating for an item by leveraging ratings from similar users, based on the principle that users with similar past preferences are likely to share future preferences. The method consists of three main steps: normalization, similarity computation, and prediction. First, user-mean normalization was applied to remove individual rating biases, centering each user's ratings around zero to make users with different rating scales directly comparable. Second, confidence-weighted cosine similarity was computed between users, where each co-rated item was weighted by its confidence value (1.0 for explicit ratings, 0.5 for implicit ratings). This approach identifies users with comparable rating patterns while accounting for the reliability of different feedback types. Third, predictions employed k-nearest neighbors ($k=50$) filtering to aggregate ratings from the most similar users while reducing noise from weakly similar users. Predicted ratings combine the target user's mean rating with a weighted average of normalized ratings from the most similar users, where weights reflect both user similarity and rating confidence.

```{r ubcf-similarity}
# ============================================================================
# UBCF: USER SIMILARITY COMPUTATION
# ============================================================================

compute_user_similarity_matrix <- function(user_item_matrix_normalized,
                                            confidence_matrix) {
  
  # Prepare matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  
  # Weighted ratings
  weighted_mat <- mat * conf
  
  # Compute cosine similarity
  numerator <- weighted_mat %*% t(weighted_mat)
  magnitudes <- sqrt(rowSums(weighted_mat^2))
  denominator <- outer(magnitudes, magnitudes)
  
  user_similarity_matrix <- numerator / denominator
  user_similarity_matrix[is.nan(user_similarity_matrix)] <- 0
  diag(user_similarity_matrix) <- 0
  
  rownames(user_similarity_matrix) <- rownames(user_item_matrix_normalized)
  colnames(user_similarity_matrix) <- rownames(user_item_matrix_normalized)
  
  return(user_similarity_matrix)
}
```

```{r ubcf-recommendations}
# ============================================================================
# UBCF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_ubcf <- function(target_user,
                                     user_item_matrix,
                                     user_item_matrix_normalized,
                                     confidence_matrix,
                                     user_sim_matrix,
                                     user_means,
                                     book_info,
                                     n_recommendations = 10,
                                     k = NULL) {
  
  target_user <- as.character(target_user)
  
  # Find unrated books
  unrated_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get similarities
  sims <- user_sim_matrix[target_user, ]
  sims[is.na(sims) | is.nan(sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < length(sims)) {
    non_zero_count <- sum(sims != 0)
    if (non_zero_count > 0) {
      k_actual <- min(k, non_zero_count)
      top_k_users <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
      sims_filtered <- rep(0, length(sims))
      names(sims_filtered) <- names(sims)
      sims_filtered[top_k_users] <- sims[top_k_users]
      sims <- sims_filtered
    }
  }
  
  if (sum(abs(sims) > 0) == 0) return(data.frame())
  
  # Prepare weighted matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  weighted_mat <- mat * conf
  
  # Calculate predictions
  weighted_ratings <- t(weighted_mat[, unrated_books, drop = FALSE]) %*% sims
  
  rated_mask <- !is.na(confidence_matrix[, unrated_books, drop = FALSE])
  conf_weighted_sims <- sweep(
    conf[, unrated_books, drop = FALSE] * rated_mask,
    1,
    abs(sims),
    "*"
  )
  sum_weighted_sims <- colSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + user_means[target_user]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Get top recommendations
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  top_books <- sort(preds_valid, decreasing = TRUE)[1:min(n_recommendations, length(preds_valid))]
  
  recommendations <- data.frame(
    ISBN = names(top_books),
    Predicted_Rating = as.numeric(top_books)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# UBCF: RECOMMENDATION FOR NEW USERS (COLD-START)
# ============================================================================

recommend_for_new_user_ubcf <- function(new_user_ratings,
                                         user_item_matrix,
                                         user_item_matrix_normalized,
                                         confidence_matrix,
                                         user_means,
                                         book_info,
                                         n_recommendations = 10,
                                         k = NULL) {
  
  if (length(new_user_ratings) == 0) return(data.frame())
  
  # Create new user vector
  new_user_vector <- rep(NA, ncol(user_item_matrix))
  names(new_user_vector) <- colnames(user_item_matrix)
  
  matched_books <- intersect(names(new_user_ratings), names(new_user_vector))
  if (length(matched_books) == 0) return(data.frame())
  
  new_user_vector[matched_books] <- new_user_ratings[matched_books]
  
  # Normalize
  new_user_mean <- mean(new_user_vector, na.rm = TRUE)
  new_user_normalized <- new_user_vector - new_user_mean
  new_user_normalized[is.na(new_user_normalized)] <- 0
  new_user_conf <- ifelse(is.na(new_user_vector), 0, 1)
  new_user_weighted <- new_user_normalized * new_user_conf
  
  # Prepare matrices
  mat_normalized <- user_item_matrix_normalized
  mat_normalized[is.na(mat_normalized)] <- 0
  conf <- confidence_matrix
  conf[is.na(conf)] <- 0
  mat_weighted <- mat_normalized * conf
  
  # Compute similarities
  existing_magnitudes <- sqrt(rowSums(mat_weighted^2))
  new_user_magnitude <- sqrt(sum(new_user_weighted^2))
  
  if (new_user_magnitude == 0) return(data.frame())
  
  new_user_sims <- as.vector(
    (mat_weighted %*% new_user_weighted) /
      (existing_magnitudes * new_user_magnitude)
  )
  new_user_sims[!is.finite(new_user_sims)] <- 0
  names(new_user_sims) <- rownames(user_item_matrix_normalized)
  
  # Apply k-NN
  if (!is.null(k) && k > 0 && k < length(new_user_sims)) {
    threshold <- sort(abs(new_user_sims), decreasing = TRUE)[min(k, sum(new_user_sims != 0))]
    new_user_sims[abs(new_user_sims) < threshold] <- 0
  }
  
  if (all(new_user_sims == 0)) return(data.frame())
  
  # Make predictions
  unrated_books <- names(new_user_vector)[is.na(new_user_vector)]
  if (length(unrated_books) == 0) return(data.frame())
  
  weighted_ratings <- as.vector(t(mat_weighted[, unrated_books, drop = FALSE]) %*% new_user_sims)
  
  rated_mask <- !is.na(user_item_matrix[, unrated_books, drop = FALSE])
  conf_for_unrated <- conf[, unrated_books, drop = FALSE] * rated_mask
  sum_weighted_sims <- as.vector(t(conf_for_unrated) %*% abs(new_user_sims))
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- (weighted_ratings / sum_weighted_sims) + new_user_mean
  preds <- pmin(pmax(preds, 1), 10)
  names(preds) <- unrated_books
  
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  n_actual <- min(n_recommendations, length(preds_valid))
  top_indices <- order(preds_valid, decreasing = TRUE)[1:n_actual]
  
  recommendations <- data.frame(
    ISBN = names(preds_valid)[top_indices],
    Predicted_Rating = round(preds_valid[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```

```{r ubcf-cv}
# ============================================================================
# UBCF: CROSS-VALIDATION
# ============================================================================

cross_validate_ubcf <- function(user_item_matrix,
                                 confidence_matrix,
                                 original_ratings_matrix,
                                 n_folds = 5,
                                 k = 30) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    test_obs <- observed[test_indices, , drop = FALSE]
    
    # Create training matrices
    train_matrix <- user_item_matrix
    train_confidence <- confidence_matrix
    train_matrix[test_obs] <- NA
    train_confidence[test_obs] <- NA
    
    # Normalize and compute similarity
    norm_result <- normalize_matrix_user(train_matrix)
    train_normalized <- norm_result$normalized
    user_means <- norm_result$means
    user_sim_matrix <- compute_user_similarity_matrix(train_normalized, train_confidence)
    
    # Prepare for prediction
    train_norm_filled <- train_normalized
    train_norm_filled[is.na(train_norm_filled)] <- 0
    train_conf_filled <- train_confidence
    train_conf_filled[is.na(train_conf_filled)] <- 0
    
    # Make predictions
    predictions <- numeric(nrow(test_obs))
    
    for (i in 1:nrow(test_obs)) {
      user_idx <- test_obs[i, 1]
      item_idx <- test_obs[i, 2]
      
      sims <- user_sim_matrix[user_idx, ]
      sims[is.na(sims) | is.nan(sims)] <- 0
      
      # Apply k-NN
      if (k < length(sims) && sum(abs(sims) > 0) > 0) {
        k_actual <- min(k, sum(abs(sims) > 0))
        top_k_idx <- order(abs(sims), decreasing = TRUE)[1:k_actual]
        sims_filtered <- rep(0, length(sims))
        sims_filtered[top_k_idx] <- sims[top_k_idx]
        sims <- sims_filtered
      }
      
      rated_mask <- !is.na(train_matrix[, item_idx])
      
      if (sum(abs(sims[rated_mask])) == 0) {
        predictions[i] <- user_means[user_idx]
        next
      }
      
      # Confidence-weighted prediction
      weighted_ratings <- train_norm_filled[rated_mask, item_idx] *
        train_conf_filled[rated_mask, item_idx]
      weighted_sims <- abs(sims[rated_mask]) *
        train_conf_filled[rated_mask, item_idx]
      
      weighted_sum <- sum(sims[rated_mask] * weighted_ratings)
      sum_weighted_sims <- sum(weighted_sims)
      
      if (sum_weighted_sims > 0) {
        predictions[i] <- (weighted_sum / sum_weighted_sims) + user_means[user_idx]
      } else {
        predictions[i] <- user_means[user_idx]
      }
    }
    
    predictions <- pmin(pmax(predictions, 1), 10)
    actual <- user_item_matrix[test_obs]
    original_ratings <- original_ratings_matrix[test_obs]
    metrics <- evaluate_predictions(predictions, actual, original_ratings)
    
    cv_results[fold, ] <- list(fold, metrics$rmse_explicit)
  }
  
  return(cv_results)
}
```

```{r ubcf-implementation}
# ============================================================================
# UBCF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Normalize and compute similarity
normalized_result_ubcf <- normalize_matrix_user(user_item_matrix)
user_item_matrix_normalized_ubcf <- normalized_result_ubcf$normalized
user_means_ubcf <- normalized_result_ubcf$means

user_similarity_matrix <- compute_user_similarity_matrix(
  user_item_matrix_normalized_ubcf,
  confidence_matrix
)

# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ubcf <- recommend_for_user_ubcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_sim_matrix = user_similarity_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)


# Generate recommendations for new user (cold-start)
sample_books <- colnames(user_item_matrix)[1:5]
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), sample_books)

new_user_recs_ubcf <- recommend_for_new_user_ubcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)


```



### 3.3.2 Item-Based Collaborative Filtering (IBCF)

IBCF estimates ratings based on item similarity rather than user similarity, operating under the assumption that users tend to prefer items similar to those they have previously liked. This approach is often more stable than UBCF in sparse datasets, as item relationships change less frequently than individual user preferences. The method follows a three-step process similar to UBCF, but applied to items. First, item-mean normalization was used to remove item-specific popularity biases, centering ratings for each item around zero. Second, confidence-weighted cosine similarity was computed between items based on co-rated users, with confidence values weighting each user's contribution to the similarity measure. Third, predictions employed k-nearest neighbors ($k=50$) filtering to aggregate the target user's ratings for the most similar items. Predicted ratings combine the target item's mean rating with a weighted average of the target user's normalized ratings for similar items, where weights reflect both item similarity and rating confidence.

```{r ibcf-similarity}
# ============================================================================
# IBCF: ITEM SIMILARITY COMPUTATION
# ============================================================================

compute_item_similarity_matrix <- function(user_item_matrix_normalized,
                                            confidence_matrix) {
  
  # Prepare matrices
  mat <- user_item_matrix_normalized
  mat[is.na(mat)] <- 0
  conf <- confidence_matrix
  conf[is.na(conf)] <- 0
  
  # Transpose for item-item similarity
  mat_t <- t(mat)
  conf_t <- t(conf)
  
  # Weighted ratings
  weighted_mat_t <- mat_t * conf_t
  
  # Compute cosine similarity
  numerator <- weighted_mat_t %*% t(weighted_mat_t)
  magnitudes <- sqrt(rowSums(weighted_mat_t^2))
  denominator <- outer(magnitudes, magnitudes)
  
  item_similarity_matrix <- numerator / denominator
  item_similarity_matrix[is.nan(item_similarity_matrix)] <- 0
  diag(item_similarity_matrix) <- 0
  
  rownames(item_similarity_matrix) <- colnames(user_item_matrix_normalized)
  colnames(item_similarity_matrix) <- colnames(user_item_matrix_normalized)
  
  return(item_similarity_matrix)
}
```

```{r ibcf-recommendations}
# ============================================================================
# IBCF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_ibcf <- function(target_user,
                                     user_item_matrix,
                                     user_item_matrix_normalized,
                                     confidence_matrix,
                                     item_sim_matrix,
                                     item_means,
                                     book_info,
                                     n_recommendations = 10,
                                     k = NULL) {
  
  target_user <- as.character(target_user)
  
  # Find unrated books
  unrated_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get item similarities
  item_sims <- item_sim_matrix[unrated_books, , drop = FALSE]
  item_sims[is.na(item_sims) | is.nan(item_sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < ncol(item_sims)) {
    for (i in 1:nrow(item_sims)) {
      sims <- item_sims[i, ]
      non_zero_count <- sum(sims != 0)
      if (non_zero_count > 0) {
        k_actual <- min(k, non_zero_count)
        top_k_items <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
        sims_filtered <- rep(0, length(sims))
        names(sims_filtered) <- names(sims)
        sims_filtered[top_k_items] <- sims[top_k_items]
        item_sims[i, ] <- sims_filtered
      }
    }
  }
  
  if (sum(abs(item_sims) > 0) == 0) return(data.frame())
  
  # Prepare weighted matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  weighted_mat <- mat * conf
  
  # Calculate predictions
  weighted_ratings <- item_sims %*% weighted_mat[target_user, ]
  
  rated_mask <- !is.na(user_item_matrix[target_user, ])
  rated_item_mask <- matrix(
    rated_mask,
    nrow = nrow(item_sims),
    ncol = length(rated_mask),
    byrow = TRUE
  )
  
  conf_weighted_sims <- rated_item_mask * abs(item_sims) * conf[target_user, ]
  sum_weighted_sims <- rowSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + item_means[unrated_books]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Get top recommendations
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  top_books <- sort(preds_valid, decreasing = TRUE)[1:min(n_recommendations, length(preds_valid))]
  
  recommendations <- data.frame(
    ISBN = names(top_books),
    Predicted_Rating = as.numeric(top_books)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# IBCF: RECOMMENDATION FOR NEW USERS (COLD-START)
# ============================================================================

recommend_for_new_user_ibcf <- function(new_user_ratings,
                                         user_item_matrix,
                                         user_item_matrix_normalized,
                                         confidence_matrix,
                                         item_means,
                                         item_sim_matrix,
                                         book_info,
                                         n_recommendations = 10,
                                         k = NULL) {
  
  if (length(new_user_ratings) == 0) return(data.frame())
  
  # Create new user vector
  new_user_vector <- rep(NA, ncol(user_item_matrix))
  names(new_user_vector) <- colnames(user_item_matrix)
  
  matched_books <- intersect(names(new_user_ratings), names(new_user_vector))
  if (length(matched_books) == 0) return(data.frame())
  
  new_user_vector[matched_books] <- new_user_ratings[matched_books]
  
  # Normalize
  new_user_normalized <- new_user_vector - item_means
  new_user_normalized[is.na(new_user_normalized)] <- 0
  new_user_conf <- ifelse(is.na(new_user_vector), 0, 1)
  new_user_weighted <- new_user_normalized * new_user_conf
  
  # Find unrated books
  unrated_books <- names(new_user_vector)[is.na(new_user_vector)]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get item similarities
  item_sims <- item_sim_matrix[unrated_books, , drop = FALSE]
  item_sims[is.na(item_sims) | is.nan(item_sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < ncol(item_sims)) {
    for (i in 1:nrow(item_sims)) {
      sims <- item_sims[i, ]
      non_zero_count <- sum(sims != 0)
      if (non_zero_count > 0) {
        k_actual <- min(k, non_zero_count)
        top_k_items <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
        sims_filtered <- rep(0, length(sims))
        names(sims_filtered) <- names(sims)
        sims_filtered[top_k_items] <- sims[top_k_items]
        item_sims[i, ] <- sims_filtered
      }
    }
  }
  
  if (sum(abs(item_sims) > 0) == 0) return(data.frame())
  
  # Make predictions
  weighted_ratings <- item_sims %*% new_user_weighted
  
  rated_mask <- !is.na(new_user_vector)
  rated_item_mask <- matrix(
    rated_mask,
    nrow = nrow(item_sims),
    ncol = length(rated_mask),
    byrow = TRUE
  )
  
  conf_weighted_sims <- rated_item_mask * abs(item_sims) * new_user_conf
  sum_weighted_sims <- rowSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + item_means[unrated_books]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  n_actual <- min(n_recommendations, length(preds_valid))
  top_indices <- order(preds_valid, decreasing = TRUE)[1:n_actual]
  
  recommendations <- data.frame(
    ISBN = names(preds_valid)[top_indices],
    Predicted_Rating = round(preds_valid[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```

```{r ibcf-cv}
# ============================================================================
# IBCF: CROSS-VALIDATION
# ============================================================================

cross_validate_ibcf <- function(user_item_matrix,
                                 confidence_matrix,
                                 original_ratings_matrix,
                                 n_folds = 5,
                                 k = 30) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    test_obs <- observed[test_indices, , drop = FALSE]
    
    # Create training matrices
    train_matrix <- user_item_matrix
    train_confidence <- confidence_matrix
    train_matrix[test_obs] <- NA
    train_confidence[test_obs] <- NA
    
    # Normalize and compute similarity
    norm_result <- normalize_matrix_item(train_matrix)
    train_normalized <- norm_result$normalized
    item_means <- norm_result$means
    item_sim_matrix <- compute_item_similarity_matrix(train_normalized, train_confidence)
    
    # Prepare for prediction
    train_norm_filled <- train_normalized
    train_norm_filled[is.na(train_norm_filled)] <- 0
    train_conf_filled <- train_confidence
    train_conf_filled[is.na(train_conf_filled)] <- 0
    
    # Make predictions
    predictions <- numeric(nrow(test_obs))
    
    for (i in 1:nrow(test_obs)) {
      user_idx <- test_obs[i, 1]
      item_idx <- test_obs[i, 2]
      
      rated_items <- which(!is.na(train_matrix[user_idx, ]))
      
      if (length(rated_items) == 0) {
        predictions[i] <- item_means[item_idx]
        next
      }
      
      sims <- item_sim_matrix[item_idx, rated_items]
      sims[is.na(sims) | is.nan(sims)] <- 0
      
      # Apply k-NN
      if (sum(abs(sims) > 0) > 0 && k < length(sims)) {
        k_actual <- min(k, sum(abs(sims) > 0))
        top_k_idx <- order(abs(sims), decreasing = TRUE)[1:k_actual]
        sims_filtered <- rep(0, length(sims))
        sims_filtered[top_k_idx] <- sims[top_k_idx]
        sims <- sims_filtered
      }
      
      # Confidence-weighted prediction
      if (sum(abs(sims)) > 0) {
        weighted_ratings <- train_norm_filled[user_idx, rated_items] *
          train_conf_filled[user_idx, rated_items]
        weighted_sims <- abs(sims) * train_conf_filled[user_idx, rated_items]
        
        weighted_sum <- sum(sims * weighted_ratings)
        sum_weighted_sims <- sum(weighted_sims)
        
        if (sum_weighted_sims > 0) {
          predictions[i] <- (weighted_sum / sum_weighted_sims) + item_means[item_idx]
        } else {
          predictions[i] <- item_means[item_idx]
        }
      } else {
        predictions[i] <- item_means[item_idx]
      }
    }
    
    predictions <- pmin(pmax(predictions, 1), 10)
    actual <- user_item_matrix[test_obs]
    original_ratings <- original_ratings_matrix[test_obs]
    metrics <- evaluate_predictions(predictions, actual, original_ratings)
    
    cv_results[fold, ] <- list(fold, metrics$rmse_explicit)
    # cat("RMSE:", round(metrics$rmse_explicit, 4), "\n")
  }
  
  return(cv_results)
}
```

```{r ibcf-implementation}
# ============================================================================
# IBCF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Normalize and compute similarity
normalized_result_ibcf <- normalize_matrix_item(user_item_matrix)
user_item_matrix_normalized_ibcf <- normalized_result_ibcf$normalized
item_means_ibcf <- normalized_result_ibcf$means

item_similarity_matrix <- compute_item_similarity_matrix(
  user_item_matrix_normalized_ibcf,
  confidence_matrix
)

# cat("Similarity matrix computed\n\n")

# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ibcf <- recommend_for_user_ibcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_sim_matrix = item_similarity_matrix,
  item_means = item_means_ibcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)


# Generate recommendations for new user (cold-start)
new_user_recs_ibcf <- recommend_for_new_user_ibcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_means = item_means_ibcf,
  item_sim_matrix = item_similarity_matrix,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)


```



### 3.3.3 Matrix Factorization (MF)

Matrix factorization decomposes the user-item rating matrix into lower-dimensional user and item latent factor matrices, capturing hidden preference patterns that may not be evident from raw ratings. The recosystem package was employed, implementing parallel stochastic gradient descent for efficient factorization. Hyperparameter tuning was performed via grid search over latent factor dimensions (10, 20, 30), learning rates (0.01, 0.05, 0.1), and L2 regularization parameters (0.01, 0.1) for both user and item factors. The optimal configuration was selected based on cross-validation performance. Models were trained for 100 iterations using 4 threads for parallel computation. Only explicit ratings (excluding 0) were used to ensure meaningful latent representations.

```{r mf-data-prep}
# ============================================================================
# MF: DATA PREPARATION (EXPLICIT RATINGS ONLY)
# ============================================================================

create_explicit_matrix <- function(ratings_data,
                                    min_ratings_per_book = 3,
                                    min_ratings_per_user = 2) {
  
  # Filter explicit ratings only
  explicit_ratings <- ratings_data %>%
    filter(Book.Rating > 0) %>%
    select(User.ID, ISBN, Book.Rating)
  
  # Create user-item matrix
  user_item_matrix <- explicit_ratings %>%
    pivot_wider(
      names_from = ISBN,
      values_from = Book.Rating,
      values_fill = NA
    )
  
  # Convert to matrix
  user_ids <- user_item_matrix$User.ID
  user_item_matrix <- as.matrix(user_item_matrix[, -1])
  rownames(user_item_matrix) <- user_ids
  
  # Filter by minimum ratings
  books_to_keep <- colSums(!is.na(user_item_matrix)) >= min_ratings_per_book
  user_item_matrix <- user_item_matrix[, books_to_keep]
  
  users_to_keep <- rowSums(!is.na(user_item_matrix)) >= min_ratings_per_user
  user_item_matrix <- user_item_matrix[users_to_keep, ]
  
  return(user_item_matrix)
}

prepare_recosystem_data <- function(user_item_matrix, original_ratings_matrix) {
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  
  train_data <- data.frame(
    user_index = observed[, 1] - 1,
    item_index = observed[, 2] - 1,
    rating = user_item_matrix[observed],
    original_rating = original_ratings_matrix[observed]
  )
  
  # Filter explicit ratings only
  train_data_explicit <- train_data[train_data$original_rating != 0, ]
  
  user_ids <- data.frame(
    user_index = 0:(nrow(user_item_matrix) - 1),
    user_id = rownames(user_item_matrix)
  )
  
  item_ids <- data.frame(
    item_index = 0:(ncol(user_item_matrix) - 1),
    item_id = colnames(user_item_matrix)
  )
  
  list(
    train_data = train_data_explicit,
    user_ids = user_ids,
    item_ids = item_ids,
    n_users = nrow(user_item_matrix),
    n_items = ncol(user_item_matrix)
  )
}
```

```{r mf-training}
# ============================================================================
# MF: HYPERPARAMETER TUNING
# ============================================================================

tune_reco_model <- function(train_data,
                             n_factors = c(10, 20, 30),
                             learning_rate = c(0.1, 0.05, 0.01),
                             costp_l2 = c(0.01, 0.1),
                             costq_l2 = c(0.01, 0.1),
                             n_iter = 50,
                             verbose = TRUE) {
  
  train_set <- data_memory(
    user_index = train_data$user_index,
    item_index = train_data$item_index,
    rating = train_data$rating,
    index1 = FALSE
  )
  
  rs <- Reco()
  
  opts <- rs$tune(train_set, opts = list(
    dim = n_factors,
    lrate = learning_rate,
    costp_l2 = costp_l2,
    costq_l2 = costq_l2,
    niter = n_iter,
    nthread = 4,
    verbose = verbose
  ))
  
  return(opts)
}

# ============================================================================
# MF: MODEL TRAINING
# ============================================================================

train_reco_model <- function(train_data,
                              n_factors = 20,
                              learning_rate = 0.1,
                              costp_l2 = 0.01,
                              costq_l2 = 0.01,
                              n_iter = 100,
                              n_threads = 4,
                              verbose = TRUE) {
  
  train_set <- data_memory(
    user_index = train_data$user_index,
    item_index = train_data$item_index,
    rating = train_data$rating,
    index1 = FALSE
  )
  
  model <- Reco()
  model$train(train_set, opts = list(
    dim = n_factors,
    lrate = learning_rate,
    costp_l2 = costp_l2,
    costq_l2 = costq_l2,
    niter = n_iter,
    nthread = n_threads,
    verbose = verbose
  ))
  
  return(model)
}
```

```{r mf-recommendations}
# ============================================================================
# MF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_mf <- function(model,
                                   user_item_matrix,
                                   user_id,
                                   n_recommendations = 10,
                                   user_ids_map,
                                   item_ids_map,
                                   book_info) {
  
  if (!user_id %in% user_ids_map$user_id) stop("User not found")
  
  user_idx <- user_ids_map$user_index[user_ids_map$user_id == user_id]
  user_row <- user_item_matrix[as.character(user_id), ]
  unrated_items <- which(is.na(user_row))
  
  if (length(unrated_items) == 0) return(data.frame())
  
  item_indices <- unrated_items - 1
  pred_set <- data_memory(
    user_index = rep(user_idx, length(item_indices)),
    item_index = item_indices,
    index1 = FALSE
  )
  
  pred_ratings <- model$predict(pred_set, out_memory())
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  recommended_items <- colnames(user_item_matrix)[unrated_items[top_indices]]
  
  data.frame(
    ISBN = recommended_items,
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
}

# ============================================================================
# MF: RECOMMENDATION FOR NEW USERS (COLD-START)
# ============================================================================

recommend_for_new_user_mf <- function(model,
                                       new_user_ratings,
                                       user_item_matrix,
                                       item_ids_map,
                                       book_info,
                                       n_recommendations = 10) {
  
  # Hybrid approach for cold-start
  ratings_df <- if (is.numeric(new_user_ratings) && !is.null(names(new_user_ratings))) {
    data.frame(item_id = names(new_user_ratings), rating = as.numeric(new_user_ratings))
  } else {
    new_user_ratings
  }
  
  ratings_df <- ratings_df[ratings_df$item_id %in% item_ids_map$item_id, ]
  if (nrow(ratings_df) == 0) stop("No rated items exist in training data")
  
  # Calculate item means for fallback
  item_means <- colMeans(user_item_matrix, na.rm = TRUE)
  
  # Get unrated items
  all_items <- item_ids_map$item_id
  rated_items <- ratings_df$item_id
  unrated_items <- setdiff(all_items, rated_items)
  
  if (length(unrated_items) == 0) return(data.frame())
  
  # Use item means adjusted by user bias
  user_mean_rating <- mean(ratings_df$rating)
  global_mean_rating <- mean(user_item_matrix, na.rm = TRUE)
  user_bias <- user_mean_rating - global_mean_rating
  
  pred_ratings <- item_means[unrated_items] + user_bias
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  recommended_items <- names(pred_ratings)[top_indices]
  
  data.frame(
    ISBN = recommended_items,
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
}
```

```{r mf-cv}
# ============================================================================
# MF: CROSS-VALIDATION
# ============================================================================

cross_validate_mf <- function(user_item_matrix,
                               original_ratings_matrix,
                               optimal_params,
                               n_folds = 5,
                               n_iter = 50,
                               verbose = TRUE) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  if (n_ratings < n_folds * 2) {
    stop("Not enough ratings to perform cross-validation.")
  }
  
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    Fold = integer(n_folds),
    RMSE = numeric(n_folds)
  )
  
  if (verbose) {
    # cat("Starting", n_folds, "-fold cross-validation...\n")
  }
  
  for (fold in 1:n_folds) {
    # if (verbose) cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_idx <- which(fold_indices == fold)
    train_idx <- which(fold_indices != fold)
    
    train_obs <- observed[train_idx, , drop = FALSE]
    test_obs <- observed[test_idx, , drop = FALSE]
    
    # Prepare training data (explicit only)
    train_data <- data.frame(
      user_index = train_obs[, 1] - 1,
      item_index = train_obs[, 2] - 1,
      rating = user_item_matrix[train_obs],
      original_rating = original_ratings_matrix[train_obs]
    )
    
    train_data_explicit <- train_data[train_data$original_rating != 0, ]
    
    # Prepare test data (explicit only)
    test_data <- data.frame(
      user_index = test_obs[, 1] - 1,
      item_index = test_obs[, 2] - 1,
      rating = user_item_matrix[test_obs],
      original_rating = original_ratings_matrix[test_obs]
    )
    
    test_data_explicit <- test_data[test_data$original_rating != 0, ]
    
    if (nrow(test_data_explicit) == 0) {
      cv_results[fold, ] <- c(fold, NA)
      next
    }
    
    # Train model
    train_set <- data_memory(
      user_index = train_data_explicit$user_index,
      item_index = train_data_explicit$item_index,
      rating = train_data_explicit$rating,
      index1 = FALSE
    )
    
    test_set <- data_memory(
      user_index = test_data_explicit$user_index,
      item_index = test_data_explicit$item_index,
      rating = test_data_explicit$rating,
      index1 = FALSE
    )
    
    r <- Reco()
    r$train(train_set, opts = list(
      dim = optimal_params$min$dim,
      lrate = optimal_params$min$lrate,
      costp_l2 = optimal_params$min$costp_l2,
      costq_l2 = optimal_params$min$costq_l2,
      niter = n_iter,
      nthread = 4,
      verbose = FALSE
    ))
    
    preds <- r$predict(test_set, out_memory())
    preds <- pmax(pmin(preds, 10), 1)
    
    # Calculate RMSE
    actual <- test_data_explicit$rating
    original <- test_data_explicit$original_rating
    metrics <- evaluate_predictions(preds, actual, original)
    
    cv_results[fold, ] <- c(fold, metrics$rmse_explicit)
    # cat("RMSE:", round(metrics$rmse_explicit, 4), "\n")
  }
  
  return(cv_results)
}
```

```{r mf-implementation}
# ============================================================================
# MF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Prepare data for recosystem
prepared <- prepare_recosystem_data(user_item_matrix, original_ratings_matrix)

# Hyperparameter tuning
# optimal_params_mf <- tune_reco_model(
#   train_data = prepared$train_data,
#   n_factors = c(10, 20, 30),
#   learning_rate = c(0.1, 0.05, 0.01),
#   costp_l2 = c(0.01, 0.1),
#   costq_l2 = c(0.01, 0.1),
#   n_iter = 50,
#   verbose = FALSE
# )

# save(optimal_params_mf, file = 'mf_tuning_results.Rdata')

# Train final model
# model_mf <- train_reco_model(
#   train_data = prepared$train_data,
#   n_factors = optimal_params_mf$min$dim,
#   learning_rate = optimal_params_mf$min$lrate,
#   costp_l2 = optimal_params_mf$min$costp_l2,
#   costq_l2 = optimal_params_mf$min$costq_l2,
#   n_iter = 100,
#   n_threads = 4,
#   verbose = FALSE
# )

# save(model_mf, file = 'mf_final_model_mf.Rdata')


# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_mf <- recommend_for_user_mf(
  model = model_mf,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids_map = prepared$user_ids,
  item_ids_map = prepared$item_ids,
  book_info = book_info
)


# Generate recommendations for new user (cold-start)
recs_new_mf <- recommend_for_new_user_mf(
  model = model_mf,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  item_ids_map = prepared$item_ids,
  book_info = book_info,
  n_recommendations = 10
)


# Run cross-validation
cv_results_mf <- cross_validate_mf(
  user_item_matrix = user_item_matrix,
  original_ratings_matrix = original_ratings_matrix,
  optimal_params = optimal_params_mf,
  n_folds = 5,
  n_iter = 50,
  verbose = TRUE
)


```



### 3.3.4 Neural Network-Based Collaborative Filtering (NN)

Neural network-based collaborative filtering was implemented using H2O's deep learning framework, with user-mean normalization applied to input ratings. The model learned complex non-linear relationships between users and items through dense embeddings of categorical User ID and Book ID features. Training employed adaptive learning rates with early stopping based on validation RMSE. Hyperparameter tuning was performed via grid search over 24 configurations, evaluating layer depths (1–3 layers), hidden unit counts (16–128), activation functions (Tanh, ReLU), and regularization strengths (0.00001–0.0001). The best configuration was selected based on five-fold cross-validation performance. Only explicit rating data were used for training.

```{r nn-data-prep}
# ============================================================================
# NN: DATA PREPARATION
# ============================================================================

prepare_h2o_data <- function(user_item_matrix, original_ratings_matrix = NULL) {
  
  # Apply user-mean normalization
  user_means <- rowMeans(user_item_matrix, na.rm = TRUE)
  normalized_matrix <- sweep(user_item_matrix, 1, user_means, FUN = "-")
  
  # Get observed ratings
  observed <- which(!is.na(normalized_matrix), arr.ind = TRUE)
  
  # Create training data
  train_data <- data.frame(
    user_id = rownames(user_item_matrix)[observed[, 1]],
    book_id = colnames(user_item_matrix)[observed[, 2]],
    rating = as.numeric(normalized_matrix[observed])
  )
  
  # Add original rating information
  if (!is.null(original_ratings_matrix)) {
    train_data$original_rating <- as.numeric(original_ratings_matrix[observed])
    train_data$is_implicit <- (train_data$original_rating == 0)
  } else {
    train_data$original_rating <- train_data$rating + user_means[observed[, 1]]
    train_data$is_implicit <- FALSE
  }
  
  return(list(
    train_data = train_data,
    user_ids = data.frame(user_id = rownames(user_item_matrix)),
    book_ids = data.frame(book_id = colnames(user_item_matrix)),
    n_users = nrow(user_item_matrix),
    n_items = ncol(user_item_matrix),
    user_means = user_means,
    original_matrix = user_item_matrix,
    original_ratings_matrix = original_ratings_matrix
  ))
}
```

```{r nn-tuning}
# ============================================================================
# NN: HYPERPARAMETER TUNING WITH GRID SEARCH
# ============================================================================

h2o.init(ip = "localhost", port = 54321, nthreads = 4)

# # Prepare data
# h2o_data_tuning <- prepare_h2o_data(user_item_matrix, original_ratings_matrix)
# 
# # Filter explicit ratings only
# train_data_explicit <- h2o_data_tuning$train_data[!h2o_data_tuning$train_data$is_implicit, ]
# 
# # Convert to H2O frame
# train_h2o_tuning <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
# train_h2o_tuning$user_id <- as.factor(train_h2o_tuning$user_id)
# train_h2o_tuning$book_id <- as.factor(train_h2o_tuning$book_id)
# 
# # Define hyperparameter grid
# hyper_params <- list(
#   activation = c("Tanh", "Rectifier"),
#   hidden = list(
#     c(16),
#     c(32, 16),
#     c(64, 32),
#     c(64, 32, 16),
#     c(128, 64, 32)
#   ),
#   l1 = c(0.00001, 0.0001),
#   l2 = c(0.00001, 0.0001)
# )
# 
# cat("Total models to train:", 
#     length(hyper_params$activation) * length(hyper_params$hidden) * 
#     length(hyper_params$l1) * length(hyper_params$l2), "\n\n")
# 
# set.seed(123)

# # Run grid search
# model_grid <- h2o.grid(
#   algorithm = "deeplearning",
#   grid_id = "nn_grid_book_recommendations",
#   hyper_params = hyper_params,
#   x = c("user_id", "book_id"),
#   y = "rating",
#   training_frame = train_h2o_tuning,
#   nfolds = 5,
#   stopping_metric = "RMSE",
#   stopping_rounds = 5,
#   stopping_tolerance = 0.001,
#   seed = 123,
#   adaptive_rate = TRUE,
#   epochs = 50,
#   max_w2 = 10,
#   initial_weight_distribution = "UniformAdaptive"
# )
# 
# # Get results
# grid_results <- h2o.getGrid(
#   "nn_grid_book_recommendations",
#   sort_by = "rmse",
#   decreasing = FALSE)

# all_models_df <- as.data.frame(grid_results@summary_table)

# Select best model
# best_model_id <- grid_results@model_ids[[1]]
# best_model <- h2o.getModel(best_model_id)
# cv_rmse <- h2o.rmse(best_model, xval = TRUE)

# best_nn_results <- data.frame(
#   Activation = best_model@allparameters$activation,
#   Hidden_Layers = paste(best_model@allparameters$hidden, collapse = ", "),
#   N_Layers = length(best_model@allparameters$hidden),
#   L1_Regularization = best_model@allparameters$l1,
#   L2_Regularization = best_model@allparameters$l2,
#   CV_RMSE = round(cv_rmse, 4),
#   Data_Type = "Explicit Only"
# )


# Grid search results
kable(all_models_df %>% select(-model_ids),
      format = "latex",
      booktabs = TRUE,
      caption = "NN Grid Search Results: Hyperparameter optimization results across 24 configurations showing activation functions, hidden layer architectures, and regularization parameters",
      align = c("c", "c", "c", "c", "c"),
      digits = 3) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Best model results
kable(best_nn_results,
      format = "latex",
      booktabs = TRUE,
      caption = "Optimal NN Configuration: Best performing model parameters selected from grid search including activation function, hidden layer architecture, regularization settings, and cross-validation RMSE",
      align = c("c", "c", "c", "c", "c", "c"),
      digits = 3) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )



# save(best_nn_results, grid_results, all_models_df, file = 'nn_tuning_results.Rdata') 
```

```{r nn-training}
# ============================================================================
# NN: MODEL TRAINING FUNCTION
# ============================================================================

train_h2o_model <- function(train_data,
                             hidden = c(64, 32),
                             epochs = 30,
                             activation = "Tanh",
                             l1 = 0.0001,
                             l2 = 0.0001,
                             seed = 123,
                             verbose = FALSE) {
  
  if (!verbose) h2o.no_progress()
  
  model <- h2o.deeplearning(
    x = c("user_id", "book_id"),
    y = "rating",
    training_frame = train_data,
    hidden = hidden,
    epochs = epochs,
    activation = activation,
    l1 = l1,
    l2 = l2,
    adaptive_rate = TRUE,
    seed = seed,
    stopping_rounds = 5,
    stopping_metric = "RMSE",
    stopping_tolerance = 0.001,
    max_w2 = 10,
    initial_weight_distribution = "UniformAdaptive",
    initial_weight_scale = 0.5
  )
  
  return(model)
}
```

```{r nn-recommendations}
# ============================================================================
# NN: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_nn <- function(model,
                                   user_item_matrix,
                                   user_id,
                                   n_recommendations = 10,
                                   user_ids,
                                   book_ids,
                                   user_means,
                                   book_info) {
  
  if (!user_id %in% user_ids$user_id) {
    stop("User ID not found in training data")
  }
  
  # Find unrated books
  user_row <- user_item_matrix[as.character(user_id), , drop = TRUE]
  unrated_books <- which(is.na(user_row))
  
  if (length(unrated_books) == 0) {
    message("User has rated all available books")
    return(data.frame())
  }
  
  # Prepare prediction input
  pred_data <- data.frame(
    user_id = rep(user_id, length(unrated_books)),
    book_id = colnames(user_item_matrix)[unrated_books]
  )
  
  # Convert to H2O frame
  pred_h2o <- as.h2o(pred_data)
  pred_h2o$user_id <- as.factor(pred_h2o$user_id)
  pred_h2o$book_id <- as.factor(pred_h2o$book_id)
  
  # Predict ratings
  predictions <- h2o.predict(model, pred_h2o)
  predictions_df <- as.data.frame(predictions)
  
  # Denormalize predictions
  user_mean <- user_means[as.character(user_id)]
  pred_ratings <- predictions_df$predict + user_mean
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  # Select top recommendations
  n_to_recommend <- min(n_recommendations, length(pred_ratings))
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:n_to_recommend]
  
  recommendations <- data.frame(
    ISBN = pred_data$book_id[top_indices],
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  )
  
  recommendations <- recommendations %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# NN: RECOMMENDATION FOR NEW USERS (COLD-START)
# ============================================================================

recommend_for_new_user_nn <- function(model,
                                       new_user_ratings,
                                       user_item_matrix,
                                       book_ids,
                                       book_info,
                                       n_recommendations = 10,
                                       user_means = NULL) {
  
  if (length(new_user_ratings) == 0) {
    stop("No ratings provided for new user")
  }
  
  rated_books <- names(new_user_ratings)
  valid_books <- rated_books[rated_books %in% book_ids$book_id]
  
  if (length(valid_books) == 0) {
    stop("None of the rated books exist in training data")
  }
  
  # Find similar users based on rated books
  new_user_mean <- mean(new_user_ratings)
  new_user_normalized <- new_user_ratings - new_user_mean
  
  user_similarities <- numeric(nrow(user_item_matrix))
  names(user_similarities) <- rownames(user_item_matrix)
  
  for (i in 1:nrow(user_item_matrix)) {
    user_ratings <- user_item_matrix[i, valid_books]
    valid_overlap <- !is.na(user_ratings)
    
    if (sum(valid_overlap) == 0) {
      user_similarities[i] <- 0
      next
    }
    
    overlap_books <- valid_books[valid_overlap]
    existing_mean <- mean(user_ratings[valid_overlap], na.rm = TRUE)
    existing_normalized <- user_ratings[valid_overlap] - existing_mean
    new_normalized <- new_user_normalized[overlap_books]
    
    # Cosine similarity
    numerator <- sum(existing_normalized * new_normalized)
    denominator <- sqrt(sum(existing_normalized^2)) * sqrt(sum(new_normalized^2))
    
    if (denominator > 0) {
      user_similarities[i] <- numerator / denominator
    } else {
      user_similarities[i] <- 0
    }
  }
  
  # Use top-K similar users
  k <- min(30, sum(user_similarities > 0))
  
  if (k == 0) {
    # Fallback to item popularity
    item_means <- colMeans(user_item_matrix, na.rm = TRUE)
    all_books <- book_ids$book_id
    unrated_books <- setdiff(all_books, rated_books)
    pred_ratings <- item_means[unrated_books] + (new_user_mean - mean(user_means))
  } else {
    top_k_users <- names(sort(user_similarities, decreasing = TRUE)[1:k])
    similarities <- user_similarities[top_k_users]
    
    all_books <- book_ids$book_id
    unrated_books <- setdiff(all_books, rated_books)
    
    # Weighted average of similar users' ratings
    weighted_preds <- numeric(length(unrated_books))
    names(weighted_preds) <- unrated_books
    
    for (book in unrated_books) {
      similar_user_ratings <- user_item_matrix[top_k_users, book]
      valid_ratings <- !is.na(similar_user_ratings)
      
      if (sum(valid_ratings) > 0) {
        weighted_sum <- sum(similarities[valid_ratings] * similar_user_ratings[valid_ratings])
        sum_weights <- sum(abs(similarities[valid_ratings]))
        weighted_preds[book] <- weighted_sum / sum_weights
      } else {
        weighted_preds[book] <- mean(user_item_matrix[, book], na.rm = TRUE)
      }
    }
    
    pred_ratings <- weighted_preds + new_user_mean - mean(user_means)
  }
  
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  
  recommendations <- data.frame(
    ISBN = names(pred_ratings)[top_indices],
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```

```{r nn-cv}
# ============================================================================
# NN: CROSS-VALIDATION
# ============================================================================

cross_validate_nn <- function(user_item_matrix,
                               original_ratings_matrix = NULL,
                               n_folds = 5,
                               seed = 123,
                               hidden = c(64, 32),
                               epochs = 20,
                               activation = "Tanh",
                               l1 = 0.0001,
                               l2 = 0.0001) {
  
  set.seed(seed)
  h2o.no_progress()
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  if (n_ratings < n_folds * 2) {
    stop("Not enough ratings to perform cross-validation")
  }
  
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  # cat("Starting", n_folds, "-fold cross-validation for NN...\n")
  # cat("Training on EXPLICIT ratings only\n\n")
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    train_indices <- which(fold_indices != fold)
    
    test_obs <- observed[test_indices, , drop = FALSE]
    train_obs <- observed[train_indices, , drop = FALSE]
    
    # Create training matrix
    train_matrix <- user_item_matrix
    train_matrix[test_obs] <- NA
    
    # Normalize
    train_user_means <- rowMeans(train_matrix, na.rm = TRUE)
    train_normalized <- sweep(train_matrix, 1, train_user_means, FUN = "-")
    
    # Create training data
    train_data <- data.frame(
      user_id = rownames(user_item_matrix)[train_obs[, 1]],
      book_id = colnames(user_item_matrix)[train_obs[, 2]],
      rating = as.numeric(train_normalized[train_obs])
    )
    
    # Filter explicit ratings
    if (!is.null(original_ratings_matrix)) {
      train_data$original_rating <- original_ratings_matrix[train_obs]
      train_data_explicit <- train_data[train_data$original_rating != 0, ]
    } else {
      train_data_explicit <- train_data
    }
    
    if (nrow(train_data_explicit) == 0) {
      cv_results[fold, ] <- c(fold, NA)
      # cat("No explicit ratings - skipping\n")
      next
    }
    
    # Convert to H2O
    train_h2o <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
    train_h2o$user_id <- as.factor(train_h2o$user_id)
    train_h2o$book_id <- as.factor(train_h2o$book_id)
    
    # Train model
    model <- h2o.deeplearning(
      x = c("user_id", "book_id"),
      y = "rating",
      training_frame = train_h2o,
      hidden = hidden,
      epochs = epochs,
      activation = activation,
      l1 = l1,
      l2 = l2,
      adaptive_rate = TRUE,
      seed = seed,
      stopping_rounds = 3,
      stopping_tolerance = 0.001,
      max_w2 = 10,
      initial_weight_distribution = "UniformAdaptive",
      initial_weight_scale = 0.5,
      verbose = FALSE
    )
    
    # Prepare test data (explicit only)
    test_user_means <- train_user_means[rownames(user_item_matrix)[test_obs[, 1]]]
    test_data <- data.frame(
      user_id = rownames(user_item_matrix)[test_obs[, 1]],
      book_id = colnames(user_item_matrix)[test_obs[, 2]],
      rating = as.numeric(user_item_matrix[test_obs]) - test_user_means
    )
    
    if (!is.null(original_ratings_matrix)) {
      test_data$original_rating <- original_ratings_matrix[test_obs]
      test_data_explicit <- test_data[test_data$original_rating != 0, ]
      test_user_means_explicit <- test_user_means[test_data$original_rating != 0]
      actual_explicit <- user_item_matrix[test_obs][test_data$original_rating != 0]
    } else {
      test_data_explicit <- test_data
      test_user_means_explicit <- test_user_means
      actual_explicit <- user_item_matrix[test_obs]
    }
    
    # Convert to H2O
    test_h2o <- as.h2o(test_data_explicit[, c("user_id", "book_id", "rating")])
    test_h2o$user_id <- as.factor(test_h2o$user_id)
    test_h2o$book_id <- as.factor(test_h2o$book_id)
    
    # Predict
    predictions <- as.data.frame(h2o.predict(model, test_h2o))
    predictions_denorm <- predictions$predict + test_user_means_explicit
    predictions_denorm <- pmax(pmin(predictions_denorm, 10), 1)
    
    # Calculate RMSE
    rmse_value <- sqrt(mean((predictions_denorm - actual_explicit)^2, na.rm = TRUE))
    
    cv_results[fold, ] <- list(fold, rmse_value)
    # cat("RMSE:", round(rmse_value, 4), "\n")
    
    # Cleanup
    h2o.rm(model)
    h2o.rm(train_h2o)
    h2o.rm(test_h2o)
    gc()
  }
  
  return(cv_results)
}
```

```{r nn-implementation}
# ============================================================================
# NN: MODEL TRAINING AND EVALUATION
# ============================================================================

# Initialize H2O cluster
suppressMessages(h2o.init(ip = "localhost", port = 54321))
h2o.no_progress()

# Prepare data for final model
prepared <- prepare_h2o_data(user_item_matrix, original_ratings_matrix)

optimal_params_nn <- list(
  hidden = as.numeric(unlist(strsplit(as.character(best_nn_results$Hidden_Layers), ", "))),
  epochs = 30,  # Default value
  activation = as.character(best_nn_results$Activation),
  l1 = as.numeric(best_nn_results$L1_Regularisation),
  l2 = as.numeric(best_nn_results$L2_Regularisation)
)

# Train final model
train_data_explicit <- prepared$train_data[!prepared$train_data$is_implicit, ]

train_h2o <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
train_h2o$user_id <- as.factor(train_h2o$user_id)
train_h2o$book_id <- as.factor(train_h2o$book_id)


# Train new model
model_nn <- train_h2o_model(
  train_data = train_h2o,
  hidden = optimal_params_nn$hidden,
  epochs = optimal_params_nn$epochs,
  activation = optimal_params_nn$activation,
  l1 = optimal_params_nn$l1,
  l2 = optimal_params_nn$l2,
  verbose = FALSE)
  
# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_nn <- recommend_for_user_nn(
  model = model_nn,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids = prepared$user_ids,
  book_ids = prepared$book_ids,
  user_means = prepared$user_means,
  book_info = book_info
)


# Generate recommendations for new user (cold-start)
new_user_recs_nn <- recommend_for_new_user_nn(
  model = model_nn,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  book_ids = prepared$book_ids,
  book_info = book_info,
  n_recommendations = 10,
  user_means = prepared$user_means
)


```



## 3.4 Cold-Start Evaluation

To assess cold start performance, a synthetic new user profile was created by selecting the first five books from the user-item matrix. The new user provided ratings of [7, 8, 6, 5, 7] for these books, representing a typical mix of positive evaluations with moderate variation. Each method then generated top-10 book recommendations based solely on this limited profile. 

For memory-based methods, the cold-start implementation remained consistent with their core algorithms. Both UBCF and IBCF rely on similarity calculations, which naturally extend to new users by computing similarities on-the-fly from their available ratings. As a result, these models behaved the same way for new and existing users, maintaining methodological consistency. For cold-start users with five or fewer initial ratings, similarity to all existing users was computed based on overlap in rated items. Predictions for unrated items were generated using a weighted average of the $k=50$ most similar users' ratings, adjusted by the new user's mean rating to account for individual rating tendencies. 

In contrast, the Matrix Factorization and Neural Network models could not leverage their learned representations for unseen users, as no latent factors or embeddings existed. To enable predictions, both relied on fallback strategies. For cold-start users, MF used a hybrid approach where the new user's bias was calculated as the difference between their mean rating and the global mean rating. Predictions for unrated items adjusted item mean ratings by this user bias, effectively assuming that new users exhibit average preferences modulated by their individual rating tendencies. This allows reasonable recommendations without requiring user-specific latent factors. The NN model applied a similarity-based fallback strategy for new users without learned embeddings. This method computes cosine similarities between the new user's normalized ratings and all existing user profiles, using the top 30 most similar users to generate weighted predictions for unrated books. This hybrid approach allows the neural network to provide plausible recommendations for cold-start users without retraining the model.




## 3.5 Catalog Size Impact Analysis

To examine the effect of catalog size on prediction accuracy, the number of books used for training was systematically varied, selecting the most popular titles ranked by total number of ratings for $N \in {5, 10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100, 110, 120, 130, 145}$. Users were filtered to include only those with at least three ratings within each catalog subset. For each catalog, per-user 80/20 train-test splits were applied, ensuring that the same books appeared in both sets and that each user contributed proportionally to training and testing. Matrix factorization models were trained on each catalog using previously identified hyperparameters. The analysis was repeated across five random seeds to assess robustness, with mean RMSE and standard error computed for each catalog size.




# 4. Results and Evaluation

## 4.1 Cross-Validation Performance

Comparison The 5-fold cross-validation analysis revealed substantial differences in performance across the four collaborative filtering methods (Table 2, Figure 2). Matrix Factorization achieved the lowest RMSE of 1.820 (SD = 0.020), making it the most accurate approach for this dataset. Neural Networks performed comparably, with an RMSE of 1.899 (SD = 0.053), demonstrating that model-based methods substantially outperform memory-based techniques. Item-Based Collaborative Filtering achieved an RMSE of 2.746 (SD = 0.018), while User-Based Collaborative Filtering recorded 3.182 (SD = 0.047). The gap between the best (MF) and worst (UBCF) methods corresponds to a 43% reduction in prediction error, highlighting the importance of algorithm choice in sparse datasets. Prediction consistency, measured by standard deviation across folds, was generally low for all methods, indicating stable performance.

```{r comprehensive-comparison}
# ============================================================================
# COMPREHENSIVE CROSS-VALIDATION COMPARISON
# ============================================================================

run_comprehensive_cv_comparison <- function(user_item_matrix,
                                            confidence_matrix,
                                            original_ratings_matrix,
                                            optimal_params_mf,
                                            optimal_params_nn,
                                            n_folds = 5,
                                            k = 50) {
  
  
  all_results <- data.frame(
    Method = character(4),
    CV_RMSE_Mean = numeric(4),
    CV_RMSE_SD = numeric(4),
    stringsAsFactors = FALSE
  )
  
  # UBCF
  # cat("=== Running UBCF ===\n")
  cv_results_ubcf <- cross_validate_ubcf(
    user_item_matrix,
    confidence_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    k = k
  )
  
  all_results[1, ] <- list(
    "UBCF",
    round(mean(cv_results_ubcf$rmse), 3),
    round(sd(cv_results_ubcf$rmse), 3)
  )
  
  # IBCF
  # cat("\n=== Running IBCF ===\n")
  cv_results_ibcf <- cross_validate_ibcf(
    user_item_matrix,
    confidence_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    k = k
  )
  
  all_results[2, ] <- list(
    "IBCF",
    round(mean(cv_results_ibcf$rmse), 3),
    round(sd(cv_results_ibcf$rmse), 3)
  )
  
  # MF
  # cat("\n=== Running MF ===\n")
  cv_results_mf <- cross_validate_mf(
    user_item_matrix,
    original_ratings_matrix,
    optimal_params_mf,
    n_folds = n_folds,
    n_iter = 50,
    verbose = TRUE
  )
  
  all_results[3, ] <- list(
    "MF",
    round(mean(cv_results_mf$RMSE, na.rm = TRUE), 3),
    round(sd(cv_results_mf$RMSE, na.rm = TRUE), 3)
  )
  
  # NN
  # cat("\n=== Running NN ===\n")
  cv_results_nn <- cross_validate_nn(
    user_item_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    hidden = optimal_params_nn$hidden,
    epochs = optimal_params_nn$epochs,
    activation = optimal_params_nn$activation,
    l1 = optimal_params_nn$l1,
    l2 = optimal_params_nn$l2
  )
  
  all_results[4, ] <- list(
    "NN",
    round(mean(cv_results_nn$rmse, na.rm = TRUE), 3),
    round(sd(cv_results_nn$rmse, na.rm = TRUE), 3)
  )
  

  # Return all results including individual CV results
  return(list(
    summary = all_results,
    cv_results_ubcf = cv_results_ubcf,
    cv_results_ibcf = cv_results_ibcf,
    cv_results_mf = cv_results_mf,
    cv_results_nn = cv_results_nn
  ))
}

# # Run comprehensive comparison
# comparison_results <- run_comprehensive_cv_comparison(
#   user_item_matrix = user_item_matrix,
#   confidence_matrix = confidence_matrix,
#   original_ratings_matrix = original_ratings_matrix,
#   optimal_params_mf = optimal_params_mf,
#   optimal_params_nn = optimal_params_nn,
#   n_folds = 5,
#   k = 50
# )
# 
# # Extract results
# cv_comparison_results <- comparison_results$summary
# cv_results_ubcf <- comparison_results$cv_results_ubcf
# cv_results_ibcf <- comparison_results$cv_results_ibcf
# cv_results_mf <- comparison_results$cv_results_mf
# cv_results_nn <- comparison_results$cv_results_nn
# 
# save(cv_comparison_results, cv_results_ubcf, cv_results_ibcf, cv_results_mf, cv_results_nn, 
#      file = "all_cv_results.Rdata")

# Display results
kable(
  cv_comparison_results,
  format = "latex",
  booktabs = TRUE,
  caption = "Cross-Validation Performance Comparison: RMSE results across all four collaborative filtering methods showing mean performance and standard deviation from 5-fold cross-validation",
  col.names = c("Method", "Mean RMSE", "SD RMSE"),
  align = c("l", "c", "c"),
  digits = 3
) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )



```



## 4.2 Cold-Start Analysis

To evaluate recommendation diversity and algorithmic convergence, 40 total recommendations (10 per method) were analyzed for a synthetic new user. The results revealed substantial divergence across collaborative filtering approaches. Only 7 unique titles (17.5%) appeared in multiple recommendation sets, while 33 titles (82.5%) were unique to a single method. This limited overlap indicates that each algorithm identifies distinct items as suitable for the same user profile, despite all operating on identical rating data.

The strongest consensus emerged around Harry Potter and the Order of the Phoenix by J.K. Rowling, which appeared in three methods with varying predicted ratings: UBCF (7.34), IBCF (7.03), and NN (6.50). This represents the highest level of agreement observed across models. Beyond this single title, no other book appeared in more than two methods, underscoring fundamental differences in how collaborative filtering approaches model user preferences.

Six additional titles appeared in exactly two methods, accounting for 15% of total recommendations: The Hobbit (IBCF: 6.49, MF: 5.54), Fahrenheit 451 (MF: 5.35, NN: 6.71), Tuesdays with Morrie (IBCF: 6.10, NN: 6.50), Watership Down (UBCF: 5.91, MF: 5.54), The Fellowship of the Ring (IBCF: 6.25, MF: 5.52), and The Da Vinci Code (IBCF: 6.30, MF: 7.47). These pairwise overlaps revealed rating discrepancies of up to 1.36 points, suggesting different confidence calibrations across algorithms. Memory-based methods generally assigned higher predicted ratings than model-based approaches for shared items, reflecting distinct estimation biases and confidence behaviors.

Overall, the low overlap highlights that UBCF emphasizes user similarity patterns, IBCF captures item co-occurrence structures, MF extracts linear latent factors, and NN models non-linear interactions. This diversity suggests that ensemble approaches combining multiple methods could improve coverage and robustness in cold-start scenarios. Since the synthetic user had no historical ratings, this analysis focuses on the plausibility and diversity of recommendations rather than predictive accuracy. Predicted ratings should therefore be interpreted as relative confidence scores rather than direct performance measures. Complete recommendation listings are provided in Appendix B.


```{r results-visualization, fig.cap="Cross-Validation Performance Comparison: Mean RMSE with ±1 SD error bars across four collaborative filtering methods."}
# ============================================================================
# VISUALIZATION: PERFORMANCE COMPARISON
# ============================================================================

# Ensure H2O is available for any remaining operations
if (!h2o::h2o.clusterIsUp()) {
  suppressMessages(h2o::h2o.init(ip = "localhost", port = 54321))
  h2o::h2o.no_progress()
}

# Create bar plot for RMSE comparison
ggplot(cv_comparison_results, aes(x = reorder(Method, CV_RMSE_Mean), y = CV_RMSE_Mean)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_errorbar(
    aes(ymin = CV_RMSE_Mean - CV_RMSE_SD, ymax = CV_RMSE_Mean + CV_RMSE_SD),
    width = 0.2
  ) +
  geom_text(aes(label = round(CV_RMSE_Mean, 3)), vjust = -0.5) +
  labs(
    title = "Cross-Validation Performance Comparison",
    subtitle = "Lower RMSE indicates better performance",
    x = "Method",
    y = "Mean RMSE"
  ) +
  theme_cowplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```



## 4.3 Catalog Size Impact

Analysis of catalog size reveals a non-monotonic relationship with prediction accuracy (Figure 3). Matrix factorization achieved optimal performance at 50 books, reaching an RMSE of 1.815 ± 0.027, representing a 10.4% improvement over the smallest catalog of 5 books (RMSE = 2.027). Performance remained stable across the 40–60 book range (RMSE 1.815–1.849, SE < 0.030), suggesting a "sweet spot" where the balance between data richness and model complexity is most favorable, with consistent predictions across multiple random seeds.

Sharp improvements occurred in the early phase: expanding from 5 to 10 books reduced RMSE by 7.3% (2.027 to 1.879), highlighting the critical importance of minimum data density, while increasing from 10 to 20 books added another 2.8% improvement (1.879 to 1.826). Performance continued to improve modestly through 30 books (RMSE = 1.841), after which gains became more variable. Beyond 50 books, performance plateaued and slightly declined, with the full 145-book catalog yielding an RMSE of 1.833 ± 0.011, indicating that expanding beyond 50–60 items yields diminishing returns and may introduce noise from less-rated titles.

This non-monotonic pattern reflects two competing effects. In the early phase, increasing the number of books improves latent factor estimation by providing richer training data. In the later phase, however, adding less-rated titles introduces noise that degrades model performance. Analysis of standard errors supports this interpretation: smaller catalogs (5–30 books) exhibit higher variability (SE 0.0153–0.0895), while larger catalogs (50+ books) demonstrate more consistent performance (SE 0.0057–0.0266). This increased stability reflects the model's ability to learn robust patterns with sufficient training data, even when absolute accuracy plateaus.

Sparsity increases steadily from 32.3% at 5 books to 94.0% at 145 books (Appendix C.2..........). Notably, the optimal performance at 50 books occurs at 86.7% sparsity, demonstrating that matrix factorization effectively handles highly sparse environments through latent factor modeling. This robustness highlights a key advantage of model-based collaborative filtering in sparse recommendation settings.





```{r}
# ============================================================================
# DATASET SIZE IMPACT ANALYSIS - FULLY CORRECTED
# ============================================================================
# Research Question: How many book titles does a recommender system need
# to achieve good prediction accuracy?
# 
# Methodology:
# 1. For each catalog size (5, 10, 20, 30, 50, 75, 100, 145 books)
# 2. Select TOP N most popular books
# 3. For EACH USER: Split their ratings 80/20 into train/test
# 4. Train MF model on 80%, test on 20%
# 5. Run 5 times with different random seeds for robustness
# 6. CRITICAL: Only include users who have rated enough books to split fairly
#
# Key: This ensures the same books appear in both train and test,
# and each user contributes proportionally to both sets
# ============================================================================

analyze_catalog_sizes_fair <- function(data, 
                                      catalog_sizes = c(5, 10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100, 110, 120, 130, 145),
                                      optimal_params_mf,
                                      test_fraction = 0.2,
                                      min_user_ratings = 3,  # Minimum ratings per user to participate
                                      n_seeds = 5,
                                      base_seed = 123) {
  
  set.seed(base_seed)
  all_results <- data.frame()
  
  # Get book popularity rankings
  book_popularity <- data %>%
    group_by(ISBN) %>%
    summarise(n_ratings = n(), .groups = "drop") %>%
    arrange(desc(n_ratings))
  
  # cat("=== Fair Catalog Size Analysis ===\n")
  # cat("Method: Per-user 80/20 split on TOP N popular books\n")
  # cat("Minimum ratings per user:", min_user_ratings, "\n")
  # cat("Testing catalog sizes:", paste(catalog_sizes, collapse = ", "), "\n")
  # cat("Total catalog sizes to test:", length(catalog_sizes), "\n\n")
  
  # Verify we have enough books
  if (nrow(book_popularity) < max(catalog_sizes)) {
    stop("Not enough books in dataset. Available: ", nrow(book_popularity), 
         ", Requested: ", max(catalog_sizes))
  }
  
  # Process each catalog size
  for (n_books in catalog_sizes) {
    # cat("=== Catalog Size:", n_books, "books ===\n")
    
    # Select TOP N most popular books
    selected_books <- head(book_popularity$ISBN, n_books)
    data_subset <- data %>% filter(ISBN %in% selected_books)
    
    # Count ratings per user in this subset
    user_rating_counts <- data_subset %>%
      group_by(User.ID) %>%
      summarise(n_ratings = n(), .groups = "drop")
    
    # Only keep users with enough ratings to split
    eligible_users <- user_rating_counts %>%
      filter(n_ratings >= min_user_ratings) %>%
      pull(User.ID)
    
    data_subset_filtered <- data_subset %>%
      filter(User.ID %in% eligible_users)
    
    n_eligible <- length(eligible_users)
    n_total_ratings <- nrow(data_subset_filtered)
    
    # cat("Selected", length(selected_books), "books\n")
    # cat("Eligible users (>=", min_user_ratings, "ratings):", n_eligible, "\n")
    # cat("Total ratings from eligible users:", n_total_ratings, "\n")
    
    if (n_eligible < 10) {
      # cat("SKIP: Not enough eligible users\n\n")
      
      all_results <- rbind(all_results, data.frame(
        catalog_size = n_books,
        seed = NA,
        n_users = 0,
        n_items = length(selected_books),
        n_ratings_total = n_total_ratings,
        n_explicit = NA,
        sparsity_pct = NA,
        n_train = NA,
        n_test = NA,
        rmse = NA,
        reason = "Insufficient eligible users"
      ))
      next
    }
    
    # Run multiple seeds
    for (seed_num in 1:n_seeds) {
      current_seed <- base_seed + seed_num
      set.seed(current_seed)
      
      # cat("  Seed", seed_num, "of", n_seeds, "...")
      
      # ================================================================
      # CRITICAL: PER-USER 80/20 SPLIT
      # ================================================================
      # For each user, randomly select 20% of their ratings for testing
      # This ensures EVERY user contributes to both train and test sets
      
      train_data_list <- list()
      test_data_list <- list()
      
      for (user in eligible_users) {
        user_ratings <- data_subset_filtered %>% 
          filter(User.ID == user)
        
        n_user_ratings <- nrow(user_ratings)
        n_test_user <- max(1, floor(test_fraction * n_user_ratings))
        
        # Randomly select test indices for this user
        test_indices <- sample(1:n_user_ratings, n_test_user)
        
        test_data_list[[as.character(user)]] <- user_ratings[test_indices, ]
        train_data_list[[as.character(user)]] <- user_ratings[-test_indices, ]
      }
      
      # Combine all users' train and test data
      train_data_final <- bind_rows(train_data_list)
      test_data_final <- bind_rows(test_data_list)
      
      # cat("Train:", nrow(train_data_final), "ratings, ",
      #     "Test:", nrow(test_data_final), "ratings...")
      
      # ================================================================
      # CREATE MATRICES
      # ================================================================
      
      # Create FULL matrix first (to get consistent dimensions)
      matrix_result_full <- create_user_item_matrix(
        data_subset_filtered,
        min_ratings_per_book = 1,  # Don't filter - we already selected books
        min_ratings_per_user = 1,  # Don't filter - we already selected users
        implicit_rating = 4,
        implicit_confidence = 0.5
      )
      
      full_matrix <- matrix_result_full$ratings
      full_original <- matrix_result_full$original_ratings
      
      # Create TRAIN matrix (mask test ratings)
      train_matrix <- full_matrix
      train_original <- full_original
      
      # Mask out test ratings
      for (i in 1:nrow(test_data_final)) {
        user_id <- as.character(test_data_final$User.ID[i])
        book_id <- test_data_final$ISBN[i]
        
        if (user_id %in% rownames(train_matrix) && 
            book_id %in% colnames(train_matrix)) {
          train_matrix[user_id, book_id] <- NA
          train_original[user_id, book_id] <- NA
        }
      }
      
      # Calculate metrics
      n_users <- nrow(full_matrix)
      n_items <- ncol(full_matrix)
      n_ratings_total <- sum(!is.na(full_matrix))
      n_explicit_total <- sum(full_original[!is.na(full_original)] > 0)
      sparsity <- mean(is.na(full_matrix)) * 100
      
      # Prepare train data for MF (explicit only)
      train_observed <- which(!is.na(train_matrix), arr.ind = TRUE)
      
      train_data_mf <- data.frame(
        user_index = train_observed[, 1] - 1,
        item_index = train_observed[, 2] - 1,
        rating = train_matrix[train_observed],
        original_rating = train_original[train_observed]
      )
      
      # Filter to explicit ratings only
      train_data_explicit <- train_data_mf %>%
        filter(original_rating > 0)
      
      n_train_explicit <- nrow(train_data_explicit)
      
      if (n_train_explicit < 10) {
        # cat(" SKIP (only", n_train_explicit, "explicit training ratings)\n")
        
        all_results <- rbind(all_results, data.frame(
          catalog_size = n_books,
          seed = seed_num,
          n_users = n_users,
          n_items = n_items,
          n_ratings_total = n_ratings_total,
          n_explicit = n_explicit_total,
          sparsity_pct = sparsity,
          n_train = n_train_explicit,
          n_test = NA,
          rmse = NA,
          reason = "Insufficient training data"
        ))
        next
      }
      
      # ================================================================
      # TRAIN MODEL
      # ================================================================
      
      # Create mapping for users and items
      user_ids_map <- data.frame(
        user_index = 0:(n_users - 1),
        user_id = rownames(train_matrix)
      )
      
      item_ids_map <- data.frame(
        item_index = 0:(n_items - 1),
        item_id = colnames(train_matrix)
      )
      
      # Train MF model
      train_set <- data_memory(
        user_index = train_data_explicit$user_index,
        item_index = train_data_explicit$item_index,
        rating = train_data_explicit$rating,
        index1 = FALSE
      )
      
      model <- Reco()
      suppressWarnings({
        model$train(train_set, opts = list(
          dim = optimal_params_mf$min$dim,
          lrate = optimal_params_mf$min$lrate,
          costp_l2 = optimal_params_mf$min$costp_l2,
          costq_l2 = optimal_params_mf$min$costq_l2,
          niter = 50,
          nthread = 4,
          verbose = FALSE
        ))
      })
      
      # ================================================================
      # MAKE PREDICTIONS ON TEST SET
      # ================================================================
      
      predictions <- numeric(nrow(test_data_final))
      test_actuals <- numeric(nrow(test_data_final))
      test_original_ratings <- numeric(nrow(test_data_final))
      valid_predictions <- rep(TRUE, nrow(test_data_final))
      
      for (i in 1:nrow(test_data_final)) {
        user_id <- as.character(test_data_final$User.ID[i])
        book_id <- test_data_final$ISBN[i]
        
        # Check if user and book exist in matrices
        if (!(user_id %in% user_ids_map$user_id) || 
            !(book_id %in% item_ids_map$item_id)) {
          valid_predictions[i] <- FALSE
          next
        }
        
        # Get indices
        user_idx <- user_ids_map$user_index[user_ids_map$user_id == user_id]
        item_idx <- item_ids_map$item_index[item_ids_map$item_id == book_id]
        
        # Make prediction
        pred_set <- data_memory(
          user_index = user_idx,
          item_index = item_idx,
          index1 = FALSE
        )
        
        pred <- model$predict(pred_set, out_memory())
        predictions[i] <- pmax(pmin(pred, 10), 1)
        
        # Get actual rating
        test_actuals[i] <- full_matrix[user_id, book_id]
        test_original_ratings[i] <- full_original[user_id, book_id]
      }
      
      # Filter to valid predictions and explicit ratings only
      valid_mask <- valid_predictions & (test_original_ratings > 0)
      n_test_explicit <- sum(valid_mask)
      
      if (n_test_explicit == 0) {
        # cat(" SKIP (no valid explicit test ratings)\n")
        
        all_results <- rbind(all_results, data.frame(
          catalog_size = n_books,
          seed = seed_num,
          n_users = n_users,
          n_items = n_items,
          n_ratings_total = n_ratings_total,
          n_explicit = n_explicit_total,
          sparsity_pct = sparsity,
          n_train = n_train_explicit,
          n_test = 0,
          rmse = NA,
          reason = "No valid test ratings"
        ))
        next
      }
      
      # Calculate RMSE on explicit test ratings
      pred_explicit <- predictions[valid_mask]
      actual_explicit <- test_actuals[valid_mask]
      
      rmse <- sqrt(mean((pred_explicit - actual_explicit)^2, na.rm = TRUE))
      
      # Store results
      all_results <- rbind(all_results, data.frame(
        catalog_size = n_books,
        seed = seed_num,
        n_users = n_users,
        n_items = n_items,
        n_ratings_total = n_ratings_total,
        n_explicit = n_explicit_total,
        sparsity_pct = sparsity,
        n_train = n_train_explicit,
        n_test = n_test_explicit,
        rmse = rmse,
        reason = "Success"
      ))
      
      # cat(" RMSE =", round(rmse, 4), "\n")
    }
    
    # cat("\n")
  }
  
  # cat("=== Analysis Complete ===\n\n")
  
  # ================================================================
  # AGGREGATE RESULTS
  # ================================================================
  
  summary_results <- all_results %>%
    group_by(catalog_size) %>%
    summarise(
      n_valid_runs = sum(!is.na(rmse)),
      rmse_mean = mean(rmse, na.rm = TRUE),
      rmse_sd = sd(rmse, na.rm = TRUE),
      rmse_se = sd(rmse, na.rm = TRUE) / sqrt(sum(!is.na(rmse))),
      rmse_min = min(rmse, na.rm = TRUE),
      rmse_max = max(rmse, na.rm = TRUE),
      n_users_mean = round(mean(n_users, na.rm = TRUE), 0),
      n_items_mean = round(mean(n_items, na.rm = TRUE), 0),
      sparsity_mean = round(mean(sparsity_pct, na.rm = TRUE), 1),
      n_explicit_mean = round(mean(n_explicit, na.rm = TRUE), 0),
      n_train_mean = round(mean(n_train, na.rm = TRUE), 0),
      n_test_mean = round(mean(n_test, na.rm = TRUE), 0),
      .groups = "drop"
    ) %>%
    arrange(catalog_size) %>%
    mutate(
      rmse_change = rmse_mean - lag(rmse_mean),
      improvement = -rmse_change,
      percent_improvement = -rmse_change / lag(rmse_mean) * 100
    )
  
  return(list(
    detailed = all_results,
    summary = summary_results
  ))
}

# ============================================================================
# RUN THE ANALYSIS
# ============================================================================

# catalog_analysis_results <- analyze_catalog_sizes_fair(
#   data = data,
#   catalog_sizes = c(5, 10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100, 110, 120, 130, 145),
#   optimal_params_mf = optimal_params_mf,
#   test_fraction = 0.2,
#   min_user_ratings = 3,  
#   n_seeds = 5,
#   base_seed = 123
# )
# 
# # Extract results
# size_impact_results <- catalog_analysis_results$summary
# size_impact_results_detailed <- catalog_analysis_results$detailed
# 
# save(size_impact_results, size_impact_results_detailed, 
#      file = "dataset_size_analysis_complete.Rdata")


# Table moved to Appendix C.2 for detailed reference

```


```{r dataset-size-visualization, fig.cap="Catalog Size Impact on Prediction Accuracy: Mean RMSE (±1 SE) across catalog sizes from 5 to 145 books. Red diamond marks optimal size (50 books, RMSE=1.815)."}

# RMSE Trend with Error Bars
ggplot(size_impact_results, aes(x = catalog_size, y = rmse_mean)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  geom_errorbar(aes(ymin = rmse_mean - rmse_se, ymax = rmse_mean + rmse_se),
                width = 3, color = "steelblue", alpha = 0.7) +
  geom_point(data = size_impact_results[which.min(size_impact_results$rmse_mean), ],
             color = "red", size = 5, shape = 18) +
  labs(title = "A. RMSE by Catalog Size",
       subtitle = "Error bars show ±1 SE across 5 seeds. Red diamond indicates optimal size.",
       x = "Number of Books", y = "Mean RMSE") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

```






```{r cleanup, include=FALSE}
# Clean up H2O cluster at the end
if (exists("h2o") && h2o::h2o.clusterIsUp()) {
  h2o::h2o.shutdown(prompt = FALSE)
}
```






# 5. Discussion

This study compared four collaborative filtering approaches for book recommendation, two memory-based (UBCF, IBCF) and two model-based (MF, NN), to understand how algorithm design, data sparsity, and catalog size influence accuracy and recommendation diversity.

## 5.1 Comparative Performance

Model-based methods substantially outperformed memory-based ones, confirming their superior robustness to data sparsity. Matrix Factorization achieved the best overall balance of accuracy and stability, while the Neural Network model performed comparably, reinforcing the advantage of latent representation learning in capturing hidden user–item structures. These results align with established findings that latent factor models generalize better than similarity-based techniques when data are sparse. In contrast, the memory-based algorithms (UBCF and IBCF) performed notably worse due to their reliance on explicit pairwise similarities, which become unreliable when users have limited rating overlap. The 43% performance gap between the best (MF) and worst (UBCF) models underscores how algorithmic structure critically shapes predictive accuracy under sparse conditions. Between the two, IBCF performed better—consistent with the notion that item–item relationships are more stable than user–user ones, as items exhibit recurring co-occurrence patterns even when individual user histories are incomplete.




## 5.2 Catalog Size Dynamics


The catalog size experiment revealed a non-monotonic relationship between dataset breadth and accuracy, with MF performance improving sharply up to 50 books (optimal RMSE = 1.815 ± 0.027) before plateauing and slightly worsening beyond that point. This indicates diminishing returns in predictive performance as the catalog expands beyond a critical density threshold. The improvement phase (from 5 to 50 books) coincides with rapidly increasing interaction diversity, enabling better estimation of user and item latent factors. However, once sufficient diversity is achieved, adding less-rated titles introduces noise that slightly degrades accuracy. The optimal performance at 86.7% sparsity demonstrates that MF effectively manages high sparsity but still benefits from a balanced ratio of users to items. The decreasing variability in RMSE beyond 50 books (SE < 0.03) also reflects greater model stability once sufficient data density is reached. Collectively, these findings suggest that recommender systems benefit most from moderate catalog sizes with adequate interaction density, rather than from maximizing the number of items without ensuring sufficient ratings per item.






## 5.3 Recommendation Divergence


The 82.5% uniqueness among recommended titles highlights fundamental differences in how collaborative filtering approaches model user preferences. Memory-based methods (UBCF and IBCF) rely on explicit similarity metrics but operate on orthogonal principles: UBCF forms predictions from similar users, while IBCF uses item co-occurrence patterns. This explains the minimal overlap between them—only two shared titles—and UBCF's higher uniqueness rate (70%), reflecting its sensitivity to individual rating behavior and tendency toward more personalized but less generalizable recommendations. Model-based approaches instead infer latent preference structures. Matrix Factorization captures linear relationships within a compact embedding space, producing more conservative, globally optimized predictions. Neural networks extend this to nonlinear mappings, revealing higher-order interactions that MF cannot represent directly. Their outputs overlapped on only one title, illustrating how different representational biases influence recommendation diversity even within model-based methods. The single title common to three models - Harry Potter and the Order of the Phoenix - likely reflects a strong global preference signal detectable across both similarity-based and latent approaches. Rating discrepancies for shared titles (up to 1.36 points) further suggest distinct confidence calibrations across algorithms, an insight that could inform ensemble methods aiming to blend complementary model behaviors.

////////////////// Detailed recommendation examples demonstrating these differences are provided in Appendix B. 



## 5.4 Cold-Start Limitations

The cold-start analysis highlights a key methodological limitation in comparing collaborative filtering approaches within this study. Model-based methods, including Matrix Factorization and Neural Networks, were disadvantaged by the need to use simplified fallback strategies that differ fundamentally from their core learning mechanisms. This explains their limited interpretability and weaker relevance to real-world cold-start scenarios. The findings underscore a fundamental trade-off, with model-based methods achieving superior accuracy for existing users but requiring auxiliary strategies for new ones, while memory-based methods, though less accurate overall, remain algorithmically consistent across both cases. In practical systems, effective cold-start handling typically relies on hybrid solutions such as integrating content-based features, transfer learning, or embedding initialization to restore the representational advantages of model-based algorithms. Therefore, the cold-start results should not be interpreted as evidence that memory-based methods outperform model-based ones. Rather, they reveal the need for hybrid or extended designs when latent representation models encounter users with no prior interaction data.


# 6. Conclusion

The study implemented and rigorously evaluated four collaborative filtering approaches for book recommendation using a modified Book-Crossing dataset, providing both practical guidance and theoretical insights for recommender system design. Cross-validation under consistent conditions revealed clear differences in predictive performance, highlighted algorithm-specific strengths and limitations, and offered evidence on how catalog size and cold-start strategies influence recommendation quality.

MF was the most accurate method (RMSE = 1.820), while confidence-weighted similarity effectively leveraged implicit feedback to enhance memory-based methods. Catalog size analysis showed that accuracy gains plateaued beyond 50–60 books, indicating that near-optimal performance can be achieved without expansive catalogs.

Limitations include the 145-book dataset, which may not capture full recommendation diversity, the need for refined handling of implicit feedback, and suboptimal cold-start strategies for model-based systems that do not fully exploit learned representations. Future work should explore hybrid ensembles combining complementary methods, advanced deep learning architectures (autoencoders, attention mechanisms), context-aware metadata (genres, authors, temporal patterns), and online learning for real-time adaptation. Overall, this study offers both theoretical insights and practical guidance for developing production-ready recommendation systems across domains.










# Appendix: Detailed Results and Supporting Analysis

This appendix contains detailed statistical summaries, individual method results, and technical details that support the main analysis but are not essential to the core narrative.

## Appendix A: Detailed Exploratory Data Analysis

```{r appendix-detailed-eda}
# ============================================================================
# DETAILED USER AND BOOK ACTIVITY ANALYSIS
# ============================================================================

# User activity analysis
user_activity <- data %>%
  group_by(User.ID) %>%
  summarise(
    num_ratings = n(),
    avg_rating = mean(Book.Rating, na.rm = TRUE),
    .groups = "drop"
  )

# Book popularity analysis
book_popularity <- data %>%
  group_by(ISBN) %>%
  summarise(
    num_ratings = n(),
    avg_rating = mean(Book.Rating, na.rm = TRUE),
    .groups = "drop"
  )

# User activity summary table
user_activity_summary <- data.frame(
  Metric = c(
    "Total Users",
    "Min Ratings per User", 
    "Max Ratings per User",
    "Mean Ratings per User",
    "Users with 1-2 Ratings",
    "Users with 3-5 Ratings", 
    "Users with 6-10 Ratings",
    "Users with 11-20 Ratings",
    "Users with >20 Ratings"
  ),
  Value = c(
    nrow(user_activity),
    min(user_activity$num_ratings),
    max(user_activity$num_ratings),
    round(mean(user_activity$num_ratings), 2),
    sum(user_activity$num_ratings <= 2),
    sum(user_activity$num_ratings >= 3 & user_activity$num_ratings <= 5),
    sum(user_activity$num_ratings >= 6 & user_activity$num_ratings <= 10),
    sum(user_activity$num_ratings >= 11 & user_activity$num_ratings <= 20),
    sum(user_activity$num_ratings > 20)
  )
)

kable(user_activity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed User Activity Distribution: Statistical breakdown of user rating patterns showing distribution of ratings per user and user activity categories",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Book popularity summary table
book_popularity_summary <- data.frame(
  Metric = c(
    "Total Books",
    "Min Ratings per Book",
    "Max Ratings per Book", 
    "Mean Ratings per Book",
    "Books with 1-4 Ratings",
    "Books with 5-9 Ratings",
    "Books with 10-19 Ratings", 
    "Books with 20-49 Ratings",
    "Books with ≥50 Ratings"
  ),
  Value = c(
    nrow(book_popularity),
    min(book_popularity$num_ratings),
    max(book_popularity$num_ratings),
    round(mean(book_popularity$num_ratings), 2),
    sum(book_popularity$num_ratings <= 4),
    sum(book_popularity$num_ratings >= 5 & book_popularity$num_ratings <= 9),
    sum(book_popularity$num_ratings >= 10 & book_popularity$num_ratings <= 19),
    sum(book_popularity$num_ratings >= 20 & book_popularity$num_ratings <= 49),
    sum(book_popularity$num_ratings >= 50)
  )
)

kable(book_popularity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Book Popularity Distribution: Statistical analysis of item rating patterns showing distribution of ratings per book and popularity categories",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Rating distribution (numeric breakdown)
rating_dist <- table(data$Book.Rating)
rating_dist_df <- data.frame(
  Rating = names(rating_dist),
  Count = as.numeric(rating_dist),
  Percentage = round(as.numeric(rating_dist) / sum(rating_dist) * 100, 1)
)

kable(rating_dist_df, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Rating Distribution: Frequency analysis of explicit and implicit feedback showing count and percentage for each rating value",
      align = c("c", "c", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Dataset sparsity analysis
total_possible_ratings <- length(unique(data$User.ID)) * length(unique(data$ISBN))
actual_ratings <- nrow(data)
sparsity_percentage <- round((1 - actual_ratings / total_possible_ratings) * 100, 2)

sparsity_summary <- data.frame(
  Metric = c(
    "Total Possible User-Book Pairs",
    "Actual Ratings",
    "Sparsity Percentage"
  ),
  Value = c(
    format(total_possible_ratings, big.mark = ","),
    format(actual_ratings, big.mark = ","),
    paste0(sparsity_percentage, "%")
  )
)

kable(sparsity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Dataset Sparsity Analysis: Statistical summary of matrix sparsity showing total possible user-book pairs, actual ratings, and sparsity percentage",
      align = c("l", "c")
) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

## Appendix B: Method-Specific Recommendations

The following tables show detailed recommendation outputs for each method, demonstrating the practical application of each approach.

### B.1 UBCF Recommendations

```{r appendix-ubcf-recommendations, include=FALSE}
# UBCF recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ubcf <- recommend_for_user_ubcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_sim_matrix = user_similarity_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

recs_ubcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "UBCF Recommendations: Top 10 book suggestions for existing user based on user similarity with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# UBCF recommendations for new user
sample_books <- colnames(user_item_matrix)[1:5]
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), sample_books)

new_user_recs_ubcf <- recommend_for_new_user_ubcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

new_user_recs_ubcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "UBCF Cold-Start Recommendations: Top 10 book suggestions for new user using similarity-based approach with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

### B.2 IBCF Recommendations

```{r appendix-ibcf-recommendations, include=FALSE}
# IBCF recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ibcf <- recommend_for_user_ibcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_sim_matrix = item_similarity_matrix,
  item_means = item_means_ibcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

recs_ibcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "IBCF Recommendations: Top 10 book suggestions for existing user based on item similarity with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# IBCF recommendations for new user
new_user_recs_ibcf <- recommend_for_new_user_ibcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_means = item_means_ibcf,
  item_sim_matrix = item_similarity_matrix,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

new_user_recs_ibcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "IBCF Cold-Start Recommendations: Top 10 book suggestions for new user using item-based approach with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

### B.3 MF Recommendations

```{r appendix-mf-recommendations, include=FALSE}
# MF recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_mf <- recommend_for_user_mf(
  model = model_mf,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids_map = prepared$user_ids,
  item_ids_map = prepared$item_ids,
  book_info = book_info
)

recs_mf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF Recommendations: Top 10 book suggestions for existing user using matrix factorization with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# MF recommendations for new user
recs_new_mf <- recommend_for_new_user_mf(
  model = model_mf,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  item_ids_map = prepared$item_ids,
  book_info = book_info,
  n_recommendations = 10
)

recs_new_mf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF Cold-Start Recommendations: Top 10 book suggestions for new user using hybrid approach with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

### B.4 NN Recommendations

```{r appendix-nn-recommendations, include=FALSE}
# NN recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_nn <- recommend_for_user_nn(
  model = model_nn,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids = prepared$user_ids,
  book_ids = prepared$book_ids,
  user_means = prepared$user_means,
  book_info = book_info
)

recs_nn %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "NN Recommendations: Top 10 book suggestions for existing user using neural network with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# NN recommendations for new user
new_user_recs_nn <- recommend_for_new_user_nn(
  model = model_nn,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  book_ids = prepared$book_ids,
  book_info = book_info,
  n_recommendations = 10,
  user_means = prepared$user_means
)

new_user_recs_nn %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "NN Cold-Start Recommendations: Top 10 book suggestions for new user using similarity-based fallback with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

## Appendix C: Model Training Details

### C.1 MF Cross-Validation Folds

```{r appendix-mf-cv-details, include=FALSE}
cv_results_mf %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF Cross-Validation Results: Individual fold performance across 5-fold cross-validation showing RMSE for each fold",
    digits = 3,
    col.names = c("Fold", "RMSE"),
    align = c("c", "c")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

### C.2 Catalog Size Impact Analysis

```{r appendix-catalog-size-table, include=FALSE}
kable(
  size_impact_results %>% 
    select(catalog_size, rmse_mean, rmse_se, sparsity_mean),
  format = "latex",
  booktabs = TRUE,
  caption = "Catalog Size Impact Analysis: Detailed RMSE performance across different catalog sizes showing mean RMSE, standard error, and sparsity percentage",
  digits = 3,
  col.names = c("Catalog Size", "Mean RMSE", "SE", "Sparsity %"),
  align = c("c", "c", "c", "c")
) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```
