---
title: "Ensemble Recommender System for Book Recommendations: A Comparative Analysis of Collaborative Filtering Approaches"
author: "Hope Hennessy"
date: "2025-10-01"
output:
  pdf_document:
    latex_engine: xelatex
---

# Abstract

This study compares four collaborative filtering approaches for book recommendation systems using a modified Book-Crossing dataset (45,370 ratings, 10,000 users, 145 books). We implement item-based (IBCF) and user-based (UBCF) collaborative filtering from scratch, alongside matrix factorization (MF) and neural network (NN) models. Results show matrix factorization achieves the lowest prediction error (RMSE = 1.820), while incorporating confidence-weighted similarity improves memory-based methods by leveraging implicit feedback. Accuracy gains plateau around 50 titles, suggesting diminishing returns beyond this range. These findings offer practical guidance for designing efficient hybrid recommenders in high-sparsity environments.

# 1. Introduction

Recommender systems help users navigate information overload by predicting preferences based on patterns in user–item interactions. Collaborative filtering (CF), a widely used approach, infers user interests from historical behavior and powers major platforms such as Amazon and Netflix, enhancing engagement through personalized suggestions.

This study employs a modified Book-Crossing dataset containing 45,370 ratings from 10,000 users across 145 books. Ratings range from 0 to 10, with 0 representing implicit feedback ("read but not rated") and 1–10 representing explicit evaluations. The combination of explicit and implicit ratings presents challenges for collaborative filtering in sparse environments, requiring strategies to extract meaningful signals from limited data.

The objectives of this study are to: (1) implement and evaluate four collaborative filtering methods—user-based (UBCF), item-based (IBCF), matrix factorization (MF), and neural network-based (NN); (2) compare their predictive performance using cross-validation; (3) assess robustness to cold-start scenarios; and (4) investigate how catalog size affects prediction accuracy.

Memory-based methods (UBCF and IBCF) rely on similarity computations directly from the user–item matrix, while model-based approaches (MF and NN) learn latent representations that capture hidden preference patterns in lower-dimensional spaces. To address sparsity and incorporate implicit feedback, we employ confidence-weighted similarity measures, enabling more accurate and robust recommendations across diverse scenarios.

# 2. Data Understanding and Preparation

## 2.1 Data Description

The dataset consists of three components: book_ratings, which records user–book interactions with variables User.ID, ISBN, and Book.Rating on a 0–10 scale; book_info, which provides metadata such as Book.Title and Book.Author; and user_info, which contains limited demographic details like age but is excluded from this analysis to focus on collaborative filtering. Ratings from 1–10 represent explicit feedback, where users actively rated books, while a rating of 0 indicates implicit feedback, meaning the user read the book but did not provide a score.

\newpage


```{r setup, include=FALSE}
# ============================================================================
# GLOBAL SETTINGS AND CONFIGURATION
# ============================================================================

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6,
  fig.show = "hold",
  out.width = "80%",
  dpi = 300,
  crop = NULL
)

# Set random seed for reproducibility
set.seed(123)

# ============================================================================
# LIBRARY IMPORTS
# ============================================================================

library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(cowplot)

library(knitr)
library(kableExtra)

library(caret)
library(recosystem)
library(h2o)

options(
  lintr.linter_file = "none",
  lintr.exclude_linters = c(
    "object_usage_linter",
    "object_name_linter",
    "cyclocomp_linter",
    "line_length_linter"
  )
)


# ============================================================================
# HELPER FUNCTIONS FOR TABLE FORMATTING
# ============================================================================

# Function to truncate long text for table display
truncate_text <- function(text, max_length = 50) {
  ifelse(nchar(text) > max_length, 
         paste0(substr(text, 1, max_length - 3), "..."), 
         text)
}

# Function to format recommendation tables consistently
format_recommendation_table <- function(data, caption, method_name) {
  data %>%
    mutate(Rank = row_number()) %>%
    mutate(
      Book.Title = truncate_text(Book.Title, 50),
      Book.Author = truncate_text(Book.Author, 25)
    ) %>%
    select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
    kable(
      format = "latex",
      booktabs = TRUE,
      caption = paste(method_name, caption),
      digits = 2,
      align = c("c", "l", "l", "c"),
      col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
    ) %>%
    kable_styling(
      latex_options = c("HOLD_position", "scale_down"),
      font_size = 10
    )
}

# Load saved results
load('nn_tuning_results.Rdata')
load('mf_tuning_results.Rdata')
load('mf_final_model_mf.Rdata')
load('mf_tuning_results.Rdata')
load('all_cv_results.Rdata')
load('dataset_size_analysis_complete.Rdata')

```

## 2.2 Exploratory Data Analysis

As shown in Table 1, the dataset contains 10,000 users, 145 books, and 45,370 interactions, with an average rating of 3.16 (including implicit ratings) and 7.8 for explicit ratings. Matrix sparsity is high (96.9% before filtering, 99.2% after), posing challenges for collaborative filtering and motivating our confidence-weighting approach. 

The dataset exhibits typical recommendation system characteristics (Figure 1). Panel A shows a right-skewed distribution of explicit ratings (mean 7.8), reflecting a positivity bias. Panel B shows a long-tail user activity distribution, with most users rating fewer than 10 books. Panel C indicates no strong correlation between book popularity and average rating, reducing popularity bias. Panel D reveals 40.3% explicit and 59.7% implicit ratings. 

To address the challenge of mixed feedback, we employ confidence weighting where explicit ratings receive weight 1.0 and implicit ratings (converted to a rating of 4) receive weight 0.5. In this approach, the neutral rating of 4 was used to represent moderate interest, while the 0.5 confidence weight down-weights less informative signals yet still incorporates implicit interactions. Detailed statistical breakdowns of user activity, book popularity, and rating distributions are provided in Appendix A.

```{r data-loading, include=FALSE}
# ============================================================================
# DATA LOADING
# ============================================================================

load("book_ratings.Rdata")

head(book_info)
str(book_info)
dim(book_info)
head(book_ratings)
str(book_ratings)
dim(book_ratings)
head(user_info) # will not be using age in this report
str(user_info)
dim(user_info)

# Check for missing values
missing_summary <- data.frame(
  Dataset = c("Book Info", "Book Ratings", "User Info"),
  Missing_Values = c(
    sum(is.na(book_info)),
    sum(is.na(book_ratings)),
    sum(is.na(user_info))   # missing age values
  )
)

```


```{r data-cleaning}
# ============================================================================
# DUPLICATE DETECTION AND HANDLING
# ============================================================================

# Initial data merging
initial_data <- book_ratings %>%
  left_join(book_info, by = "ISBN")

# Clean book titles and authors for duplicate detection
book_info_clean <- book_info %>%
  mutate(
    clean_title = Book.Title %>%
      tolower() %>%
      gsub('\\(.*?\\)', '', .) %>%
      gsub('\\s*:\\s*.*$', '', .) %>%
      gsub('^(the|a|an)\\s+', '', .) %>%
      gsub('[^a-z0-9\\s-]', '', .) %>%
      gsub('\\s+', ' ', .) %>%
      trimws(),
    clean_author = Book.Author %>%
      tolower() %>%
      gsub('\\s+jr\\.*$', '', .) %>%
      gsub('\\s+sr\\.*$', '', .) %>%
      gsub('\\s+(iii|ii)$', '', .) %>%
      gsub('\\s+', ' ', .) %>%
      trimws()
  )

# Identify duplicates
duplicates <- book_info_clean %>%
  group_by(clean_title, clean_author) %>%
  filter(n() > 1) %>%
  arrange(clean_title, clean_author, ISBN) %>%
  ungroup()

# ============================================================================
# MERGE RATINGS FUNCTION
# ============================================================================

merge_ratings <- function(ratings_data, book_info_data, duplicates_data) {
  
  if (nrow(duplicates_data) > 0) {
    # Create canonical ISBN mapping
    duplicate_mapping <- duplicates_data %>%
      group_by(clean_title, clean_author) %>%
      arrange(ISBN) %>%
      mutate(canonical_ISBN = first(ISBN)) %>%
      ungroup() %>%
      select(ISBN, canonical_ISBN)
    
    # Create complete mapping
    all_books_mapping <- book_info_data %>%
      select(ISBN) %>%
      left_join(duplicate_mapping, by = "ISBN") %>%
      mutate(canonical_ISBN = ifelse(is.na(canonical_ISBN), ISBN, canonical_ISBN))
    
    # Apply mapping to ratings
    ratings_mapped <- ratings_data %>%
      left_join(all_books_mapping, by = "ISBN") %>%
      mutate(
        original_ISBN = ISBN,
        ISBN = canonical_ISBN
      ) %>%
      select(-canonical_ISBN, -original_ISBN)
    
    # Keep highest rating per user-book combination
    ratings_merged <- ratings_mapped %>%
      group_by(User.ID, ISBN) %>%
      slice_max(Book.Rating, n = 1, with_ties = FALSE) %>%
      ungroup()
    
    # Update book info with cleaned titles and authors
    book_info_updated <- book_info_data %>%
      left_join(all_books_mapping, by = "ISBN") %>%
      filter(ISBN == canonical_ISBN) %>%
      mutate(
        Book.Title = Book.Title %>%
          gsub('\\(.*?\\)', '', .) %>%  # Remove parenthetical info
          gsub('\\s*:\\s*.*$', '', .) %>%  # Remove subtitle after colon
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws(),  # Remove leading/trailing whitespace
        Book.Author = Book.Author %>%
          gsub('\\s+jr\\.*$', '', .) %>%  # Remove "Jr." suffixes
          gsub('\\s+sr\\.*$', '', .) %>%  # Remove "Sr." suffixes
          gsub('\\s+(iii|ii)$', '', .) %>%  # Remove "II", "III" suffixes
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws()  # Remove leading/trailing whitespace
      ) %>%
      select(ISBN, Book.Title, Book.Author)
    
  } else {
    ratings_merged <- ratings_data
    # Clean all book titles and authors even when no duplicates
    book_info_updated <- book_info_data %>%
      mutate(
        Book.Title = Book.Title %>%
          gsub('\\(.*?\\)', '', .) %>%  # Remove parenthetical info
          gsub('\\s*:\\s*.*$', '', .) %>%  # Remove subtitle after colon
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws(),  # Remove leading/trailing whitespace
        Book.Author = Book.Author %>%
          gsub('\\s+jr\\.*$', '', .) %>%  # Remove "Jr." suffixes
          gsub('\\s+sr\\.*$', '', .) %>%  # Remove "Sr." suffixes
          gsub('\\s+(iii|ii)$', '', .) %>%  # Remove "II", "III" suffixes
          gsub('\\s+', ' ', .) %>%  # Replace multiple spaces with single space
          trimws()  # Remove leading/trailing whitespace
      ) %>%
      select(ISBN, Book.Title, Book.Author)
    all_books_mapping <- NULL
  }
  
  return(list(
    ratings_merged = ratings_merged,
    book_info_updated = book_info_updated,
    mapping = all_books_mapping
  ))
}

# Execute merge
merged_result <- merge_ratings(book_ratings, book_info, duplicates)

# Create final clean dataset
data <- merged_result$ratings_merged %>%
  left_join(merged_result$book_info_updated, by = "ISBN") %>%
  select(User.ID, ISBN, Book.Rating, Book.Title, Book.Author)

book_info <- merged_result$book_info_updated
```

## 2.3 Data Preprocessing and Filtering

Duplicate book entries, where the same title and author appeared under different ISBNs, were identified using normalized string matching and consolidated to a canonical ISBN, with each user's highest rating retained.

To ensure reliable collaborative filtering, books with ≥5 ratings and users with ≥3 ratings were retained, resulting in 4,200 active users (58.0% of users excluded) and all 145 books included.




```{r eda-summary}
# ============================================================================
# DATASET SUMMARY STATISTICS
# ============================================================================

# Basic statistics
data_summary <- data.frame(
  Metric = c(
    "Total Users",
    "Total Books",
    "Total Ratings",
    "Average Rating",
    "Rating Range",
    "Matrix Sparsity"
  ),
  Value = c(
    length(unique(data$User.ID)),
    length(unique(data$ISBN)),
    nrow(data),
    round(mean(data$Book.Rating, na.rm = TRUE), 2),
    paste(min(data$Book.Rating), "-", max(data$Book.Rating)),
    paste0(
      round(
        (1 - nrow(data) / 
          (length(unique(data$User.ID)) * length(unique(data$ISBN)))) * 100,
        2
      ),
      "%"
    )
  )
)

kable(data_summary,
      format = "latex",
      booktabs = TRUE,
      caption = "Dataset Overview: Summary statistics for book recommendation analysis including user count, book count, rating statistics, and matrix sparsity",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

```




```{r eda-visualizations, fig.cap="Key Dataset Characteristics: (A) Explicit rating distribution (mean 7.8), (B) User activity patterns showing long-tail distribution, (C) Book popularity versus average rating, (D) Explicit (40.3%) versus implicit (59.7%) feedback proportions. Confidence weighting (1.0 for explicit, 0.5 for implicit) enables leveraging all interactions while appropriately weighting signal quality."}

  
# ============================================================================
# FOCUSED VISUALIZATIONS: MOST INFORMATIVE FOR RECOMMENDER SYSTEMS
# ============================================================================

# 1. EXPLICIT RATING DISTRIBUTION (Most Important)
explicit_data <- data %>% filter(Book.Rating > 0)

p1 <- ggplot(explicit_data, aes(x = Book.Rating)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(explicit_data$Book.Rating), 
             linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = mean(explicit_data$Book.Rating), y = max(table(explicit_data$Book.Rating)) * 0.9, 
           label = paste("Mean =", round(mean(explicit_data$Book.Rating), 1)), 
           color = "red", hjust = -0.1, size = 3.5) +
  labs(
    title = "Distribution of Explicit Ratings",
    x = "Rating",
    y = "Frequency"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 2. SPARSITY ANALYSIS (Critical for CF)
sparsity_data <- data %>%
  group_by(User.ID) %>%
  summarise(
    total_books_rated = n(),
    explicit_ratings = sum(Book.Rating > 0),
    implicit_ratings = sum(Book.Rating == 0),
    .groups = "drop"
  ) %>%
  mutate(
    explicit_ratio = explicit_ratings / total_books_rated,
    user_type = case_when(
      explicit_ratio == 1 ~ "Only Explicit",
      explicit_ratio == 0 ~ "Only Implicit", 
      explicit_ratio > 0.5 ~ "Mostly Explicit",
      TRUE ~ "Mostly Implicit"
    )
  )

p2 <- ggplot(sparsity_data, aes(x = total_books_rated)) +
  geom_histogram(binwidth = 1, fill = "lightcoral", color = "white", alpha = 0.8) +
  scale_x_continuous(limits = c(0, quantile(sparsity_data$total_books_rated, 0.95))) +
  labs(
    title = "User Activity Distribution",
    x = "Total Books per User",
    y = "Number of Users"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 3. BOOK POPULARITY (Important for Cold-Start)
book_popularity <- data %>%
  group_by(ISBN) %>%
  summarise(
    total_interactions = n(),
    explicit_ratings = sum(Book.Rating > 0),
    avg_rating = mean(Book.Rating[Book.Rating > 0], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(avg_rating))  # Only books with explicit ratings

p3 <- ggplot(book_popularity, aes(x = explicit_ratings, y = avg_rating)) +
  geom_point(alpha = 0.6, color = "darkgreen", size = 2) +
  labs(
    title = "Book Popularity vs Average Rating",
    x = "Number of Explicit Ratings",
    y = "Average Rating"
  ) +
  theme_cowplot() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# 4. EXPLICIT VS IMPLICIT RATIO (Key Insight)
interaction_summary <- data %>%
  summarise(
    total_interactions = n(),
    explicit_count = sum(Book.Rating > 0),
    implicit_count = sum(Book.Rating == 0),
    explicit_pct = round(mean(Book.Rating > 0) * 100, 1),
    implicit_pct = round(mean(Book.Rating == 0) * 100, 1)
  )

# Create a simple bar chart for the ratio
ratio_data <- data.frame(
  Type = c("Explicit Ratings", "Implicit Ratings"),
  Count = c(interaction_summary$explicit_count, interaction_summary$implicit_count),
  Percentage = c(interaction_summary$explicit_pct, interaction_summary$implicit_pct)
)

p4 <- ggplot(ratio_data, aes(x = Type, y = Count, fill = Type)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(Percentage, "%")), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Explicit Ratings" = "steelblue", "Implicit Ratings" = "lightcoral")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Explicit vs Implicit Feedback",
    x = "Feedback Type",
    y = "Count"
  ) +
  theme_cowplot() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )

# Create the combined plot with legend
combined_plot <- plot_grid(
  p1, p2,
  p3, p4,
  ncol = 2, nrow = 2,
  labels = c("A", "B", "C", "D"),
  label_size = 12
)

# Display the plot
combined_plot
```

# 3. Methodology

## 3.1 Evaluation Framework

Performance was evaluated using 5-fold cross-validation with RMSE as the primary metric, calculated only on explicit ratings (1–10) to assess true prediction accuracy. In each fold, 80% of ratings were used for training and 20% for testing, with test values temporarily set to NA to prevent leakage. Confidence weighting was applied consistently to memory-based methods for fair comparison.

## 3.2 User-Item Matrix Construction and Implicit Feedback

Three matrices managed mixed feedback. Implicit ratings were converted to 4 in the ratings matrix, a confidence matrix weighted explicit ratings 1.0 and implicit 0.5 for similarity calculations, and the original matrix retained raw values for RMSE evaluation, enabling sparsity handling while leveraging all interactions.

```{r matrix-construction}
# ============================================================================
# USER-ITEM MATRIX CREATION FUNCTION
# ============================================================================

create_user_item_matrix <- function(ratings_data,
                                     min_ratings_per_book = 3,
                                     min_ratings_per_user = 2,
                                     implicit_rating = 4,
                                     implicit_confidence = 0.5) {
  
  # Handle implicit ratings (rating = 0)
  ratings_clean <- ratings_data %>%
    mutate(
      is_implicit = (Book.Rating == 0),
      original_rating = Book.Rating,
      Book.Rating = ifelse(Book.Rating == 0, implicit_rating, Book.Rating),
      confidence = ifelse(is_implicit, implicit_confidence, 1.0)
    )
  
  # Create matrices
  user_item_matrix <- ratings_clean %>%
    select(User.ID, ISBN, Book.Rating) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = Book.Rating,
      values_fill = NA
    )
  
  confidence_matrix <- ratings_clean %>%
    select(User.ID, ISBN, confidence) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = confidence,
      values_fill = NA
    )
  
  original_ratings_matrix <- ratings_clean %>%
    select(User.ID, ISBN, original_rating) %>%
    pivot_wider(
      names_from = ISBN,
      values_from = original_rating,
      values_fill = NA
    )
  
  # Convert to matrix format
  user_ids <- user_item_matrix$User.ID
  user_item_matrix <- as.matrix(user_item_matrix[, -1])
  confidence_matrix <- as.matrix(confidence_matrix[, -1])
  original_ratings_matrix <- as.matrix(original_ratings_matrix[, -1])
  
  rownames(user_item_matrix) <- user_ids
  rownames(confidence_matrix) <- user_ids
  rownames(original_ratings_matrix) <- user_ids
  
  # Filter by minimum ratings
  books_to_keep <- colSums(!is.na(user_item_matrix)) >= min_ratings_per_book
  user_item_matrix <- user_item_matrix[, books_to_keep]
  confidence_matrix <- confidence_matrix[, books_to_keep]
  original_ratings_matrix <- original_ratings_matrix[, books_to_keep]
  
  users_to_keep <- rowSums(!is.na(user_item_matrix)) >= min_ratings_per_user
  user_item_matrix <- user_item_matrix[users_to_keep, ]
  confidence_matrix <- confidence_matrix[users_to_keep, ]
  original_ratings_matrix <- original_ratings_matrix[users_to_keep, ]
  
  return(list(
    ratings = user_item_matrix,
    confidence = confidence_matrix,
    original_ratings = original_ratings_matrix
  ))
}

# ============================================================================
# EVALUATION FUNCTION
# ============================================================================

evaluate_predictions <- function(pred, actual, original_ratings) {
  # Separate explicit and implicit ratings
  is_implicit <- (original_ratings == 0)
  explicit_mask <- !is_implicit & !is.na(actual)
  
  # Calculate RMSE on explicit ratings only
  rmse_explicit <- sqrt(mean((pred[explicit_mask] - actual[explicit_mask])^2, na.rm = TRUE))
  rmse_overall <- sqrt(mean((pred - actual)^2, na.rm = TRUE))
  
  return(list(
    rmse_explicit = rmse_explicit,
    rmse_overall = rmse_overall,
    n_explicit = sum(explicit_mask),
    n_implicit = sum(is_implicit)
  ))
}

# ============================================================================
# UNIFIED MATRIX CREATION FOR ALL METHODS
# ============================================================================

# Create matrices once with consistent parameters
matrix_result <- create_user_item_matrix(
  data,
  min_ratings_per_book = 5,
  min_ratings_per_user = 3,
  implicit_rating = 4,
  implicit_confidence = 0.5
)

user_item_matrix <- matrix_result$ratings
confidence_matrix <- matrix_result$confidence
original_ratings_matrix <- matrix_result$original_ratings

```


```{r common-functions}
# ============================================================================
# NORMALIZATION FUNCTIONS
# ============================================================================

# User-mean normalization (for UBCF and NN)
normalize_matrix_user <- function(user_item_matrix) {
  user_means <- rowMeans(user_item_matrix, na.rm = TRUE)
  user_item_matrix_normalized <- sweep(user_item_matrix, 1, user_means, FUN = "-")
  
  return(list(
    normalized = user_item_matrix_normalized,
    means = user_means
  ))
}

# Item-mean normalization (for IBCF)
normalize_matrix_item <- function(user_item_matrix) {
  item_means <- colMeans(user_item_matrix, na.rm = TRUE)
  user_item_matrix_normalized <- sweep(user_item_matrix, 2, item_means, FUN = "-")
  
  return(list(
    normalized = user_item_matrix_normalized,
    means = item_means
  ))
}
```





## 3.3 Collaborative Filtering Approaches

### 3.3.1 User-Based Collaborative Filtering (UBCF)

UBCF estimates ratings by leveraging similar users' preferences. User-mean normalization removed individual biases, confidence-weighted cosine similarity identified similar users, and k=50 nearest neighbors generated predictions combining target user's mean with weighted averages from similar users.

```{r ubcf-similarity}
# ============================================================================
# UBCF: USER SIMILARITY COMPUTATION
# ============================================================================

compute_user_similarity_matrix <- function(user_item_matrix_normalized,
                                            confidence_matrix) {
  
  # Prepare matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  
  # Weighted ratings
  weighted_mat <- mat * conf
  
  # Compute cosine similarity
  numerator <- weighted_mat %*% t(weighted_mat)
  magnitudes <- sqrt(rowSums(weighted_mat^2))
  denominator <- outer(magnitudes, magnitudes)
  
  user_similarity_matrix <- numerator / denominator
  user_similarity_matrix[is.nan(user_similarity_matrix)] <- 0
  diag(user_similarity_matrix) <- 0
  
  rownames(user_similarity_matrix) <- rownames(user_item_matrix_normalized)
  colnames(user_similarity_matrix) <- rownames(user_item_matrix_normalized)
  
  return(user_similarity_matrix)
}
```



```{r ubcf-recommendations}
# ============================================================================
# UBCF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_ubcf <- function(target_user,
                                     user_item_matrix,
                                     user_item_matrix_normalized,
                                     confidence_matrix,
                                     user_sim_matrix,
                                     user_means,
                                     book_info,
                                     n_recommendations = 10,
                                     k = NULL) {
  
  target_user <- as.character(target_user)
  
  # Find unrated books
  unrated_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get similarities
  sims <- user_sim_matrix[target_user, ]
  sims[is.na(sims) | is.nan(sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < length(sims)) {
    non_zero_count <- sum(sims != 0)
    if (non_zero_count > 0) {
      k_actual <- min(k, non_zero_count)
      top_k_users <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
      sims_filtered <- rep(0, length(sims))
      names(sims_filtered) <- names(sims)
      sims_filtered[top_k_users] <- sims[top_k_users]
      sims <- sims_filtered
    }
  }
  
  if (sum(abs(sims) > 0) == 0) return(data.frame())
  
  # Prepare weighted matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  weighted_mat <- mat * conf
  
  # Calculate predictions
  weighted_ratings <- t(weighted_mat[, unrated_books, drop = FALSE]) %*% sims
  
  rated_mask <- !is.na(confidence_matrix[, unrated_books, drop = FALSE])
  conf_weighted_sims <- sweep(
    conf[, unrated_books, drop = FALSE] * rated_mask,
    1,
    abs(sims),
    "*"
  )
  sum_weighted_sims <- colSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + user_means[target_user]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Get top recommendations
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  top_books <- sort(preds_valid, decreasing = TRUE)[1:min(n_recommendations, length(preds_valid))]
  
  recommendations <- data.frame(
    ISBN = names(top_books),
    Predicted_Rating = as.numeric(top_books)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# UBCF: RECOMMENDATION FOR NEW USERS (COLD-START)
# ============================================================================

recommend_for_new_user_ubcf <- function(new_user_ratings,
                                         user_item_matrix,
                                         user_item_matrix_normalized,
                                         confidence_matrix,
                                         user_means,
                                         book_info,
                                         n_recommendations = 10,
                                         k = NULL) {
  
  if (length(new_user_ratings) == 0) return(data.frame())
  
  # Create new user vector
  new_user_vector <- rep(NA, ncol(user_item_matrix))
  names(new_user_vector) <- colnames(user_item_matrix)
  
  matched_books <- intersect(names(new_user_ratings), names(new_user_vector))
  if (length(matched_books) == 0) return(data.frame())
  
  new_user_vector[matched_books] <- new_user_ratings[matched_books]
  
  # Normalize
  new_user_mean <- mean(new_user_vector, na.rm = TRUE)
  new_user_normalized <- new_user_vector - new_user_mean
  new_user_normalized[is.na(new_user_normalized)] <- 0
  new_user_conf <- ifelse(is.na(new_user_vector), 0, 1)
  new_user_weighted <- new_user_normalized * new_user_conf
  
  # Prepare matrices
  mat_normalized <- user_item_matrix_normalized
  mat_normalized[is.na(mat_normalized)] <- 0
  conf <- confidence_matrix
  conf[is.na(conf)] <- 0
  mat_weighted <- mat_normalized * conf
  
  # Compute similarities
  existing_magnitudes <- sqrt(rowSums(mat_weighted^2))
  new_user_magnitude <- sqrt(sum(new_user_weighted^2))
  
  if (new_user_magnitude == 0) return(data.frame())
  
  new_user_sims <- as.vector(
    (mat_weighted %*% new_user_weighted) /
      (existing_magnitudes * new_user_magnitude)
  )
  new_user_sims[!is.finite(new_user_sims)] <- 0
  names(new_user_sims) <- rownames(user_item_matrix_normalized)
  
  # Apply k-NN
  if (!is.null(k) && k > 0 && k < length(new_user_sims)) {
    threshold <- sort(abs(new_user_sims), decreasing = TRUE)[min(k, sum(new_user_sims != 0))]
    new_user_sims[abs(new_user_sims) < threshold] <- 0
  }
  
  if (all(new_user_sims == 0)) return(data.frame())
  
  # Make predictions
  unrated_books <- names(new_user_vector)[is.na(new_user_vector)]
  if (length(unrated_books) == 0) return(data.frame())
  
  weighted_ratings <- as.vector(t(mat_weighted[, unrated_books, drop = FALSE]) %*% new_user_sims)
  
  rated_mask <- !is.na(user_item_matrix[, unrated_books, drop = FALSE])
  conf_for_unrated <- conf[, unrated_books, drop = FALSE] * rated_mask
  sum_weighted_sims <- as.vector(t(conf_for_unrated) %*% abs(new_user_sims))
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- (weighted_ratings / sum_weighted_sims) + new_user_mean
  preds <- pmin(pmax(preds, 1), 10)
  names(preds) <- unrated_books
  
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  n_actual <- min(n_recommendations, length(preds_valid))
  top_indices <- order(preds_valid, decreasing = TRUE)[1:n_actual]
  
  recommendations <- data.frame(
    ISBN = names(preds_valid)[top_indices],
    Predicted_Rating = round(preds_valid[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```





```{r ubcf-cv}
# ============================================================================
# UBCF: CROSS-VALIDATION
# ============================================================================

cross_validate_ubcf <- function(user_item_matrix,
                                 confidence_matrix,
                                 original_ratings_matrix,
                                 n_folds = 5,
                                 k = 30) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    test_obs <- observed[test_indices, , drop = FALSE]
    
    # Create training matrices
    train_matrix <- user_item_matrix
    train_confidence <- confidence_matrix
    train_matrix[test_obs] <- NA
    train_confidence[test_obs] <- NA
    
    # Normalize and compute similarity
    norm_result <- normalize_matrix_user(train_matrix)
    train_normalized <- norm_result$normalized
    user_means <- norm_result$means
    user_sim_matrix <- compute_user_similarity_matrix(train_normalized, train_confidence)
    
    # Prepare for prediction
    train_norm_filled <- train_normalized
    train_norm_filled[is.na(train_norm_filled)] <- 0
    train_conf_filled <- train_confidence
    train_conf_filled[is.na(train_conf_filled)] <- 0
    
    # Make predictions
    predictions <- numeric(nrow(test_obs))
    
    for (i in 1:nrow(test_obs)) {
      user_idx <- test_obs[i, 1]
      item_idx <- test_obs[i, 2]
      
      sims <- user_sim_matrix[user_idx, ]
      sims[is.na(sims) | is.nan(sims)] <- 0
      
      # Apply k-NN
      if (k < length(sims) && sum(abs(sims) > 0) > 0) {
        k_actual <- min(k, sum(abs(sims) > 0))
        top_k_idx <- order(abs(sims), decreasing = TRUE)[1:k_actual]
        sims_filtered <- rep(0, length(sims))
        sims_filtered[top_k_idx] <- sims[top_k_idx]
        sims <- sims_filtered
      }
      
      rated_mask <- !is.na(train_matrix[, item_idx])
      
      if (sum(abs(sims[rated_mask])) == 0) {
        predictions[i] <- user_means[user_idx]
        next
      }
      
      # Confidence-weighted prediction
      weighted_ratings <- train_norm_filled[rated_mask, item_idx] *
        train_conf_filled[rated_mask, item_idx]
      weighted_sims <- abs(sims[rated_mask]) *
        train_conf_filled[rated_mask, item_idx]
      
      weighted_sum <- sum(sims[rated_mask] * weighted_ratings)
      sum_weighted_sims <- sum(weighted_sims)
      
      if (sum_weighted_sims > 0) {
        predictions[i] <- (weighted_sum / sum_weighted_sims) + user_means[user_idx]
      } else {
        predictions[i] <- user_means[user_idx]
      }
    }
    
    predictions <- pmin(pmax(predictions, 1), 10)
    actual <- user_item_matrix[test_obs]
    original_ratings <- original_ratings_matrix[test_obs]
    metrics <- evaluate_predictions(predictions, actual, original_ratings)
    
    cv_results[fold, ] <- list(fold, metrics$rmse_explicit)
  }
  
  return(cv_results)
}
```



```{r ubcf-implementation}
# ============================================================================
# UBCF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Normalize and compute similarity
normalized_result_ubcf <- normalize_matrix_user(user_item_matrix)
user_item_matrix_normalized_ubcf <- normalized_result_ubcf$normalized
user_means_ubcf <- normalized_result_ubcf$means

user_similarity_matrix <- compute_user_similarity_matrix(
  user_item_matrix_normalized_ubcf,
  confidence_matrix
)

# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ubcf <- recommend_for_user_ubcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_sim_matrix = user_similarity_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)


# Generate recommendations for new user (cold-start)
sample_books <- colnames(user_item_matrix)[1:5]
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), sample_books)

new_user_recs_ubcf <- recommend_for_new_user_ubcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)


```




### 3.3.2 Item-Based Collaborative Filtering (IBCF)

IBCF uses item similarity rather than user similarity. Item-mean normalization removed popularity biases, confidence-weighted cosine similarity computed item relationships, and k=50 neighbors aggregated the target user's ratings for similar items.

```{r ibcf-similarity}
# ============================================================================
# IBCF: ITEM SIMILARITY COMPUTATION
# ============================================================================

compute_item_similarity_matrix <- function(user_item_matrix_normalized,
                                            confidence_matrix) {
  
  # Prepare matrices
  mat <- user_item_matrix_normalized
  mat[is.na(mat)] <- 0
  conf <- confidence_matrix
  conf[is.na(conf)] <- 0
  
  # Transpose for item-item similarity
  mat_t <- t(mat)
  conf_t <- t(conf)
  
  # Weighted ratings
  weighted_mat_t <- mat_t * conf_t
  
  # Compute cosine similarity
  numerator <- weighted_mat_t %*% t(weighted_mat_t)
  magnitudes <- sqrt(rowSums(weighted_mat_t^2))
  denominator <- outer(magnitudes, magnitudes)
  
  item_similarity_matrix <- numerator / denominator
  item_similarity_matrix[is.nan(item_similarity_matrix)] <- 0
  diag(item_similarity_matrix) <- 0
  
  rownames(item_similarity_matrix) <- colnames(user_item_matrix_normalized)
  colnames(item_similarity_matrix) <- colnames(user_item_matrix_normalized)
  
  return(item_similarity_matrix)
}
```



```{r ibcf-recommendations}
# ============================================================================
# IBCF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_ibcf <- function(target_user,
                                     user_item_matrix,
                                     user_item_matrix_normalized,
                                     confidence_matrix,
                                     item_sim_matrix,
                                     item_means,
                                     book_info,
                                     n_recommendations = 10,
                                     k = NULL) {
  
  target_user <- as.character(target_user)
  
  # Find unrated books
  unrated_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get item similarities
  item_sims <- item_sim_matrix[unrated_books, , drop = FALSE]
  item_sims[is.na(item_sims) | is.nan(item_sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < ncol(item_sims)) {
    for (i in 1:nrow(item_sims)) {
      sims <- item_sims[i, ]
      non_zero_count <- sum(sims != 0)
      if (non_zero_count > 0) {
        k_actual <- min(k, non_zero_count)
        top_k_items <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
        sims_filtered <- rep(0, length(sims))
        names(sims_filtered) <- names(sims)
        sims_filtered[top_k_items] <- sims[top_k_items]
        item_sims[i, ] <- sims_filtered
      }
    }
  }
  
  if (sum(abs(item_sims) > 0) == 0) return(data.frame())
  
  # Prepare weighted matrices
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  weighted_mat <- mat * conf
  
  # Calculate predictions
  weighted_ratings <- item_sims %*% weighted_mat[target_user, ]
  
  rated_mask <- !is.na(user_item_matrix[target_user, ])
  rated_item_mask <- matrix(
    rated_mask,
    nrow = nrow(item_sims),
    ncol = length(rated_mask),
    byrow = TRUE
  )
  
  conf_weighted_sims <- rated_item_mask * abs(item_sims) * conf[target_user, ]
  sum_weighted_sims <- rowSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + item_means[unrated_books]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Get top recommendations
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  top_books <- sort(preds_valid, decreasing = TRUE)[1:min(n_recommendations, length(preds_valid))]
  
  recommendations <- data.frame(
    ISBN = names(top_books),
    Predicted_Rating = as.numeric(top_books)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# IBCF: RECOMMENDATION FOR NEW USERS (COLD-START)
# ============================================================================

recommend_for_new_user_ibcf <- function(new_user_ratings,
                                         user_item_matrix,
                                         user_item_matrix_normalized,
                                         confidence_matrix,
                                         item_means,
                                         item_sim_matrix,
                                         book_info,
                                         n_recommendations = 10,
                                         k = NULL) {
  
  if (length(new_user_ratings) == 0) return(data.frame())
  
  # Create new user vector
  new_user_vector <- rep(NA, ncol(user_item_matrix))
  names(new_user_vector) <- colnames(user_item_matrix)
  
  matched_books <- intersect(names(new_user_ratings), names(new_user_vector))
  if (length(matched_books) == 0) return(data.frame())
  
  new_user_vector[matched_books] <- new_user_ratings[matched_books]
  
  # Normalize
  new_user_normalized <- new_user_vector - item_means
  new_user_normalized[is.na(new_user_normalized)] <- 0
  new_user_conf <- ifelse(is.na(new_user_vector), 0, 1)
  new_user_weighted <- new_user_normalized * new_user_conf
  
  # Find unrated books
  unrated_books <- names(new_user_vector)[is.na(new_user_vector)]
  if (length(unrated_books) == 0) return(data.frame())
  
  # Get item similarities
  item_sims <- item_sim_matrix[unrated_books, , drop = FALSE]
  item_sims[is.na(item_sims) | is.nan(item_sims)] <- 0
  
  # Apply k-NN filtering
  if (!is.null(k) && k < ncol(item_sims)) {
    for (i in 1:nrow(item_sims)) {
      sims <- item_sims[i, ]
      non_zero_count <- sum(sims != 0)
      if (non_zero_count > 0) {
        k_actual <- min(k, non_zero_count)
        top_k_items <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
        sims_filtered <- rep(0, length(sims))
        names(sims_filtered) <- names(sims)
        sims_filtered[top_k_items] <- sims[top_k_items]
        item_sims[i, ] <- sims_filtered
      }
    }
  }
  
  if (sum(abs(item_sims) > 0) == 0) return(data.frame())
  
  # Make predictions
  weighted_ratings <- item_sims %*% new_user_weighted
  
  rated_mask <- !is.na(new_user_vector)
  rated_item_mask <- matrix(
    rated_mask,
    nrow = nrow(item_sims),
    ncol = length(rated_mask),
    byrow = TRUE
  )
  
  conf_weighted_sims <- rated_item_mask * abs(item_sims) * new_user_conf
  sum_weighted_sims <- rowSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + item_means[unrated_books]
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) return(data.frame())
  
  n_actual <- min(n_recommendations, length(preds_valid))
  top_indices <- order(preds_valid, decreasing = TRUE)[1:n_actual]
  
  recommendations <- data.frame(
    ISBN = names(preds_valid)[top_indices],
    Predicted_Rating = round(preds_valid[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```


```{r ibcf-cv}
# ============================================================================
# IBCF: CROSS-VALIDATION
# ============================================================================

cross_validate_ibcf <- function(user_item_matrix,
                                 confidence_matrix,
                                 original_ratings_matrix,
                                 n_folds = 5,
                                 k = 30) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    test_obs <- observed[test_indices, , drop = FALSE]
    
    # Create training matrices
    train_matrix <- user_item_matrix
    train_confidence <- confidence_matrix
    train_matrix[test_obs] <- NA
    train_confidence[test_obs] <- NA
    
    # Normalize and compute similarity
    norm_result <- normalize_matrix_item(train_matrix)
    train_normalized <- norm_result$normalized
    item_means <- norm_result$means
    item_sim_matrix <- compute_item_similarity_matrix(train_normalized, train_confidence)
    
    # Prepare for prediction
    train_norm_filled <- train_normalized
    train_norm_filled[is.na(train_norm_filled)] <- 0
    train_conf_filled <- train_confidence
    train_conf_filled[is.na(train_conf_filled)] <- 0
    
    # Make predictions
    predictions <- numeric(nrow(test_obs))
    
    for (i in 1:nrow(test_obs)) {
      user_idx <- test_obs[i, 1]
      item_idx <- test_obs[i, 2]
      
      rated_items <- which(!is.na(train_matrix[user_idx, ]))
      
      if (length(rated_items) == 0) {
        predictions[i] <- item_means[item_idx]
        next
      }
      
      sims <- item_sim_matrix[item_idx, rated_items]
      sims[is.na(sims) | is.nan(sims)] <- 0
      
      # Apply k-NN
      if (sum(abs(sims) > 0) > 0 && k < length(sims)) {
        k_actual <- min(k, sum(abs(sims) > 0))
        top_k_idx <- order(abs(sims), decreasing = TRUE)[1:k_actual]
        sims_filtered <- rep(0, length(sims))
        sims_filtered[top_k_idx] <- sims[top_k_idx]
        sims <- sims_filtered
      }
      
      # Confidence-weighted prediction
      if (sum(abs(sims)) > 0) {
        weighted_ratings <- train_norm_filled[user_idx, rated_items] *
          train_conf_filled[user_idx, rated_items]
        weighted_sims <- abs(sims) * train_conf_filled[user_idx, rated_items]
        
        weighted_sum <- sum(sims * weighted_ratings)
        sum_weighted_sims <- sum(weighted_sims)
        
        if (sum_weighted_sims > 0) {
          predictions[i] <- (weighted_sum / sum_weighted_sims) + item_means[item_idx]
        } else {
          predictions[i] <- item_means[item_idx]
        }
      } else {
        predictions[i] <- item_means[item_idx]
      }
    }
    
    predictions <- pmin(pmax(predictions, 1), 10)
    actual <- user_item_matrix[test_obs]
    original_ratings <- original_ratings_matrix[test_obs]
    metrics <- evaluate_predictions(predictions, actual, original_ratings)
    
    cv_results[fold, ] <- list(fold, metrics$rmse_explicit)
    # cat("RMSE:", round(metrics$rmse_explicit, 4), "\n")
  }
  
  return(cv_results)
}
```


```{r ibcf-implementation}
# ============================================================================
# IBCF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Normalize and compute similarity
normalized_result_ibcf <- normalize_matrix_item(user_item_matrix)
user_item_matrix_normalized_ibcf <- normalized_result_ibcf$normalized
item_means_ibcf <- normalized_result_ibcf$means

item_similarity_matrix <- compute_item_similarity_matrix(
  user_item_matrix_normalized_ibcf,
  confidence_matrix
)

# cat("Similarity matrix computed\n\n")

# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ibcf <- recommend_for_user_ibcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_sim_matrix = item_similarity_matrix,
  item_means = item_means_ibcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)


# Generate recommendations for new user (cold-start)
new_user_recs_ibcf <- recommend_for_new_user_ibcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_means = item_means_ibcf,
  item_sim_matrix = item_similarity_matrix,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)


```




### 3.3.3 Matrix Factorization (MF)

MF decomposes the rating matrix into lower-dimensional latent factors using recosystem with parallel stochastic gradient descent. Grid search optimized latent dimensions (10, 20, 30), learning rates (0.01-0.1), and L2 regularization (0.01-0.1). Models trained for 100 iterations using 4 threads on explicit ratings only.

```{r mf-data-prep}
# ============================================================================
# MF: DATA PREPARATION (EXPLICIT RATINGS ONLY)
# ============================================================================

create_explicit_matrix <- function(ratings_data,
                                    min_ratings_per_book = 3,
                                    min_ratings_per_user = 2) {
  
  # Filter explicit ratings only
  explicit_ratings <- ratings_data %>%
    filter(Book.Rating > 0) %>%
    select(User.ID, ISBN, Book.Rating)
  
  # Create user-item matrix
  user_item_matrix <- explicit_ratings %>%
    pivot_wider(
      names_from = ISBN,
      values_from = Book.Rating,
      values_fill = NA
    )
  
  # Convert to matrix
  user_ids <- user_item_matrix$User.ID
  user_item_matrix <- as.matrix(user_item_matrix[, -1])
  rownames(user_item_matrix) <- user_ids
  
  # Filter by minimum ratings
  books_to_keep <- colSums(!is.na(user_item_matrix)) >= min_ratings_per_book
  user_item_matrix <- user_item_matrix[, books_to_keep]
  
  users_to_keep <- rowSums(!is.na(user_item_matrix)) >= min_ratings_per_user
  user_item_matrix <- user_item_matrix[users_to_keep, ]
  
  return(user_item_matrix)
}

prepare_recosystem_data <- function(user_item_matrix, original_ratings_matrix) {
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  
  train_data <- data.frame(
    user_index = observed[, 1] - 1,
    item_index = observed[, 2] - 1,
    rating = user_item_matrix[observed],
    original_rating = original_ratings_matrix[observed]
  )
  
  # Filter explicit ratings only
  train_data_explicit <- train_data[train_data$original_rating != 0, ]
  
  user_ids <- data.frame(
    user_index = 0:(nrow(user_item_matrix) - 1),
    user_id = rownames(user_item_matrix)
  )
  
  item_ids <- data.frame(
    item_index = 0:(ncol(user_item_matrix) - 1),
    item_id = colnames(user_item_matrix)
  )
  
  list(
    train_data = train_data_explicit,
    user_ids = user_ids,
    item_ids = item_ids,
    n_users = nrow(user_item_matrix),
    n_items = ncol(user_item_matrix)
  )
}
```



```{r mf-training}
# ============================================================================
# MF: HYPERPARAMETER TUNING
# ============================================================================

tune_reco_model <- function(train_data,
                             n_factors = c(10, 20, 30),
                             learning_rate = c(0.1, 0.05, 0.01),
                             costp_l2 = c(0.01, 0.1),
                             costq_l2 = c(0.01, 0.1),
                             n_iter = 50,
                             verbose = TRUE) {
  
  train_set <- data_memory(
    user_index = train_data$user_index,
    item_index = train_data$item_index,
    rating = train_data$rating,
    index1 = FALSE
  )
  
  rs <- Reco()
  
  opts <- rs$tune(train_set, opts = list(
    dim = n_factors,
    lrate = learning_rate,
    costp_l2 = costp_l2,
    costq_l2 = costq_l2,
    niter = n_iter,
    nthread = 4,
    verbose = verbose
  ))
  
  return(opts)
}

# ============================================================================
# MF: MODEL TRAINING
# ============================================================================

train_reco_model <- function(train_data,
                              n_factors = 20,
                              learning_rate = 0.1,
                              costp_l2 = 0.01,
                              costq_l2 = 0.01,
                              n_iter = 100,
                              n_threads = 4,
                              verbose = TRUE) {
  
  train_set <- data_memory(
    user_index = train_data$user_index,
    item_index = train_data$item_index,
    rating = train_data$rating,
    index1 = FALSE
  )
  
  model <- Reco()
  model$train(train_set, opts = list(
    dim = n_factors,
    lrate = learning_rate,
    costp_l2 = costp_l2,
    costq_l2 = costq_l2,
    niter = n_iter,
    nthread = n_threads,
    verbose = verbose
  ))
  
  return(model)
}
```


```{r mf-recommendations}
# ============================================================================
# MF: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_mf <- function(model,
                                   user_item_matrix,
                                   user_id,
                                   n_recommendations = 10,
                                   user_ids_map,
                                   item_ids_map,
                                   book_info) {
  
  if (!user_id %in% user_ids_map$user_id) stop("User not found")
  
  user_idx <- user_ids_map$user_index[user_ids_map$user_id == user_id]
  user_row <- user_item_matrix[as.character(user_id), ]
  unrated_items <- which(is.na(user_row))
  
  if (length(unrated_items) == 0) return(data.frame())
  
  item_indices <- unrated_items - 1
  pred_set <- data_memory(
    user_index = rep(user_idx, length(item_indices)),
    item_index = item_indices,
    index1 = FALSE
  )
  
  pred_ratings <- model$predict(pred_set, out_memory())
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  recommended_items <- colnames(user_item_matrix)[unrated_items[top_indices]]
  
  data.frame(
    ISBN = recommended_items,
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
}

# ============================================================================
# MF: RECOMMENDATION FOR NEW USERS (COLD-START)
# ============================================================================

recommend_for_new_user_mf <- function(model,
                                       new_user_ratings,
                                       user_item_matrix,
                                       item_ids_map,
                                       book_info,
                                       n_recommendations = 10) {
  
  # Hybrid approach for cold-start
  ratings_df <- if (is.numeric(new_user_ratings) && !is.null(names(new_user_ratings))) {
    data.frame(item_id = names(new_user_ratings), rating = as.numeric(new_user_ratings))
  } else {
    new_user_ratings
  }
  
  ratings_df <- ratings_df[ratings_df$item_id %in% item_ids_map$item_id, ]
  if (nrow(ratings_df) == 0) stop("No rated items exist in training data")
  
  # Calculate item means for fallback
  item_means <- colMeans(user_item_matrix, na.rm = TRUE)
  
  # Get unrated items
  all_items <- item_ids_map$item_id
  rated_items <- ratings_df$item_id
  unrated_items <- setdiff(all_items, rated_items)
  
  if (length(unrated_items) == 0) return(data.frame())
  
  # Use item means adjusted by user bias
  user_mean_rating <- mean(ratings_df$rating)
  global_mean_rating <- mean(user_item_matrix, na.rm = TRUE)
  user_bias <- user_mean_rating - global_mean_rating
  
  pred_ratings <- item_means[unrated_items] + user_bias
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  recommended_items <- names(pred_ratings)[top_indices]
  
  data.frame(
    ISBN = recommended_items,
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
}
```



```{r mf-cv}
# ============================================================================
# MF: CROSS-VALIDATION
# ============================================================================

cross_validate_mf <- function(user_item_matrix,
                               original_ratings_matrix,
                               optimal_params,
                               n_folds = 5,
                               n_iter = 50,
                               verbose = TRUE) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  if (n_ratings < n_folds * 2) {
    stop("Not enough ratings to perform cross-validation.")
  }
  
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    Fold = integer(n_folds),
    RMSE = numeric(n_folds)
  )
  
  if (verbose) {
    # cat("Starting", n_folds, "-fold cross-validation...\n")
  }
  
  for (fold in 1:n_folds) {
    # if (verbose) cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_idx <- which(fold_indices == fold)
    train_idx <- which(fold_indices != fold)
    
    train_obs <- observed[train_idx, , drop = FALSE]
    test_obs <- observed[test_idx, , drop = FALSE]
    
    # Prepare training data (explicit only)
    train_data <- data.frame(
      user_index = train_obs[, 1] - 1,
      item_index = train_obs[, 2] - 1,
      rating = user_item_matrix[train_obs],
      original_rating = original_ratings_matrix[train_obs]
    )
    
    train_data_explicit <- train_data[train_data$original_rating != 0, ]
    
    # Prepare test data (explicit only)
    test_data <- data.frame(
      user_index = test_obs[, 1] - 1,
      item_index = test_obs[, 2] - 1,
      rating = user_item_matrix[test_obs],
      original_rating = original_ratings_matrix[test_obs]
    )
    
    test_data_explicit <- test_data[test_data$original_rating != 0, ]
    
    if (nrow(test_data_explicit) == 0) {
      cv_results[fold, ] <- c(fold, NA)
      next
    }
    
    # Train model
    train_set <- data_memory(
      user_index = train_data_explicit$user_index,
      item_index = train_data_explicit$item_index,
      rating = train_data_explicit$rating,
      index1 = FALSE
    )
    
    test_set <- data_memory(
      user_index = test_data_explicit$user_index,
      item_index = test_data_explicit$item_index,
      rating = test_data_explicit$rating,
      index1 = FALSE
    )
    
    r <- Reco()
    r$train(train_set, opts = list(
      dim = optimal_params$min$dim,
      lrate = optimal_params$min$lrate,
      costp_l2 = optimal_params$min$costp_l2,
      costq_l2 = optimal_params$min$costq_l2,
      niter = n_iter,
      nthread = 4,
      verbose = FALSE
    ))
    
    preds <- r$predict(test_set, out_memory())
    preds <- pmax(pmin(preds, 10), 1)
    
    # Calculate RMSE
    actual <- test_data_explicit$rating
    original <- test_data_explicit$original_rating
    metrics <- evaluate_predictions(preds, actual, original)
    
    cv_results[fold, ] <- c(fold, metrics$rmse_explicit)
    # cat("RMSE:", round(metrics$rmse_explicit, 4), "\n")
  }
  
  return(cv_results)
}
```



```{r mf-implementation}
# ============================================================================
# MF: MODEL TRAINING AND EVALUATION
# ============================================================================

# Prepare data for recosystem
prepared <- prepare_recosystem_data(user_item_matrix, original_ratings_matrix)

# Hyperparameter tuning
# optimal_params_mf <- tune_reco_model(
#   train_data = prepared$train_data,
#   n_factors = c(10, 20, 30),
#   learning_rate = c(0.1, 0.05, 0.01),
#   costp_l2 = c(0.01, 0.1),
#   costq_l2 = c(0.01, 0.1),
#   n_iter = 50,
#   verbose = FALSE
# )

# save(optimal_params_mf, file = 'mf_tuning_results.Rdata')

# Train final model
# model_mf <- train_reco_model(
#   train_data = prepared$train_data,
#   n_factors = optimal_params_mf$min$dim,
#   learning_rate = optimal_params_mf$min$lrate,
#   costp_l2 = optimal_params_mf$min$costp_l2,
#   costq_l2 = optimal_params_mf$min$costq_l2,
#   n_iter = 100,
#   n_threads = 4,
#   verbose = FALSE
# )

# save(model_mf, file = 'mf_final_model_mf.Rdata')


# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_mf <- recommend_for_user_mf(
  model = model_mf,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids_map = prepared$user_ids,
  item_ids_map = prepared$item_ids,
  book_info = book_info
)


# Generate recommendations for new user (cold-start)
recs_new_mf <- recommend_for_new_user_mf(
  model = model_mf,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  item_ids_map = prepared$item_ids,
  book_info = book_info,
  n_recommendations = 10
)


# Run cross-validation
cv_results_mf <- cross_validate_mf(
  user_item_matrix = user_item_matrix,
  original_ratings_matrix = original_ratings_matrix,
  optimal_params = optimal_params_mf,
  n_folds = 5,
  n_iter = 50,
  verbose = TRUE
)


```




### 3.3.4 Neural Network-Based Collaborative Filtering (NN)

NN-based CF used H2O's deep learning framework with user-mean normalization. Grid search evaluated 24 configurations across layer depths (1-3), hidden units (16-128), activation functions (Tanh, ReLU), and regularization strengths (0.00001-0.0001). Results showed optimal configuration with Tanh activation, 2 hidden layers (64, 32), and moderate regularization (Table 3, Table 4).

```{r nn-data-prep}
# ============================================================================
# NN: DATA PREPARATION
# ============================================================================

prepare_h2o_data <- function(user_item_matrix, original_ratings_matrix = NULL) {
  
  # Apply user-mean normalization
  user_means <- rowMeans(user_item_matrix, na.rm = TRUE)
  normalized_matrix <- sweep(user_item_matrix, 1, user_means, FUN = "-")
  
  # Get observed ratings
  observed <- which(!is.na(normalized_matrix), arr.ind = TRUE)
  
  # Create training data
  train_data <- data.frame(
    user_id = rownames(user_item_matrix)[observed[, 1]],
    book_id = colnames(user_item_matrix)[observed[, 2]],
    rating = as.numeric(normalized_matrix[observed])
  )
  
  # Add original rating information
  if (!is.null(original_ratings_matrix)) {
    train_data$original_rating <- as.numeric(original_ratings_matrix[observed])
    train_data$is_implicit <- (train_data$original_rating == 0)
  } else {
    train_data$original_rating <- train_data$rating + user_means[observed[, 1]]
    train_data$is_implicit <- FALSE
  }
  
  return(list(
    train_data = train_data,
    user_ids = data.frame(user_id = rownames(user_item_matrix)),
    book_ids = data.frame(book_id = colnames(user_item_matrix)),
    n_users = nrow(user_item_matrix),
    n_items = ncol(user_item_matrix),
    user_means = user_means,
    original_matrix = user_item_matrix,
    original_ratings_matrix = original_ratings_matrix
  ))
}
```



```{r nn-tuning}
# ============================================================================
# NN: HYPERPARAMETER TUNING WITH GRID SEARCH
# ============================================================================

# Initialize H2O cluster (check if already running)
if (!h2o::h2o.clusterIsUp()) {
  suppressMessages(h2o::h2o.init(ip = "localhost", port = 54321, nthreads = 4))
}
h2o::h2o.no_progress()

# # Prepare data
# h2o_data_tuning <- prepare_h2o_data(user_item_matrix, original_ratings_matrix)
# 
# # Filter explicit ratings only
# train_data_explicit <- h2o_data_tuning$train_data[!h2o_data_tuning$train_data$is_implicit, ]
# 
# # Convert to H2O frame
# train_h2o_tuning <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
# train_h2o_tuning$user_id <- as.factor(train_h2o_tuning$user_id)
# train_h2o_tuning$book_id <- as.factor(train_h2o_tuning$book_id)
# 
# # Define hyperparameter grid
# hyper_params <- list(
#   activation = c("Tanh", "Rectifier"),
#   hidden = list(
#     c(16),
#     c(32, 16),
#     c(64, 32),
#     c(64, 32, 16),
#     c(128, 64, 32)
#   ),
#   l1 = c(0.00001, 0.0001),
#   l2 = c(0.00001, 0.0001)
# )
# 
# cat("Total models to train:", 
#     length(hyper_params$activation) * length(hyper_params$hidden) * 
#     length(hyper_params$l1) * length(hyper_params$l2), "\n\n")
# 
# set.seed(123)

# # Run grid search
# model_grid <- h2o.grid(
#   algorithm = "deeplearning",
#   grid_id = "nn_grid_book_recommendations",
#   hyper_params = hyper_params,
#   x = c("user_id", "book_id"),
#   y = "rating",
#   training_frame = train_h2o_tuning,
#   nfolds = 5,
#   stopping_metric = "RMSE",
#   stopping_rounds = 5,
#   stopping_tolerance = 0.001,
#   seed = 123,
#   adaptive_rate = TRUE,
#   epochs = 50,
#   max_w2 = 10,
#   initial_weight_distribution = "UniformAdaptive"
# )
# 
# # Get results
# grid_results <- h2o.getGrid(
#   "nn_grid_book_recommendations",
#   sort_by = "rmse",
#   decreasing = FALSE)

# all_models_df <- as.data.frame(grid_results@summary_table)

# Select best model
# best_model_id <- grid_results@model_ids[[1]]
# best_model <- h2o.getModel(best_model_id)
# cv_rmse <- h2o.rmse(best_model, xval = TRUE)

# best_nn_results <- data.frame(
#   Activation = best_model@allparameters$activation,
#   Hidden_Layers = paste(best_model@allparameters$hidden, collapse = ", "),
#   N_Layers = length(best_model@allparameters$hidden),
#   L1_Regularization = best_model@allparameters$l1,
#   L2_Regularization = best_model@allparameters$l2,
#   CV_RMSE = round(cv_rmse, 4),
#   Data_Type = "Explicit Only"
# )


# Grid search results
kable(all_models_df %>% select(-model_ids),
      format = "latex",
      booktabs = TRUE,
      caption = "NN Grid Search Results: Hyperparameter optimization results across 24 configurations showing activation functions, hidden layer architectures, and regularization parameters",
      align = c("c", "c", "c", "c", "c"),
      digits = 3) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Best model results
kable(best_nn_results,
      format = "latex",
      booktabs = TRUE,
      caption = "Optimal NN Configuration: Best performing model parameters selected from grid search including activation function, hidden layer architecture, regularization settings, and cross-validation RMSE",
      align = c("c", "c", "c", "c", "c", "c"),
      digits = 3) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )



# save(best_nn_results, grid_results, all_models_df, file = 'nn_tuning_results.Rdata') 
```


```{r nn-training}
# ============================================================================
# NN: MODEL TRAINING FUNCTION
# ============================================================================

train_h2o_model <- function(train_data,
                             hidden = c(64, 32),
                             epochs = 30,
                             activation = "Tanh",
                             l1 = 0.0001,
                             l2 = 0.0001,
                             seed = 123,
                             verbose = FALSE) {
  
  if (!verbose) h2o.no_progress()
  
  model <- h2o.deeplearning(
    x = c("user_id", "book_id"),
    y = "rating",
    training_frame = train_data,
    hidden = hidden,
    epochs = epochs,
    activation = activation,
    l1 = l1,
    l2 = l2,
    adaptive_rate = TRUE,
    seed = seed,
    stopping_rounds = 5,
    stopping_metric = "RMSE",
    stopping_tolerance = 0.001,
    max_w2 = 10,
    initial_weight_distribution = "UniformAdaptive",
    initial_weight_scale = 0.5
  )
  
  return(model)
}
```



```{r nn-recommendations}
# ============================================================================
# NN: RECOMMENDATION FOR EXISTING USERS
# ============================================================================

recommend_for_user_nn <- function(model,
                                   user_item_matrix,
                                   user_id,
                                   n_recommendations = 10,
                                   user_ids,
                                   book_ids,
                                   user_means,
                                   book_info) {
  
  if (!user_id %in% user_ids$user_id) {
    stop("User ID not found in training data")
  }
  
  # Find unrated books
  user_row <- user_item_matrix[as.character(user_id), , drop = TRUE]
  unrated_books <- which(is.na(user_row))
  
  if (length(unrated_books) == 0) {
    message("User has rated all available books")
    return(data.frame())
  }
  
  # Prepare prediction input
  pred_data <- data.frame(
    user_id = rep(user_id, length(unrated_books)),
    book_id = colnames(user_item_matrix)[unrated_books]
  )
  
  # Convert to H2O frame
  pred_h2o <- as.h2o(pred_data)
  pred_h2o$user_id <- as.factor(pred_h2o$user_id)
  pred_h2o$book_id <- as.factor(pred_h2o$book_id)
  
  # Predict ratings
  predictions <- h2o.predict(model, pred_h2o)
  predictions_df <- as.data.frame(predictions)
  
  # Denormalize predictions
  user_mean <- user_means[as.character(user_id)]
  pred_ratings <- predictions_df$predict + user_mean
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  # Select top recommendations
  n_to_recommend <- min(n_recommendations, length(pred_ratings))
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:n_to_recommend]
  
  recommendations <- data.frame(
    ISBN = pred_data$book_id[top_indices],
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  )
  
  recommendations <- recommendations %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}

# ============================================================================
# NN: RECOMMENDATION FOR NEW USERS (COLD-START)
# ============================================================================

recommend_for_new_user_nn <- function(model,
                                       new_user_ratings,
                                       user_item_matrix,
                                       book_ids,
                                       book_info,
                                       n_recommendations = 10,
                                       user_means = NULL) {
  
  if (length(new_user_ratings) == 0) {
    stop("No ratings provided for new user")
  }
  
  rated_books <- names(new_user_ratings)
  valid_books <- rated_books[rated_books %in% book_ids$book_id]
  
  if (length(valid_books) == 0) {
    stop("None of the rated books exist in training data")
  }
  
  # Find similar users based on rated books
  new_user_mean <- mean(new_user_ratings)
  new_user_normalized <- new_user_ratings - new_user_mean
  
  user_similarities <- numeric(nrow(user_item_matrix))
  names(user_similarities) <- rownames(user_item_matrix)
  
  for (i in 1:nrow(user_item_matrix)) {
    user_ratings <- user_item_matrix[i, valid_books]
    valid_overlap <- !is.na(user_ratings)
    
    if (sum(valid_overlap) == 0) {
      user_similarities[i] <- 0
      next
    }
    
    overlap_books <- valid_books[valid_overlap]
    existing_mean <- mean(user_ratings[valid_overlap], na.rm = TRUE)
    existing_normalized <- user_ratings[valid_overlap] - existing_mean
    new_normalized <- new_user_normalized[overlap_books]
    
    # Cosine similarity
    numerator <- sum(existing_normalized * new_normalized)
    denominator <- sqrt(sum(existing_normalized^2)) * sqrt(sum(new_normalized^2))
    
    if (denominator > 0) {
      user_similarities[i] <- numerator / denominator
    } else {
      user_similarities[i] <- 0
    }
  }
  
  # Use top-K similar users
  k <- min(30, sum(user_similarities > 0))
  
  if (k == 0) {
    # Fallback to item popularity
    item_means <- colMeans(user_item_matrix, na.rm = TRUE)
    all_books <- book_ids$book_id
    unrated_books <- setdiff(all_books, rated_books)
    pred_ratings <- item_means[unrated_books] + (new_user_mean - mean(user_means))
  } else {
    top_k_users <- names(sort(user_similarities, decreasing = TRUE)[1:k])
    similarities <- user_similarities[top_k_users]
    
    all_books <- book_ids$book_id
    unrated_books <- setdiff(all_books, rated_books)
    
    # Weighted average of similar users' ratings
    weighted_preds <- numeric(length(unrated_books))
    names(weighted_preds) <- unrated_books
    
    for (book in unrated_books) {
      similar_user_ratings <- user_item_matrix[top_k_users, book]
      valid_ratings <- !is.na(similar_user_ratings)
      
      if (sum(valid_ratings) > 0) {
        weighted_sum <- sum(similarities[valid_ratings] * similar_user_ratings[valid_ratings])
        sum_weights <- sum(abs(similarities[valid_ratings]))
        weighted_preds[book] <- weighted_sum / sum_weights
      } else {
        weighted_preds[book] <- mean(user_item_matrix[, book], na.rm = TRUE)
      }
    }
    
    pred_ratings <- weighted_preds + new_user_mean - mean(user_means)
  }
  
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  
  recommendations <- data.frame(
    ISBN = names(pred_ratings)[top_indices],
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```



```{r nn-cv}
# ============================================================================
# NN: CROSS-VALIDATION
# ============================================================================

cross_validate_nn <- function(user_item_matrix,
                               original_ratings_matrix = NULL,
                               n_folds = 5,
                               seed = 123,
                               hidden = c(64, 32),
                               epochs = 20,
                               activation = "Tanh",
                               l1 = 0.0001,
                               l2 = 0.0001) {
  
  set.seed(seed)
  h2o.no_progress()
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  if (n_ratings < n_folds * 2) {
    stop("Not enough ratings to perform cross-validation")
  }
  
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  # cat("Starting", n_folds, "-fold cross-validation for NN...\n")
  # cat("Training on EXPLICIT ratings only\n\n")
  
  for (fold in 1:n_folds) {
    # cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    train_indices <- which(fold_indices != fold)
    
    test_obs <- observed[test_indices, , drop = FALSE]
    train_obs <- observed[train_indices, , drop = FALSE]
    
    # Create training matrix
    train_matrix <- user_item_matrix
    train_matrix[test_obs] <- NA
    
    # Normalize
    train_user_means <- rowMeans(train_matrix, na.rm = TRUE)
    train_normalized <- sweep(train_matrix, 1, train_user_means, FUN = "-")
    
    # Create training data
    train_data <- data.frame(
      user_id = rownames(user_item_matrix)[train_obs[, 1]],
      book_id = colnames(user_item_matrix)[train_obs[, 2]],
      rating = as.numeric(train_normalized[train_obs])
    )
    
    # Filter explicit ratings
    if (!is.null(original_ratings_matrix)) {
      train_data$original_rating <- original_ratings_matrix[train_obs]
      train_data_explicit <- train_data[train_data$original_rating != 0, ]
    } else {
      train_data_explicit <- train_data
    }
    
    if (nrow(train_data_explicit) == 0) {
      cv_results[fold, ] <- c(fold, NA)
      # cat("No explicit ratings - skipping\n")
      next
    }
    
    # Convert to H2O
    train_h2o <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
    train_h2o$user_id <- as.factor(train_h2o$user_id)
    train_h2o$book_id <- as.factor(train_h2o$book_id)
    
    # Train model
    model <- h2o.deeplearning(
      x = c("user_id", "book_id"),
      y = "rating",
      training_frame = train_h2o,
      hidden = hidden,
      epochs = epochs,
      activation = activation,
      l1 = l1,
      l2 = l2,
      adaptive_rate = TRUE,
      seed = seed,
      stopping_rounds = 3,
      stopping_tolerance = 0.001,
      max_w2 = 10,
      initial_weight_distribution = "UniformAdaptive",
      initial_weight_scale = 0.5,
      verbose = FALSE
    )
    
    # Prepare test data (explicit only)
    test_user_means <- train_user_means[rownames(user_item_matrix)[test_obs[, 1]]]
    test_data <- data.frame(
      user_id = rownames(user_item_matrix)[test_obs[, 1]],
      book_id = colnames(user_item_matrix)[test_obs[, 2]],
      rating = as.numeric(user_item_matrix[test_obs]) - test_user_means
    )
    
    if (!is.null(original_ratings_matrix)) {
      test_data$original_rating <- original_ratings_matrix[test_obs]
      test_data_explicit <- test_data[test_data$original_rating != 0, ]
      test_user_means_explicit <- test_user_means[test_data$original_rating != 0]
      actual_explicit <- user_item_matrix[test_obs][test_data$original_rating != 0]
    } else {
      test_data_explicit <- test_data
      test_user_means_explicit <- test_user_means
      actual_explicit <- user_item_matrix[test_obs]
    }
    
    # Convert to H2O
    test_h2o <- as.h2o(test_data_explicit[, c("user_id", "book_id", "rating")])
    test_h2o$user_id <- as.factor(test_h2o$user_id)
    test_h2o$book_id <- as.factor(test_h2o$book_id)
    
    # Predict
    predictions <- as.data.frame(h2o.predict(model, test_h2o))
    predictions_denorm <- predictions$predict + test_user_means_explicit
    predictions_denorm <- pmax(pmin(predictions_denorm, 10), 1)
    
    # Calculate RMSE
    rmse_value <- sqrt(mean((predictions_denorm - actual_explicit)^2, na.rm = TRUE))
    
    cv_results[fold, ] <- list(fold, rmse_value)
    # cat("RMSE:", round(rmse_value, 4), "\n")
    
    # Cleanup
    h2o.rm(model)
    h2o.rm(train_h2o)
    h2o.rm(test_h2o)
    gc()
  }
  
  return(cv_results)
}
```



```{r nn-implementation}
# ============================================================================
# NN: MODEL TRAINING AND EVALUATION
# ============================================================================

# Initialize H2O cluster
suppressMessages(h2o.init(ip = "localhost", port = 54321))
h2o.no_progress()

# Prepare data for final model
prepared <- prepare_h2o_data(user_item_matrix, original_ratings_matrix)

optimal_params_nn <- list(
  hidden = as.numeric(unlist(strsplit(as.character(best_nn_results$Hidden_Layers), ", "))),
  epochs = 30,  # Default value
  activation = as.character(best_nn_results$Activation),
  l1 = as.numeric(best_nn_results$L1_Regularisation),
  l2 = as.numeric(best_nn_results$L2_Regularisation)
)

# Train final model
train_data_explicit <- prepared$train_data[!prepared$train_data$is_implicit, ]

train_h2o <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
train_h2o$user_id <- as.factor(train_h2o$user_id)
train_h2o$book_id <- as.factor(train_h2o$book_id)


# Train new model
model_nn <- train_h2o_model(
  train_data = train_h2o,
  hidden = optimal_params_nn$hidden,
  epochs = optimal_params_nn$epochs,
  activation = optimal_params_nn$activation,
  l1 = optimal_params_nn$l1,
  l2 = optimal_params_nn$l2,
  verbose = FALSE)
  
# Generate recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_nn <- recommend_for_user_nn(
  model = model_nn,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids = prepared$user_ids,
  book_ids = prepared$book_ids,
  user_means = prepared$user_means,
  book_info = book_info
)


# Generate recommendations for new user (cold-start)
new_user_recs_nn <- recommend_for_new_user_nn(
  model = model_nn,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  book_ids = prepared$book_ids,
  book_info = book_info,
  n_recommendations = 10,
  user_means = prepared$user_means
)


```



## 3.4 Cold-Start Evaluation

A synthetic new user profile was created by selecting the first five books from the user-item matrix and assigning ratings [7, 8, 6, 5, 7]. Memory-based methods computed similarities on-the-fly from available ratings. 

Memory-based methods (UBCF and IBCF) computed similarities on-the-fly from the new user's ratings to existing users or items in the training set. MF employed a hybrid approach using user bias, calculated as the difference between the new user's mean rating and the global mean rating, combined with item popularity estimates. NN methods applied a similarity-based fallback strategy, identifying the top 30 most similar existing users based on overlapping ratings and generating predictions through weighted averaging of these users' preferences.

This evaluation demonstrates each method's ability to generate meaningful recommendations with minimal user history, a critical requirement for practical recommender systems facing continuous user acquisition.


## 3.5 Catalog Size Impact Analysis

Catalog size (N ∈ {5, 10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100, 110, 120, 130, 145}) was varied using the most popular books, with users required to have at least three ratings per catalog. The analysis employed the best-performing recommender system, MF, and was repeated across five random seeds, with mean RMSE and standard error computed for each catalog size.


# 4. Results and Evaluation

## 4.1 Cross-Validation Performance

MF achieved the lowest RMSE of 1.820 (SD = 0.020), followed by NN at 1.899 (SD = 0.053), IBCF at 2.746 (SD = 0.018), and UBCF at 3.182 (SD = 0.047). Table 2 and Figure 2 present these results.

The 43.0% reduction in error between the best method (MF) and worst method (UBCF) highlights the substantial impact of algorithm selection in sparse datasets. Model-based approaches (MF and NN) demonstrated significantly superior performance compared to memory-based methods (IBCF and UBCF), confirming their advantage in learning latent patterns from limited data.

```{r comprehensive-comparison}
# ============================================================================
# COMPREHENSIVE CROSS-VALIDATION COMPARISON
# ============================================================================

run_comprehensive_cv_comparison <- function(user_item_matrix,
                                            confidence_matrix,
                                            original_ratings_matrix,
                                            optimal_params_mf,
                                            optimal_params_nn,
                                            n_folds = 5,
                                            k = 50) {
  
  
  all_results <- data.frame(
    Method = character(4),
    CV_RMSE_Mean = numeric(4),
    CV_RMSE_SD = numeric(4),
    stringsAsFactors = FALSE
  )
  
  # UBCF
  # cat("=== Running UBCF ===\n")
  cv_results_ubcf <- cross_validate_ubcf(
    user_item_matrix,
    confidence_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    k = k
  )
  
  all_results[1, ] <- list(
    "UBCF",
    round(mean(cv_results_ubcf$rmse), 3),
    round(sd(cv_results_ubcf$rmse), 3)
  )
  
  # IBCF
  # cat("\n=== Running IBCF ===\n")
  cv_results_ibcf <- cross_validate_ibcf(
    user_item_matrix,
    confidence_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    k = k
  )
  
  all_results[2, ] <- list(
    "IBCF",
    round(mean(cv_results_ibcf$rmse), 3),
    round(sd(cv_results_ibcf$rmse), 3)
  )
  
  # MF
  # cat("\n=== Running MF ===\n")
  cv_results_mf <- cross_validate_mf(
    user_item_matrix,
    original_ratings_matrix,
    optimal_params_mf,
    n_folds = n_folds,
    n_iter = 50,
    verbose = TRUE
  )
  
  all_results[3, ] <- list(
    "MF",
    round(mean(cv_results_mf$RMSE, na.rm = TRUE), 3),
    round(sd(cv_results_mf$RMSE, na.rm = TRUE), 3)
  )
  
  # NN
  # cat("\n=== Running NN ===\n")
  cv_results_nn <- cross_validate_nn(
    user_item_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    hidden = optimal_params_nn$hidden,
    epochs = optimal_params_nn$epochs,
    activation = optimal_params_nn$activation,
    l1 = optimal_params_nn$l1,
    l2 = optimal_params_nn$l2
  )
  
  all_results[4, ] <- list(
    "NN",
    round(mean(cv_results_nn$rmse, na.rm = TRUE), 3),
    round(sd(cv_results_nn$rmse, na.rm = TRUE), 3)
  )
  

  # Return all results including individual CV results
  return(list(
    summary = all_results,
    cv_results_ubcf = cv_results_ubcf,
    cv_results_ibcf = cv_results_ibcf,
    cv_results_mf = cv_results_mf,
    cv_results_nn = cv_results_nn
  ))
}

# # Run comprehensive comparison
# comparison_results <- run_comprehensive_cv_comparison(
#   user_item_matrix = user_item_matrix,
#   confidence_matrix = confidence_matrix,
#   original_ratings_matrix = original_ratings_matrix,
#   optimal_params_mf = optimal_params_mf,
#   optimal_params_nn = optimal_params_nn,
#   n_folds = 5,
#   k = 50
# )
# 
# # Extract results
# cv_comparison_results <- comparison_results$summary
# cv_results_ubcf <- comparison_results$cv_results_ubcf
# cv_results_ibcf <- comparison_results$cv_results_ibcf
# cv_results_mf <- comparison_results$cv_results_mf
# cv_results_nn <- comparison_results$cv_results_nn
# 
# save(cv_comparison_results, cv_results_ubcf, cv_results_ibcf, cv_results_mf, cv_results_nn, 
#      file = "all_cv_results.Rdata")

# Display results
kable(
  cv_comparison_results,
  format = "latex",
  booktabs = TRUE,
  caption = "Cross-Validation Performance Comparison: RMSE results across all four collaborative filtering methods showing mean performance and standard deviation from 5-fold cross-validation",
  col.names = c("Method", "Mean RMSE", "SD RMSE"),
  align = c("l", "c", "c"),
  digits = 3
) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )



```

## 4.2 Cold-Start Analysis

Analyzing the 40 total recommendations (10 per method) revealed substantial algorithmic divergence. Of these 40 recommendations, only 7 unique titles (17.5%) appeared in multiple methods, meaning 33 titles (82.5%) were recommended by only one method. This demonstrates that different algorithms identify fundamentally different items as suitable for the same user profile.

Among the shared recommendations, "Harry Potter and the Order of the Phoenix" appeared in three methods with varying predicted ratings: UBCF (7.34), IBCF (7.03), and NN (6.50). This represents the highest consensus across methods. Six additional titles appeared in exactly two methods, showing notable rating discrepancies of up to 1.36 points. These variations suggest different confidence calibrations across algorithms, with memory-based methods generally predicting higher ratings than model-based approaches for the same items.

The low overlap highlights that each algorithm leverages distinct aspects of user-item relationships: UBCF emphasizes user similarity patterns, IBCF focuses on item co-occurrence, MF captures linear latent factors, and NN models non-linear interactions. This diversity suggests potential value in ensemble approaches that combine multiple methods. Detailed recommendation tables showing all 40 suggestions are provided in Appendix B.




```{r results-visualization, fig.cap="Cross-Validation Performance Comparison: Mean RMSE with ±1 SD error bars across four collaborative filtering methods."}
# ============================================================================
# VISUALIZATION: PERFORMANCE COMPARISON
# ============================================================================

# Ensure H2O is available for any remaining operations
if (!h2o::h2o.clusterIsUp()) {
  suppressMessages(h2o::h2o.init(ip = "localhost", port = 54321))
  h2o::h2o.no_progress()
}

# Create bar plot for RMSE comparison
ggplot(cv_comparison_results, aes(x = reorder(Method, CV_RMSE_Mean), y = CV_RMSE_Mean)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_errorbar(
    aes(ymin = CV_RMSE_Mean - CV_RMSE_SD, ymax = CV_RMSE_Mean + CV_RMSE_SD),
    width = 0.2
  ) +
  geom_text(aes(label = round(CV_RMSE_Mean, 3)), vjust = -0.5) +
  labs(
    title = "Cross-Validation Performance Comparison",
    subtitle = "Lower RMSE indicates better performance",
    x = "Method",
    y = "Mean RMSE"
  ) +
  theme_cowplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```



## 4.3 Catalog Size Impact

MF achieved optimal performance with 50 books, reaching an RMSE of 1.815 ± 0.027, a 10.4% improvement over the 5-book catalog, which had an RMSE of 2.027. Performance remained stable across the 40–60 book range, with standard errors below 0.030, indicating consistent predictions across multiple random seeds.

Sharp improvements occurred in the early phase, as expanding from 5 to 10 books reduced RMSE by 7.3% (2.027 to 1.879), and increasing from 10 to 20 books added another 2.8% improvement (1.879 to 1.826). Beyond 50 books, performance plateaued and slightly declined, with the full 145-book catalog yielding an RMSE of 1.833, slightly worse than the 50-book optimum.

This non-monotonic pattern reflects two competing effects. Increasing the number of books improves latent factor estimation in the early phase, while adding less-rated titles introduces noise in the later phase. The optimal catalog size of 50 books occurs at 86.7% sparsity, demonstrating MF's effectiveness in highly sparse environments.

These results suggest that recommendation systems can achieve near-optimal accuracy using carefully curated catalogs of 50–60 items rather than exhaustive inventories, improving both computational efficiency and recommendation quality. Figure 3 illustrates these trends, and detailed results for all 17 catalog sizes are provided in Appendix C.2.



```{r}
# ============================================================================
# DATASET SIZE IMPACT ANALYSIS - FULLY CORRECTED
# ============================================================================
# Research Question: How many book titles does a recommender system need
# to achieve good prediction accuracy?
# 
# Methodology:
# 1. For each catalog size (5, 10, 20, 30, 50, 75, 100, 145 books)
# 2. Select TOP N most popular books
# 3. For EACH USER: Split their ratings 80/20 into train/test
# 4. Train MF model on 80%, test on 20%
# 5. Run 5 times with different random seeds for robustness
# 6. CRITICAL: Only include users who have rated enough books to split fairly
#
# Key: This ensures the same books appear in both train and test,
# and each user contributes proportionally to both sets
# ============================================================================

analyze_catalog_sizes_fair <- function(data, 
                                      catalog_sizes = c(5, 10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100, 110, 120, 130, 145),
                                      optimal_params_mf,
                                      test_fraction = 0.2,
                                      min_user_ratings = 3,  # Minimum ratings per user to participate
                                      n_seeds = 5,
                                      base_seed = 123) {
  
  set.seed(base_seed)
  all_results <- data.frame()
  
  # Get book popularity rankings
  book_popularity <- data %>%
    group_by(ISBN) %>%
    summarise(n_ratings = n(), .groups = "drop") %>%
    arrange(desc(n_ratings))
  
  # cat("=== Fair Catalog Size Analysis ===\n")
  # cat("Method: Per-user 80/20 split on TOP N popular books\n")
  # cat("Minimum ratings per user:", min_user_ratings, "\n")
  # cat("Testing catalog sizes:", paste(catalog_sizes, collapse = ", "), "\n")
  # cat("Total catalog sizes to test:", length(catalog_sizes), "\n\n")
  
  # Verify we have enough books
  if (nrow(book_popularity) < max(catalog_sizes)) {
    stop("Not enough books in dataset. Available: ", nrow(book_popularity), 
         ", Requested: ", max(catalog_sizes))
  }
  
  # Process each catalog size
  for (n_books in catalog_sizes) {
    # cat("=== Catalog Size:", n_books, "books ===\n")
    
    # Select TOP N most popular books
    selected_books <- head(book_popularity$ISBN, n_books)
    data_subset <- data %>% filter(ISBN %in% selected_books)
    
    # Count ratings per user in this subset
    user_rating_counts <- data_subset %>%
      group_by(User.ID) %>%
      summarise(n_ratings = n(), .groups = "drop")
    
    # Only keep users with enough ratings to split
    eligible_users <- user_rating_counts %>%
      filter(n_ratings >= min_user_ratings) %>%
      pull(User.ID)
    
    data_subset_filtered <- data_subset %>%
      filter(User.ID %in% eligible_users)
    
    n_eligible <- length(eligible_users)
    n_total_ratings <- nrow(data_subset_filtered)
    
    # cat("Selected", length(selected_books), "books\n")
    # cat("Eligible users (>=", min_user_ratings, "ratings):", n_eligible, "\n")
    # cat("Total ratings from eligible users:", n_total_ratings, "\n")
    
    if (n_eligible < 10) {
      # cat("SKIP: Not enough eligible users\n\n")
      
      all_results <- rbind(all_results, data.frame(
        catalog_size = n_books,
        seed = NA,
        n_users = 0,
        n_items = length(selected_books),
        n_ratings_total = n_total_ratings,
        n_explicit = NA,
        sparsity_pct = NA,
        n_train = NA,
        n_test = NA,
        rmse = NA,
        reason = "Insufficient eligible users"
      ))
      next
    }
    
    # Run multiple seeds
    for (seed_num in 1:n_seeds) {
      current_seed <- base_seed + seed_num
      set.seed(current_seed)
      
      # cat("  Seed", seed_num, "of", n_seeds, "...")
      
      # ================================================================
      # CRITICAL: PER-USER 80/20 SPLIT
      # ================================================================
      # For each user, randomly select 20% of their ratings for testing
      # This ensures EVERY user contributes to both train and test sets
      
      train_data_list <- list()
      test_data_list <- list()
      
      for (user in eligible_users) {
        user_ratings <- data_subset_filtered %>% 
          filter(User.ID == user)
        
        n_user_ratings <- nrow(user_ratings)
        n_test_user <- max(1, floor(test_fraction * n_user_ratings))
        
        # Randomly select test indices for this user
        test_indices <- sample(1:n_user_ratings, n_test_user)
        
        test_data_list[[as.character(user)]] <- user_ratings[test_indices, ]
        train_data_list[[as.character(user)]] <- user_ratings[-test_indices, ]
      }
      
      # Combine all users' train and test data
      train_data_final <- bind_rows(train_data_list)
      test_data_final <- bind_rows(test_data_list)
      
      # cat("Train:", nrow(train_data_final), "ratings, ",
      #     "Test:", nrow(test_data_final), "ratings...")
      
      # ================================================================
      # CREATE MATRICES
      # ================================================================
      
      # Create FULL matrix first (to get consistent dimensions)
      matrix_result_full <- create_user_item_matrix(
        data_subset_filtered,
        min_ratings_per_book = 1,  # Don't filter - we already selected books
        min_ratings_per_user = 1,  # Don't filter - we already selected users
        implicit_rating = 4,
        implicit_confidence = 0.5
      )
      
      full_matrix <- matrix_result_full$ratings
      full_original <- matrix_result_full$original_ratings
      
      # Create TRAIN matrix (mask test ratings)
      train_matrix <- full_matrix
      train_original <- full_original
      
      # Mask out test ratings
      for (i in 1:nrow(test_data_final)) {
        user_id <- as.character(test_data_final$User.ID[i])
        book_id <- test_data_final$ISBN[i]
        
        if (user_id %in% rownames(train_matrix) && 
            book_id %in% colnames(train_matrix)) {
          train_matrix[user_id, book_id] <- NA
          train_original[user_id, book_id] <- NA
        }
      }
      
      # Calculate metrics
      n_users <- nrow(full_matrix)
      n_items <- ncol(full_matrix)
      n_ratings_total <- sum(!is.na(full_matrix))
      n_explicit_total <- sum(full_original[!is.na(full_original)] > 0)
      sparsity <- mean(is.na(full_matrix)) * 100
      
      # Prepare train data for MF (explicit only)
      train_observed <- which(!is.na(train_matrix), arr.ind = TRUE)
      
      train_data_mf <- data.frame(
        user_index = train_observed[, 1] - 1,
        item_index = train_observed[, 2] - 1,
        rating = train_matrix[train_observed],
        original_rating = train_original[train_observed]
      )
      
      # Filter to explicit ratings only
      train_data_explicit <- train_data_mf %>%
        filter(original_rating > 0)
      
      n_train_explicit <- nrow(train_data_explicit)
      
      if (n_train_explicit < 10) {
        # cat(" SKIP (only", n_train_explicit, "explicit training ratings)\n")
        
        all_results <- rbind(all_results, data.frame(
          catalog_size = n_books,
          seed = seed_num,
          n_users = n_users,
          n_items = n_items,
          n_ratings_total = n_ratings_total,
          n_explicit = n_explicit_total,
          sparsity_pct = sparsity,
          n_train = n_train_explicit,
          n_test = NA,
          rmse = NA,
          reason = "Insufficient training data"
        ))
        next
      }
      
      # ================================================================
      # TRAIN MODEL
      # ================================================================
      
      # Create mapping for users and items
      user_ids_map <- data.frame(
        user_index = 0:(n_users - 1),
        user_id = rownames(train_matrix)
      )
      
      item_ids_map <- data.frame(
        item_index = 0:(n_items - 1),
        item_id = colnames(train_matrix)
      )
      
      # Train MF model
      train_set <- data_memory(
        user_index = train_data_explicit$user_index,
        item_index = train_data_explicit$item_index,
        rating = train_data_explicit$rating,
        index1 = FALSE
      )
      
      model <- Reco()
      suppressWarnings({
        model$train(train_set, opts = list(
          dim = optimal_params_mf$min$dim,
          lrate = optimal_params_mf$min$lrate,
          costp_l2 = optimal_params_mf$min$costp_l2,
          costq_l2 = optimal_params_mf$min$costq_l2,
          niter = 50,
          nthread = 4,
          verbose = FALSE
        ))
      })
      
      # ================================================================
      # MAKE PREDICTIONS ON TEST SET
      # ================================================================
      
      predictions <- numeric(nrow(test_data_final))
      test_actuals <- numeric(nrow(test_data_final))
      test_original_ratings <- numeric(nrow(test_data_final))
      valid_predictions <- rep(TRUE, nrow(test_data_final))
      
      for (i in 1:nrow(test_data_final)) {
        user_id <- as.character(test_data_final$User.ID[i])
        book_id <- test_data_final$ISBN[i]
        
        # Check if user and book exist in matrices
        if (!(user_id %in% user_ids_map$user_id) || 
            !(book_id %in% item_ids_map$item_id)) {
          valid_predictions[i] <- FALSE
          next
        }
        
        # Get indices
        user_idx <- user_ids_map$user_index[user_ids_map$user_id == user_id]
        item_idx <- item_ids_map$item_index[item_ids_map$item_id == book_id]
        
        # Make prediction
        pred_set <- data_memory(
          user_index = user_idx,
          item_index = item_idx,
          index1 = FALSE
        )
        
        pred <- model$predict(pred_set, out_memory())
        predictions[i] <- pmax(pmin(pred, 10), 1)
        
        # Get actual rating
        test_actuals[i] <- full_matrix[user_id, book_id]
        test_original_ratings[i] <- full_original[user_id, book_id]
      }
      
      # Filter to valid predictions and explicit ratings only
      valid_mask <- valid_predictions & (test_original_ratings > 0)
      n_test_explicit <- sum(valid_mask)
      
      if (n_test_explicit == 0) {
        # cat(" SKIP (no valid explicit test ratings)\n")
        
        all_results <- rbind(all_results, data.frame(
          catalog_size = n_books,
          seed = seed_num,
          n_users = n_users,
          n_items = n_items,
          n_ratings_total = n_ratings_total,
          n_explicit = n_explicit_total,
          sparsity_pct = sparsity,
          n_train = n_train_explicit,
          n_test = 0,
          rmse = NA,
          reason = "No valid test ratings"
        ))
        next
      }
      
      # Calculate RMSE on explicit test ratings
      pred_explicit <- predictions[valid_mask]
      actual_explicit <- test_actuals[valid_mask]
      
      rmse <- sqrt(mean((pred_explicit - actual_explicit)^2, na.rm = TRUE))
      
      # Store results
      all_results <- rbind(all_results, data.frame(
        catalog_size = n_books,
        seed = seed_num,
        n_users = n_users,
        n_items = n_items,
        n_ratings_total = n_ratings_total,
        n_explicit = n_explicit_total,
        sparsity_pct = sparsity,
        n_train = n_train_explicit,
        n_test = n_test_explicit,
        rmse = rmse,
        reason = "Success"
      ))
      
      # cat(" RMSE =", round(rmse, 4), "\n")
    }
    
    # cat("\n")
  }
  
  # cat("=== Analysis Complete ===\n\n")
  
  # ================================================================
  # AGGREGATE RESULTS
  # ================================================================
  
  summary_results <- all_results %>%
    group_by(catalog_size) %>%
    summarise(
      n_valid_runs = sum(!is.na(rmse)),
      rmse_mean = mean(rmse, na.rm = TRUE),
      rmse_sd = sd(rmse, na.rm = TRUE),
      rmse_se = sd(rmse, na.rm = TRUE) / sqrt(sum(!is.na(rmse))),
      rmse_min = min(rmse, na.rm = TRUE),
      rmse_max = max(rmse, na.rm = TRUE),
      n_users_mean = round(mean(n_users, na.rm = TRUE), 0),
      n_items_mean = round(mean(n_items, na.rm = TRUE), 0),
      sparsity_mean = round(mean(sparsity_pct, na.rm = TRUE), 1),
      n_explicit_mean = round(mean(n_explicit, na.rm = TRUE), 0),
      n_train_mean = round(mean(n_train, na.rm = TRUE), 0),
      n_test_mean = round(mean(n_test, na.rm = TRUE), 0),
      .groups = "drop"
    ) %>%
    arrange(catalog_size) %>%
    mutate(
      rmse_change = rmse_mean - lag(rmse_mean),
      improvement = -rmse_change,
      percent_improvement = -rmse_change / lag(rmse_mean) * 100
    )
  
  return(list(
    detailed = all_results,
    summary = summary_results
  ))
}

# ============================================================================
# RUN THE ANALYSIS
# ============================================================================

# catalog_analysis_results <- analyze_catalog_sizes_fair(
#   data = data,
#   catalog_sizes = c(5, 10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100, 110, 120, 130, 145),
#   optimal_params_mf = optimal_params_mf,
#   test_fraction = 0.2,
#   min_user_ratings = 3,  
#   n_seeds = 5,
#   base_seed = 123
# )
# 
# # Extract results
# size_impact_results <- catalog_analysis_results$summary
# size_impact_results_detailed <- catalog_analysis_results$detailed
# 
# save(size_impact_results, size_impact_results_detailed, 
#      file = "dataset_size_analysis_complete.Rdata")


# Table moved to Appendix C.2 for detailed reference

```


```{r dataset-size-visualization, fig.cap="Catalog Size Impact on Prediction Accuracy: Mean RMSE (±1 SE) across catalog sizes from 5 to 145 books. Red diamond marks optimal size (50 books, RMSE=1.815)."}

# RMSE Trend with Error Bars
ggplot(size_impact_results, aes(x = catalog_size, y = rmse_mean)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  geom_errorbar(aes(ymin = rmse_mean - rmse_se, ymax = rmse_mean + rmse_se),
                width = 3, color = "steelblue", alpha = 0.7) +
  geom_point(data = size_impact_results[which.min(size_impact_results$rmse_mean), ],
             color = "red", size = 5, shape = 18) +
  labs(title = "A. RMSE by Catalog Size",
       subtitle = "Error bars show ±1 SE across 5 seeds. Red diamond indicates optimal size.",
       x = "Number of Books", y = "Mean RMSE") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

```






```{r cleanup, include=FALSE}
# Clean up H2O cluster at the end
if (exists("h2o") && h2o::h2o.clusterIsUp()) {
  h2o::h2o.shutdown(prompt = FALSE)
}
```






# 5. Discussion

## 5.1 Comparative Performance

Model-based methods substantially outperformed memory-based ones, confirming superior robustness to sparsity. MF achieved best accuracy-stability balance. The 43% performance gap underscores how algorithmic structure shapes accuracy under sparse conditions. IBCF outperformed UBCF as item relationships prove more stable than user relationships.

## 5.2 Catalog Size Dynamics

Non-monotonic relationship revealed optimal performance at 50 books before plateauing. Improvement phase (5→50 books) enabled better latent factor estimation. Beyond this threshold, less-rated titles introduced noise. Decreasing variability beyond 50 books (SE < 0.030) reflects greater stability with sufficient data density.

## 5.3 Recommendation Divergence

82.5% uniqueness highlights fundamental differences in preference modeling. Memory-based methods operate on orthogonal principles (user similarity vs. item co-occurrence), explaining minimal overlap. Model-based approaches infer latent structures differently—MF captures linear relationships while NN reveals nonlinear interactions. Rating discrepancies for shared titles (up to 1.36 points) suggest distinct confidence calibrations. Detailed recommendation examples demonstrating these differences are provided in Appendix B.

## 5.4 Cold-Start Limitations

Model-based methods required simplified fallback strategies differing from core mechanisms, limiting interpretability. This reveals a fundamental trade-off: model-based methods achieve superior accuracy for existing users but need auxiliary strategies for new ones, while memory-based methods remain algorithmically consistent. Practical systems require hybrid solutions integrating content-based features or transfer learning.

# 6. Conclusion

This study implemented and evaluated four collaborative filtering approaches for book recommendations, providing key insights for system design. MF was the most accurate method (RMSE = 1.820), while confidence-weighted similarity effectively leveraged implicit feedback to enhance memory-based methods. Catalog size analysis showed that accuracy gains plateaued beyond 50–60 books, indicating that near-optimal performance can be achieved without expansive catalogs.

Limitations include the 145-book dataset, which may not capture full recommendation diversity, the need for refined handling of implicit feedback, and suboptimal cold-start strategies for model-based systems that do not fully exploit learned representations. Future work should explore hybrid ensembles combining complementary methods, advanced deep learning architectures (autoencoders, attention mechanisms), context-aware metadata (genres, authors, temporal patterns), and online learning for real-time adaptation. Overall, this study offers both theoretical insights and practical guidance for developing production-ready recommendation systems across domains.

# Appendix: Detailed Results and Supporting Analysis

This appendix contains detailed statistical summaries, individual method results, and technical details that support the main analysis but are not essential to the core narrative.

## Appendix A: Detailed Exploratory Data Analysis

```{r appendix-detailed-eda}
# ============================================================================
# DETAILED USER AND BOOK ACTIVITY ANALYSIS
# ============================================================================

# User activity analysis
user_activity <- data %>%
  group_by(User.ID) %>%
  summarise(
    num_ratings = n(),
    avg_rating = mean(Book.Rating, na.rm = TRUE),
    .groups = "drop"
  )

# Book popularity analysis
book_popularity <- data %>%
  group_by(ISBN) %>%
  summarise(
    num_ratings = n(),
    avg_rating = mean(Book.Rating, na.rm = TRUE),
    .groups = "drop"
  )

# User activity summary table
user_activity_summary <- data.frame(
  Metric = c(
    "Total Users",
    "Min Ratings per User", 
    "Max Ratings per User",
    "Mean Ratings per User",
    "Users with 1-2 Ratings",
    "Users with 3-5 Ratings", 
    "Users with 6-10 Ratings",
    "Users with 11-20 Ratings",
    "Users with >20 Ratings"
  ),
  Value = c(
    nrow(user_activity),
    min(user_activity$num_ratings),
    max(user_activity$num_ratings),
    round(mean(user_activity$num_ratings), 2),
    sum(user_activity$num_ratings <= 2),
    sum(user_activity$num_ratings >= 3 & user_activity$num_ratings <= 5),
    sum(user_activity$num_ratings >= 6 & user_activity$num_ratings <= 10),
    sum(user_activity$num_ratings >= 11 & user_activity$num_ratings <= 20),
    sum(user_activity$num_ratings > 20)
  )
)

kable(user_activity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed User Activity Distribution: Statistical breakdown of user rating patterns showing distribution of ratings per user and user activity categories",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Book popularity summary table
book_popularity_summary <- data.frame(
  Metric = c(
    "Total Books",
    "Min Ratings per Book",
    "Max Ratings per Book", 
    "Mean Ratings per Book",
    "Books with 1-4 Ratings",
    "Books with 5-9 Ratings",
    "Books with 10-19 Ratings", 
    "Books with 20-49 Ratings",
    "Books with ≥50 Ratings"
  ),
  Value = c(
    nrow(book_popularity),
    min(book_popularity$num_ratings),
    max(book_popularity$num_ratings),
    round(mean(book_popularity$num_ratings), 2),
    sum(book_popularity$num_ratings <= 4),
    sum(book_popularity$num_ratings >= 5 & book_popularity$num_ratings <= 9),
    sum(book_popularity$num_ratings >= 10 & book_popularity$num_ratings <= 19),
    sum(book_popularity$num_ratings >= 20 & book_popularity$num_ratings <= 49),
    sum(book_popularity$num_ratings >= 50)
  )
)

kable(book_popularity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Book Popularity Distribution: Statistical analysis of item rating patterns showing distribution of ratings per book and popularity categories",
      align = c("l", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Rating distribution (numeric breakdown)
rating_dist <- table(data$Book.Rating)
rating_dist_df <- data.frame(
  Rating = names(rating_dist),
  Count = as.numeric(rating_dist),
  Percentage = round(as.numeric(rating_dist) / sum(rating_dist) * 100, 1)
)

kable(rating_dist_df, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Rating Distribution: Frequency analysis of explicit and implicit feedback showing count and percentage for each rating value",
      align = c("c", "c", "c")) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# Dataset sparsity analysis
total_possible_ratings <- length(unique(data$User.ID)) * length(unique(data$ISBN))
actual_ratings <- nrow(data)
sparsity_percentage <- round((1 - actual_ratings / total_possible_ratings) * 100, 2)

sparsity_summary <- data.frame(
  Metric = c(
    "Total Possible User-Book Pairs",
    "Actual Ratings",
    "Sparsity Percentage"
  ),
  Value = c(
    format(total_possible_ratings, big.mark = ","),
    format(actual_ratings, big.mark = ","),
    paste0(sparsity_percentage, "%")
  )
)

kable(sparsity_summary, 
      format = "latex",
      booktabs = TRUE,
      caption = "Detailed Dataset Sparsity Analysis: Statistical summary of matrix sparsity showing total possible user-book pairs, actual ratings, and sparsity percentage",
      align = c("l", "c")
) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

## Appendix B: Method-Specific Recommendations

The following tables show detailed recommendation outputs for each method, demonstrating the practical application of each approach.

### B.1 UBCF Recommendations

```{r appendix-ubcf-recommendations, include=FALSE}
# UBCF recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ubcf <- recommend_for_user_ubcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_sim_matrix = user_similarity_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

recs_ubcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "UBCF Recommendations: Top 10 book suggestions for existing user based on user similarity with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# UBCF recommendations for new user
sample_books <- colnames(user_item_matrix)[1:5]
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), sample_books)

new_user_recs_ubcf <- recommend_for_new_user_ubcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix,
  user_means = user_means_ubcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

new_user_recs_ubcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "UBCF Cold-Start Recommendations: Top 10 book suggestions for new user using similarity-based approach with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

### B.2 IBCF Recommendations

```{r appendix-ibcf-recommendations, include=FALSE}
# IBCF recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_ibcf <- recommend_for_user_ibcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_sim_matrix = item_similarity_matrix,
  item_means = item_means_ibcf,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

recs_ibcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "IBCF Recommendations: Top 10 book suggestions for existing user based on item similarity with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# IBCF recommendations for new user
new_user_recs_ibcf <- recommend_for_new_user_ibcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix,
  item_means = item_means_ibcf,
  item_sim_matrix = item_similarity_matrix,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

new_user_recs_ibcf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "IBCF Cold-Start Recommendations: Top 10 book suggestions for new user using item-based approach with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

### B.3 MF Recommendations

```{r appendix-mf-recommendations, include=FALSE}
# MF recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_mf <- recommend_for_user_mf(
  model = model_mf,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids_map = prepared$user_ids,
  item_ids_map = prepared$item_ids,
  book_info = book_info
)

recs_mf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF Recommendations: Top 10 book suggestions for existing user using matrix factorization with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# MF recommendations for new user
recs_new_mf <- recommend_for_new_user_mf(
  model = model_mf,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  item_ids_map = prepared$item_ids,
  book_info = book_info,
  n_recommendations = 10
)

recs_new_mf %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF Cold-Start Recommendations: Top 10 book suggestions for new user using hybrid approach with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

### B.4 NN Recommendations

```{r appendix-nn-recommendations, include=FALSE}
# NN recommendations for existing user
sample_user <- rownames(user_item_matrix)[3]

recs_nn <- recommend_for_user_nn(
  model = model_nn,
  user_item_matrix = user_item_matrix,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids = prepared$user_ids,
  book_ids = prepared$book_ids,
  user_means = prepared$user_means,
  book_info = book_info
)

recs_nn %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "NN Recommendations: Top 10 book suggestions for existing user using neural network with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )

# NN recommendations for new user
new_user_recs_nn <- recommend_for_new_user_nn(
  model = model_nn,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix,
  book_ids = prepared$book_ids,
  book_info = book_info,
  n_recommendations = 10,
  user_means = prepared$user_means
)

new_user_recs_nn %>%
  mutate(Rank = row_number()) %>%
  mutate(
    Book.Title = ifelse(nchar(Book.Title) > 50, 
                      paste0(substr(Book.Title, 1, 47), "..."), 
                      Book.Title),
    Book.Author = ifelse(nchar(Book.Author) > 25, 
                        paste0(substr(Book.Author, 1, 22), "..."), 
                        Book.Author)
  ) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "NN Cold-Start Recommendations: Top 10 book suggestions for new user using similarity-based fallback with predicted ratings",
    digits = 2,
    align = c("c", "l", "l", "c"),
    col.names = c("Rank", "Book Title", "Author", "Predicted Rating")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

## Appendix C: Model Training Details

### C.1 MF Cross-Validation Folds

```{r appendix-mf-cv-details, include=FALSE}
cv_results_mf %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "MF Cross-Validation Results: Individual fold performance across 5-fold cross-validation showing RMSE for each fold",
    digits = 3,
    col.names = c("Fold", "RMSE"),
    align = c("c", "c")
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```

### C.2 Catalog Size Impact Analysis

```{r appendix-catalog-size-table, include=FALSE}
kable(
  size_impact_results %>% 
    select(catalog_size, rmse_mean, rmse_se, sparsity_mean),
  format = "latex",
  booktabs = TRUE,
  caption = "Catalog Size Impact Analysis: Detailed RMSE performance across different catalog sizes showing mean RMSE, standard error, and sparsity percentage",
  digits = 3,
  col.names = c("Catalog Size", "Mean RMSE", "SE", "Sparsity %"),
  align = c("c", "c", "c", "c")
) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 10
  )
```
