---
title: "Ensemble Recommender System for Book Recommendations: A Comparative Analysis of Collaborative Filtering Approaches"
author: "Hope Hennessy"
date: "2025-10-01"
output: pdf_document
---

\newpage

# Abstract

This study presents a comprehensive comparative analysis of four collaborative filtering approaches for book recommendation systems using a modified Book-Crossing dataset. We implement item-based collaborative filtering, user-based collaborative filtering, matrix factorization, and neural network-based methods to build an ensemble recommender system. Our analysis includes cross-validation performance evaluation, cold start problem handling, and investigation of dataset size effects on predictive accuracy. The results demonstrate the relative strengths and limitations of each approach, providing insights for practical recommendation system deployment.

## Assignment Overview

Build an ensemble recommender system for book recommendations using a modified "Book-Crossing" dataset containing ratings (0-10 scale) from 10,000 users on 150 books.

### Core Requirements

1. **Build Four Types of Recommender Systems:**
   - Item-based collaborative filtering (code from scratch)
   - User-based collaborative filtering (code from scratch)
   - Matrix factorization-based collaborative filtering
   - Neural network-based collaborative filtering

2. **System Capabilities:**
   - Recommend books to existing users
   - Handle new users (assuming they provide ratings for ≤5 books initially)

3. **Evaluation and Analysis:**
   - Compare accuracy across all four methods using cross-validation
   - Investigate the relationship between dataset size and accuracy
   - Determine if there's a point where adding more titles doesn't improve accuracy

4. **Data Analysis:**
   - Conduct exploratory data analysis (EDA)
   - Use findings to inform train/test data splitting

# 1. Introduction

## 1.1 Background

Recommender systems have become essential components of modern digital platforms, helping users discover relevant content from vast catalogs. Collaborative filtering approaches, which leverage user-item interaction patterns, remain among the most effective recommendation techniques. This study focuses on book recommendation systems, which face unique challenges including high sparsity, diverse user preferences, and the cold start problem for new users.

## 1.2 Objectives

The primary objectives of this research are:

1. **Implement Four Collaborative Filtering Methods**: Develop item-based, user-based, matrix factorization, and neural network-based recommendation systems
2. **Comparative Performance Analysis**: Evaluate and compare the accuracy of each method using cross-validation
3. **Cold Start Problem Investigation**: Develop strategies for handling new users with limited rating history
4. **Dataset Size Impact Analysis**: Examine how the number of available titles affects predictive accuracy
5. **Ensemble System Development**: Create a unified recommendation framework combining multiple approaches

# Data set up & EDA

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.width = 8, 
  fig.height = 6, 
  fig.align = "center", 
  warning = FALSE, 
  message = FALSE, 
  fig.show = 'hold', 
  out.width = '70%',
  dpi = 300)

set.seed(123)

# Load libraries
library(tidyverse)
library(patchwork)
library(caret)
library(kableExtra)
library(recosystem)
library(h2o)
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)


# Relax linting rules for academic work
options(lintr.linter_file = "none")  # Disable linting entirely
options(lintr.exclude_linters = c("object_usage_linter", "object_name_linter", 
                                  "cyclocomp_linter", "line_length_linter"))

```

Data Loading and Initial Exploration


```{r}
# Data Loading
load("book_ratings.Rdata")


head(book_info)
str(book_info)
dim(book_info)
head(book_ratings)
str(book_ratings)
dim(book_ratings)
head(user_info) # don't need Age to build recommender, but can include this info if want to go further
str(user_info)
dim(user_info)


# Check for missing values
missing_values <- data.frame(
  Dataset = c("Book Info", "Book Ratings", "User Info"),
  Missing_Values = c(
    sum(is.na(book_info)),
    sum(is.na(book_ratings)),
    sum(is.na(user_info))
  )
)

```

## Duplicate Book Detection & Handling

```{r}
# Initial data merging (but don't use 'data' yet as we need to handle duplicates first)
initial_data <- book_ratings %>%
  left_join(book_info, by = "ISBN")

# Duplicate Book Detection & Handling
# Simple duplicate detection by exact title and author match
book_info_clean <- book_info %>%
  mutate(
    clean_title = (Book.Title %>%
      tolower() %>%
      gsub('\\(.*?\\)', '', .) %>%
      gsub('\\s*:\\s*.*$', '', .) %>%
      gsub('^(the|a|an)\\s+', '', .) %>%
      gsub('[^a-z0-9\\s-]', '', .) %>%
      gsub('\\s+', ' ', .) %>%
      trimws()),
    
    clean_author = (Book.Author %>%
      tolower() %>%
      gsub('\\s+jr\\.*$', '', .) %>%
      gsub('\\s+sr\\.*$', '', .) %>%
      gsub('\\s+iii$', '', .) %>%
      gsub('\\s+ii$', '', .) %>%
      gsub('\\s+', ' ', .) %>%
      trimws())
  )

# Find duplicates by exact title and author match
duplicates <- book_info_clean %>%
  group_by(clean_title, clean_author) %>%
  filter(n() > 1) %>%
  arrange(clean_title, clean_author, ISBN) %>%
  ungroup()

# FIXED: Improved Merge Ratings Function
improved_merge_ratings <- function(ratings_data, book_info_data, duplicates_data) {
  
  if (nrow(duplicates_data) > 0) {
    # Create canonical mapping for duplicates
    duplicate_mapping <- duplicates_data %>%
      group_by(clean_title, clean_author) %>%
      arrange(ISBN) %>%
      mutate(canonical_ISBN = first(ISBN)) %>%
      ungroup() %>%
      select(ISBN, canonical_ISBN)
    
    # Create complete mapping (non-duplicates map to themselves)
    all_books_mapping <- book_info_data %>%
      select(ISBN) %>%
      left_join(duplicate_mapping, by = 'ISBN') %>%
      mutate(canonical_ISBN = ifelse(is.na(canonical_ISBN), ISBN, canonical_ISBN))
    
    # Apply mapping to ratings data
    ratings_mapped <- ratings_data %>%
      left_join(all_books_mapping, by = 'ISBN') %>%
      mutate(original_ISBN = ISBN,
             ISBN = canonical_ISBN) %>%
      select(-canonical_ISBN, -original_ISBN)
    
    # Keep highest rating per user-book combination
    ratings_merged <- ratings_mapped %>%
      group_by(User.ID, ISBN) %>%
      slice_max(Book.Rating, n = 1, with_ties = FALSE) %>%
      ungroup()
    
    # Update book_info to keep only canonical ISBNs
    book_info_updated <- book_info_data %>%
      left_join(all_books_mapping, by = 'ISBN') %>%
      filter(ISBN == canonical_ISBN) %>%
      select(ISBN, Book.Title, Book.Author)
    
  } else {
    ratings_merged <- ratings_data
    book_info_updated <- book_info_data
    all_books_mapping <- NULL
  }
  
  return(list(
    ratings_merged = ratings_merged,
    book_info_updated = book_info_updated,
    mapping = all_books_mapping
  ))
}

# Apply the merge strategy
if (nrow(duplicates) > 0) {
  cat('=== APPLYING DUPLICATE MERGE STRATEGY ===\n')
  cat('Found', nrow(duplicates), 'duplicate books\n')
  cat('Strategy: Keep highest rating per user for duplicate books\n\n')
  
  # Apply merging
  merged_result <- improved_merge_ratings(book_ratings, book_info, duplicates)
  
  # Statistics
  cat('MERGE RESULTS:\n')
  cat('- Original ratings:', nrow(book_ratings), '\n')
  cat('- Merged ratings:', nrow(merged_result$ratings_merged), '\n')
  cat('- Ratings removed:', nrow(book_ratings) - nrow(merged_result$ratings_merged), '\n')
  cat('- Original unique books:', length(unique(book_info$ISBN)), '\n')
  cat('- Final unique books:', length(unique(merged_result$book_info_updated$ISBN)), '\n\n')
  
  # Create final data object with required columns
  data <- merged_result$ratings_merged %>%
    left_join(merged_result$book_info_updated, by = "ISBN") %>%
    select(User.ID, ISBN, Book.Rating, Book.Title, Book.Author)
  
  # Update book_info for later use
  book_info <- merged_result$book_info_updated
  
} else {
  cat('No duplicates found - no merge needed.\n')
  
  # Create final data object with required columns
  data <- book_ratings %>%
    left_join(book_info, by = "ISBN") %>%
    select(User.ID, ISBN, Book.Rating, Book.Title, Book.Author)
}

# Verify final data structure
cat('\nFINAL DATA STRUCTURE:\n')
str(data)

cat('\nDATA SUMMARY:\n')
cat('- Total ratings:', nrow(data), '\n')
cat('- Unique users:', length(unique(data$User.ID)), '\n')
cat('- Unique books:', length(unique(data$ISBN)), '\n')
cat('- Rating range:', paste(range(data$Book.Rating), collapse = ' to '), '\n')

# Check for any remaining duplicates
remaining_duplicates <- data %>%
  group_by(User.ID, ISBN) %>%
  filter(n() > 1)

if (nrow(remaining_duplicates) > 0) {
  cat('\nWARNING: Still have duplicate user-book pairs!\n')
  print(remaining_duplicates)
} else {
  cat('\nVerification: No duplicate user-book pairs remain.\n')
}


```


```{r}
summary(data) # can clearly see age has some impossible outliers
head(data)
dim(data)

sapply(data, function(x) if(is.numeric(x)) range(x, na.rm = TRUE)) # check var ranges

# Check rating distribution
table(data$Book.Rating)

length(unique(data$User.ID))
length(data$User.ID)
```




```{r matrix-construction}
# -------------------------------------------------------------------
# Count ratings per book
# -------------------------------------------------------------------
counts_per_book <- data %>%
  group_by(ISBN) %>%
  summarise(num_ratings = n(), .groups = "drop")

# Plot distribution
ggplot(counts_per_book, aes(x = num_ratings)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  scale_x_continuous(limits = c(0, quantile(counts_per_book$num_ratings, 0.95))) +
  labs(title = "Distribution of Ratings per Book",
       x = "Number of Ratings",
       y = "Count of Books")


# -------------------------------------------------------------------
# Count ratings per user
# -------------------------------------------------------------------
counts_per_user <- data %>%
  group_by(User.ID) %>%
  summarise(num_ratings = n(), .groups = "drop")

users_per_count <- counts_per_user %>%
  count(num_ratings, name = "num_users")

# Plot distribution
ggplot(counts_per_user, aes(x = num_ratings)) +
  geom_histogram(binwidth = 1, fill = "lightgreen", color = "black") +
  scale_x_continuous(limits = c(0, quantile(counts_per_user$num_ratings, 0.95))) +
  labs(title = "Distribution of Ratings per User",
       x = "Number of Ratings",
       y = "Count of Users")

# Total unique users
length(unique(data$User.ID))
```


# User-Item Matrix Construction

```{r unified-matrix-creation}

# User-item matrix creation function with implicit rating handling
create_user_item_matrix <- function(ratings_data, 
                                    min_ratings_per_book = 3, 
                                    min_ratings_per_user = 2,
                                    implicit_rating = 4,
                                    implicit_confidence = 0.5) {
  
  # CHANGE: Don't convert 0 to NA anymore
  # Instead, convert 0 to implicit rating value
  ratings_clean <- ratings_data %>%
    mutate(
      # Track original zeros for confidence weighting
      is_implicit = (Book.Rating == 0),
      # Store original rating for evaluation purposes
      original_rating = Book.Rating,
      # Convert implicit (0) to assumed rating
      Book.Rating = ifelse(Book.Rating == 0, implicit_rating, Book.Rating),
      # Add confidence scores
      confidence = ifelse(is_implicit, implicit_confidence, 1.0)
    )
  
  # Create three matrices: ratings, confidence, and original ratings
  user_item_matrix <- ratings_clean %>%
    select(User.ID, ISBN, Book.Rating) %>%
    pivot_wider(names_from = ISBN, values_from = Book.Rating, values_fill = NA)
  
  confidence_matrix <- ratings_clean %>%
    select(User.ID, ISBN, confidence) %>%
    pivot_wider(names_from = ISBN, values_from = confidence, values_fill = NA)
  
  # Store original ratings for evaluation
  original_ratings_matrix <- ratings_clean %>%
    select(User.ID, ISBN, original_rating) %>%
    pivot_wider(names_from = ISBN, values_from = original_rating, values_fill = NA)
  
  # Convert to matrix format
  user_ids <- user_item_matrix$User.ID
  user_item_matrix <- as.matrix(user_item_matrix[, -1])
  confidence_matrix <- as.matrix(confidence_matrix[, -1])
  original_ratings_matrix <- as.matrix(original_ratings_matrix[, -1])
  rownames(user_item_matrix) <- user_ids
  rownames(confidence_matrix) <- user_ids
  rownames(original_ratings_matrix) <- user_ids
  
  # Filter books and users as before
  books_to_keep <- colSums(!is.na(user_item_matrix)) >= min_ratings_per_book
  user_item_matrix <- user_item_matrix[, books_to_keep]
  confidence_matrix <- confidence_matrix[, books_to_keep]
  original_ratings_matrix <- original_ratings_matrix[, books_to_keep]
  
  users_to_keep <- rowSums(!is.na(user_item_matrix)) >= min_ratings_per_user
  user_item_matrix <- user_item_matrix[users_to_keep, ]
  confidence_matrix <- confidence_matrix[users_to_keep, ]
  original_ratings_matrix <- original_ratings_matrix[users_to_keep, ]
  
  return(list(
    ratings = user_item_matrix,
    confidence = confidence_matrix,
    original_ratings = original_ratings_matrix
  ))
}

```


Evaluation function

```{r}
# Evaluation metrics function
# Modified evaluation to handle implicit ratings properly
evaluate_predictions <- function(pred, actual, original_ratings) {
  # Separate metrics for explicit and implicit
  is_implicit <- (original_ratings == 0)
  
  # RMSE for explicit ratings only
  explicit_mask <- !is_implicit & !is.na(actual)
  rmse_explicit <- sqrt(mean((pred[explicit_mask] - actual[explicit_mask])^2, na.rm = TRUE))
  
  # Overall RMSE (with imputed values)
  rmse_overall <- sqrt(mean((pred - actual)^2, na.rm = TRUE))
  
  return(list(
    rmse_explicit = rmse_explicit,
    rmse_overall = rmse_overall,
    n_explicit = sum(explicit_mask),
    n_implicit = sum(is_implicit)
  ))
}

```


# 4. Collaborative Filtering Methods Implementation

## 4.1 User-Based Collaborative Filtering (UBCF)


```{r ubcf-functions}

# User-mean normalization function
normalize_matrix_user <- function(user_item_matrix) {
  
  # Center ratings by subtracting user mean
  user_means <- rowMeans(user_item_matrix, na.rm = TRUE)
  user_item_matrix_normalized <- sweep(user_item_matrix, 1, user_means, FUN = "-")
  
  return(list(normalized = user_item_matrix_normalized, means = user_means))
}
```


```{r}
# Weighted cosine similarity matrix computation for users
compute_user_similarity_matrix <- function(user_item_matrix_normalized, confidence_matrix) {
  
  n_users <- nrow(user_item_matrix_normalized)
  
  # Weight the normalized ratings by confidence
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  
  # Replace NA with 0 for computation
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  
  # Weighted ratings (rating × confidence)
  weighted_mat <- mat * conf
  
  # Weighted dot products (incorporating confidence)
  numerator <- weighted_mat %*% t(weighted_mat)
  
  # Weighted magnitudes
  magnitudes <- sqrt(rowSums(weighted_mat^2))
  denominator <- outer(magnitudes, magnitudes)
  
  # Cosine similarity
  user_similarity_matrix <- numerator / denominator
  user_similarity_matrix[is.nan(user_similarity_matrix)] <- 0
  diag(user_similarity_matrix) <- 0
  
  rownames(user_similarity_matrix) <- rownames(user_item_matrix_normalized)
  colnames(user_similarity_matrix) <- rownames(user_item_matrix_normalized)
  
  return(user_similarity_matrix)
}
```

```{r ubcf-recommendation-functions}
# Recommendation function for existing users (UBCF)
recommend_for_user_ubcf <- function(target_user, user_item_matrix, 
                                    user_item_matrix_normalized, 
                                    confidence_matrix,
                                    user_sim_matrix,
                                    user_means, book_info, 
                                    n_recommendations = 10, 
                                    k = NULL) {
  
  target_user <- as.character(target_user)
  
  # Identify truly unrated books (NA values)
  unrated_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  
  if (length(unrated_books) == 0) {
    return(data.frame())
  }
  
  # Get user similarities
  sims <- user_sim_matrix[target_user, ]
  sims[is.na(sims) | is.nan(sims)] <- 0
  
  # k-NN filtering
  if (!is.null(k) && k < length(sims)) {
    non_zero_count <- sum(sims != 0)
    if (non_zero_count > 0) {
      k_actual <- min(k, non_zero_count)
      top_k_users <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
      sims_filtered <- rep(0, length(sims))
      names(sims_filtered) <- names(sims)
      sims_filtered[top_k_users] <- sims[top_k_users]
      sims <- sims_filtered
    }
  }
  
  if (sum(abs(sims) > 0) == 0) {
    return(data.frame())
  }
  
  # Weight predictions by confidence
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  
  # Weight ratings by confidence when making predictions
  weighted_mat <- mat * conf
  
  # Predict for unrated books
  weighted_ratings <- t(weighted_mat[, unrated_books, drop = FALSE]) %*% sims
  
  # VECTORIZED: Calculate confidence-weighted similarity sums
  # Create mask for which users have rated each unrated book
  rated_mask <- !is.na(confidence_matrix[, unrated_books, drop = FALSE])
  
  # Confidence-weighted similarities for normalization
  conf_weighted_sims <- sweep(conf[, unrated_books, drop = FALSE] * rated_mask, 
                              1, abs(sims), "*")
  sum_weighted_sims <- colSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  # Calculate predictions
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + user_means[target_user]
  
  # Convert to vector and clip
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Get top N recommendations
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) {
    return(data.frame())
  }
  
  top_books <- sort(preds_valid, decreasing = TRUE)[1:min(n_recommendations, length(preds_valid))]
  
  recommendations <- data.frame(
    ISBN = names(top_books),
    Predicted_Rating = as.numeric(top_books),
    stringsAsFactors = FALSE
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```

```{r}
# Improved recommendation function for new users (cold start - UBCF)
recommend_for_new_user_ubcf <- function(new_user_ratings, 
                                        user_item_matrix, 
                                        user_item_matrix_normalized,
                                        confidence_matrix,
                                        user_means,
                                        book_info, 
                                        n_recommendations = 10, 
                                        k = NULL) {
  
  # Input validation
  if (length(new_user_ratings) == 0) {
    return(data.frame())
  }
  
  # Create new user vector aligned with existing matrix
  new_user_vector <- rep(NA, ncol(user_item_matrix))
  names(new_user_vector) <- colnames(user_item_matrix)
  
  # Check for valid books
  matched_books <- intersect(names(new_user_ratings), names(new_user_vector))
  if (length(matched_books) == 0) {
    return(data.frame())
  }
  
  # Fill in ratings (all explicit with confidence = 1.0)
  new_user_vector[matched_books] <- new_user_ratings[matched_books]
  
  # User-mean normalization
  new_user_mean <- mean(new_user_vector, na.rm = TRUE)
  new_user_normalized <- new_user_vector - new_user_mean
  new_user_normalized[is.na(new_user_normalized)] <- 0
  
  # Since new user ratings are all explicit, confidence = 1.0 for rated items
  new_user_conf <- ifelse(is.na(new_user_vector), 0, 1)
  
  # Weight the new user's normalized ratings by confidence
  new_user_weighted <- new_user_normalized * new_user_conf
  
  # Prepare existing users' matrices
  mat_normalized <- user_item_matrix_normalized
  mat_normalized[is.na(mat_normalized)] <- 0
  conf <- confidence_matrix
  conf[is.na(conf)] <- 0
  
  # Weight existing users' ratings by their confidence
  mat_weighted <- mat_normalized * conf
  
  # Compute similarities with all existing users
  # Using vectorized operations instead of loops
  existing_magnitudes <- sqrt(rowSums(mat_weighted^2))
  new_user_magnitude <- sqrt(sum(new_user_weighted^2))
  
  # Check for zero magnitude
  if (new_user_magnitude == 0) {
    return(data.frame())
  }
  
  # Calculate cosine similarities
  new_user_sims <- as.vector((mat_weighted %*% new_user_weighted) / 
                             (existing_magnitudes * new_user_magnitude))
  
  # Handle special values
  new_user_sims[!is.finite(new_user_sims)] <- 0
  names(new_user_sims) <- rownames(user_item_matrix_normalized)
  
  # Apply k-NN filtering if specified
  if (!is.null(k) && k > 0 && k < length(new_user_sims)) {
    # Keep only top k most similar users
    threshold <- sort(abs(new_user_sims), decreasing = TRUE)[min(k, sum(new_user_sims != 0))]
    new_user_sims[abs(new_user_sims) < threshold] <- 0
  }
  
  # Check if we have similar users
  if (all(new_user_sims == 0)) {
    return(data.frame())
  }
  
  # Get unrated books
  unrated_books <- names(new_user_vector)[is.na(new_user_vector)]
  
  if (length(unrated_books) == 0) {
    return(data.frame())
  }
  
  # Make predictions for unrated books
  # Vectorized computation for all unrated books at once
  weighted_ratings <- as.vector(t(mat_weighted[, unrated_books, drop = FALSE]) %*% new_user_sims)
  
  # Calculate normalization factors
  # Only count users who rated each book, weighted by confidence
  rated_mask <- !is.na(user_item_matrix[, unrated_books, drop = FALSE])
  conf_for_unrated <- conf[, unrated_books, drop = FALSE] * rated_mask
  
  # Sum of confidence-weighted similarities
  sum_weighted_sims <- as.vector(t(conf_for_unrated) %*% abs(new_user_sims))
  sum_weighted_sims[sum_weighted_sims == 0] <- 1  # Avoid division by zero
  
  # Calculate final predictions
  preds <- (weighted_ratings / sum_weighted_sims) + new_user_mean
  
  # Clip to valid range
  preds <- pmin(pmax(preds, 1), 10)
  names(preds) <- unrated_books
  
  # Remove NA predictions and get top N
  preds_valid <- preds[!is.na(preds)]
  
  if (length(preds_valid) == 0) {
    return(data.frame())
  }
  
  # Select top recommendations
  n_actual <- min(n_recommendations, length(preds_valid))
  top_indices <- order(preds_valid, decreasing = TRUE)[1:n_actual]
  
  # Create output dataframe
  recommendations <- data.frame(
    ISBN = names(preds_valid)[top_indices],
    Predicted_Rating = round(preds_valid[top_indices], 2),
    stringsAsFactors = FALSE
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```


```{r ubcf-implementation}
# ================================================================
# UBCF IMPLEMENTATION
# ================================================================


# Create matrices with implicit feedback handling
matrix_result <- create_user_item_matrix(
  data,
  min_ratings_per_book = 5,  
  min_ratings_per_user = 3,
  implicit_rating = 4,
  implicit_confidence = 0.5
)

user_item_matrix_ubcf <- matrix_result$ratings
confidence_matrix_ubcf <- matrix_result$confidence
original_ratings_matrix_ubcf <- matrix_result$original_ratings

# Normalize matrix (user-mean)
normalized_result_ubcf <- normalize_matrix_user(user_item_matrix_ubcf)
user_item_matrix_normalized_ubcf <- normalized_result_ubcf$normalized
user_means <- normalized_result_ubcf$means

# Compute similarity with confidence weighting
user_similarity_matrix <- compute_user_similarity_matrix(
  user_item_matrix_normalized_ubcf, 
  confidence_matrix_ubcf
)

# Make recommendations (now confidence-aware)
sample_user <- rownames(user_item_matrix_ubcf)[3]

recs_ubcf <- recommend_for_user_ubcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix_ubcf,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix_ubcf,  # NEW parameter
  user_sim_matrix = user_similarity_matrix,
  user_means = user_means,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
) 

recs_ubcf %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(caption = "Top 10 Recommendations for Existing User (UBCF)", 
        digits = 2, align = c("c", "l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

# Recommendations for new user (cold start)
sample_books <- colnames(user_item_matrix_ubcf)[1:5]
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), sample_books)

# Display new user's ratings
new_user_ratings_df <- data.frame(
  ISBN = names(new_user_ratings),
  Rating = as.numeric(new_user_ratings)) %>%
  left_join(book_info, by = "ISBN") %>%
  select(Book.Title, Book.Author, Rating)

new_user_ratings_df %>%
  kable(caption = "New User's Initial Ratings", 
        digits = 0, align = c("l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

new_user_recs <- recommend_for_new_user_ubcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix_ubcf,
  user_item_matrix_normalized = user_item_matrix_normalized_ubcf,
  confidence_matrix = confidence_matrix_ubcf,  # Pass confidence matrix
  user_means = user_means,
  book_info = book_info,
  n_recommendations = 10,
  k = 50
)

# Display recommendations
new_user_recs %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(caption = "Top 10 Recommendations for New User (UBCF with Confidence)", 
        digits = 2, align = c("c", "l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)
```


```{r cross-validation-ubcf}
# =====================================
# CROSS-VALIDATION FOR USER-BASED CF 
# =====================================

cross_validate_ubcf <- function(user_item_matrix, confidence_matrix, original_ratings_matrix, n_folds = 5, k = 30) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  # Create folds
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  # Pre-allocate results dataframe
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds))
  

  for (fold in 1:n_folds) {
    cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    test_obs <- observed[test_indices, , drop = FALSE]
    
    # Create train matrices (both ratings and confidence)
    train_matrix <- user_item_matrix
    train_confidence <- confidence_matrix
    train_matrix[test_obs] <- NA
    train_confidence[test_obs] <- NA
    
    # Normalize (user-mean)
    norm_result <- normalize_matrix_user(train_matrix)
    train_normalized <- norm_result$normalized
    user_means <- norm_result$means
    
    # Compute user similarity with confidence weighting
    user_sim_matrix <- compute_user_similarity_matrix(train_normalized, train_confidence)
    
    # Prepare matrices for prediction
    train_norm_filled <- train_normalized
    train_norm_filled[is.na(train_norm_filled)] <- 0
    train_conf_filled <- train_confidence
    train_conf_filled[is.na(train_conf_filled)] <- 0
    
    # Make predictions for test set
    predictions <- numeric(nrow(test_obs))
    
    for (i in 1:nrow(test_obs)) {
      user_idx <- test_obs[i, 1]
      item_idx <- test_obs[i, 2]
      
      # Get similarities to all other users
      sims <- user_sim_matrix[user_idx, ]
      sims[is.na(sims) | is.nan(sims)] <- 0
      
      # Apply k-NN filtering
      if (k < length(sims) && sum(abs(sims) > 0) > 0) {
        k_actual <- min(k, sum(abs(sims) > 0))
        top_k_idx <- order(abs(sims), decreasing = TRUE)[1:k_actual]
        sims_filtered <- rep(0, length(sims))
        sims_filtered[top_k_idx] <- sims[top_k_idx]
        sims <- sims_filtered
      }
      
      # Find which users (among similar ones) rated this item
      rated_mask <- !is.na(train_matrix[, item_idx])
      
      # If no similar user rated this item, use user mean
      if (sum(abs(sims[rated_mask])) == 0) {
        predictions[i] <- user_means[user_idx]
        next
      }
      
      # Confidence-weighted prediction
      # Weight ratings by confidence
      weighted_ratings <- train_norm_filled[rated_mask, item_idx] * train_conf_filled[rated_mask, item_idx]
      
      # Weight similarities by confidence
      weighted_sims <- abs(sims[rated_mask]) * train_conf_filled[rated_mask, item_idx]
      
      # Weighted average
      weighted_sum <- sum(sims[rated_mask] * weighted_ratings)
      sum_weighted_sims <- sum(weighted_sims)
      
      if (sum_weighted_sims > 0) {
        predictions[i] <- (weighted_sum / sum_weighted_sims) + user_means[user_idx]
      } else {
        predictions[i] <- user_means[user_idx]
      }
    }
    
    # Clip to valid range
    predictions <- pmin(pmax(predictions, 1), 10)
    
    # Get actual ratings (converted ratings)
    actual <- user_item_matrix[test_obs]
    
    # Get original ratings for proper evaluation
    original_ratings <- original_ratings_matrix[test_obs]
    metrics <- evaluate_predictions(predictions, actual, original_ratings)
    
    cv_results[fold, ] <- list(fold, metrics$rmse_explicit)  # Use explicit-only RMSE
    
  }
  
  return(cv_results)
}

# Run cross-validation
cv_results_ubcf <- cross_validate_ubcf(
  user_item_matrix_ubcf, 
  confidence_matrix_ubcf,
  original_ratings_matrix_ubcf,
  n_folds = 5, 
  k = 50)

# Display results
cv_results_ubcf %>%
  kable(caption = "UBCF Cross-Validation Results", 
        digits = 4, 
        col.names = c("Fold", "RMSE"),
        align = c("c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)
```




## 4.2 Item-Based Collaborative Filtering (IBCF)

Item-based collaborative filtering identifies items with similar rating patterns and recommends items similar to those a user has rated highly. This approach is more stable than user-based methods as item preferences change less frequently than user preferences.

```{r ibcf-functions}

# Item-mean normalization function
normalize_matrix_item <- function(user_item_matrix) {
  
  # Center ratings by subtracting item mean
  item_means <- colMeans(user_item_matrix, na.rm = TRUE)
  user_item_matrix_normalized <- sweep(user_item_matrix, 2, item_means, FUN = "-")
  
  return(list(normalized = user_item_matrix_normalized, means = item_means))
}


# Weighted cosine similarity matrix computation for items
compute_item_similarity_matrix <- function(user_item_matrix_normalized, confidence_matrix) {
  
  # Replace NA with 0 for matrix operations
  mat <- user_item_matrix_normalized
  mat[is.na(mat)] <- 0
  
  conf <- confidence_matrix
  conf[is.na(conf)] <- 0
  
  # Transpose to work with items as rows (not users)
  mat_t <- t(mat)
  conf_t <- t(conf)
  
  # Weight the ratings by confidence
  weighted_mat_t <- mat_t * conf_t
  
  # Weighted dot products (numerator)
  numerator <- weighted_mat_t %*% t(weighted_mat_t)
  
  # Weighted magnitudes (denominator)
  magnitudes <- sqrt(rowSums(weighted_mat_t^2))
  denominator <- outer(magnitudes, magnitudes)
  
  # Cosine similarity calculation
  item_similarity_matrix <- numerator / denominator
  
  # Replace NaN values with 0
  item_similarity_matrix[is.nan(item_similarity_matrix)] <- 0
  
  # Set self-similarity to 0
  diag(item_similarity_matrix) <- 0

  rownames(item_similarity_matrix) <- colnames(user_item_matrix_normalized)
  colnames(item_similarity_matrix) <- colnames(user_item_matrix_normalized)
  
  return(item_similarity_matrix)
}
```


```{r ibcf-recommendation-functions}
# Recommendation function for existing users (IBCF) - Final Optimized
recommend_for_user_ibcf <- function(target_user, user_item_matrix,
                                    user_item_matrix_normalized, 
                                    confidence_matrix,
                                    item_sim_matrix,
                                    item_means, book_info, 
                                    n_recommendations = 10,
                                    k = NULL) {
  
  target_user <- as.character(target_user)
  
  # Identify truly unrated books (NA values)
  unrated_books <- colnames(user_item_matrix)[is.na(user_item_matrix[target_user, ])]
  
  if (length(unrated_books) == 0) {
    return(data.frame())
  }
  
  # Get item similarities for unrated books
  item_sims <- item_sim_matrix[unrated_books, , drop = FALSE]
  item_sims[is.na(item_sims) | is.nan(item_sims)] <- 0
  
  # k-NN filtering
  if (!is.null(k) && k < ncol(item_sims)) {
    for (i in 1:nrow(item_sims)) {
      sims <- item_sims[i, ]
      non_zero_count <- sum(sims != 0)
      if (non_zero_count > 0) {
        k_actual <- min(k, non_zero_count)
        top_k_items <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
        sims_filtered <- rep(0, length(sims))
        names(sims_filtered) <- names(sims)
        sims_filtered[top_k_items] <- sims[top_k_items]
        item_sims[i, ] <- sims_filtered
      }
    }
  }
  
  if (sum(abs(item_sims) > 0) == 0) {
    return(data.frame())
  }
  
  # Weight predictions by confidence
  mat <- user_item_matrix_normalized
  conf <- confidence_matrix
  mat[is.na(mat)] <- 0
  conf[is.na(conf)] <- 0
  
  # Weight ratings by confidence when making predictions
  weighted_mat <- mat * conf
  
  # Predict for unrated books
  weighted_ratings <- item_sims %*% weighted_mat[target_user, ]
  
  # VECTORIZED: Calculate confidence-weighted similarity sums
  rated_mask <- !is.na(user_item_matrix[target_user, ])
  rated_item_mask <- matrix(rated_mask, nrow = nrow(item_sims), 
                            ncol = length(rated_mask), byrow = TRUE)
  
  # Confidence-weighted similarities for normalization
  conf_weighted_sims <- rated_item_mask * abs(item_sims) * conf[target_user, ]
  sum_weighted_sims <- rowSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  # Calculate predictions
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + item_means[unrated_books]
  
  # Convert to vector and clip
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Get top N recommendations
  preds_valid <- preds[!is.na(preds)]
  if (length(preds_valid) == 0) {
    return(data.frame())
  }
  
  top_books <- sort(preds_valid, decreasing = TRUE)[1:min(n_recommendations, length(preds_valid))]
  
  recommendations <- data.frame(
    ISBN = names(top_books),
    Predicted_Rating = as.numeric(top_books),
    stringsAsFactors = FALSE
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```


```{r}
# Recommendation function for new users (cold start - IBCF) - Final Optimized
recommend_for_new_user_ibcf <- function(new_user_ratings, 
                                        user_item_matrix,
                                        user_item_matrix_normalized, 
                                        confidence_matrix,
                                        item_means,
                                        item_sim_matrix, 
                                        book_info,
                                        n_recommendations = 10, 
                                        k = NULL) {
  
  # Input validation
  if (length(new_user_ratings) == 0) {
    return(data.frame())
  }
  
  # Create new user vector aligned with existing matrix
  new_user_vector <- rep(NA, ncol(user_item_matrix))
  names(new_user_vector) <- colnames(user_item_matrix)
  
  # Check for valid books
  matched_books <- intersect(names(new_user_ratings), names(new_user_vector))
  if (length(matched_books) == 0) {
    return(data.frame())
  }
  
  # Fill in ratings (all explicit with confidence = 1.0)
  new_user_vector[matched_books] <- new_user_ratings[matched_books]
  
  # Item-mean normalization
  new_user_normalized <- new_user_vector - item_means
  new_user_normalized[is.na(new_user_normalized)] <- 0
  
  # Since new user ratings are all explicit, confidence = 1.0 for rated items
  new_user_conf <- ifelse(is.na(new_user_vector), 0, 1)
  
  # Weight the new user's normalized ratings by confidence
  new_user_weighted <- new_user_normalized * new_user_conf
  
  # Get unrated books
  unrated_books <- names(new_user_vector)[is.na(new_user_vector)]
  
  if (length(unrated_books) == 0) {
    return(data.frame())
  }
  
  # Get item similarities for unrated books
  item_sims <- item_sim_matrix[unrated_books, , drop = FALSE]
  item_sims[is.na(item_sims) | is.nan(item_sims)] <- 0
  
  # Apply k-NN filtering if specified
  if (!is.null(k) && k < ncol(item_sims)) {
    for (i in 1:nrow(item_sims)) {
      sims <- item_sims[i, ]
      non_zero_count <- sum(sims != 0)
      if (non_zero_count > 0) {
        k_actual <- min(k, non_zero_count)
        top_k_items <- names(sort(abs(sims), decreasing = TRUE)[1:k_actual])
        sims_filtered <- rep(0, length(sims))
        names(sims_filtered) <- names(sims)
        sims_filtered[top_k_items] <- sims[top_k_items]
        item_sims[i, ] <- sims_filtered
      }
    }
  }
  
  # Check if we have similar items
  if (sum(abs(item_sims) > 0) == 0) {
    return(data.frame())
  }
  
  # Make predictions for unrated books (vectorized)
  weighted_ratings <- item_sims %*% new_user_weighted
  
  # Calculate normalization factors
  rated_mask <- !is.na(new_user_vector)
  rated_item_mask <- matrix(rated_mask, nrow = nrow(item_sims), 
                            ncol = length(rated_mask), byrow = TRUE)
  
  # Confidence-weighted similarities
  conf_weighted_sims <- rated_item_mask * abs(item_sims) * new_user_conf
  sum_weighted_sims <- rowSums(conf_weighted_sims)
  sum_weighted_sims[sum_weighted_sims == 0] <- 1
  
  # Calculate final predictions
  preds <- weighted_ratings / sum_weighted_sims
  preds[is.nan(preds)] <- NA
  preds <- preds + item_means[unrated_books]
  
  # Convert to vector and clip
  preds <- as.vector(preds)
  names(preds) <- unrated_books
  preds <- pmin(pmax(preds, 1), 10)
  
  # Remove NA predictions and get top N
  preds_valid <- preds[!is.na(preds)]
  
  if (length(preds_valid) == 0) {
    return(data.frame())
  }
  
  # Select top recommendations
  n_actual <- min(n_recommendations, length(preds_valid))
  top_indices <- order(preds_valid, decreasing = TRUE)[1:n_actual]
  
  # Create output dataframe
  recommendations <- data.frame(
    ISBN = names(preds_valid)[top_indices],
    Predicted_Rating = round(preds_valid[top_indices], 2),
    stringsAsFactors = FALSE
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```


```{r ibcf-implementation}
# ================================================================
# IBCF IMPLEMENTATION - OPTIMIZED VERSION
# ================================================================
# Key optimizations: Vectorized predictions, enhanced confidence weighting, 
# improved k-NN filtering, robust cross-validation
# ================================================================

# Create matrices with implicit feedback handling for IBCF
matrix_result_ibcf <- create_user_item_matrix(
  data,
  min_ratings_per_book = 5,  
  min_ratings_per_user = 3,
  implicit_rating = 4,        # Treat implicit as "slightly below neutral"
  implicit_confidence = 0.5    # Lower confidence for imputed values
)

user_item_matrix_ibcf <- matrix_result_ibcf$ratings
confidence_matrix_ibcf <- matrix_result_ibcf$confidence
original_ratings_matrix_ibcf <- matrix_result_ibcf$original_ratings

# Normalize matrix (item-mean)
normalized_result_ibcf <- normalize_matrix_item(user_item_matrix_ibcf)
user_item_matrix_normalized_ibcf <- normalized_result_ibcf$normalized
item_means <- normalized_result_ibcf$means

# Compute item-item similarity matrix with confidence weighting
item_similarity_matrix <- compute_item_similarity_matrix(
  user_item_matrix_normalized_ibcf, 
  confidence_matrix_ibcf
)

# Recommendations for existing user
sample_user <- rownames(user_item_matrix_ibcf)[3]

recs_ibcf <- recommend_for_user_ibcf(
  target_user = sample_user,
  user_item_matrix = user_item_matrix_ibcf,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix_ibcf,
  item_sim_matrix = item_similarity_matrix,
  item_means = item_means,
  book_info = book_info,
  n_recommendations = 10,
  k = 50)

recs_ibcf %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(caption = "Top 10 Recommendations for Existing User (IBCF)", 
        digits = 2, align = c("c", "l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

# Recommendations for new user (cold start)
sample_books <- colnames(user_item_matrix_ibcf)[1:5]
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), sample_books)

# Display new user's ratings
new_user_ratings_df <- data.frame(
  ISBN = names(new_user_ratings),
  Rating = as.numeric(new_user_ratings)) %>%
  left_join(book_info, by = "ISBN") %>%
  select(Book.Title, Book.Author, Rating)

new_user_ratings_df %>%
  kable(caption = "New User's Initial Ratings", 
        digits = 0, align = c("l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

new_user_recs_ibcf <- recommend_for_new_user_ibcf(
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix_ibcf,
  user_item_matrix_normalized = user_item_matrix_normalized_ibcf,
  confidence_matrix = confidence_matrix_ibcf,
  item_means = item_means,
  item_sim_matrix = item_similarity_matrix,
  book_info = book_info,
  n_recommendations = 10,
  k = 50)  

new_user_recs_ibcf %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(caption = "Top 10 Recommendations for New User (IBCF)", 
        digits = 2, align = c("c", "l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)
```


```{r cross-validation-ibcf}
# ================================================================
# CROSS-VALIDATION FOR ITEM-BASED CF
# ================================================================

cross_validate_ibcf <- function(user_item_matrix, confidence_matrix, original_ratings_matrix, n_folds = 5, k = 30) {
  
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  # Create folds
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  # Pre-allocate results dataframe
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds))
  
  for (fold in 1:n_folds) {
    cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    test_obs <- observed[test_indices, , drop = FALSE]
    
    # Create train matrices (both ratings and confidence)
    train_matrix <- user_item_matrix
    train_confidence <- confidence_matrix
    train_matrix[test_obs] <- NA
    train_confidence[test_obs] <- NA
    
    # Normalize (item-mean)
    norm_result <- normalize_matrix_item(train_matrix)
    train_normalized <- norm_result$normalized
    item_means <- norm_result$means
    
    # Compute item similarity with confidence weighting
    item_sim_matrix <- compute_item_similarity_matrix(train_normalized, train_confidence)
    
    # Prepare matrices for prediction
    train_norm_filled <- train_normalized
    train_norm_filled[is.na(train_norm_filled)] <- 0
    train_conf_filled <- train_confidence
    train_conf_filled[is.na(train_conf_filled)] <- 0
    
    # Make predictions for test set
    predictions <- numeric(nrow(test_obs))
    
    for (i in 1:nrow(test_obs)) {
      user_idx <- test_obs[i, 1]
      item_idx <- test_obs[i, 2]
      
      # Get user's rated items in training set
      rated_items <- which(!is.na(train_matrix[user_idx, ]))
      
      if (length(rated_items) == 0) {
        predictions[i] <- item_means[item_idx]
        next
      }
      
      # Get similarities between target item and user's rated items
      sims <- item_sim_matrix[item_idx, rated_items]
      sims[is.na(sims) | is.nan(sims)] <- 0
      
      # IMPROVED: Better k-NN filtering (similar to UBCF)
      if (sum(abs(sims) > 0) > 0 && k < length(sims)) {
        k_actual <- min(k, sum(abs(sims) > 0))
        top_k_idx <- order(abs(sims), decreasing = TRUE)[1:k_actual]
        sims_filtered <- rep(0, length(sims))
        sims_filtered[top_k_idx] <- sims[top_k_idx]
        sims <- sims_filtered
      }
      
      # IMPROVED: Better confidence-weighted prediction (similar to UBCF)
      if (sum(abs(sims)) > 0) {
        # Weight ratings by confidence
        weighted_ratings <- train_norm_filled[user_idx, rated_items] * train_conf_filled[user_idx, rated_items]
        
        # Weight similarities by confidence
        weighted_sims <- abs(sims) * train_conf_filled[user_idx, rated_items]
        
        # Weighted average
        weighted_sum <- sum(sims * weighted_ratings)
        sum_weighted_sims <- sum(weighted_sims)
        
        if (sum_weighted_sims > 0) {
          predictions[i] <- (weighted_sum / sum_weighted_sims) + item_means[item_idx]
        } else {
          predictions[i] <- item_means[item_idx]
        }
      } else {
        predictions[i] <- item_means[item_idx]
      }
    }
    
    # Clip to valid range
    predictions <- pmin(pmax(predictions, 1), 10)
    
    # Get actual ratings
    actual <- user_item_matrix[test_obs]
    
    # Calculate metrics with original ratings for implicit/explicit separation
    original_ratings <- original_ratings_matrix[test_obs]  # These are the true original ratings
    metrics <- evaluate_predictions(predictions, actual, original_ratings)
    
    cv_results[fold, ] <- list(fold, metrics$rmse_explicit)  # Use explicit-only RMSE
    
  }
  
  return(cv_results)
}

# Run cross-validation
cv_results_ibcf <- cross_validate_ibcf(
  user_item_matrix_ibcf, 
  confidence_matrix_ibcf,
  original_ratings_matrix_ibcf,
  n_folds = 5, 
  k = 50)

# Display results
cv_results_ibcf %>%
  kable(caption = "IBCF Cross-Validation Results", 
        digits = 4, 
        col.names = c("Fold", "RMSE"),
        align = c("c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)
```



## 4.3 Matrix Factorization-Based Collaborative Filtering

Matrix factorization decomposes the user-item rating matrix into lower-dimensional user and item factor matrices. This approach captures latent factors that explain user preferences and item characteristics, enabling more accurate predictions even with sparse data.

```{r mf-data-preparation}
# ================================================================
# DATA PREPARATION FOR RECOSYSTEM (EXPLICIT RATINGS ONLY)
# ================================================================

# Create matrix with only explicit ratings for Matrix Factorization
create_explicit_matrix <- function(ratings_data, 
                                   min_ratings_per_book = 3, 
                                   min_ratings_per_user = 2) {
  
  # Filter out implicit ratings (0s) - only keep explicit ratings
  explicit_ratings <- ratings_data %>%
    filter(Book.Rating > 0) %>%  # Only explicit ratings
    select(User.ID, ISBN, Book.Rating)
  
  # Create user-item matrix
  user_item_matrix <- explicit_ratings %>%
    pivot_wider(names_from = ISBN, values_from = Book.Rating, values_fill = NA)
  
  # Convert to matrix format
  user_ids <- user_item_matrix$User.ID
  user_item_matrix <- as.matrix(user_item_matrix[, -1])
  rownames(user_item_matrix) <- user_ids
  
  # Filter books and users
  books_to_keep <- colSums(!is.na(user_item_matrix)) >= min_ratings_per_book
  user_item_matrix <- user_item_matrix[, books_to_keep]
  
  users_to_keep <- rowSums(!is.na(user_item_matrix)) >= min_ratings_per_user
  user_item_matrix <- user_item_matrix[users_to_keep, ]
  
  return(user_item_matrix)
}

prepare_recosystem_data <- function(user_item_matrix, original_ratings_matrix) {
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  
  # Get all ratings
  train_data <- data.frame(
    user_index = observed[,1] - 1,   # 0-based indexing
    item_index = observed[,2] - 1,
    rating = user_item_matrix[observed],
    original_rating = original_ratings_matrix[observed]
  )
  
  # FILTER TO ONLY EXPLICIT RATINGS (original_rating != 0)
  train_data_explicit <- train_data[train_data$original_rating != 0, ]
  
  user_ids <- data.frame(
    user_index = 0:(nrow(user_item_matrix) - 1),
    user_id = rownames(user_item_matrix)
  )
  
  item_ids <- data.frame(
    item_index = 0:(ncol(user_item_matrix) - 1),
    item_id = colnames(user_item_matrix)
  )
  
  list(
    train_data = train_data_explicit,
    user_ids = user_ids,
    item_ids = item_ids,
    n_users = nrow(user_item_matrix),
    n_items = ncol(user_item_matrix)
  )
}
```


```{r mf-tuning}
# ================================================================
# HYPERPARAMETER TUNING
# ================================================================

tune_reco_model <- function(train_data,
                            n_factors = c(10, 20, 30),
                            learning_rate = c(0.1, 0.05, 0.01),
                            costp_l2 = c(0.01, 0.1),
                            costq_l2 = c(0.01, 0.1),
                            n_iter = 50,
                            verbose = TRUE) {
  
  train_set <- data_memory(
    user_index = train_data$user_index,
    item_index = train_data$item_index,
    rating = train_data$rating,
    index1 = FALSE
  )
  
  # Create model
  rs <- Reco()
  
  # Tune hyperparameters
  opts <- rs$tune(train_set, opts = list(
    dim = n_factors,
    lrate = learning_rate,
    costp_l2 = costp_l2,
    costq_l2 = costq_l2,
    niter = n_iter,
    nthread = 4,
    verbose = verbose
  ))
  
  return(opts)
}
```


```{r mf-training}
# ================================================================
# MODEL TRAINING 
# ================================================================

train_reco_model <- function(train_data, 
                             n_factors = 20,
                             learning_rate = 0.1,
                             costp_l2 = 0.01,
                             costq_l2 = 0.01,
                             n_iter = 100,
                             n_threads = 4,
                             verbose = TRUE) {
  
  train_set <- data_memory(
    user_index = train_data$user_index,
    item_index = train_data$item_index,
    rating = train_data$rating,
    index1 = FALSE
  )
  
  model <- Reco()
  model$train(train_set, opts = list(
    dim = n_factors,
    lrate = learning_rate,
    costp_l2 = costp_l2,
    costq_l2 = costq_l2,
    niter = n_iter,
    nthread = n_threads,
    verbose = verbose
  ))
  
  return(model)
}
```


```{r mf-recommendation-functions}
# ================================================================
# RECOMMENDATION FUNCTIONS
# ================================================================

# Recommend for existing users
recommend_for_user_mf <- function(model, user_item_matrix, user_id, n_recommendations = 10,
                                  user_ids_map, item_ids_map, book_info) {
  
  if (!user_id %in% user_ids_map$user_id) stop("User not found")
  
  user_idx <- user_ids_map$user_index[user_ids_map$user_id == user_id]
  user_row <- user_item_matrix[as.character(user_id), ]
  unrated_items <- which(is.na(user_row))
  if (length(unrated_items) == 0) return(data.frame())
  
  item_indices <- unrated_items - 1
  pred_set <- data_memory(user_index = rep(user_idx, length(item_indices)),
                          item_index = item_indices,
                          index1 = FALSE)
  pred_ratings <- model$predict(pred_set, out_memory())
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)  # Clip to valid range
  
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  recommended_items <- colnames(user_item_matrix)[unrated_items[top_indices]]
  
  data.frame(
    ISBN = recommended_items,
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
}

# Recommend for new users (cold start) - Hybrid approach
recommend_for_new_user_mf <- function(model, new_user_ratings, user_item_matrix, 
                                      item_ids_map, book_info, n_recommendations = 10) {
  
  # For new users, use item-based collaborative filtering as fallback
  # since Matrix Factorization doesn't handle cold start well
  
  ratings_df <- if (is.numeric(new_user_ratings) && !is.null(names(new_user_ratings))) {
    data.frame(item_id = names(new_user_ratings), rating = as.numeric(new_user_ratings), stringsAsFactors = FALSE)
  } else { new_user_ratings }
  
  ratings_df <- ratings_df[ratings_df$item_id %in% item_ids_map$item_id, ]
  if (nrow(ratings_df) == 0) stop("No rated items exist in training data")
  
  # Calculate item means for fallback approach
  item_means <- colMeans(user_item_matrix, na.rm = TRUE)
  
  # Get unrated items
  all_items <- item_ids_map$item_id
  rated_items <- ratings_df$item_id
  unrated_items <- setdiff(all_items, rated_items)
  
  if (length(unrated_items) == 0) return(data.frame())
  
  # Simple item-based approach: use item means as baseline
  # Then adjust based on user's rating pattern
  user_mean_rating <- mean(ratings_df$rating)
  global_mean_rating <- mean(user_item_matrix, na.rm = TRUE)
  
  # Predict ratings using item means adjusted by user bias
  user_bias <- user_mean_rating - global_mean_rating
  pred_ratings <- item_means[unrated_items] + user_bias
  
  # Clip to valid range
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  # Get top recommendations
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  recommended_items <- names(pred_ratings)[top_indices]
  
  data.frame(
    ISBN = recommended_items,
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
}
```





```{r mf-implementation}
# ================================================================
# MATRIX FACTORIZATION IMPLEMENTATION
# ================================================================

# Create matrices with implicit feedback handling
matrix_result_mf <- create_user_item_matrix(
  data,
  min_ratings_per_book = 5,  
  min_ratings_per_user = 3,
  implicit_rating = 4,
  implicit_confidence = 0.5
)

user_item_matrix_mf <- matrix_result_mf$ratings
confidence_matrix_mf <- matrix_result_mf$confidence
original_ratings_matrix_mf <- matrix_result_mf$original_ratings

# Prepare data for recosystem - FILTERS TO EXPLICIT ONLY
prepared <- prepare_recosystem_data(user_item_matrix_mf, original_ratings_matrix_mf)

cat("Matrix Factorization using EXPLICIT ratings only\n")
cat("- Total ratings in matrix:", sum(!is.na(user_item_matrix_mf)), "\n")
cat("- Explicit ratings used:", nrow(prepared$train_data), "\n")
cat("- Implicit ratings excluded:", 
    sum(!is.na(user_item_matrix_mf)) - nrow(prepared$train_data), "\n\n")

# Hyperparameter Optimization
optimal_params_mf <- tune_reco_model(
  train_data = prepared$train_data,
  n_factors = c(10, 20, 30),
  learning_rate = c(0.1, 0.05, 0.01),
  costp_l2 = c(0.01, 0.1),
  costq_l2 = c(0.01, 0.1),
  n_iter = 50,
  verbose = FALSE
)

cat("Optimal hyperparameters:\n")
cat("- Factors:", optimal_params_mf$min$dim, "\n")
cat("- Learning rate:", optimal_params_mf$min$lrate, "\n")
cat("- User regularization:", optimal_params_mf$min$costp_l2, "\n")
cat("- Item regularization:", optimal_params_mf$min$costq_l2, "\n\n")

# Train Final Model with optimal parameters
model_mf <- train_reco_model(
  train_data = prepared$train_data,
  n_factors = optimal_params_mf$min$dim,
  learning_rate = optimal_params_mf$min$lrate,
  costp_l2 = optimal_params_mf$min$costp_l2,
  costq_l2 = optimal_params_mf$min$costq_l2,
  n_iter = 100,
  n_threads = 4,
  verbose = FALSE
)

model_mf
summary(model_mf)
```



```{r mf-recommendations}
# ================================================================
# GENERATE RECOMMENDATIONS
# ================================================================

# Existing user
sample_user <- rownames(user_item_matrix_mf)[3]

recs_existing <- recommend_for_user_mf(
  model = model_mf,
  user_item_matrix = user_item_matrix_mf,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids_map = prepared$user_ids,
  item_ids_map = prepared$item_ids,
  book_info = book_info
)

recs_existing %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(caption = "Top 10 Recommendations for Existing User", digits = 2) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover"))

# New user
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), colnames(user_item_matrix_mf)[1:5])

recs_new <- recommend_for_new_user_mf(
  model = model_mf,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix_mf,
  item_ids_map = prepared$item_ids,
  book_info = book_info,
  n_recommendations = 10
)

recs_new %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(caption = "Top 10 Recommendations for New User", digits = 2) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped","hover"))
```


```{r cross-validation-mf}
# ================================================================
# CROSS-VALIDATION FOR MATRIX FACTORIZATION
# ================================================================

cross_validate_mf <- function(user_item_matrix, 
                              original_ratings_matrix,
                              optimal_params,  # FIXED: Consistent parameter name
                              n_folds = 5, 
                              n_iter = 50, 
                              verbose = TRUE) {
  set.seed(123)
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  if (n_ratings < n_folds * 2) {
    stop("Not enough ratings to perform cross-validation.")
  }
  
  # Assign fold indices randomly
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  # Store RMSE results
  cv_results <- data.frame(
    Fold = integer(n_folds),
    RMSE = numeric(n_folds)
  )
  
  if (verbose) {
    cat("Starting", n_folds, "-fold cross-validation with optimal parameters...\n")
  }
  
  for (fold in 1:n_folds) {
    
    if (verbose) cat("Fold", fold, "of", n_folds, "... ")
    
    # Split train/test by folds
    test_idx <- which(fold_indices == fold)
    train_idx <- which(fold_indices != fold)
    
    train_obs <- observed[train_idx, , drop = FALSE]
    test_obs  <- observed[test_idx, , drop = FALSE]
    
    # Prepare training data (explicit only)
    train_data <- data.frame(
      user_index = train_obs[, 1] - 1,
      item_index = train_obs[, 2] - 1,
      rating = user_item_matrix[train_obs],
      original_rating = original_ratings_matrix[train_obs]
    )
    
    # Filter to explicit ratings only
    train_data_explicit <- train_data[train_data$original_rating != 0, ]
    
    # Prepare test data (explicit only for fair evaluation)
    test_data <- data.frame(
      user_index = test_obs[, 1] - 1,
      item_index = test_obs[, 2] - 1,
      rating = user_item_matrix[test_obs],
      original_rating = original_ratings_matrix[test_obs]
    )
    
    test_data_explicit <- test_data[test_data$original_rating != 0, ]
    
    if (nrow(test_data_explicit) == 0) {
      cv_results[fold, ] <- c(fold, NA)
      next
    }
    
    # Train model
    train_set <- data_memory(
      user_index = train_data_explicit$user_index,
      item_index = train_data_explicit$item_index,
      rating = train_data_explicit$rating,
      index1 = FALSE
    )
    
    test_set <- data_memory(
      user_index = test_data_explicit$user_index,
      item_index = test_data_explicit$item_index,
      rating = test_data_explicit$rating,
      index1 = FALSE
    )
    
    r <- Reco()
    r$train(train_set, opts = list(
      dim = optimal_params$min$dim,
      lrate = optimal_params$min$lrate,
      costp_l2 = optimal_params$min$costp_l2,
      costq_l2 = optimal_params$min$costq_l2,
      niter = n_iter,
      nthread = 4,
      verbose = FALSE
    ))
    
    # Predict
    preds <- r$predict(test_set, out_memory())
    preds <- pmax(pmin(preds, 10), 1)
    
    # Calculate RMSE (explicit only)
    actual <- test_data_explicit$rating
    original <- test_data_explicit$original_rating
    
    metrics <- evaluate_predictions(preds, actual, original)
    cv_results[fold, ] <- c(fold, metrics$rmse_explicit)
    
    cat("RMSE:", round(metrics$rmse_explicit, 4), "\n")
  }
  
  return(cv_results)
}



```




```{r}
# Run cross-validation
cv_results_mf <- cross_validate_mf(
  user_item_matrix = user_item_matrix_mf,
  original_ratings_matrix = original_ratings_matrix_mf,
  optimal_params = optimal_params_mf,
  n_folds = 5,
  n_iter = 50,
  verbose = TRUE
)

# Display results
cv_results_mf %>%
  kable(caption = "MF Cross-Validation Results", 
        digits = 4, 
        col.names = c("Fold", "RMSE"),
        align = c("c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

```





## 4.4 Neural Network-Based Collaborative Filtering

Neural networks can learn complex non-linear relationships between users and items through deep learning architectures. H2O's Deep Learning implementation provides scalable neural network training with hyperparameter optimization capabilities.

```{r nn-data-preparation}
# ================================================================
# DATA PREPARATION FOR H2O WITH IMPLICIT RATINGS
# ================================================================

prepare_h2o_data <- function(user_item_matrix, original_ratings_matrix = NULL) {
  
  # Apply user-mean normalization
  user_means <- rowMeans(user_item_matrix, na.rm = TRUE)
  normalized_matrix <- sweep(user_item_matrix, 1, user_means, FUN = "-")

  # Get observed ratings
  observed <- which(!is.na(normalized_matrix), arr.ind = TRUE)

  # Create training data
  train_data <- data.frame(
    user_id = rownames(user_item_matrix)[observed[, 1]],
    book_id = colnames(user_item_matrix)[observed[, 2]],
    rating = as.numeric(normalized_matrix[observed])
  )
  
  # Add original rating information if provided
  if (!is.null(original_ratings_matrix)) {
    train_data$original_rating <- as.numeric(original_ratings_matrix[observed])
    train_data$is_implicit <- (train_data$original_rating == 0)
  } else {
    train_data$original_rating <- train_data$rating + user_means[observed[, 1]]
    train_data$is_implicit <- FALSE
  }

  return(list(
    train_data = train_data,
    user_ids = data.frame(user_id = rownames(user_item_matrix)),
    book_ids = data.frame(book_id = colnames(user_item_matrix)),
    n_users = nrow(user_item_matrix),
    n_items = ncol(user_item_matrix),
    user_means = user_means,
    original_matrix = user_item_matrix,
    original_ratings_matrix = original_ratings_matrix
  ))
}
```


```{r nn-tuning}
# ================================================================
# HYPERPARAMETER TUNING - COMPREHENSIVE APPROACH
# ================================================================

# Initialize H2O cluster
h2o.init(nthreads = -1, max_mem_size = "4G")

# Create matrices with implicit feedback handling for Neural Network
matrix_result_nn <- create_user_item_matrix(
  data,
  min_ratings_per_book = 5,  
  min_ratings_per_user = 3,
  implicit_rating = 4,
  implicit_confidence = 0.5
)

user_item_matrix_nn <- matrix_result_nn$ratings
original_ratings_matrix_nn <- matrix_result_nn$original_ratings

# Prepare data for hyperparameter tuning
h2o_data_tuning <- prepare_h2o_data(user_item_matrix_nn, original_ratings_matrix_nn)

# Filter to explicit ratings only
train_data_explicit <- h2o_data_tuning$train_data[!h2o_data_tuning$train_data$is_implicit, ]

cat("Neural Network training on EXPLICIT ratings only\n")
cat("- Total ratings in matrix:", sum(!is.na(user_item_matrix_nn)), "\n")
cat("- Explicit ratings used:", nrow(train_data_explicit), "\n")
cat("- Implicit ratings excluded:", sum(h2o_data_tuning$train_data$is_implicit), "\n\n")

# Convert training data to H2O frame
train_h2o_tuning <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
train_h2o_tuning$user_id <- as.factor(train_h2o_tuning$user_id)
train_h2o_tuning$book_id <- as.factor(train_h2o_tuning$book_id)

# ================================================================
# SMART GRID SEARCH: Test network architectures systematically
# ================================================================

cat("Grid Search Strategy:\n")
cat("1. Architecture: Test shallow to deep networks\n")
cat("2. Activation: Test Tanh (stable) vs Rectifier (powerful)\n")
cat("3. Regularization: Test weak to strong L1/L2\n")
cat("4. No dropout (requires specific activation functions)\n\n")

# IMPROVED: More comprehensive architecture search
hyper_params <- list(
  # Test 2 activation functions
  activation = c("Tanh", "Rectifier"),
  
  # Test 5 different architectures (shallow → deep, narrow → wide)
  hidden = list(
    c(16),           # Very shallow: just 1 layer (baseline)
    c(32, 16),       # Shallow: 2 layers, decreasing
    c(64, 32),       # Medium: 2 layers, decreasing  
    c(64, 32, 16),   # Deep: 3 layers, decreasing
    c(128, 64, 32)   # Very deep: 3 layers, wider
  ),
  
  # Test 2 levels of L1 regularization (feature selection)
  l1 = c(0.00001, 0.0001),
  
  # Test 2 levels of L2 regularization (weight decay)
  l2 = c(0.00001, 0.0001)
  
  # NOTE: Removed hidden_dropout_ratios as it requires specific activation functions
  # like RectifierWithDropout, TanhWithDropout, etc.
)

# Calculate total number of models to train
n_models <- length(hyper_params$activation) * 
            length(hyper_params$hidden) * 
            length(hyper_params$l1) * 
            length(hyper_params$l2)

cat(sprintf("Total models to train: %d\n", n_models))
cat("Estimated time: ~%d minutes (depending on your hardware)\n\n", n_models * 2)

set.seed(123)

# Run grid search
model_grid <- h2o.grid(
  algorithm = "deeplearning",
  grid_id = "nn_grid_book_recommendations",
  hyper_params = hyper_params,
  x = c("user_id", "book_id"),
  y = "rating",
  training_frame = train_h2o_tuning,
  nfolds = 5,
  stopping_metric = "RMSE",
  stopping_rounds = 5,              # Increased for deeper networks
  stopping_tolerance = 0.001,
  seed = 123,
  adaptive_rate = TRUE,
  epochs = 50,                      # Increase max epochs for deeper nets
  max_w2 = 10,                      # Weight constraint for stability
  initial_weight_distribution = "UniformAdaptive"
)

# ================================================================
# ANALYZE RESULTS
# ================================================================

cat("\n=== GRID SEARCH RESULTS ===\n")

# Get all results sorted by RMSE
grid_results <- h2o.getGrid(
  "nn_grid_book_recommendations", 
  sort_by = "rmse", 
  decreasing = FALSE
)

# Convert to data frame for analysis
all_models_df <- as.data.frame(grid_results@summary_table)

# Display top 10 models
cat("\nTop 10 Best Models:\n")
kable(head(all_models_df, 10), 
      caption = "Top 10 Neural Network Models (H2O Grid Search)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ================================================================
# ARCHITECTURE ANALYSIS: Which network depth works best?
# ================================================================

cat("\n=== ARCHITECTURE ANALYSIS ===\n")

# Extract number of layers from each model
all_models_df$n_layers <- sapply(strsplit(as.character(all_models_df$hidden), ","), length)

# Average RMSE by number of layers
architecture_analysis <- all_models_df %>%
  group_by(n_layers) %>%
  summarise(
    n_models = n(),
    mean_rmse = mean(as.numeric(rmse), na.rm = TRUE),
    sd_rmse = sd(as.numeric(rmse), na.rm = TRUE),
    best_rmse = min(as.numeric(rmse), na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(mean_rmse)

cat("\nPerformance by Network Depth:\n")
kable(architecture_analysis, 
      caption = "RMSE by Number of Hidden Layers",
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ================================================================
# ACTIVATION FUNCTION ANALYSIS
# ================================================================

activation_analysis <- all_models_df %>%
  group_by(activation) %>%
  summarise(
    n_models = n(),
    mean_rmse = mean(as.numeric(rmse), na.rm = TRUE),
    best_rmse = min(as.numeric(rmse), na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(mean_rmse)

cat("\nPerformance by Activation Function:\n")
kable(activation_analysis, 
      caption = "RMSE by Activation Function",
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ================================================================
# REGULARIZATION ANALYSIS
# ================================================================

regularization_analysis <- all_models_df %>%
  group_by(l1, l2) %>%
  summarise(
    n_models = n(),
    mean_rmse = mean(as.numeric(rmse), na.rm = TRUE),
    best_rmse = min(as.numeric(rmse), na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(mean_rmse)

cat("\nPerformance by Regularization:\n")
kable(regularization_analysis, 
      caption = "RMSE by L1/L2 Regularization",
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ================================================================
# SELECT BEST MODEL
# ================================================================

# Best model
best_model_id <- grid_results@model_ids[[1]]
best_model <- h2o.getModel(best_model_id)
cv_rmse <- h2o.rmse(best_model, xval = TRUE)

cat("\n=== BEST MODEL SELECTED ===\n")

best_nn_results <- data.frame(
  Activation = best_model@allparameters$activation,
  Hidden_Layers = paste(best_model@allparameters$hidden, collapse = ", "),
  N_Layers = length(best_model@allparameters$hidden),
  L1_Regularisation = best_model@allparameters$l1,
  L2_Regularisation = best_model@allparameters$l2,
  CV_RMSE = round(cv_rmse, 4),
  Data_Type = "Explicit Only"
)

kable(best_nn_results, 
      caption = "Best Neural Network Model Specifications") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# ================================================================
# VISUALIZE ARCHITECTURE IMPACT
# ================================================================

# Create visualization of architecture performance
arch_plot <- ggplot(all_models_df, aes(x = factor(n_layers), y = as.numeric(rmse))) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(
    title = "Neural Network Performance by Depth",
    subtitle = "Distribution of RMSE across different network architectures",
    x = "Number of Hidden Layers",
    y = "Cross-Validation RMSE"
  ) +
  theme_minimal()

print(arch_plot)

# Save results
save(best_nn_results, grid_results, all_models_df, file = 'nn_tuning_results.Rdata')

cat("\n✓ Grid search complete! Results saved to 'nn_tuning_results.Rdata'\n")
```


```{r nn-training}
# ================================================================
# MODEL TRAINING FUNCTION - SIMPLE APPROACH
# ================================================================

train_h2o_model <- function(train_data,
                            hidden = c(64, 32),
                            epochs = 30,
                            activation = "Tanh",
                            l1 = 0.0001,
                            l2 = 0.0001,
                            seed = 123,
                            verbose = FALSE) {
  
  if (!verbose) h2o.no_progress()
  
  # Simple neural network training on explicit ratings
  model <- h2o.deeplearning(
    x = c("user_id", "book_id"),
    y = "rating",
    training_frame = train_data,
    hidden = hidden,
    epochs = epochs,
    activation = activation,
    l1 = l1,
    l2 = l2,
    adaptive_rate = TRUE,
    seed = seed,
    stopping_rounds = 5,
    stopping_metric = "RMSE",
    stopping_tolerance = 0.001,
    max_w2 = 10,  # Weight constraint for stability
    initial_weight_distribution = "UniformAdaptive",
    initial_weight_scale = 0.5
  )
  
  return(model)
}
```







```{r nn-recommendation-functions}
# ================================================================
# RECOMMENDATION FUNCTIONS
# ================================================================

# Function to recommend for existing users
recommend_for_user_nn <- function(model, user_item_matrix, user_id,
                                  n_recommendations = 10,
                                  user_ids, book_ids, user_means, book_info) {
  
  # Validate user_id
  if (!user_id %in% user_ids$user_id) {
    stop("User ID not found in the training data.")
  }
  
  # Extract unrated books for the given user
  user_row <- user_item_matrix[as.character(user_id), , drop = TRUE]
  unrated_books <- which(is.na(user_row))
  
  # If the user has rated all books, return empty data frame
  if (length(unrated_books) == 0) {
    message("User has rated all available books — no new recommendations.")
    return(data.frame())
  }
  
  # Prepare prediction input
  pred_data <- data.frame(
    user_id = rep(user_id, length(unrated_books)),
    book_id = colnames(user_item_matrix)[unrated_books],
    stringsAsFactors = FALSE
  )
  
  # Convert to H2O frame
  pred_h2o <- as.h2o(pred_data)
  pred_h2o$user_id <- as.factor(pred_h2o$user_id)
  pred_h2o$book_id <- as.factor(pred_h2o$book_id)
  
  # Predict ratings
  predictions <- h2o.predict(model, pred_h2o)
  predictions_df <- as.data.frame(predictions)
  
  # Denormalize predictions (add back user mean)
  user_mean <- user_means[as.character(user_id)]
  pred_ratings <- predictions_df$predict + user_mean
  
  # Standard neural network prediction - no confidence post-processing needed
  
  # Ensure valid rating bounds
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  # Select top-N recommendations
  n_to_recommend <- min(n_recommendations, length(pred_ratings))
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:n_to_recommend]
  
  # Create final recommendation table
  recommendations <- data.frame(
    ISBN = pred_data$book_id[top_indices],
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  )
  
  recommendations <- recommendations %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}



# Function to recommend for new users (cold start problem)

recommend_for_new_user_nn <- function(model, 
                                              new_user_ratings, 
                                              user_item_matrix,
                                              book_ids,
                                              book_info, 
                                              n_recommendations = 10,
                                              user_means = NULL) {
  
  # Input validation
  if (length(new_user_ratings) == 0) {
    stop("No ratings provided for new user")
  }
  
  rated_books <- names(new_user_ratings)
  valid_books <- rated_books[rated_books %in% book_ids$book_id]
  
  if (length(valid_books) == 0) {
    stop("None of the rated books exist in training data")
  }
  
  # Strategy: Find similar users based on overlap in rated books
  # Then use their preferences weighted by similarity
  
  # 1. Calculate new user's normalized profile
  new_user_mean <- mean(new_user_ratings)
  new_user_normalized <- new_user_ratings - new_user_mean
  
  # 2. Find existing users who rated similar books
  user_similarities <- numeric(nrow(user_item_matrix))
  names(user_similarities) <- rownames(user_item_matrix)
  
  for (i in 1:nrow(user_item_matrix)) {
    # Get overlap in rated books
    user_ratings <- user_item_matrix[i, valid_books]
    valid_overlap <- !is.na(user_ratings)
    
    if (sum(valid_overlap) == 0) {
      user_similarities[i] <- 0
      next
    }
    
    # Calculate cosine similarity on overlapping items
    overlap_books <- valid_books[valid_overlap]
    
    # Normalize existing user's ratings
    existing_mean <- mean(user_ratings[valid_overlap], na.rm = TRUE)
    existing_normalized <- user_ratings[valid_overlap] - existing_mean
    new_normalized <- new_user_normalized[overlap_books]
    
    # Cosine similarity
    numerator <- sum(existing_normalized * new_normalized)
    denominator <- sqrt(sum(existing_normalized^2)) * sqrt(sum(new_normalized^2))
    
    if (denominator > 0) {
      user_similarities[i] <- numerator / denominator
    } else {
      user_similarities[i] <- 0
    }
  }
  
  # 3. Use top-K similar users to make predictions via the neural network
  k <- min(30, sum(user_similarities > 0))
  
  if (k == 0) {
    # Fallback to item popularity
    item_means <- colMeans(user_item_matrix, na.rm = TRUE)
    all_books <- book_ids$book_id
    unrated_books <- setdiff(all_books, rated_books)
    pred_ratings <- item_means[unrated_books] + (new_user_mean - mean(user_means))
  } else {
    # Get top-K similar users
    top_k_users <- names(sort(user_similarities, decreasing = TRUE)[1:k])
    similarities <- user_similarities[top_k_users]
    
    # Get their predictions for unrated books
    all_books <- book_ids$book_id
    unrated_books <- setdiff(all_books, rated_books)
    
    # Weighted average of similar users' ratings for unrated items
    weighted_preds <- numeric(length(unrated_books))
    names(weighted_preds) <- unrated_books
    
    for (book in unrated_books) {
      # Get ratings from similar users
      similar_user_ratings <- user_item_matrix[top_k_users, book]
      valid_ratings <- !is.na(similar_user_ratings)
      
      if (sum(valid_ratings) > 0) {
        weighted_sum <- sum(similarities[valid_ratings] * similar_user_ratings[valid_ratings])
        sum_weights <- sum(abs(similarities[valid_ratings]))
        weighted_preds[book] <- weighted_sum / sum_weights
      } else {
        # Use item mean
        weighted_preds[book] <- mean(user_item_matrix[, book], na.rm = TRUE)
      }
    }
    
    # Adjust for user bias
    pred_ratings <- weighted_preds + new_user_mean - mean(user_means)
  }
  
  # Clip predictions
  pred_ratings <- pmax(pmin(pred_ratings, 10), 1)
  
  # Get top recommendations
  top_indices <- order(pred_ratings, decreasing = TRUE)[1:min(n_recommendations, length(pred_ratings))]
  
  recommendations <- data.frame(
    ISBN = names(pred_ratings)[top_indices],
    Predicted_Rating = round(pred_ratings[top_indices], 2)
  ) %>%
    left_join(book_info, by = "ISBN") %>%
    select(ISBN, Book.Title, Book.Author, Predicted_Rating)
  
  return(recommendations)
}
```


```{r nn-cross-validation}
# ================================================================
# CROSS-VALIDATION FOR NEURAL NETWORK
# ================================================================

cross_validate_nn <- function(user_item_matrix, 
                              original_ratings_matrix = NULL,
                              n_folds = 5, 
                              seed = 123, 
                              hidden = c(64, 32), 
                              epochs = 20, 
                              activation = "Tanh",
                              l1 = 0.0001, 
                              l2 = 0.0001) {
  
  set.seed(seed)
  h2o.no_progress()
  
  # Get observed ratings
  observed <- which(!is.na(user_item_matrix), arr.ind = TRUE)
  n_ratings <- nrow(observed)
  
  if (n_ratings < n_folds * 2) {
    stop("Not enough ratings to perform cross-validation.")
  }
  
  # Create folds
  fold_indices <- sample(rep(1:n_folds, length.out = n_ratings))
  
  cv_results <- data.frame(
    fold = integer(n_folds),
    rmse = numeric(n_folds)
  )
  
  cat("Starting", n_folds, "-fold cross-validation for Neural Network...\n")
  cat("Training on EXPLICIT ratings only (no confidence weighting)\n\n")
  
  for (fold in 1:n_folds) {
    cat("Fold", fold, "of", n_folds, "... ")
    
    # Split data
    test_indices <- which(fold_indices == fold)
    train_indices <- which(fold_indices != fold)
    
    test_obs <- observed[test_indices, , drop = FALSE]
    train_obs <- observed[train_indices, , drop = FALSE]
    
    # Create train matrix
    train_matrix <- user_item_matrix
    train_matrix[test_obs] <- NA
    
    # Normalize training data (prevents data leakage)
    train_user_means <- rowMeans(train_matrix, na.rm = TRUE)
    train_normalized <- sweep(train_matrix, 1, train_user_means, FUN = "-")
    
    # Create training data
    train_data <- data.frame(
      user_id = rownames(user_item_matrix)[train_obs[, 1]],
      book_id = colnames(user_item_matrix)[train_obs[, 2]],
      rating = as.numeric(train_normalized[train_obs])
    )
    
    # Filter to explicit ratings only
    if (!is.null(original_ratings_matrix)) {
      train_data$original_rating <- original_ratings_matrix[train_obs]
      train_data_explicit <- train_data[train_data$original_rating != 0, ]
    } else {
      train_data_explicit <- train_data
    }
    
    # Skip if no training data
    if (nrow(train_data_explicit) == 0) {
      cv_results[fold, ] <- c(fold, NA)
      cat("No explicit ratings - skipping\n")
      next
    }
    
    # Convert to H2O (no confidence column)
    train_h2o <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
    train_h2o$user_id <- as.factor(train_h2o$user_id)
    train_h2o$book_id <- as.factor(train_h2o$book_id)
    
    # Train model (no confidence weighting)
    model <- h2o.deeplearning(
      x = c("user_id", "book_id"),
      y = "rating",
      training_frame = train_h2o,
      hidden = hidden,
      epochs = epochs,
      activation = activation,
      l1 = l1,
      l2 = l2,
      adaptive_rate = TRUE,
      seed = seed,
      stopping_rounds = 3,
      stopping_tolerance = 0.001,
      max_w2 = 10,
      initial_weight_distribution = "UniformAdaptive",
      initial_weight_scale = 0.5,
      verbose = FALSE
    )
    
    # Prepare test data (explicit only)
    test_user_means <- train_user_means[rownames(user_item_matrix)[test_obs[, 1]]]
    test_data <- data.frame(
      user_id = rownames(user_item_matrix)[test_obs[, 1]],
      book_id = colnames(user_item_matrix)[test_obs[, 2]],
      rating = as.numeric(user_item_matrix[test_obs]) - test_user_means
    )
    
    if (!is.null(original_ratings_matrix)) {
      test_data$original_rating <- original_ratings_matrix[test_obs]
      test_data_explicit <- test_data[test_data$original_rating != 0, ]
      test_user_means_explicit <- test_user_means[test_data$original_rating != 0]
      actual_explicit <- user_item_matrix[test_obs][test_data$original_rating != 0]
    } else {
      test_data_explicit <- test_data
      test_user_means_explicit <- test_user_means
      actual_explicit <- user_item_matrix[test_obs]
    }
    
    # Convert to H2O
    test_h2o <- as.h2o(test_data_explicit[, c("user_id", "book_id", "rating")])
    test_h2o$user_id <- as.factor(test_h2o$user_id)
    test_h2o$book_id <- as.factor(test_h2o$book_id)
    
    # Predict
    predictions <- as.data.frame(h2o.predict(model, test_h2o))
    predictions_denorm <- predictions$predict + test_user_means_explicit
    predictions_denorm <- pmax(pmin(predictions_denorm, 10), 1)
    
    # Calculate RMSE on explicit ratings only
    rmse_value <- sqrt(mean((predictions_denorm - actual_explicit)^2, na.rm = TRUE))
    
    cv_results[fold, ] <- list(fold, rmse_value)
    cat("RMSE:", round(rmse_value, 4), "\n")
    
    # Cleanup
    h2o.rm(model)
    h2o.rm(train_h2o)
    h2o.rm(test_h2o)
    gc()
  }
  
  return(cv_results)
}

```



```{r nn-implementation}
# ================================================================
# NEURAL NETWORK IMPLEMENTATION
# ================================================================

# Prepare data for final model
prepared <- prepare_h2o_data(user_item_matrix_nn, original_ratings_matrix_nn)

cat("Data prepared for Neural Network:\n")
cat("Users:", prepared$n_users, "\n")
cat("Items:", prepared$n_items, "\n")
cat("Total ratings:", nrow(prepared$train_data), "\n")
cat("Explicit ratings:", sum(!prepared$train_data$is_implicit), "\n")
cat("Implicit ratings:", sum(prepared$train_data$is_implicit), "\n\n")  


# Extract optimal parameters from grid search
# Check if best_model exists, otherwise use default parameters
if (exists("best_model") && !is.null(best_model)) {
  optimal_params_nn <- list(
    hidden = best_model@allparameters$hidden,
    epochs = ifelse(is.null(best_model@allparameters$epochs), 30, best_model@allparameters$epochs),
    activation = best_model@allparameters$activation,
    l1 = best_model@allparameters$l1,
    l2 = best_model@allparameters$l2
  )
} else {
  # Use default parameters if grid search failed
  cat("Warning: best_model not found, using default parameters\n")
  optimal_params_nn <- list(
    hidden = c(64, 32),
    epochs = 30,
    activation = "Tanh",
    l1 = 0.0001,
    l2 = 0.0001
  )
}


cat("Using optimal parameters:\n")
print(optimal_params_nn)

# Cross-Validation with optimal parameters
cat("Running cross-validation...\n")
nn_cv_results <- cross_validate_nn(
  user_item_matrix_nn,
  original_ratings_matrix_nn,
  n_folds = 5,
  hidden = optimal_params_nn$hidden,
  epochs = optimal_params_nn$epochs,
  activation = optimal_params_nn$activation,
  l1 = optimal_params_nn$l1,
  l2 = optimal_params_nn$l2
)

# Display cross-validation results
nn_cv_results %>%
  kable(caption = "Neural Network Cross-Validation Results (Explicit Ratings Only)", 
        digits = 4, 
        col.names = c("Fold", "RMSE"),
        align = c("c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

# Summary statistics
cat("\nCross-validation summary:\n")
cat("Mean RMSE:", round(mean(nn_cv_results$rmse, na.rm = TRUE), 4), "\n")
cat("SD RMSE:", round(sd(nn_cv_results$rmse, na.rm = TRUE), 4), "\n\n")

# Train Final Model
cat("Training final model with optimal parameters...\n")

# Use explicit ratings only (consistent with CV and MF)
train_data_explicit <- prepared$train_data[!prepared$train_data$is_implicit, ]

cat("Final model training data:\n")
cat("- Explicit ratings:", nrow(train_data_explicit), "\n")
cat("- No confidence weighting (all explicit ratings weighted equally)\n\n")

# Convert to H2O (no confidence column)
train_h2o <- as.h2o(train_data_explicit[, c("user_id", "book_id", "rating")])
train_h2o$user_id <- as.factor(train_h2o$user_id)
train_h2o$book_id <- as.factor(train_h2o$book_id)

# Train final model
model_nn <- train_h2o_model(
  train_data = train_h2o,
  hidden = optimal_params_nn$hidden,
  epochs = optimal_params_nn$epochs,
  activation = optimal_params_nn$activation,
  l1 = optimal_params_nn$l1,
  l2 = optimal_params_nn$l2,
  verbose = FALSE
)

cat("Model training complete!\n\n")

# Display model summary
cat("Final model summary:\n")
print(model_nn)
```


```{r nn-recommendations}
# ================================================================
# GENERATE RECOMMENDATIONS
# ================================================================

### RECOMMENDATIONS FOR EXISTING USER
sample_user <- rownames(user_item_matrix_nn)[3]

recs_nn <- recommend_for_user_nn(
  model = model_nn,
  user_item_matrix = user_item_matrix_nn,
  user_id = sample_user,
  n_recommendations = 10,
  user_ids = prepared$user_ids,
  book_ids = prepared$book_ids,
  user_means = prepared$user_means,
  book_info = book_info
)

recs_nn %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(caption = "Top 10 Recommendations for Existing User (Neural Network)", 
        digits = 2, align = c("c", "l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)


### RECOMMENDATIONS FOR NEW USER

# Simulate new user with 5 book ratings
sample_books <- colnames(user_item_matrix_nn)[1:5]
new_user_ratings <- setNames(c(7, 8, 6, 5, 7), sample_books)

# Display new user's ratings
new_user_ratings_df <- data.frame(
  ISBN = names(new_user_ratings),
  Rating = as.numeric(new_user_ratings)
) %>%
  left_join(book_info, by = "ISBN") %>%
  select(Book.Title, Book.Author, Rating)

new_user_ratings_df %>%
  kable(caption = "New User's Initial Ratings", 
        digits = 0, align = c("l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

# Get recommendations using similarity-based cold start approach
new_user_recs_nn <- recommend_for_new_user_nn(
  model = model_nn,
  new_user_ratings = new_user_ratings,
  user_item_matrix = user_item_matrix_nn,
  book_ids = prepared$book_ids,
  book_info = book_info,
  n_recommendations = 10,
  user_means = prepared$user_means
)

new_user_recs_nn %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Book.Title, Book.Author, Predicted_Rating) %>%
  kable(caption = "Top 10 Recommendations for New User (Similarity-Based Cold Start)", 
        digits = 2, align = c("c", "l", "l", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

# Clean up H2O cluster
h2o.shutdown(prompt = FALSE)
```




# 5. Results and Analysis

## 5.1 Cross-Validation Performance Analysis
```{r}
# ================================================================
# PART 3: COMPREHENSIVE CROSS-VALIDATION COMPARISON
# ================================================================

run_comprehensive_cv_comparison <- function(user_item_matrix, 
                                           original_ratings_matrix,
                                           n_folds = 5, 
                                           optimal_params_mf, 
                                           optimal_params_nn, 
                                           k = 50) {
  
  # Initialize H2O cluster
  h2o.init(nthreads = -1, max_mem_size = "4G")
  h2o.no_progress()
  
  # Pre-allocate results
  all_results <- data.frame(
    Method = character(4),
    CV_RMSE_Mean = numeric(4),
    CV_RMSE_SD = numeric(4),
    stringsAsFactors = FALSE
  )
  
  # User-Based CF
  cat("Running User-Based CF cross-validation...\n")
  # Create confidence matrix for CF methods (simple approach: 1.0 for explicit, 0.5 for implicit)
  confidence_matrix_cf <- ifelse(original_ratings_matrix == 0, 0.5, 1.0)
  confidence_matrix_cf[is.na(confidence_matrix_cf)] <- 0.5
  
  ubcf_cv <- cross_validate_ubcf(user_item_matrix, confidence_matrix_cf, original_ratings_matrix, n_folds = n_folds, k = k)
  all_results[1, ] <- list(
    "User-Based CF",
    round(mean(ubcf_cv$rmse), 3),
    round(sd(ubcf_cv$rmse), 3)
  )
  
  # Item-Based CF
  cat("Running Item-Based CF cross-validation...\n")
  ibcf_cv <- cross_validate_ibcf(user_item_matrix, confidence_matrix_cf, original_ratings_matrix, n_folds = n_folds, k = k)
  all_results[2, ] <- list(
    "Item-Based CF",
    round(mean(ibcf_cv$rmse), 3),
    round(sd(ibcf_cv$rmse), 3)
  )
  
  # Matrix Factorization (explicit ratings only)
  cat("Running Matrix Factorization cross-validation...\n")
  # Create explicit-only matrix for MF
  user_item_matrix_mf <- create_explicit_matrix(
    data,
    min_ratings_per_book = 5,
    min_ratings_per_user = 3
  )
  
  # Create original ratings matrix for MF (explicit only)
  original_ratings_matrix_mf <- user_item_matrix_mf  # For explicit-only, original = current
  
  mf_cv_results <- cross_validate_mf(
    user_item_matrix_mf,
    original_ratings_matrix_mf,
    optimal_params_mf,
    n_folds = n_folds,
    n_iter = 50,
    verbose = FALSE
  )
  all_results[3, ] <- list(
    "Matrix Factorization",
    round(mean(mf_cv_results$RMSE), 3),
    round(sd(mf_cv_results$RMSE), 3)
  )
  
  # Neural Network
  cat("Running Neural Network cross-validation...\n")
  nn_cv_results <- cross_validate_nn(
    user_item_matrix,
    original_ratings_matrix,
    n_folds = n_folds,
    hidden = optimal_params_nn$hidden,
    epochs = optimal_params_nn$epochs,
    activation = optimal_params_nn$activation,
    l1 = optimal_params_nn$l1,
    l2 = optimal_params_nn$l2
  )
  
  all_results[4, ] <- list(
    "Neural Network",
    round(mean(nn_cv_results$rmse), 3),
    round(sd(nn_cv_results$rmse), 3)
  )
  
  # Clean up H2O
  h2o.shutdown(prompt = FALSE)
  
  return(all_results)
}

# ================================================================
# PREPARE DATA AND RUN COMPARISON
# ================================================================

# Create unified matrices ONCE for all methods
if (!exists("user_item_matrix") || !exists("original_ratings_matrix")) {
  cat("Creating unified matrices for comparison...\n")
  matrix_result_unified <- create_user_item_matrix(
    data,
    min_ratings_per_book = 5,
    min_ratings_per_user = 3,
    implicit_rating = 4,
    implicit_confidence = 0.5
  )
  
  user_item_matrix <- matrix_result_unified$ratings
  original_ratings_matrix <- matrix_result_unified$original_ratings
}

# Display matrix information
cat("Using unified matrix for comprehensive comparison:\n")
cat("Matrix dimensions:", nrow(user_item_matrix), "users x", ncol(user_item_matrix), "items\n")
cat("Matrix sparsity:", round(mean(is.na(user_item_matrix)) * 100, 2), "%\n\n")

# Ensure optimal parameters are available for comparison
if (!exists("optimal_params_mf")) {
  optimal_params_mf <- list(
    min = list(
      dim = 20,
      lrate = 0.1,
      costp_l2 = 0.01,
      costq_l2 = 0.01
    )
  )
  cat("Using default Matrix Factorization parameters\n")
}

if (!exists("optimal_params_nn")) {
  optimal_params_nn <- list(
    hidden = c(64, 32),
    epochs = 20,
    activation = "Tanh",
    l1 = 0.00001,
    l2 = 0.00001
  )
  cat("Using default Neural Network parameters\n")
}

cat("Matrix Factorization parameters:\n")
print(optimal_params_mf$min)  # Note: access the 'min' element
cat("\nNeural Network parameters:\n")
print(optimal_params_nn)
cat("\n")

# Run the comprehensive comparison
cat("Starting comprehensive cross-validation comparison...\n\n")
cv_comparison_results <- run_comprehensive_cv_comparison(
  user_item_matrix = user_item_matrix,
  original_ratings_matrix = original_ratings_matrix,
  n_folds = 5,
  optimal_params_mf = optimal_params_mf,
  optimal_params_nn = optimal_params_nn,
  k = 50
)

# Display results in formatted table
kable(cv_comparison_results, 
      caption = "Cross-Validation Performance Comparison - All Methods",
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "center") %>%
  column_spec(2:3, color = "blue") 

```


## 5.2 Dataset Size Impact Analysis

```{r dataset-size-analysis}
# ================================================================
# PART 4: DATASET SIZE VS ACCURACY ANALYSIS (REQUIREMENT 3)
# ================================================================

# First, identify the best performing method from cross-validation
if (exists("cv_comparison_results")) {
  best_method <- cv_comparison_results$Method[which.min(cv_comparison_results$CV_RMSE_Mean)]
  best_rmse <- min(cv_comparison_results$CV_RMSE_Mean)
  
  cat("Best performing method:", best_method, "(RMSE:", best_rmse, ")\n")
  cat("Using", best_method, "for dataset size analysis\n\n")
  
  analyze_dataset_size_impact_best <- function(data, book_sizes = c(30, 60, 90, 120, 150), 
                                              n_folds = 3, best_method, optimal_params_mf, optimal_params_nn) {
    
    # Initialize H2O cluster only if Neural Network is the best method
    if (best_method == "Neural Network") {
      h2o.init(nthreads = -1, max_mem_size = "4G")
      h2o.no_progress()
    }
    
    # Pre-allocate results dataframe for efficiency
    n_total_results <- length(book_sizes)
    results <- data.frame(
      Dataset_Size = integer(n_total_results),
      Method = character(n_total_results),
      RMSE = numeric(n_total_results),
      Users = integer(n_total_results),
      Books = integer(n_total_results),
      Sparsity = numeric(n_total_results),
      stringsAsFactors = FALSE
    )
    
    result_idx <- 1
    
    for (n_books in book_sizes) {
      # Sample books
      available_books <- unique(data$ISBN)
      sampled_books <- sample(available_books, min(n_books, length(available_books)))
      subset_data <- data %>% filter(ISBN %in% sampled_books)
      
      # Create matrix
      subset_matrix <- create_user_item_matrix(
        subset_data, 
        min_ratings_per_book = 3, 
        min_ratings_per_user = 2
      )
      
      n_users <- nrow(subset_matrix)
      n_items <- ncol(subset_matrix)
      sparsity <- round(mean(is.na(subset_matrix)) * 100, 2)
      
      # Evaluate only the best method
      if (best_method == "Item-Based CF") {
        cv_results <- cross_validate_ibcf(subset_matrix, n_folds = n_folds, k = 30)
      } else if (best_method == "User-Based CF") {
        cv_results <- cross_validate_ubcf(subset_matrix, n_folds = n_folds, k = 30)
      } else if (best_method == "Matrix Factorization") {
        cv_results <- cross_validate_mf(subset_matrix, n_folds = n_folds,
                                       n_factors = optimal_params_mf$min$dim,
                                       learning_rate = optimal_params_mf$min$lrate,
                                       costp_l2 = optimal_params_mf$min$costp_l2,
                                       costq_l2 = optimal_params_mf$min$costq_l2,
                                       n_iter = 50)
      } else if (best_method == "Neural Network") {
        cv_results <- cross_validate_nn(subset_matrix, 
                                       original_ratings_matrix = NULL,  # No original ratings matrix in subset
                                       n_folds = n_folds, 
                                       hidden = optimal_params_nn$hidden,
                                       epochs = optimal_params_nn$epochs,
                                       activation = optimal_params_nn$activation,
                                       l1 = optimal_params_nn$l1,
                                       l2 = optimal_params_nn$l2)
      }
      
      results[result_idx, ] <- list(n_books, best_method, round(mean(cv_results$rmse), 3), n_users, n_items, sparsity)
      result_idx <- result_idx + 1
    }
    
    # Clean up H2O cluster if it was initialized
    if (best_method == "Neural Network") {
      h2o.shutdown(prompt = FALSE)
    }
    
    return(results)
  }
  
  # Run dataset size analysis for best method only
  dataset_size_results <- analyze_dataset_size_impact_best(data, 
                                                           book_sizes = c(30, 60, 90, 120, 150),
                                                           n_folds = 3, 
                                                           best_method = best_method,
                                                           optimal_params_mf = optimal_params_mf,
                                                           optimal_params_nn = optimal_params_nn)
  
  # Display results in formatted table
  kable(dataset_size_results, 
        caption = paste("Dataset Size Impact on Predictive Accuracy -", best_method),
        format = "html") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE,
                  position = "center") %>%
    column_spec(3, color = "red") %>%
    add_footnote("Note: Lower RMSE values indicate better performance")
  
} else {
  cat("Cross-validation results not available. Please run the comprehensive comparison first.\n")
}
```



# Discussion

* Further work- seperat models for implicit and explicit ratings - combine intellegently





# EDA stufff

User Activity Analysis

```{r user-activity-preliminary}
# Group and count the number of records per User.ID
hist_data <- data %>%
  group_by(User.ID) %>%
  count(name = "count")
hist_data

# Check the maximum count
max(hist_data$count)

```


Demographic Data Integration

```{r age-analysis}
# Merge with user_info to include age 
full_data <- data %>%
  left_join(user_info, by = "User.ID")

boxplot(full_data$Age) # Age has some very large outliers
full_data <- full_data %>% filter(full_data$Age < 110) 
# data %>% filter(Age < 5)
```


Exploratory Data Analysis

Dataset Characteristics

```{r data-summary}
# Basic summary statistics
data_summary <- data.frame(
  Metric = c("Total Users", "Total Books", "Total Ratings", 
             "Average Rating", "Rating Range"),
  Value = c(
    length(unique(data$User.ID)),
    length(unique(data$ISBN)),
    nrow(data),
    round(mean(data$Book.Rating, na.rm = TRUE), 2),
    paste(min(data$Book.Rating, na.rm = TRUE), "-", max(data$Book.Rating, na.rm = TRUE))
  )
)

kable(data_summary, caption = "Dataset Summary") %>%
  kable_styling(latex_options = "HOLD_position")

# Rating distribution
rating_dist <- table(data$Book.Rating)
rating_dist_df <- data.frame(
  Rating = names(rating_dist),
  Count = as.numeric(rating_dist),
  Percentage = round(as.numeric(rating_dist) / sum(rating_dist) * 100, 1)
)

kable(rating_dist_df, caption = "Rating Distribution") %>%
  kable_styling(latex_options = "HOLD_position")

# Rating histogram
hist(data$Book.Rating, main = "Distribution of Book Ratings",
     col = "steelblue", xlab = "Rating", ylab = "Frequency")

# ================================================================
# USER AND BOOK ACTIVITY ANALYSIS
# ================================================================

# User activity analysis
user_activity <- data %>%
  group_by(User.ID) %>%
  summarise(
    num_ratings = n(),
    avg_rating = mean(Book.Rating, na.rm = TRUE),
    .groups = "drop"
  )

# Book popularity analysis
book_popularity <- data %>%
  group_by(ISBN) %>%
  summarise(
    num_ratings = n(),
    avg_rating = mean(Book.Rating, na.rm = TRUE),
    .groups = "drop"
  )

# User activity summary table
user_activity_summary <- data.frame(
  Metric = c(
    "Total Users",
    "Min Ratings per User", 
    "Max Ratings per User",
    "Mean Ratings per User",
    "Users with 1-2 Ratings",
    "Users with 3-5 Ratings", 
    "Users with 6-10 Ratings",
    "Users with 11-20 Ratings",
    "Users with >20 Ratings"
  ),
  Value = c(
    nrow(user_activity),
    min(user_activity$num_ratings),
    max(user_activity$num_ratings),
    round(mean(user_activity$num_ratings), 2),
    sum(user_activity$num_ratings <= 2),
    sum(user_activity$num_ratings >= 3 & user_activity$num_ratings <= 5),
    sum(user_activity$num_ratings >= 6 & user_activity$num_ratings <= 10),
    sum(user_activity$num_ratings >= 11 & user_activity$num_ratings <= 20),
    sum(user_activity$num_ratings > 20)
  )
)

kable(user_activity_summary, caption = "User Activity Distribution") %>%
  kable_styling(latex_options = "HOLD_position")

# Book popularity summary table
book_popularity_summary <- data.frame(
  Metric = c(
    "Total Books",
    "Min Ratings per Book",
    "Max Ratings per Book", 
    "Mean Ratings per Book",
    "Books with 1-4 Ratings",
    "Books with 5-9 Ratings",
    "Books with 10-19 Ratings", 
    "Books with 20-49 Ratings",
    "Books with ≥50 Ratings"
  ),
  Value = c(
    nrow(book_popularity),
    min(book_popularity$num_ratings),
    max(book_popularity$num_ratings),
    round(mean(book_popularity$num_ratings), 2),
    sum(book_popularity$num_ratings <= 4),
    sum(book_popularity$num_ratings >= 5 & book_popularity$num_ratings <= 9),
    sum(book_popularity$num_ratings >= 10 & book_popularity$num_ratings <= 19),
    sum(book_popularity$num_ratings >= 20 & book_popularity$num_ratings <= 49),
    sum(book_popularity$num_ratings >= 50)
  )
)

kable(book_popularity_summary, caption = "Book Popularity Distribution") %>%
  kable_styling(latex_options = "HOLD_position")

# Dataset sparsity analysis
total_possible_ratings <- length(unique(data$User.ID)) * length(unique(data$ISBN))
actual_ratings <- nrow(data)
sparsity_percentage <- round((1 - actual_ratings / total_possible_ratings) * 100, 2)

sparsity_summary <- data.frame(
  Metric = c(
    "Total Possible User-Book Pairs",
    "Actual Ratings",
    "Sparsity Percentage",
    "Data Density"
  ),
  Value = c(
    format(total_possible_ratings, big.mark = ","),
    format(actual_ratings, big.mark = ","),
    paste0(sparsity_percentage, "%"),
    paste0(round(100 - sparsity_percentage, 2), "%")
  )
)

kable(sparsity_summary, caption = "Dataset Sparsity Analysis") %>%
  kable_styling(latex_options = "HOLD_position")
```