---
title: "LaTeX Installation Test"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Testing Basic LaTeX

This is a test document to verify that LaTeX is installed correctly.

### Math Equations

Inline math: The famous equation is $E = mc^2$.

Display math:
$$\int_{0}^{\infty} e^{-x^2} dx = \frac{\sqrt{\pi}}{2}$$

### More Complex Equations

The quadratic formula:
$$x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}$$

### R Code and Output

```{r}
# Simple R code
x <- 1:10
y <- x^2
plot(x, y, main = "Simple Plot", pch = 19, col = "blue")
```

### Tables

```{r}
knitr::kable(head(mtcars[, 1:5]), caption = "Sample Data from mtcars")
```

### Greek Letters and Symbols

Some Greek letters: $\alpha, \beta, \gamma, \Delta, \Sigma$

Mathematical symbols: $\leq, \geq, \neq, \approx, \infty$

## Conclusion

If you can see this PDF with:

- Properly formatted equations
- A plot
- A table
- Mathematical symbols

Then **LaTeX is installed correctly!** ✓



## Method 1: Using knitr::kable() with LaTeX Caption

```{r}
# Create sample data
students <- data.frame(
  Name = c("Alice", "Bob", "Charlie", "Diana"),
  Math = c(85, 92, 78, 95),
  Science = c(90, 88, 85, 92),
  English = c(88, 85, 90, 89)
)

# Create table with caption
knitr::kable(students, 
             caption = "Student Grades by Subject",
             format = "latex",
             booktabs = TRUE)
```

## Method 2: Using kableExtra for Advanced Formatting

```{r message=FALSE}
library(kableExtra)

knitr::kable(students, 
             caption = "Enhanced Student Performance Table",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4472C4")
```

## Method 3: Pure LaTeX Table

You can also write pure LaTeX code directly:

\begin{table}[h]
\centering
\caption{Sales Data for Q1 2024}
\label{tab:sales}
\begin{tabular}{|l|r|r|}
\hline
\textbf{Month} & \textbf{Revenue (\$)} & \textbf{Units Sold} \\
\hline
January & 45,000 & 150 \\
February & 52,000 & 175 \\
March & 48,000 & 160 \\
\hline
\textbf{Total} & \textbf{145,000} & \textbf{485} \\
\hline
\end{tabular}
\end{table}

## Method 4: Table with Custom LaTeX Positioning

```{r}
summary_data <- data.frame(
  Statistic = c("Mean", "Median", "SD", "Min", "Max"),
  Value = c(85.5, 87.0, 6.2, 78.0, 95.0)
)

knitr::kable(summary_data,
             caption = "\\label{tab:summary}Descriptive Statistics of Test Scores",
             format = "latex",
             booktabs = TRUE,
             align = c('l', 'r'))
```

Note: The `\label{tab:summary}` allows you to reference this table elsewhere in your document using `\ref{tab:summary}`.

## Referencing Tables

You can reference Table \ref{tab:sales} in your text using LaTeX commands.

## Key LaTeX Table Options

- **caption**: Adds a title to your table
- **label**: Allows cross-referencing
- **booktabs = TRUE**: Creates professional-looking tables
- **align**: Controls column alignment ('l' = left, 'c' = center, 'r' = right)
- **format = "latex"**: Ensures proper LaTeX output


