---
title: "Sentiment Analysis"
author: "Hope Hennessy"
date: "2025-09-22"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Sentiment analysis is the study of the emotional content of a body of text. As humans, we infer emotions from text by interpreting the words used and subtle cues in how they are combined. 

Sentiment analysis seeks to replicate this process algorithmically, enabling computers to assess whether text conveys positive, negative, or neutral emotions.

## Approaches to Sentiment Analysis

One simple approach is word-level sentiment analysis. In this method, each word in a text is assigned a sentiment value—positive, negative, or neutral. The overall sentiment of the text is then aggregated from these word-level scores. For example, a document with more positive words than negative ones could be classified as having a positive sentiment.

Word-level sentiment analysis often relies on sentiment lexicons (dictionaries), which are manually compiled lists of words associated with specific sentiment values. 

Popular lexicons include:

* AFINN: Words are scored from -5 (very negative) to +5 (very positive). 
* Bing: Words are classified simply as "positive" or "negative". 
* NRC: Words are labeled with multiple sentiment categories such as "anger", "joy", "trust", as well as general "positive" or "negative".
* Loughran: Specialized for financial documents.

These lexicons form the backbone for many sentiment analysis tasks.


### Handling Negation

A common challenge in sentiment analysis is negation, which can reverse the meaning of a word. For example, "good" is positive, but "not good" is negative. Properly accounting for negation improves the accuracy of sentiment analysis.



[Chapter 2](https://www.tidytextmining.com/sentiment.html) of TMR covers sentiment analysis, and negation is handled in [Chapter 4](https://tidytextmining.com/ngrams.html#tokenizing-by-n-gram). Many of the ideas and some of the code in this workbook are drawn from these chapters. 


# Sentiment Analysis Workflow in R


```{r}
library(tidyverse)
library(tidytext)
library(textdata)

# set plot size in the notebook
options(repr.plot.width = 4, repr.plot.height = 3)
```

Tweets are loaded, cleaned, and transformed into tidy text format, with each word in a separate row.

```{r}
load("trump-tweets.RData")

# make data a tibble
tweets <- as_tibble(tweets)

# parse the date and add some date related variables
tweets <- tweets %>% 
  mutate(date = parse_datetime(str_sub(tweets$created_at, 5, 30), "%b %d %H:%M:%S %z %Y")) %>% 
  mutate(is_potus = (date > ymd(20161108))) %>%
  mutate(month = make_date(year(date), month(date)))

# turn into tidy text 
replace_reg <- "(http.*?(\\s|.$))|(www.*?(\\s|.$))|&amp;|&lt;|&gt;"
unnest_reg <- "[^\\w_#@']"
tidy_tweets <- tweets %>% 
  filter(is_retweet == FALSE) %>%                                      # remove retweets
  mutate(text = str_replace_all(text, replace_reg, '')) %>%            # remove stuff we don't want like links
  unnest_tokens(word, text, token = "regex", pattern = unnest_reg) %>% # tokenize
  filter(!word %in% stop_words$word, str_detect(word, "[A-Za-z]")) %>% # remove stop words
  select(date, word, is_potus, favorite_count, id_str, month)         # choose the variables we need
```


## Using sentiment lexicons

The lexicons can be downloaded and saved to disk for later use by running the code below in the console.

```{r, eval = FALSE}
# afinn <- get_sentiments("afinn") 
# bing <- get_sentiments("bing") 
# nrc <- get_sentiments("nrc") 
# save(afinn, bing, nrc, file = "data/dsfi-lexicons.Rdata")

load("dsfi-lexicons.Rdata")
afinn %>% head(10)
bing %>% head(10)
nrc %>% head(10)
```

* Below we use the *bing* lexicon to add a new variable indicating whether each word in our `tidy_tweets` data frame is positive or negative. 
* We use a left join here, which keeps *all* the words in `tidy_tweets`. 
* Words appearing in our tweets but not in the *bing* lexicon will appear as `NA`. We rename these "neutral", but need to be a bit careful here. 
* No sentiment lexicon contains all words, so some words that are *actually* positive or negative will be labelled as `NA` and hence "neutral". 
* We can avoid this problem by using an inner join rather than a left join, by filtering out neutral words later on, or by just keeping in mind that "neutral" doesn't really mean "neutral".

There's one last issue: in the *bing* lexicon the word "trump" is positive, which will obviously skew the sentiment of Trump's tweets, particularly bearing in mind he often tweets about himself! We (rather generously I think) manually recode the sentiment of this word to "neutral".

```{r}
tidy_tweets <- tidy_tweets %>% 
  left_join(bing) %>%                  # add sentiments (pos or neg)
  relocate(word, sentiment) %>%
  mutate(
    sentiment = case_when(
      word == "trump" ~ "neutral",  
      is.na(sentiment) ~ "neutral",
      TRUE ~ sentiment))
```

Let's look at Trump's 20 most common positive words:

```{r}
tidy_tweets %>%
  filter(sentiment == "positive") %>%
  count(word) %>%
  arrange(desc(n)) %>%
  filter(rank(desc(n)) <= 20) %>%
  ggplot(aes(reorder(word,n),n)) + geom_col() + coord_flip() + xlab("")
```

and the 20 most common negative words:


```{r}
tidy_tweets %>%
  filter(sentiment == "negative") %>%
  count(word) %>%
  arrange(desc(n)) %>%
  filter(rank(desc(n)) <= 20) %>%
  ggplot(aes(reorder(word,n),n)) + geom_col() + coord_flip() + xlab("")
```


## Changes in sentiment over time

Once we have attached sentiments to words in our data frame, we can analyze these in various ways. 

For example, we can examine trends in sentiment over time. Here we count the number of positive, negative and neutral words used each month and plot these. Because the neutral words dominate, its difficult to see any trends with them included (try this and see for yourself). We therefore remove the neutral words before plotting.

```{r}
sentiments_per_month <- tidy_tweets %>%
  group_by(month, sentiment) %>%
  summarize(n = n()) 

sentiments_per_month %>% 
  filter(sentiment != "neutral") %>% 
  ggplot(aes(x = month, y = n, fill = sentiment)) +
  geom_col() 
```

It seems to be relatively balanced throughout, although the variation in the number of tweets made each month makes it difficult to say with any certainty which sentiments dominate over time. We can improve the visualization by plotting the *proportion* of all words tweeted in a month that were positive or negative. The plot shows the raw proportions as well as smoothed versions of these. 


```{r}
sentiments_per_month <- sentiments_per_month %>% 
  group_by(month) %>% 
  mutate(total = sum(n)) %>%
  ungroup() %>% 
  mutate(freq = n/total) 

sentiments_per_month %>% filter(sentiment != "neutral") %>%
  ggplot(aes(x = month, y = freq, colour = sentiment)) +
  geom_line() + 
  geom_smooth(aes(colour = sentiment))
```

We can fit a simple linear model to check with the proportion of negative words has increased over time. Strictly speaking the linear model is not appropriate as the response is bounded to lie between 0 and 1 - you could try fitting e.g. a binomial GLM instead.


```{r}
model <- lm(freq ~ month, data = subset(sentiments_per_month, sentiment == "negative"))
summary(model)
```

## Aggregating sentiment over words

So far we've looked at the sentiment of individual words, but how can we assess the sentiment of longer sequences of text, like bigrams, sentences or entire tweets? 

One approach is to attach sentiments to each word in the longer sequence, and then add up the sentiments over words. This isn't the only way, but it is relatively easy to do and fits in nicely with the use of tidy text data. 
Suppose we want to analyze the sentiment of entire tweets. We'll measure the positivity of a tweet by the difference in the number of positive and negative words used in the tweet.


```{r}
sentiments_per_tweet <- tidy_tweets %>%
  group_by(id_str) %>%
  summarize(net_sentiment = (sum(sentiment == "positive") - sum(sentiment == "negative")),
            month = first(month))
```

To see if the measure makes sense, let's have a look at the most negative tweets.

```{r}
tweets %>% 
  left_join(sentiments_per_tweet) %>% 
  arrange(net_sentiment) %>% 
  head(10) %>%
  select(text, net_sentiment) 
```

And the most positive tweets:


```{r}
tweets %>% 
  left_join(sentiments_per_tweet) %>% 
  arrange(desc(net_sentiment)) %>% 
  head(10) %>%
  select(text, net_sentiment) 
```

We can also look at trends over time. The plot below shows the proportion of monthly tweets that were negative (i.e. where the number of negative words exceeded the number of positive ones).


```{r}
sentiments_per_tweet %>%
  group_by(month) %>%
  summarize(prop_neg = sum(net_sentiment < 0) / n()) %>%
  ggplot(aes(x = month, y = prop_neg)) +
  geom_line() + geom_smooth()
```


## Dealing with negation

One problem we haven't considered yet is what to do with terms like "not good", where a positive word is negated by the use of "not" before it. We need to reverse the sentiment of words that are preceded by negation words like not, never, *etc*.

We'll do this in the context of a sentiment analysis on bigrams. We start by creating the bigrams, and separating the two words making up each bigram. This is the same code used in the previous notebook.


```{r}
bigrams_separated  <- tweets %>%
  filter(is_retweet == FALSE) %>%
  mutate(text = str_replace_all(text, replace_reg, "")) %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  separate(bigram, c("word1", "word2"), sep = " ")
```

Then we use the *bing* sentiment dictionary to look up the sentiment of each word in each bigram. 

```{r}
bigrams_separated <- bigrams_separated %>% 
  
  # add sentiment for word 1
  left_join(bing, by = c(word1 = "word")) %>%
  rename(sentiment1 = sentiment) %>%
  mutate(sentiment1 = case_when(
    word1 == "trump" ~ "neutral",
    is.na(sentiment1) ~ "neutral",
    TRUE ~ sentiment1)) %>%
  
  # add sentiment for word 2
  left_join(bing, by = c(word2 = "word")) %>%
  rename(sentiment2 = sentiment) %>%
  mutate(sentiment2 = case_when(
    word2 == "trump" ~ "neutral",
    is.na(sentiment2) ~ "neutral",
    TRUE ~ sentiment2)) %>%
  
  relocate(month, word1, word2, sentiment1, sentiment2)

bigrams_separated
```

Now we need a list of words that we consider to be negation words. We'll use the following set, taken from TMR [Chapter 4](http://tidytextmining.com/ngrams.html), and show a few examples.

```{r}
negation_words <- c("not", "no", "never", "without")

# show a few
filter(bigrams_separated, word1 %in% negation_words) %>% 
  head(10) %>% 
  select(month, word1, word2, sentiment1, sentiment2) # for display purposes
```

We now reverse the sentiment of `word2` whenever it is preceded by a negation word, and then add up the number of positive and negative words within a bigram and take the difference. That difference (a score from -2 to +2) is the sentiment of the bigram.

We do this in two steps for illustrative purposes. First we reverse the sentiment of the second word in the bigram if the first one is a negation word. 

```{r}
bigrams_separated <- bigrams_separated %>%
  
  # create a variable that is the opposite of sentiment2
  mutate(opp_sentiment2 = recode(sentiment2, "positive" = "negative",
                                 "negative" = "positive",
                                 "neutral" = "neutral")) %>%
  
  # reverse sentiment2 if word1 is a negation word
  mutate(sentiment2 = ifelse(word1 %in% negation_words, opp_sentiment2, sentiment2)) %>%
  
  # remove the opposite sentiment variable, which we don't need any more
  select(-opp_sentiment2)
```

Next, we calculate the sentiment of each bigram and join up the words in the bigram again.

```{r}
bigrams_separated <- bigrams_separated %>%
  mutate(net_sentiment = (sentiment1 == "positive") + (sentiment2 == "positive") - 
           (sentiment1 == "negative") - (sentiment2 == "negative")) %>%
  unite(bigram, word1, word2, sep = " ", remove = FALSE)
bigrams_separated %>% select(word1, word2, sentiment1, sentiment2, net_sentiment)
```

Below we show Trump's most common positive and negative bigrams. 

```{r}
bigrams_separated %>%
  filter(net_sentiment > 0) %>% # get positive bigrams
  count(bigram, sort = TRUE) %>%
  filter(rank(desc(n)) < 20) %>%
  ggplot(aes(reorder(bigram,n),n)) + geom_col() + coord_flip() + xlab("")
```


```{r}
bigrams_separated %>%
  filter(net_sentiment < 0) %>% # get negative bigrams
  count(bigram, sort = TRUE) %>%
  filter(rank(desc(n)) < 20) %>%
  ggplot(aes(reorder(bigram,n),n)) + geom_col() + coord_flip() + xlab("")
```

None of the most common negative bigrams have negated words in them but some that are slightly less frequently used do. Notice that the joint most frequently used bigram below is "no wonder" - which is not really negative, although you can see how, using the approach we have taken, it has ended up classified as such. Cases like these would need to be handled on an individual basis.

```{r}
bigrams_separated %>%
  filter(net_sentiment < 0) %>% # get negative bigrams
  filter(word1 %in% negation_words) %>% # get bigrams where first word is negation
  count(bigram, sort = TRUE) %>%
  filter(rank(desc(n)) < 20) %>%
  ggplot(aes(reorder(bigram,n),n)) + geom_col() + coord_flip() + xlab("")
```

## Exercises

1. Adverbs like "very", "extremely", "totally", "really", "quite", "marginally", *etc* increase or decrease the magnitude of the sentiment that follows them. Think of a few qualifiers and build these into the analyses above. What are the most positive and negative tweets once these qualifiers have been added? Does it affect the assessment of whether the sentiment of Trump's tweets has changed over time?

2. Is there a relationship between the sentiment of a tweet and its popularity? Are negative tweets less popular than positive ones?
3. Is there a "day of the week" effect on the number and sentiment of tweets? Are weekends any different to weekdays?

4. Use the more nuanced *nrc* lexicon, which categorises words as "anger", "fear", *etc* to re-analyse Trump's tweets. Can you pick up anything interesting?

5. Note the appearance of the bigrams "miss usa" and "miss universe" in the plot of most common negative bigrams in the notebook. This is a clear misinterpretation of "miss". How might you remove the ambiguity in these kinds of bigrams in a general way (either just think of a way or, better, implement it!)
